### 2022-11-27 papers/general-stability.tex
cleanup to start new paper

% ****** Start of file apssamp.tex ******
%
%   This file is part of the APS files in the REVTeX 4.2 distribution.
%   Version 4.2a of REVTeX, December 2014
%
%   Copyright (c) 2014 The American Physical Society.
%
%   See the REVTeX 4 README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.2
%
% See the REVTeX 4 README file
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex apssamp.tex
%  2)  bibtex apssamp
%  3)  latex apssamp.tex
%  4)  latex apssamp.tex
%
\documentclass[%
 reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-2}

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage{hyperref}% add hypertext capabilities
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

%\usepackage[showframe,%Uncomment any one of the following lines to test 
%%scale=0.7, marginratio={1:1, 2:3}, ignoreall,% default settings
%%text={7in,10in},centering,
%%margin=1.5in,
%%total={6.5in,8.75in}, top=1.2in, left=0.9in, includefoot,
%%height=10in,a5paper,hmargin={3cm,0.8in},
%]{geometry}

\begin{document}

\title{The two faces of the complexity-stability relationship}

\author{Ann Author}
 \affiliation{MPI MiS}%Lines break automatically or can be forced with \\

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified

\begin{abstract}
Robert May famously predicted that the likelihood of an interacting system to be stable decreases with its complexity. We show that the opposite relationship is equally possible.   
\end{abstract}

\maketitle

In 1970 Robert May used a simple random matrix argument to predict that large, complex ecosystems are not generically stable. While this conclusion is not supported by empirical observation, May's prediction that "complexity begets instability" quickly became a staple of complex systems theory. OVERVIEW OF LITERATURE

May's argument goes as follows. For a dynamical system to be linearly stable, the eigenvalues of its Jacobian matrix evaluated at equilibrium must be have negative real part. Let us assume that the Jacobian can be written as $A = B - I$, where $A$ represent random interactions and $-I$

May's complexity-stability relationship---the prediction that large complex systems are unlikely to be stable, or ``complexity begets instability"---is a landmark of complex systems theory.

$A = N^\alpha B - N^\beta I$

%\tableofcontents
Consider the dynamics of $N$ degrees  
\begin{equation}
  \dot{x}_i = r_ix_i[s_i(x_i) + c_i(x_{-i})]
\end{equation}
\begin{equation}
  J_{ij}(x) = r_ix_i[\delta_{ij}  s_i'(x_i) + (1-\delta_{ij})\partial_j c_i(x_{-i})]
\end{equation}

\end{document}
%
% ****** End of file apssamp.tex ******

### 2022-11-27 scripts/macro-eco-laws.jl

@@ -1,447 +1,0 @@
-using DrWatson, Glob
-@quickactivate
-foreach(include, glob("*.jl", srcdir()))
-
-using ProgressMeter, Suppressor, ThreadsX, DataFrames
-using Plots, LaTeXStrings, DelimitedFiles, Colors, ColorSchemes, LinearAlgebra
-
-### Production statistics ###
-
-P = Dict{Symbol, Any}(
-        :scaled => false,
-        :S => 50,
-        :μ => 1, #10 .^(range(-2,stop=2,length=100)),
-        :σ => 1.,
-        :k => 1.,
-        :b0 => 1.,
-        :λ => 0,
-        :K => 10 .^(range(-2,stop=2,length=5)),
-        :dist => "normal",
-        #:dist_r => Normal(.001,.0),
-        :r => 1,
-        :z => 0,
-        :threshold => false,
-        :N => 1,
-        :symm => false,
-        :seed => 34,
-
-);
-
-l = length(P[:K])
-
-eq_distr = Matrix{Float64}(undef, P[:S], l)
-growth_r = Matrix{Float64}(undef, P[:S], l)
-carrying_cap = Matrix{Float64}(undef, P[:S], l)
-cavity_moments = Matrix{Float64}(undef, 2, l)
-richness = Vector{Float64}(undef, l)
-for (i,p) in enumerate(expand(P))
-    p[:r]=437*p[:K]^(-.25)
-    p[:μ]=p[:r]/p[:K]/p[:S]
-    #p[:μ]*=p[:r]/p[:S]/p[:b0]
-    p[:σ]*=p[:μ]
-    evolve!(p)
-    richness[i] = p[:richness] 
-    eq_distr[:, i] .= p[:equilibrium]
-    growth_r[:, i] .= p[:r]
-    carrying_cap[:, i] .= p[:K]
-#    cavity_moments[1, i] = P_n_gauss(p)[1]
-#    cavity_moments[2, i] = P_n_gauss(p)[2]
-end
-
-scatter([[mean(eq_distr[:,i]) for i in 1:l]],
-[[mean(growth_r[:,i])*P[:b0]^(1-P[:k])*mean(eq_distr[:,i].^P[:k]) for i in 1:l]],
-ylabel = L"\langle b* \ g(b^*)\rangle",
-xlabel = L"\langle b*\rangle",
-linewidth = 2,
-label = false,
[…]

### 2022-11-27 scripts/phase-diagram.jl

@@ -1,17 +1,17 @@
 using Pkg
 Pkg.add("DrWatson")
 
-Pkg.instantiate()
-
 using DrWatson
 @quickactivate
+
+Pkg.instantiate()
 
 using Glob
 foreach(include, glob("*.jl", srcdir()))
 
 
-using ProgressMeter, Suppressor, ThreadsX
-using Plots, LaTeXStrings, DelimitedFiles, Colors, ColorSchemes
+using ProgressMeter, ThreadsX
+using Plots, LaTeXStrings, DelimitedFiles
 
 P = Dict{Symbol, Any}(
         :scaled => true,
@@ -34,93 +34,31 @@ P = Dict{Symbol, Any}(
 
 ϕ = ThreadsX.collect(full_coexistence(p) for p in expand(P));
 a = reshape(ϕ, length(P[:μ]), length(P[:σ]))
-open("../papers/onofrio/fig/prob-stability-S_$(P[:S])-N_$(P[:N])-n₀_$(P[:n0])-σ_$(P[:σ])-μ_$(P[:μ]).txt", "w") do io
-    writedlm(io, a)
-end
 
 sublinear = heatmap(
-P[:μ], 
-P[:σ],
-reshape(pippa, length(P[:μ]), length(P[:σ]))',
-clims = (0,1),
-legend = :none,
-dpi = 500,
-alpha = 1.,
-c = palette([:white, COLOR_LOG49], 100),
-grid = false,
-xlabel = L"\mu S",
-ylabel = L"\sigma \sqrt{S}",
-title = L"\textrm{Stability \,\, in \,\, parameter \,\, space}"
+    P[:μ], 
+    P[:σ],
+    a',
+    clims = (0,1),
+    legend = :none,
+    dpi = 500,
+    alpha = 1.,
+    c = palette([:white, COLOR_LOG49], 100),
+    grid = false,
+    xlabel = L"\mu S",
+    ylabel = L"\sigma \sqrt{S}",
+    title = L"\textrm{Stability \,\, in \,\, parameter \,\, space}"
 )
-
-ν = sqrt(2*log(100))
-
-pippo = [x for x in .01:.005:1]
-plot(pippo, ((1-P[:k])/ν)*(pippo .- pippo.^((P[:k]-3)/(P[:k]-2))),
-color = :green,
-linestyle = :solid,
-linewidth = 4,
-alpha=1,
-ylims = [0.001,0.04],
-xlims = [.0,1]
-)
-
-open("extinction-line.txt", "w") do io
[…]

### 2022-11-27 scripts/single-run.jl

@@ -1,0 +1,52 @@
+using DrWatson, Glob
+@quickactivate
+foreach(include, glob("*.jl", srcdir()))
+
+
+using ProgressMeter, Suppressor, DataFrames
+using Plots, LaTeXStrings
+
+P = Dict{Symbol, Any}(
+    :scaled => false,
+    :S => 20,
+    :μ => .1,
+    :σ => .005,
+    :k => 1.,
+    :b0 => 1.,
+    :K => 5*ones(20),
+    :λ => 0,
+    :z => 0, #10^(-1/4),
+    :r => 1, #rand(LogNormal(1,.08),100),
+    #:dist_r => LogNormal(1,.1),
+    :N => 1,
+    :threshold => false,
+    :dist => "normal",
+    :symm => false,
+    :seed => 17,
+);
+
+#plot multi log, init. cond. [.06,6]
+
+#P[:K]=P[:r].^(-4)
+#P[:r]=P[:K].^(-1/4)
+
+#P[:μ]*=P[:r]/P[:S] #mode(P[:dist_r])/P[:S]
+#P[:σ]*=P[:μ]
+#P[:K]*=P[:r]/(P[:σ]*sqrt(P[:S]+P[:μ]))
+evolve!(P; trajectory = true);
+
+#trajectories
+plot!(P[:trajectory],
+ylabel = L"B_i(t)",
+xlabel = L"t",
+yscale = :log,
+#yticks = [10^(-3),10^(-2),10^(-1), 10^(0), 10, 10^2, 10^3],
+#ylims = [.025,.65],
+#xlims = [0,250],
+linewidth = 3,
+legend = false,
+alpha = 1,
+grid = false,
+palette = :blues, 
+#palette = :YlOrBr_5,
+)

### 2022-11-27 scripts/sinlge-run.jl

@@ -1,187 +1,0 @@
-using DrWatson, Glob
-@quickactivate
-foreach(include, glob("*.jl", srcdir()))
-
-
-using ProgressMeter, Suppressor, DataFrames
-using Plots, LaTeXStrings
-
-P = Dict{Symbol, Any}(
-    :scaled => false,
-    :S => 20,
-    :μ => .1,
-    :σ => .005,
-    :k => 1.,
-    :b0 => 1.,
-    :K => 5*ones(20),
-    :λ => 0,
-    :z => 0, #10^(-1/4),
-    :r => 1, #rand(LogNormal(1,.08),100),
-    #:dist_r => LogNormal(1,.1),
-    :N => 1,
-    :threshold => false,
-    :dist => "normal",
-    :symm => false,
-    :seed => 17,
-);
-
-#plot multi log, init. cond. [.06,6]
-
-#P[:K]=P[:r].^(-4)
-#P[:r]=P[:K].^(-1/4)
-
-#P[:μ]*=P[:r]/P[:S] #mode(P[:dist_r])/P[:S]
-#P[:σ]*=P[:μ]
-#P[:K]*=P[:r]/(P[:σ]*sqrt(P[:S]+P[:μ]))
-evolve!(P; trajectory = true);
-
-#trajectories
-plot!(P[:trajectory],
-ylabel = L"B_i(t)",
-xlabel = L"t",
-yscale = :log,
-#yticks = [10^(-3),10^(-2),10^(-1), 10^(0), 10, 10^2, 10^3],
-#ylims = [.025,.65],
-#xlims = [0,250],
-linewidth = 3,
-legend = false,
-alpha = 1,
-grid = false,
-palette = :blues, 
-#palette = :YlOrBr_5,
-)
-
-#savefig("SM-log-species-abd-sigma.svg")
-
-# plot sub multi, init. cond. [.3,.6]
-P = Dict{Symbol, Any}(
-    :scaled => false,
-    :S => 15,
-    :μ => .1,
-    :σ => .01,
-    :k => .75,
-    :b0 => 1.,
-    :K => 10^20*ones(P[:S]),
-    :λ => 0,
-    :z => 10^(-1/4),
-    :r => 1, #rand(LogNormal(1,.08),100),
-    #:dist_r => LogNormal(1,.1),
-    :N => 1,
-    :threshold => false,
-    :dist => "normal",
-    :symm => false,
-    :seed => 8,
-);
-
-# abunancies
[…]

### 2022-11-27 scripts/stability-vs-diversity.jl

@@ -1,93 +1,0 @@
-using Pkg
-Pkg.add("DrWatson")
-
-Pkg.instantiate()
-
-using DrWatson
-@quickactivate
-
-using Glob
-foreach(include, glob("*.jl", srcdir()))
-
-
-using ProgressMeter, Suppressor, ThreadsX
-using Plots, LaTeXStrings, DelimitedFiles, Colors, ColorSchemes
-
-
-
-# sublinear
-
-P = Dict{Symbol, Any}(
-        :scaled => false,
-        :S => 2:2:100,
-        :μ => .01,
-        :σ => .005,
-        :k => .75,
-        :n0 => 1e-8,
-        :λ => 0,
-        :K => 1e6,
-        :dist => "normal",
-        #:dist_r => Uniform(.01,1),
-        :N => 50,
-        :symm => false,
-    );
-
-ϕ = ThreadsX.collect(full_coexistence(p) for p in expand(P));
-
-open("../papers/onofrio/fig/SUB-stab-vs-div-N_$(P[:N])-S_$(P[:S])-n₀_$(P[:n0])-σ_$(P[:σ])-μ_$(P[:μ])-k_$(P[:k])-K_$(P[:K]).txt", "w") do io
-    writedlm(io, [P[:S] ϕ], ',')
-end 
-
-
-
-# logistic
-
-P = Dict{Symbol, Any}(
-        :scaled => false,
-        :S => 2:2:100,
-        :μ => .01,
-        :σ => .005,
-        :k => 1.,
-        :n0 => 1e-8,
-        :λ => 0,
-        :K => 20,
-        :dist => "normal",
-        #:dist_r => Uniform(.01,1),
-        :N => 1e4,
-        :symm => false,
-    );
-
-ϕ = ThreadsX.collect(full_coexistence(p) for p in expand(P));
-
-open("../papers/onofrio/fig/LOG-stab-vs-div-N_$(P[:N])-S_$(P[:S])-n₀_$(P[:n0])-σ_$(P[:σ])-μ_$(P[:μ])-k_$(P[:k])-K_$(P[:K]).txt", "w") do io
-    writedlm(io, [P[:S] ϕ], ',')
-end
-
-
-#=
[…]

### 2022-11-27 src/cavity.jl

@@ -1,353 +1,0 @@
-using StatsBase
-using Random, Distributions, Statistics
-using DifferentialEquations
-using LinearAlgebra
-using QuadGK
-using HCubature
-using NLsolve
-using SpecialFunctions
-
-# Expression for abundance distribution in sublinear case
-function P_n(n, e1_n, e2_n, p)
-    @unpack μ, σ, k, S = p
-    if !p[:scaled] μ = μ*S end
-    if !p[:scaled] σ = σ*sqrt(S) end
-    if !haskey(p, :dist_r) # unique r
-        r = haskey(p, :r) ? first(p[:r]) : 1
-        return (1-k)*n^(k-2)/(sqrt(2*π*σ^2*e2_n/r^2))*
-        exp(-(n^(k-1)-μ*e1_n/r)^2/(2*σ^2*e2_n/r^2))
-    else # uniformely distributed r
-        a, b = p[:dist_r].a, p[:dist_r].b
-        return (1-k)*n^(k-2)/(sqrt(2*π*e2_n*σ^2)*(b-a)*2*n^(2*k-2))*
-        sqrt(2*π*e2_n*σ^2)*((exp(-(a*n^(k-1)-μ*e1_n)^2/(2*σ^2*e2_n))-
-        exp(-(b*n^(k-1)-μ*e1_n)^2/(2*σ^2*e2_n)))*sqrt(2*π*e2_n*σ^2)+
-        sqrt(pi)*e1_n*μ*(erf((b*n^(k-1)-μ*e1_n)/sqrt(2*σ^2*e2_n))-
-        erf((a*n^(k-1)-μ*e1_n)/sqrt(2*σ^2*e2_n))))
-    end
-end
-
-# Expression for abundance with gaussian approximation
-function P_n_gauss(p)
-    @unpack μ, σ, k, S, b0 = p
-    μ = !p[:scaled] ? μ*S : μ
-    σ = !p[:scaled] ? σ*sqrt(S) : σ 
-    r = haskey(p, :r) ? first(p[:r]) : 1
-    e1 = (μ/r)^(1/(k-2))
-    e2 = e1^2/(1-(1/(k-1))^2*(μ*e1/r)^(2*(2-k)/(k-1))*σ^2)
-    ϕ = e2>0 ? (1+erf((e1-b0*p[:threshold])/sqrt(2*(e2-e1^2))))/2 : 1
[…]

### 2022-11-27 src/constants.jl

@@ -1,3 +1,5 @@
+using Colors, ColorSchemes
+
 COLOR_LOG = :turquoise3
 COLOR_SUB = :darkgoldenrod1
 

### 2022-11-27 src/dynamics.jl

@@ -2,27 +2,19 @@ using DifferentialEquations
 using Random, Distributions
 using LinearAlgebra
 
+## dynamical system
 
-function production(x, p)
-        return (x > p[:b0]*p[:threshold] ? x^p[:k] : 0)
-end
-function dproduction(x, p)
-    return (x > p[:b0]*p[:threshold] ? p[:k]*x^(p[:k]-1) : 0)
-end
+
 
 function F!(f, x, p)
-    f .= p[:r]*p[:b0]^(1 - p[:k]).*(production.(ppart.(x), Ref(p)) .- x.^2 ./p[:K]) .- p[:z]*x .- x.*(p[:a]*x) .+ p[:λ]
+    x .= ppart.(x)
+    f .= p[:P].(x) - 
 end
 
 function J!(j, x, p)
-    j = - x.*p[:a]
-    j[diagind(j)] .= p[:r]*p[:b0]^(1-p[:k]).*(dproduction.(x, Ref(p)) .- 2*x ./p[:K]) .-p[:z] .- p[:a]*x
-
-
-    # above_threshold = x .> p[:b0]
-    # j[diagind(j)[above_threshold]] .+= p[:k].*p[:r][above_threshold].*x[above_threshold].^(p[:k]-1)
+    j = -x .* p[:a]
+    j[diagind(j)] .= p[:r] * p[:b0]^(1 - p[:k]) .* (dproduction.(x, Ref(p)) .- 2 * x ./ p[:K]) .- p[:z] .- p[:a] * x
 end
-
 
 ## solving
 
@@ -31,31 +23,39 @@ MAX_ABUNDANCE = 1e5
 
 blowup() = DiscreteCallback((u, t, integrator) -> maximum(u) > MAX_ABUNDANCE, terminate!)
 
-function evolve!(p; trajectory = false)
+function evolve!(p; trajectory=false)
 
-    if !haskey(p, :rng) p[:rng] = MersenneTwister(p[:seed]) end
-    if !haskey(p, :a) add_interactions!(p) end
-    if !haskey(p, :r) add_growth_rates!(p) end
-    if !haskey(p, :x0) add_initial_condition!(p) end
+    if !haskey(p, :rng)
+        p[:rng] = MersenneTwister(p[:seed])
+    end
+    if !haskey(p, :a)
+        add_interactions!(p)
+    end
+    if !haskey(p, :r)
+        add_growth_rates!(p)
+    end
+    if !haskey(p, :x0)
[…]

### 2022-11-27 src/spectra.jl

@@ -1,54 +1,0 @@
-### This is a sketch, needs updating
-
-
-function spectrum(p)
-    @assert p[:converged]
-    return eigen(J(p[:equilibrium], p)).values
-end
-
-
-T(x, y, n, p) = (q = (p[:μ]-1/p[:K]^(2-p[:k])); 
-            sum( 
-                n.^2 ./ ((x .- q*n .- (p[:k] - 1).*n.^(p[:k] - 1)).^2 .+ y^2)
-            )
-            )
-
-
-
-function boundary(p; overprint = false)
-
-    if !haskey(p, :a) random_interactions!(p) end
-    if !haskey(p, :equilibrium) evolve!(p) end
-    if p[:scaled] p[:σ] = p[:σ]/sqrt(p[:S]) end
-    if p[:scaled] p[:μ] = p[:μ]/p[:S] end
-
-    s = spectrum(p)
-
-
-    n = p[:equilibrium]
-    if p[:converged] #&& all(n .> p[:n0]) 
-        s = spectrum(p)
-        
-        s = s[real.(s) .> .8*minimum(real.(s))]
-        # println(T(0, 0, n, p) - 1/(p[:σ])^2)
-        # println( p[:S]/((1-p[:k])p[:μ]p[:S])^2 - 1/(p[:σ])^2 )
-        X = range(1.1*minimum(real.(s)), .9*maximum(real.(s)); length = 100)
-        Y = range(1.1*minimum(imag.(s)), 1.1*maximum(imag.(s)); length =  100)
-
-        if overprint
-            scatter!(s, alpha = .3, legend = false,
-            aspect_ratio = 1, grid = false)
-        else
-            scatter(s, alpha = .3, legend = false,
-            aspect_ratio = 1, grid = false)
-        end
-
-        Plots.contour!(X, Y, (x,y) -> T(x, y, n, p), levels = [1/(p[:σ])^2],
-        linewidth = 2, color = :auto, colorbar = false)
-    else
-        print("The system is not feasible")
-    end
-    # println((1-p[:k])maximum(a*n))
-    # vline!([- (1-p[:k])maximum(a*n)], color = "green")
-    # vline!([- (1-p[:k])minimum(a*n)], color = "green")
-end

### 2022-11-27 scripts/phase-diagram.jl
works, first plots

@@ -6,66 +6,48 @@ using DrWatson
 
 Pkg.instantiate()
 
-using Glob
-foreach(include, glob("*.jl", srcdir()))
-
+using Glob, Revise
+foreach(includet, glob("*.jl", srcdir()))
 
 using ProgressMeter, ThreadsX
-using Plots, LaTeXStrings, DelimitedFiles
+using Plots, LaTeXStrings
 
 P = Dict{Symbol, Any}(
-        :scaled => true,
-        :S => 1000,
-        :μ => .01:.01:1.2,
-        :σ => .001:.001:.12,
-        :k => .75,
-        :b0 => 1.,
+        :scaled => false,
+        :S => 50,
+        :μ => .01:.1:2.,
+        :σ => .001:.03:.3,
+        :k => 1,
+        :a => 1,
+        :b => 1,
+        :b0 => 0.,
+        :threshold => false,
         :λ => 0,
         :z => 0,
-        :K => 1e10,
-        :threshold => false,
-        :dist => "gamma",
-        #:dist_r => Uniform(.01,1),
-        :N => 1,
+        :K => 1.,
+        :order => 2,
+        :dist => "normal",
+        :N => 5,
         :symm => false,
     );
 
-# full coexistence
 
-ϕ = ThreadsX.collect(full_coexistence(p) for p in expand(P));
-a = reshape(ϕ, length(P[:μ]), length(P[:σ]))
+begin
+    ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
+    a = reshape(ϕ, length(P[:μ]), length(P[:σ]))
 
-sublinear = heatmap(
-    P[:μ], 
-    P[:σ],
-    a',
-    clims = (0,1),
-    legend = :none,
-    dpi = 500,
-    alpha = 1.,
-    c = palette([:white, COLOR_LOG49], 100),
-    grid = false,
-    xlabel = L"\mu S",
-    ylabel = L"\sigma \sqrt{S}",
-    title = L"\textrm{Stability \,\, in \,\, parameter \,\, space}"
-)
-
-## time averaged diversity
-
-ϕ = ThreadsX.collect(time_avg_diversity(p) for p in expand(P));
-a = reshape(ϕ, length(P[:μ]), length(P[:σ]))
-
-sublinear = heatmap(
-    P[:μ], 
-    P[:σ],
-    a',
-    clims = (0,1),
-    legend = :none,
-    dpi = 500,
-    alpha = 1.,
[…]

### 2022-11-27 scripts/single-run.jl

@@ -1,52 +1,38 @@
-using DrWatson, Glob
+using DrWatson, Glob, Revise
 @quickactivate
-foreach(include, glob("*.jl", srcdir()))
+foreach(includet, glob("*.jl", srcdir()))
 
-
-using ProgressMeter, Suppressor, DataFrames
 using Plots, LaTeXStrings
 
-P = Dict{Symbol, Any}(
+p = Dict{Symbol, Any}(
     :scaled => false,
-    :S => 20,
-    :μ => .1,
-    :σ => .005,
-    :k => 1.,
-    :b0 => 1.,
-    :K => 5*ones(20),
-    :λ => 0,
-    :z => 0, #10^(-1/4),
-    :r => 1, #rand(LogNormal(1,.08),100),
-    #:dist_r => LogNormal(1,.1),
-    :N => 1,
+    :S => 50,
+    :μ => .8,
+    :σ => .05,
+    :k => 1,
+    :a => 1,
+    :b => 1,
+    :b0 => 0.,
     :threshold => false,
+    :λ => 0,
+    :z => 0,
+    :K => .3,
+    :order => 2,
     :dist => "normal",
+    :N => 5,
     :symm => false,
-    :seed => 17,
+    :seed => 1
 );
 
-#plot multi log, init. cond. [.06,6]
-
-#P[:K]=P[:r].^(-4)
-#P[:r]=P[:K].^(-1/4)
-
-#P[:μ]*=P[:r]/P[:S] #mode(P[:dist_r])/P[:S]
-#P[:σ]*=P[:μ]
-#P[:K]*=P[:r]/(P[:σ]*sqrt(P[:S]+P[:μ]))
-evolve!(P; trajectory = true);
+evolve!(p; trajectory = true);
 
 #trajectories
-plot!(P[:trajectory],
-ylabel = L"B_i(t)",
-xlabel = L"t",
-yscale = :log,
-#yticks = [10^(-3),10^(-2),10^(-1), 10^(0), 10, 10^2, 10^3],
-#ylims = [.025,.65],
-#xlims = [0,250],
-linewidth = 3,
-legend = false,
-alpha = 1,
-grid = false,
-palette = :blues, 
-#palette = :YlOrBr_5,
+plot(
+    p[:trajectory],
+    ylabel = L"B_i(t)",
+    xlabel = L"t",
+    linewidth = 1,
+    legend = false,
+    grid = false,
+    palette = :blues
 )

### 2022-11-27 src/dynamics.jl

@@ -4,22 +4,20 @@ using LinearAlgebra
 
 ## dynamical system
 
-
-
 function F!(f, x, p)
     x .= ppart.(x)
-    f .= p[:P].(x) - 
-end
+    pop = p[:r] .* x.^p[:k] - p[:z] .* x
 
-function J!(j, x, p)
-    j = -x .* p[:a]
-    j[diagind(j)] .= p[:r] * p[:b0]^(1 - p[:k]) .* (dproduction.(x, Ref(p)) .- 2 * x ./ p[:K]) .- p[:z] .- p[:a] * x
+    X = tensorpower(x.^p[:b], p[:order] - 1)
+    comm = - x.^p[:a] .* map(a -> a ⋅ X, p[:A])
+    f .= pop + comm
 end
-
 ## solving
 
 MAX_TIME = 1e4
-MAX_ABUNDANCE = 1e5
+MAX_ABUNDANCE = 1e4
+
+converged(ϵ = 1e-7) = TerminateSteadyState(ϵ)
 
 blowup() = DiscreteCallback((u, t, integrator) -> maximum(u) > MAX_ABUNDANCE, terminate!)
 
@@ -28,7 +26,7 @@ function evolve!(p; trajectory=false)
     if !haskey(p, :rng)
         p[:rng] = MersenneTwister(p[:seed])
     end
-    if !haskey(p, :a)
+    if !haskey(p, :A)
         add_interactions!(p)
     end
     if !haskey(p, :r)
@@ -40,8 +38,8 @@ function evolve!(p; trajectory=false)
 
     pb = ODEProblem(
         ODEFunction(
-            (f, x, p, t) -> F!(f, x, p); #in-place F faster
-            jac=(j, x, p, t) -> J!(j, x, p) #specify jacobian speeds things up
+            (f, x, p, t) -> F!(f, x, p) #in-place F faster
+            # jac=(j, x, p, t) -> J!(j, x, p) #specify jacobian speeds things up
         ),
         fill(0.1, p[:S]), #initial condition
         (0.0, MAX_TIME),
@@ -49,7 +47,7 @@ function evolve!(p; trajectory=false)
     )
 
     sol = solve(pb,
-        callback=CallbackSet(TerminateSteadyState(1e-3), blowup()),
+        callback=CallbackSet(converged(), blowup()),
         save_on=trajectory #don't save whole trajectory, only endpoint
     )
[…]

### 2022-11-27 src/utils.jl

@@ -1,5 +1,24 @@
 ppart(x) = max(x, 0)
 
+using Tensorial:Vec, ⊗
+function tensorpower(x::Vector, n)
+    if n == 1 
+        return x
+    elseif n == 2
+        X = Vec{length(x)}(x)
+        return X ⊗ X
+    elseif n == 3
+        X = Vec{length(x)}(x)
+        return X ⊗ X ⊗ X
+    else
+        @warn "Can't compute higher powers for now"
+    end
+end
+
+function mytensorproduct(x, y)
+    z = zeros(size(x)..., size(y)...)
+    
+end
 
 function offdiag(A::AbstractMatrix)
     [A[ι] for ι in CartesianIndices(A) if ι[1] ≠ ι[2]]

### 2022-12-06 papers/general-stability.tex
Start draft

@@
-% ****** Start of file apssamp.tex ******
-%
-%   This file is part of the APS files in the REVTeX 4.2 distribution.
-%   Version 4.2a of REVTeX, December 2014
-%
-%   Copyright (c) 2014 The American Physical Society.
-%
-%   See the REVTeX 4 README file for restrictions and more information.
-%
-% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
-% as well as the rest of the prerequisites for REVTeX 4.2
-%
-% See the REVTeX 4 README file
-% It also requires running BibTeX. The commands are as follows:
-%
-%  1)  latex apssamp.tex
-%  2)  bibtex apssamp
-%  3)  latex apssamp.tex
-%  4)  latex apssamp.tex
-%
 \documentclass[%
@@
 
-\usepackage{graphicx}% Include figure files
-\usepackage{dcolumn}% Align table columns on decimal point
-\usepackage{bm}% bold math
-%\usepackage{hyperref}% add hypertext capabilities
-%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
-%\linenumbers\relax % Commence numbering lines
-
-%\usepackage[showframe,%Uncomment any one of the following lines to test 
-%%scale=0.7, marginratio={1:1, 2:3}, ignoreall,% default settings
-%%text={7in,10in},centering,
-%%margin=1.5in,
-%%total={6.5in,8.75in}, top=1.2in, left=0.9in, includefoot,
-%%height=10in,a5paper,hmargin={3cm,0.8in},
-%]{geometry}
-
 \begin{document}
 
-\title{The two faces of the complexity-stability relationship}
+\title{Complexity-stability relationships in large, complex systems}
 
-\author{Ann Author}
- \affiliation{MPI MiS}%Lines break automatically or can be forced with \\
+\author{Onofrio Mazzarisi}
+\author{Matteo Smerlak}
+
+ \affiliation{MPI MiS}
 
@@
 \begin{abstract}
-Robert May famously predicted that the likelihood of an interacting system to be stable decreases with its complexity. We show that the opposite relationship is equally possible.   
+Fifty years ago, Robert May predicted that large, complex systems are unlikely to be stable.
+Here, we revisit May's argument to show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems: in some cases, strong interaction and large diversity are, indeed, destabilizing; in other cases, they are stabilizing.
+Which class a given system falls into depends on the relative convexity of the response to self- vs.
+cross-interactions.
+We illustrate the transition between "May" (complexity begets instability) and "anti-May" (complexity begets stability) behavior with a non-linear  generalization of the Lotka-Volterra model.
 \end{abstract}
@@
 
-In 1970 Robert May used a simple random matrix argument to predict that large, complex ecosystems are not generically stable.
-While this conclusion is not supported by empirical observation, May's prediction that "complexity begets instability" quickly became a staple of complex systems theory.
-OVERVIEW OF LITERATURE
+\section{Introduction}
 
-May's argument goes as follows.
-For a dynamical system to be linearly stable, the eigenvalues of its Jacobian matrix evaluated at equilibrium must be have negative real part.
-Let us assume that the Jacobian can be written as $A = B - I$, where $A$ represent random interactions and $-I$
+Few mathematical arguments have influenced biological thinking like May's 1970 prediction that large, complex systems cannot be stable.
+It is, indeed, a perplexing prediction.
+On the one hand, the argument is so simple and compelling that it hard not to conclude that it must hold universally.
+On the other, it is clear that at least some large, complex systems are stable---or else which stable patterns would biology be studying in the first place?
+If anything, the opposite relationship between complexity and stability seems to hold empirically: rich, strongly coupled ecosystems (eg.
+rainforests) tend to be  stable over time, while sparser ones (eg.
+arctic environments) often exhibit large fluctuations, exctinctions, and invasions.
 
-May's complexity-stability relationship---the prediction that large complex systems are unlikely to be stable, or ``complexity begets instability"---is a landmark of complex systems theory.
+May's argument can be paraphrased as follows.
+If a system with $N$ populations $x_i$ is stable, then it can be characterized by an equilibrium $x^*$.
+Near that equilibrium $x = x^* + y$, the dynamics of the system will be described by linear equations $\delta\dot{y} = A y$, and stability of these equations requires that all eigenvalues of $A$ have negative real part.
+But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions with zero mean and variance $\sigma^2$, and $-I$ corresponds to stabilizing self-interactions on some natural timescale, then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
+This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability".
+(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, to incomplete connectivity, or to correlated interactions.)
 
-$A = N^\alpha B - N^\beta I$
+In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients: spatial or food-web structure, time delays, etc.
+Here, we show that May's argument is incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
 
-%\tableofcontents
-Consider the dynamics of $N$ degrees  
-\begin{equation}
-  \dot{x}_i = r_ix_i[s_i(x_i) + c_i(x_{-i})]
+\section{Results}
+
+\subsection{Model}
+
+Consider the dynamical system in $N$ variables
+\begin{equation}\label{dynamics}
+    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j).
 \end{equation}
-\begin{equation}
-  J_{ij}(x) = r_ix_i[\delta_{ij}  s_i'(x_i) + (1-\delta_{ij})\partial_j c_i(x_{-i})]
+Here $f(x_i)$ represents the contribution of a population $i$ to its own growth (its "production function"), while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations.
+That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp.
+$a_{ji} < 0$) implies a positive (resp.
+negative) effect of $j$ on the growth of $i$.
+In the following we call $g(x_i)$ the "response function".
+
+The classic generalized Lotka-Volterra competition model corresponds to $f$, $g$, $h$ all linear.
+However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
+From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
+Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
+For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level.
+It is also common to model predator-prey interactions with a square root laws ($g(x) \sim h(x) \sim x^{1/2}$), for example.
+
+\subsection{Homogeneous interactions}
+
+Under what condition is \eqref{dynamics} linearly stable?
+We begin by considering the simple case where all self-interactions have the same strength $a_{ii} = a_{\textrm{s}}$, and similarly for cross-interactions $a_{ij} = a_{\textrm{c}}$ ($i\neq j$).
+In that case, a straightforward computation shows that an equilibrium $x_*$ will be linearly stable if
+\begin{equation}\label{homogeneous-general}
+    \left(\frac{f'_*}{f_*} - \frac{g'_*}{g_*}\right) + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g_*h'_*}{f_*} < 0. 
 \end{equation}
+Thus, the stability of a competitive equilibrium depends on the relative strength of self- and cross-interactions ($a_{\textrm{s}} - a_{\textrm{c}}$), but also on the relative convexity of the production and response functions ($f'_*/f_* - g'_*/g_*$).
+With power laws, \eqref{homogeneous-general} evaluates to $(\alpha - \beta)(N-1) < \gamma(a_{\textrm{s}}/a_{\textrm{c}}- 1) - (\alpha - \beta)(a_{\textrm{s}}/a_{\textrm{c}})$, leading to three different regimes:
+\begin{itemize}
+    \item If $\alpha = \beta$, stability requires $a_{\textrm{s}} > a_{\textrm{c}}$, i.e.
+self-interactions must be stronger than cross-interactions.
+This is the usual conclusion drawn from the Lotka-Volterra model.
+    \item If $\alpha > \beta$, stability places an upper bound on $N$: the more complex the system, the less likely to be stable. We can call this "May" behavior.
+    \item If $\alpha < \beta$, stability places an lower bound on $N$: the more complex the system, the more likely to be stable. This is "anti-May" behavior.
+\end{itemize}
+
+\subsection{Random interactions}
+
+Consider now \eqref{dynamics} with i.i.d. interactions $a_{ij}$ with mean $\mu$ and standard deviation $\sigma$.
 

### 2022-12-06 scripts/jacobian.jl

@@ -1,0 +1,45 @@
+using DrWatson, Glob, Revise
+@quickactivate
+foreach(includet, glob("*.jl", srcdir()))
+
+using Plots, LaTeXStrings
+
+P = Dict{Symbol, Any}(
+    :scaled => false,
+    :μ => .1,
+    :σ => .001,
+    :k => 1,
+    :a => 1,
+    :b => 1,
+    :b0 => .01,
+    :threshold => false,
+    :z => 0.1,
+    :K => 1.,
+    :order => 2,
+    :dist => "normal",
+    :N => 1,
+    :symm => false,
+    :seed => rand(UInt)
+);
+
+P[:S] = 50:20:150
+
+
+plot(legend = :topleft)
+for k in .8:.1:1.2
+        P[:k] = k
+        mean_diag = []
+        mean_offdiag = []
+        for p in expand(P)
+            evolve!(p)
+            if p[:converged]
+                j = J(p[:equilibrium], p)
+                push!(mean_diag, mean(diag(j)))
+                push!(mean_offdiag, mean(offdiag(j)))
+                # plot!(offdiag(j); seriestype = :stephist, label = "S = $(p[:S])")    
+            end
+        end
+        plot!(mean_diag ./ mean_offdiag, label = "k/a = $(p[:k]/p[:a])")
+end
+current()
+

### 2022-12-06 scripts/phase-diagram.jl

@@ -14,40 +14,34 @@ using Plots, LaTeXStrings
 
 P = Dict{Symbol, Any}(
         :scaled => false,
-        :S => 50,
+        :S => 20,
         :μ => .01:.1:2.,
-        :σ => .001:.03:.3,
-        :k => 1,
+        :σ => .001:.05:.5,
+        :k => .75,
         :a => 1,
         :b => 1,
-        :b0 => 0.,
+        :b0 => 10.,
         :threshold => false,
-        :λ => 0,
-        :z => 0,
-        :K => 1.,
+        :z => 0.1,
+        :K => 10.,
         :order => 2,
         :dist => "normal",
-        :N => 5,
+        :N => 4,
         :symm => false,
     );
 
-
 begin
     ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
-    a = reshape(ϕ, length(P[:μ]), length(P[:σ]))
 
     sublinear = heatmap(
-        P[:S] * P[:μ], 
-        sqrt(P[:S]) * P[:σ],
-        a',
-        clims = (0,1),
-        legend = :none,
+        P[:μ], 
+        P[:σ],
+        reshape(ϕ, length(P[:μ]), length(P[:σ]))',
         dpi = 500,
         alpha = 1.,
-        c = palette([:white, COLOR_LOG49], 100),
         grid = false,
-        xlabel = L"\mu S",
-        ylabel = L"\sigma \sqrt{S}",
+        xlabel = L"\mu",
+        ylabel = L"\sigma",
         title = L"\textrm{Stability \,\, in \,\, parameter \,\, space}"
     )
 end

### 2022-12-06 scripts/single-run.jl

@@ -7,21 +7,20 @@ using Plots, LaTeXStrings
 p = Dict{Symbol, Any}(
     :scaled => false,
     :S => 50,
-    :μ => .8,
-    :σ => .05,
-    :k => 1,
+    :μ => 0.,
+    :σ => .1,
+    :k => .75,
     :a => 1,
     :b => 1,
-    :b0 => 0.,
-    :threshold => false,
-    :λ => 0,
-    :z => 0,
-    :K => .3,
-    :order => 2,
+    :b0 => .01,
+    :threshold => true,
+    :z => 0.1,
+    :K => 1.,
+    :order => 4,
     :dist => "normal",
-    :N => 5,
+    :N => 1,
     :symm => false,
-    :seed => 1
+    :seed => rand(UInt)
 );
 
 evolve!(p; trajectory = true);

### 2022-12-06 scripts/stability-diversity.jl

@@ -1,0 +1,35 @@
+using DrWatson
+@quickactivate
+
+using Glob, Revise
+foreach(includet, glob("*.jl", srcdir()))
+
+using ProgressMeter, ThreadsX
+using Plots, LaTeXStrings
+
+P = Dict{Symbol, Any}(
+        :scaled => false,
+        :S => 2:5:100,
+        :μ => 0.,
+        :σ => .1,
+        :k => 1.,
+        :a => 1,
+        :b => 1,
+        :b0 => 1.,
+        :threshold => false,
+        :z => 0.1,
+        :K => 1.,
+        :order => 4,
+        :dist => "normal",
+        :N => 2,
+        :symm => false,
+    );
+
+ϕ = ThreadsX.collect(
+    full_coexistence!(p) for p in expand(P)
+    );
+
+plot(
+    P[:S],
+    ϕ
+)

### 2022-12-06 src/dynamics.jl

@@ -1,23 +1,37 @@
 using DifferentialEquations
 using Random, Distributions
 using LinearAlgebra
+using ForwardDiff
 
 ## dynamical system
 
 function F!(f, x, p)
     x .= ppart.(x)
-    pop = p[:r] .* x.^p[:k] - p[:z] .* x
 
-    X = tensorpower(x.^p[:b], p[:order] - 1)
-    comm = - x.^p[:a] .* map(a -> a ⋅ X, p[:A])
+    pop = p[:r] .* x.^p[:k] - p[:z] .* x
+    pop[x .< p[:b0]] .= 0
+    comm = - x.^p[:a] .* contract(p[:A], x.^p[:b])
+
     f .= pop + comm
 end
+
+function F(x, p)
+    x .= ppart.(x)
+    pop = p[:r] .* x.^p[:k] - p[:z] .* x
+    pop[x .< p[:b0]] .= 0
+    comm = - x.^p[:a] .* contract(p[:A], x.^p[:b])
+
+    return pop + comm
+end
+
+J(x, p) = ForwardDiff.jacobian(y -> F(y, p), x)
+
 ## solving
 
-MAX_TIME = 1e4
-MAX_ABUNDANCE = 1e4
+MAX_TIME = 1e3
+MAX_ABUNDANCE = 1e3
 
-converged(ϵ = 1e-7) = TerminateSteadyState(ϵ)
+converged(ϵ = 1e-3) = TerminateSteadyState(ϵ)
 
 blowup() = DiscreteCallback((u, t, integrator) -> maximum(u) > MAX_ABUNDANCE, terminate!)
 
@@ -41,7 +55,7 @@ function evolve!(p; trajectory=false)
             (f, x, p, t) -> F!(f, x, p) #in-place F faster
             # jac=(j, x, p, t) -> J!(j, x, p) #specify jacobian speeds things up
         ),
-        fill(0.1, p[:S]), #initial condition
+        ones(p[:S]), #initial condition
         (0.0, MAX_TIME),
         p
     )
@@ -117,9 +131,11 @@ function add_interactions!(p)
         dist = Gamma(m^2 / s^2, s^2 / m)
     end
 
-    A = [rand(p[:rng], dist, Tuple(fill(p[:S], p[:order] - 1))) for _ in 1:p[:S]]
+    A = rand(
+        p[:rng], dist, Tuple(fill(p[:S], p[:order]))
+    )
     for i in 1:p[:S]
[…]

### 2022-12-06 src/utils.jl

@@ -1,24 +1,10 @@
 ppart(x) = max(x, 0)
 
-using Tensorial:Vec, ⊗
-function tensorpower(x::Vector, n)
-    if n == 1 
-        return x
-    elseif n == 2
-        X = Vec{length(x)}(x)
-        return X ⊗ X
-    elseif n == 3
-        X = Vec{length(x)}(x)
-        return X ⊗ X ⊗ X
-    else
-        @warn "Can't compute higher powers for now"
-    end
-end
+using TensorOperations
+contract(A::Array{Float64, 2}, x::Vector) = @tensor y[i] := A[i, j] * x[j]
+contract(A::Array{Float64, 3}, x::Vector) = @tensor y[i] := A[i, j, k] * x[j] * x[k]
+contract(A::Array{Float64, 4}, x::Vector) = @tensor y[i] := A[i, j, k, l] * x[j] * x[k] * x[l]
 
-function mytensorproduct(x, y)
-    z = zeros(size(x)..., size(y)...)
-    
-end
 
 function offdiag(A::AbstractMatrix)
     [A[ι] for ι in CartesianIndices(A) if ι[1] ≠ ι[2]]

### 2022-12-23 papers/general-stability.tex
alpha-beta phase diagram

@@
 ]{revtex4-2}
+
+\usepackage{colortbl}
+\usepackage{graphicx} 
+\usepackage{booktabs}
+\usepackage{caption}
+\newcommand{\red}{\textcolor{red}}
 
@@
 
-Few mathematical arguments have influenced biological thinking like May's 1970 prediction that large, complex systems cannot be stable.
-It is, indeed, a perplexing prediction.
-On the one hand, the argument is so simple and compelling that it hard not to conclude that it must hold universally.
+Few mathematical arguments have influenced biological thinking like May's prediction that large, complex systems cannot be stable.
+It is, on the face of it, a perplexing conclusion.
+On the one hand, May's argument is so simple and compelling that it hard not to conclude that it must hold universally.
 On the other, it is clear that at least some large, complex systems are stable---or else which stable patterns would biology be studying in the first place?
@@
 
-May's argument can be paraphrased as follows.
+May's argument goes as follows.
 If a system with $N$ populations $x_i$ is stable, then it can be characterized by an equilibrium $x^*$.
-Near that equilibrium $x = x^* + y$, the dynamics of the system will be described by linear equations $\delta\dot{y} = A y$, and stability of these equations requires that all eigenvalues of $A$ have negative real part.
+Near that equilibrium $x = x^* + \delta x$, the dynamics of the system will be described by linear equations $d(\delta x)/dt = A (\delta x)$, and stability of these equations requires that all eigenvalues of $A$ have negative real part.
 But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions with zero mean and variance $\sigma^2$, and $-I$ corresponds to stabilizing self-interactions on some natural timescale, then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
@@
 \begin{equation}\label{dynamics}
-    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j).
+    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j) \, .
 \end{equation}
@@
 
-Under what condition is \eqref{dynamics} linearly stable?
+Under what condition does \eqref{dynamics} admit a linearly stable equilibrium?
 We begin by considering the simple case where all self-interactions have the same strength $a_{ii} = a_{\textrm{s}}$, and similarly for cross-interactions $a_{ij} = a_{\textrm{c}}$ ($i\neq j$).
-In that case, a straightforward computation shows that an equilibrium $x_*$ will be linearly stable if
+Under these assumptions, a straightforward computation shows that an equilibrium $x_*$ will be linearly stable if
 \begin{equation}\label{homogeneous-general}
@@
 Thus, the stability of a competitive equilibrium depends on the relative strength of self- and cross-interactions ($a_{\textrm{s}} - a_{\textrm{c}}$), but also on the relative convexity of the production and response functions ($f'_*/f_* - g'_*/g_*$).
-With power laws, \eqref{homogeneous-general} evaluates to $(\alpha - \beta)(N-1) < \gamma(a_{\textrm{s}}/a_{\textrm{c}}- 1) - (\alpha - \beta)(a_{\textrm{s}}/a_{\textrm{c}})$, leading to three different regimes:
+With power laws, \eqref{homogeneous-general} evaluates to
+\begin{equation}
+    (\alpha - \beta)(N-1) < \gamma(a_{\textrm{s}}/a_{\textrm{c}}- 1) - (\alpha - \beta)(a_{\textrm{s}}/a_{\textrm{c}}),
+\end{equation}
+leading to three different regimes:
 \begin{itemize}
@@
 
-Consider now \eqref{dynamics} with i.i.d. interactions $a_{ij}$ with mean $\mu$ and standard deviation $\sigma$.
+Consider now Eq.~\eqref{dynamics} with i.i.d. random 
+interactions $a_{ij}$ with mean $\mu$ and standard 
+deviation $\sigma$, with the diagonal elements $a_{ii}$ 
+extracted from a distribution with mean $\mu_s$ 
+possibly different from $\mu$ and same standard deviation
+as for the off-diagonal elements.  
+We can compute the Jacobian matrix at equilibrium
+\begin{align}
+    J_{ij}^* & = a_{ij}g_*(x_i)h_*'(x_j) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
+    J_{ii}^* & = f_*'(x_i) - \frac{g_*'(x_i)f_*(x_i)}{g_*(x_i)} - a_{ii}g_*(x_i)h_*'(x_j) \ , \label{eq: jac diag}
+\end{align}
+where we used $\sum_{j}a_{ij}g_*(x_i)h_*(x_j)=-f_*(x_i)/g_*(x_i)$.
+In order to investigates its spectral properties, 
+we follow Stone~[Stone2018] and use a
+result in random matrix theory~[Ahmadian2015]
+which generalize the classical `circular law'~[Circular].
+
+In Ref.~[Ahmadian2015] Ahmadian \emph{et al.} 
+consider matrices of the form $M + LSR$, where $M$,  
+$L$ and $R$ are deterministic matrices, and $S$ 
+is a random matrix with i.i.d. coefficients, 
+zero mean and variance $\sigma^2$. They 
+show that in the complex plane the boundary of the 
+eigenspectrum of large matrices of this form is defined by
+\begin{equation}
+    \textrm{Tr}[(M_\zeta M_\zeta^\dagger)^{-1}]\geq \sigma^{-2} \, , 
+\end{equation}
+where $M_\zeta = L^{-1}(\zeta I - M)R^{-1}$ and $\zeta\in\mathbb{C}$. 
+If $L$, $R$ and $M$ are all diagonal $N\times N$ matrices, 
+this condition simplifies to
+\begin{equation}
+    \sum_{i=1}^N\frac{(L_{i}R_{i})^2}{ \vert \zeta - M_{i}\vert^2 }\geq \sigma^{-2} \ .
+\label{eq: domain}
+\end{equation}
+
+Now we decompose the interaction matrix as
+$a = \mu \mathbf{1} + (\mu_s-\mu)I + S$,
+with $\mathbf{1}$ the rank-one matrix with all entries equal to $1$
+and $S$ a random matrix as defined above.
+We can limit ourselves to consider 
+$a = (\mu_s-\mu)I + S$ by relying on 
+rank-one perturbation theory, we would only neglect outliers
+uninfluential for the linear stbaility properties of the system by doing so
+(see Ref.~[Stone2018]).
+The Jacobian in Eqs.~\eqref{eq: jac off-diag} and \eqref{eq: jac diag} assumes therefore the form $M + LJR$ with
+\begin{align}
+    M & = \textrm{diag}\left(f_*'(\mathbf x) -
+    \frac{g_*'(\mathbf x)f_*(\mathbf x)}{g_*(\mathbf x)}
+    +(\mu_s-\mu)g_*(\mathbf x)h_*'(\mathbf x)\right) \ , \\
+    L &= \textrm{diag}(g_*(\mathbf x)) \ , \\
+    R &= \textrm{diag}(h_*'(\mathbf x)) \ .
+\end{align}
+The domain of the eigenvalues of $J^*$, defined by
+Eq.~\eqref{eq: domain}, 
+first touches the right half-plane at $\zeta = 0$, 
+hence linear stability requires   
+\begin{equation}\label{eq: random-stability}
+    \sum_{i=1}^N \cfrac{\left(\cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)}\right)^2}{
+        \Bigl |\left(\cfrac{f_*'(x_i)}{f_*(x_i)} -
+        \cfrac{g_*'(x_i)}{g_*(x_i)}\right)
+        +(\mu_s-\mu) \cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)} \Bigl |^2}
+    < \sigma^{-2}. 
+\end{equation}
+This expression can be considered the
+random interaction case equivalent of
+the expression in Eq.~\eqref{homogeneous-general} 
+which appears again in the denominator of the summand.
+It reduces indeed to the homogenous case for $\sigma\to 0$.
+\red{Discuss better.}
+
+There are straightforward results 
+that follows from this expression.
+For generalized Lotka-Volterra (GLV) models,
+i.e. $f(x_i)=x_i$, $g(x_i)=x_i$ and $h(x_i)=x_i$,
+the expression in Eq.~\eqref{eq: random-stability}
+becomes independent on the equilibria and we easily 
+recover the linear stability condition
+\begin{equation}
+    \sigma\sqrt{N} < (\mu_s-\mu) \ .
+\end{equation}
+Moreover we can define an equivalence class of models
+which share the same simple condition
+for linear stability defined by the equivalence relation
+\begin{equation}\label{eq: equiv class glv}
+    \left(\cfrac{f_*'(x_i)}{f_*(x_i)} -
+        \cfrac{g_*'(x_i)}{g_*(x_i)}\right)\Bigl/
+        \cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)}=c \ ,
+\end{equation}
+where $c$ is a constant,
+resulting in the condition $\sigma\sqrt{N} < (\mu_s-\mu) +c$.
+
+In the general case in which the dependence on the equilibria
+$\mathbf{x}_*$ does not cancel out, is not possible to
+obtain a clean complexity-stability relationship as before.
+We can make progress in the large $N$ limit by
+trasforming the sum in an integral 
+$\sum_{i=1}^N\to N\int dx_*P(x_*)$, where $P(x_*)$ 
+is the equilibrium distribution of the system
+depending on the statistic of the interactions,
+and write
+\begin{equation}\label{eq: random-stability int}
+    \int dx_* \cfrac{NP(x_*) \left(\cfrac{g_*(x)h_*'(x)}{f_*(x)}\right)^2}{
+        \Bigl |\left(\cfrac{f_*'(x)}{f_*(x)} -
+        \cfrac{g_*'(x)}{g_*(x)}\right)
+        +(\mu_s-\mu) \cfrac{g_*(x)h_*'(x)}{f_*(x)} \Bigl |^2}
+    < \sigma^{-2} \ .
+\end{equation}
+This expression allows us to obtain insights on the stability properties of
+the system in the case in which we are able to compute or estimate 
+the probability distribution function $P(x_*)$.
+
+In order to investigate the different regimes
+of complexity-stability behaviour,
+let us specialize to the power laws case, with 
+\begin{equation}
+    f(x_i)=x_i^{\alpha} \ , \quad g(x_i)=-x_i^{\beta} \ , \quad h(x_i)=x_i^{\gamma} \ .
+\end{equation} 
+Additional proportionality constants can be reabsorbed in the statistics
+of $a$ and by a rescaling of time, moreover we chose to explicitly
+factor out a minus sign in order to more straightforwardly connect
+with competitive models.
+
+We wish to consider explictly cases
+for "May" and "anti-May" regimes.
+An example of the former is given by GLV.
+In order to explore the latter, we 
+need to resort to Eq.~\eqref{eq: random-stability int} and therefore
+to compute the equilibrium probability distribution function
+for the system. This can be achieved in different ways,
+depending among other things on specific constraint that
+one can impose on the system. 
+One approach is reported below, then we specialize to a choice of
+the power laws exponens.
+
+Following, e.g., Ref.~[Roy2019] we can derive from 
+the evolution equation
+\begin{equation}
+    \dot{x_i}=x_i^{\alpha} - x_i^{\beta}\sum_{j}a_{ij}x_j^{\gamma} \, ,
+\label{eq: full abg}
+\end{equation}
+for large $N$, a dynamical mean field theory (DMFT) 
+describing the entire ensemble of $N$ variables
+thorugh the evolution of the representative random variable $x$
+\begin{equation}
+    \dot{x} = x^{\alpha}-a_sx^{\beta+\gamma}-x^{\beta}\left( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\right) \, ,
+\label{eq: dmft}
+\end{equation}
+where $a_s$ is a random variable with the statistics of $a_{ii}$,
+$\langle . \rangle$ stands for expectation value
+and $\eta$ is a gaussian variable
+with zero mean and correlation 
+$\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$. 
+The averages and correlations have to be computed self-consistently.
+If the system admits an equilibrium (i.e. is feasible), for 
+$t\to\infty$ each of the initial $N$ degrees 
+of freedom will reach a final value, 
+and the effective random variable describing their
+also become time-independent,
+$\lim_{t\to\infty}x(t)=x_*$ and we can write
+\begin{equation}
+    x_*^{\alpha-\beta} - a_s x_*^{\gamma}= \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\right) \, ,
+\end{equation} 
+where $\xi$ is a standard normal random variable (i.e. $\xi\sim\mathcal{N}(0,1)$).
+The equation above can be solved for $x_*$ in if $(\alpha-\beta)/\gamma\in\mathbb{N}$,
+or in general, if we set $a_s=0$, which amounts to include
+all the self-interaction in the nonlinear produciton term
+$x_*^{\alpha}$.
+In this case the stationary solution is given by 
+\begin{equation}
+    x_* = \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
+\end{equation}
+
+The equilibrium probability distribution function for $x_*$, $P(x_*)$,
+can be obtained as the pushforward of the distribition of $\xi$.
+The change of measure
+$\left|d \xi/d x_*\right|=|\alpha-\beta|x_*^{\alpha-\beta-1}/\sigma \sqrt{N\langle x_*^{2\gamma}\rangle}$,
+leads to
+\begin{equation}\label{eq: dist general}
+    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2 N\langle x_*^{2\gamma}\rangle}}
+    \exp{\left\{-\frac{(x_*^{\alpha-\beta}-\mu N\langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
+\end{equation}
+where the expectations have to be computed self-consistently, 
+taking also in account normalization constraint.
+
+Let us focus to the case $\alpha=1$, $\beta=1.5$ and $\gamma=1$, with
+$a_s=0$ in order to leverage the equilibrium solution in Eq.~\eqref{eq: dist general}.
+which is tested against simulations in Fig.~\ref{fig: cavity sol.}.
+
+\dots
+
+\subsection{Simulations}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
+        \caption{...}
+        \label{fig: cavity sol.}
+    \end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10.pdf}
+    \caption{...}
+    \label{fig: stability line + sims}
+\end{figure}
+
+\section{Conclusions}
 

### 2022-12-23 src/dynamics.jl

@@ -34,6 +34,10 @@ blowup() = DiscreteCallback((u, t, integrator) -> maximum(u) > MAX_ABUNDANCE, te
 
 function evolve!(p; trajectory=false)
 
+    if !haskey(p, :seed)
+        p[:seed] = rand(UInt32)
+    end
+
     if !haskey(p, :rng)
         p[:rng] = MersenneTwister(p[:seed])
     end

### 2022-12-23 scripts/phase-diagram.jl
alpha-beta phase diagram

@@ -11,37 +11,50 @@ foreach(includet, glob("*.jl", srcdir()))
 
 using ProgressMeter, ThreadsX
 using Plots, LaTeXStrings
+using Contour
+
+function plot_contour_line!(plt, P; label, level = .5)
+    
+    c = Contour.contours(
+    P[:α], 
+    P[:β],
+    Matrix(reshape(ϕ, length(P[:α]), length(P[:β]))'),
+    [level]
+    )
+    
+    for cl in levels(c)
+        for line in lines(cl)
+            plot!(plt, coordinates(line)...,
+            label = label)
+        end
+    end
+    return current()
+
+end
 
 P = Dict{Symbol, Any}(
         :scaled => false,
-        :S => 20,
-        :μ => .01:.1:2.,
-        :σ => .001:.05:.5,
-        :k => .75,
-        :a => 1,
-        :b => 1,
-        :b0 => 10.,
-        :threshold => false,
-        :z => 0.1,
-        :K => 10.,
+        :S => 100,
+        :μ => .1,
+        :σ => .1,
+        :α => 0:.1:1.5,
+        :β => 0:.1:1.5,
+        :γ => 1,
+        :b0 => 1e-2,
+        :threshold => true,
+        :z => 0.,
+        :K => 1000.,
         :order => 2,
         :dist => "normal",
-        :N => 4,
+        :N => 10,
         :symm => false,
     );
 
-begin
+plt = plot()
+for σ ∈ .01:.02:.1
+    P[:σ] = σ
     ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
-
-    sublinear = heatmap(
-        P[:μ], 
-        P[:σ],
-        reshape(ϕ, length(P[:μ]), length(P[:σ]))',
-        dpi = 500,
-        alpha = 1.,
-        grid = false,
-        xlabel = L"\mu",
-        ylabel = L"\sigma",
-        title = L"\textrm{Stability \,\, in \,\, parameter \,\, space}"
-    )
+    plot_contour_line!(plt, P; label = "σ = $(P[:σ])")
 end
+current()
+savefig(plotsdir("phase-diag"))

### 2022-12-23 src/dynamics.jl

@@ -8,18 +8,18 @@ using ForwardDiff
 function F!(f, x, p)
     x .= ppart.(x)
 
-    pop = p[:r] .* x.^p[:k] - p[:z] .* x
+    pop = p[:r] .* x.^p[:α] - p[:z] .* x
     pop[x .< p[:b0]] .= 0
-    comm = - x.^p[:a] .* contract(p[:A], x.^p[:b])
+    comm = - x.^p[:β] .* contract(p[:A], x.^p[:γ])
 
     f .= pop + comm
 end
 
 function F(x, p)
     x .= ppart.(x)
-    pop = p[:r] .* x.^p[:k] - p[:z] .* x
+    pop = p[:r] .* x.^p[:α] - p[:z] .* x
     pop[x .< p[:b0]] .= 0
-    comm = - x.^p[:a] .* contract(p[:A], x.^p[:b])
+    comm = - x.^p[:β] .* contract(p[:A], x.^p[:γ])
 
     return pop + comm
 end
@@ -36,6 +36,10 @@ converged(ϵ = 1e-3) = TerminateSteadyState(ϵ)
 blowup() = DiscreteCallback((u, t, integrator) -> maximum(u) > MAX_ABUNDANCE, terminate!)
 
 function evolve!(p; trajectory=false)
+
+    if !haskey(p, :seed)
+        p[:seed] = rand(UInt32)
+    end
 
     if !haskey(p, :rng)
         p[:rng] = MersenneTwister(p[:seed])
@@ -83,7 +87,7 @@ function equilibria!(p)
     if !haskey(p, :rng) 
         p[:rng] = MersenneTwister(p[:seed])
     end
-    if !haskey(p, :a)
+    if !haskey(p, :β)
         add_interactions!(p)
     end
     if !haskey(p, :r)

### 2023-01-11 papers/general-stability.tex
add DMFT/cavity basics

--- 2022-12-06 papers/general-stability.tex
@@
 \begin{equation}\label{homogeneous-general}
-    \left(\frac{f'_*}{f_*} - \frac{g'_*}{g_*}\right) + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g_*h'_*}{f_*} < 0. 
+    \left(\frac{f'_*}{f_*} - \frac{g'_*}{g_*}\right) + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g_*h'_*}{f_*} < 0 \, . 
 \end{equation}
@@
 
-Consider now \eqref{dynamics} with i.i.d. interactions $a_{ij}$ with mean $\mu$ and standard deviation $\sigma$.
+Consider now \eqref{dynamics} with i.i.d. interactions $a_{ij}$ with mean $\mu$ and standard deviation $\sigma$
+and let us specialize to the power laws case, with $f(x_i)=x_i^{\alpha}$, 
+$g(x_i)=-x_i^{\beta}$ and $h(x_i)=x_i^{\gamma}$. 
+Additional proportionality constants can be reabsorbed in the statistics
+of $a$ and by a rescaling of time, moreover we chose to explicitly
+factor out a minus sign in order to more straightforwardly connect
+with competitive models.
+The evolution equations read
+\begin{equation}
+    \dot{x_i}=x_i^{\alpha} - x_i^{\beta}\sum_{j}a_{ij}x_j^{\gamma} \, .
+\label{eq: full abg}
+\end{equation}
+Following, e.g., Ref.~[Roy2019] we can derive, for large $N$,
+a dynamical mean field theory (DMFT) which provide a stochastic 
+differential equation describing the entire ensemble of $N$ variables
+thorugh the evolution of the representtive random variables $x$
+\begin{equation}
+    \dot{x} = x^{\alpha}-x^{\beta}\left( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\right) \, ,
+\label{eq: dmft}
+\end{equation}
+where $\langle . \rangle$ stands for expectation value
+and $\eta$ is a gaussian variable
+with zero mean and correlation 
+$\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$. 
+The averages and correlations have to be computed self-consistently.
+The equilibrium solution can be written as
+\begin{equation}
+    x^* = \left( \mu N \langle (x^*)^{\gamma}\rangle + \sigma \sqrt{N\langle(x^*)^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, ,
+\end{equation}
+where $\xi$ is a standard normal random variable (i.e. $\xi\sim\mathcal{N}(0,1)$).
+The equilibrium probability distribution function for $x^*$, $P(x)$,
+can be obtained as the pushforward of the distribition of $\xi$.
+The change of measure gives
+\begin{equation}
+    \left|\frac{d \xi}{d x}\right|=\frac{|\alpha-\beta|x^{\alpha-\beta-1}}{\sigma \sqrt{N\langle(x^*)^{2\gamma}\rangle}} \, ,
+\end{equation}
+therefore we obtain
+\begin{equation}
+    P(x)=\frac{|\alpha-\beta|x^{\alpha-\beta-1}}{\sqrt{2\pi}\sigma \sqrt{N\langle(x^*)^{2\gamma}\rangle}}
+    \exp{\left\{-\frac{(x^{\alpha-\beta}-\mu N\langle(x^*)^{\gamma}\rangle)^2}{2\sigma^2N\langle(x^*)^{2\gamma}\rangle}\right\}} \, ,
+\end{equation}
+where the expectations have to be computed self-consistently.
 

### 2023-02-22 papers/general-stability.tex
old changes

--- 2022-12-23 papers/general-stability.tex
@@
 
-\title{Complexity-stability relationships in large, complex systems}
+\title{Beyond May: Complexity-stability relationships in high-dimensional dynamical systems}
 
@@
 May's argument goes as follows.
-If a system with $N$ populations $x_i$ is stable, then it can be characterized by an equilibrium $x^*$.
+A system with $N$ populations $x_i$ is stable is one that can be characterized by an equilibrium point $x^*$.
 Near that equilibrium $x = x^* + \delta x$, the dynamics of the system will be described by linear equations $d(\delta x)/dt = A (\delta x)$, and stability of these equations requires that all eigenvalues of $A$ have negative real part.

### 2023-02-22 scripts/jacobian.jl

@@ -5,41 +5,46 @@ foreach(includet, glob("*.jl", srcdir()))
 using Plots, LaTeXStrings
 
 P = Dict{Symbol, Any}(
-    :scaled => false,
-    :μ => .1,
-    :σ => .001,
-    :k => 1,
-    :a => 1,
-    :b => 1,
-    :b0 => .01,
-    :threshold => false,
-    :z => 0.1,
-    :K => 1.,
-    :order => 2,
-    :dist => "normal",
-    :N => 1,
-    :symm => false,
-    :seed => rand(UInt)
-);
+        :scaled => false,
+        :μ => .1,
+        :σ => .1,
+        :β => 1.,
+        :γ => 1.,
+        :b0 => 1e-5,
+        :threshold => false,
+        :z => 0.,
+        :K => 1.,
+        :order => 2,
+        :dist => "normal",
+        :N => 1,
+        :symm => false,
+    );
 
-P[:S] = 50:20:150
-
-
-plot(legend = :topleft)
-for k in .8:.1:1.2
-        P[:k] = k
-        mean_diag = []
-        mean_offdiag = []
+plot()
+P[:S] = 10:10:50
+for α in .8:.1:1.2
+        P[:α] = α
+        s = []
+        d = []
         for p in expand(P)
             evolve!(p)
-            if p[:converged]
-                j = J(p[:equilibrium], p)
-                push!(mean_diag, mean(diag(j)))
-                push!(mean_offdiag, mean(offdiag(j)))
-                # plot!(offdiag(j); seriestype = :stephist, label = "S = $(p[:S])")    
-            end
+            push!(s, survival(p))
+            push!(d, dominant(p))
         end
-        plot!(mean_diag ./ mean_offdiag, label = "k/a = $(p[:k]/p[:a])")
+        scatter!(s, d)
 end
 current()
 
+
+
+spec = plot(legend = :topleft)
+for α in .8:.1:1.2
+        P[:α] = α
+        for p in expand(P)
+            evolve!(p; trajectory = true)
+            s = spectrum(p)
+            p[:converged] && scatter!(s; label = "α = $(p[:α])", markers = :auto, marker_z = p[:richness])
+        end
+end
+vline!([0]; color = :black, label = false)
+current()

### 2023-02-22 src/dynamics.jl

@@ -1,7 +1,6 @@
 using DifferentialEquations
 using Random, Distributions
 using LinearAlgebra
-using ForwardDiff
 
 #= dynamical system =#
 
@@ -50,8 +49,7 @@ function evolve!(p; trajectory=false)
 
     pb = ODEProblem(
         ODEFunction(
-            (f, x, p, t) -> F!(f, x, p) #in-place F faster
-            # jac=(j, x, p, t) -> J!(j, x, p) #specify jacobian speeds things up
+            (f, x, p, t) -> F!(f, x, p)
         ),
         ones(p[:S]).*p[:x0], #initial condition
         (0.0, MAX_TIME),

### 2023-02-22 papers/general-stability.tex
old changes

--- 2022-12-23 papers/general-stability.tex
@@
 
-Consider now \eqref{dynamics} with i.i.d. interactions $a_{ij}$ with mean $\mu$ and standard deviation $\sigma$.
+With more general interactions coefficients $A = (a_{ij})$, the Jacobian of the system \eqref{dynamics} evaluated at a putative fixed point $x_*$ reads
+\end{eqnarray}
+\begin{equation}
+    J_* = \mathcal{D}(f'_* - f_*g'_*/g_*) + \mathcal{D}(g_*)A\mathcal{D}(h'_*)
+\end{equation}
+Let us now consider the case where $a_{ij}$ are i.i.d. with mean $\mu$ and standard deviation $\sigma$.
 

### 2023-02-22 src/dynamics.jl

@@ -1,15 +1,13 @@
 using DifferentialEquations
 using Random, Distributions
 using LinearAlgebra
-using ForwardDiff
 
 ## dynamical system
 
 function F!(f, x, p)
     x .= ppart.(x)
-
     pop = p[:r] .* x.^p[:α] - p[:z] .* x
-    pop[x .< p[:b0]] .= 0
+    if p[:threshold] pop[x .< p[:b0]] .= 0 end
     comm = - x.^p[:β] .* contract(p[:A], x.^p[:γ])
 
     f .= pop + comm
@@ -18,46 +16,31 @@ end
 function F(x, p)
     x .= ppart.(x)
     pop = p[:r] .* x.^p[:α] - p[:z] .* x
-    pop[x .< p[:b0]] .= 0
+    if p[:threshold] pop[x .< p[:b0]] .= 0 end
     comm = - x.^p[:β] .* contract(p[:A], x.^p[:γ])
 
     return pop + comm
 end
-
-J(x, p) = ForwardDiff.jacobian(y -> F(y, p), x)
 
 ## solving
 
 MAX_TIME = 1e3
 MAX_ABUNDANCE = 1e3
 
-converged(ϵ = 1e-3) = TerminateSteadyState(ϵ)
-
-blowup() = DiscreteCallback((u, t, integrator) -> maximum(u) > MAX_ABUNDANCE, terminate!)
+converged(ϵ = 1e-5) = TerminateSteadyState(ϵ)
+blowup(max_abundance = MAX_ABUNDANCE) = DiscreteCallback((u, t, integrator) -> maximum(u) > max_abundance, terminate!)
 
 function evolve!(p; trajectory=false)
 
-    if !haskey(p, :seed)
-        p[:seed] = rand(UInt32)
-    end
-
-    if !haskey(p, :rng)
-        p[:rng] = MersenneTwister(p[:seed])
-    end
-    if !haskey(p, :A)
-        add_interactions!(p)
-    end
-    if !haskey(p, :r)
-        add_growth_rates!(p)
-    end
-    if !haskey(p, :x0)
-        add_initial_condition!(p)
-    end
+    !haskey(p, :seed) && (p[:seed] = rand(UInt32))
+    !haskey(p, :rng) && (p[:rng] = MersenneTwister(p[:seed]))
+    !haskey(p, :A) && add_interactions!(p)
+    !haskey(p, :r) && add_growth_rates!(p)
[…]

### 2023-02-22 src/richness.jl

@@ -1,0 +1,1 @@
+survival(p) = sum(p[:equilibrium] .> p[:b0])/p[:S]

### 2023-02-22 src/spectrum.jl

@@ -1,0 +1,12 @@
+using ForwardDiff
+J(x, p) = ForwardDiff.jacobian(y -> F(y, p), x)
+
+function spectrum(p)
+    if p[:converged]
+        return eigen(J(p[:equilibrium], p)).values
+    else
+        @warn "Didn't converge!"
+    end
+end
+
+dominant(p) = maximum(real, spectrum(p))

### 2023-03-15 papers/general-stability.tex
not much going on

--- 2023-02-22 papers/general-stability.tex
@@
 Fifty years ago, Robert May predicted that large, complex systems are unlikely to be stable.
-Here, we revisit May's argument to show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems: in some cases, strong interaction and large diversity are, indeed, destabilizing; in other cases, they are stabilizing.
-Which class a given system falls into depends on the relative convexity of the response to self- vs.
+Here, we revisit May's argument to show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems, depending on the relative convexity of the response to self- vs.
 cross-interactions.
@@
 
-In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients: spatial or food-web structure, time delays, etc.
-Here, we show that May's argument is incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
+In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients, such as spatial or food-web structure, time delays, etc.
+Here, we show that May's argument is itself incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
 
@@
 \end{equation}
-Here $f(x_i)$ represents the contribution of a population $i$ to its own growth (its "production function"), while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations.
-That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp.
-$a_{ji} < 0$) implies a positive (resp.
-negative) effect of $j$ on the growth of $i$.
-In the following we call $g(x_i)$ the "response function".
+Here $f(x_i)$ represents the contribution of a population $i$ to its own growth, while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. 
+That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp. $a_{ji} < 0$) implies a positive (resp. negative) effect of $j$ on the growth of $i$. 
+In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''.
 
-The classic generalized Lotka-Volterra competition model corresponds to $f$, $g$, $h$ all linear.
+The classic generalized Lotka-Volterra (GLV) competition model corresponds to $f$, $g$, $h$ all linear. 
 However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$. 
@@
 This is the usual conclusion drawn from the Lotka-Volterra model.
-    \item If $\alpha > \beta$, stability places an upper bound on $N$: the more complex the system, the less likely to be stable. We can call this "May" behavior.
+    \item If $\alpha > \beta$, stability places an upper bound on $N$: the more complex the system, the less likely to be stable. 
+    We can call this "May" behavior.
     \item If $\alpha < \beta$, stability places an lower bound on $N$: the more complex the system, the more likely to be stable. This is "anti-May" behavior.

### 2023-03-15 scripts/single-run.jl

@@ -70,3 +70,5 @@ linecolor = :black,
 grid = false,
 size = (600,200)
 )
+
+p[:diversity]

### 2023-03-15 scripts/stability-diversity.jl

@@ -13,8 +13,9 @@ P = Dict{Symbol, Any}(
         :μ => 0.,
         :σ => .1,
         :k => 1.,
-        :a => 1,
-        :b => 1,
+        :α => 1,
+        :β => 1,
+        :γ => 1,
         :b0 => 1.,
         :threshold => false,
         :z => 0.1,

### 2023-03-15 src/dynamics.jl

@@ -79,7 +79,7 @@ function evolve!(p; trajectory=false)
 end
 
 function add_initial_condition!(p)
-    p[:x0] = rand(p[:rng], Uniform(2, 10), p[:S])
+    p[:b0] = rand(p[:rng], Uniform(2, 10), p[:S])
 end
 
 function add_interactions!(p)

### 2023-03-15 papers/general-stability.tex
not much going on

--- 2023-02-22 papers/general-stability.tex
@@
 
-\title{Beyond May: Complexity-stability relationships in high-dimensional dynamical systems}
+\title{Complexity-stability relationships in large, complex systems}
 
@@
 
-Few mathematical arguments have influenced biological thinking like May's prediction that large, complex systems cannot be stable.
-It is, on the face of it, a perplexing conclusion.
-On the one hand, May's argument is so simple and compelling that it hard not to conclude that it must hold universally.
-On the other, it is clear that at least some large, complex systems are stable---or else which stable patterns would biology be studying in the first place?
+Few mathematical arguments have influenced biological thinking like May's 1970 prediction that large, complex ecosystems cannot be stable. 
+The argument is straightforward and runs as follows. 
+If a system with $N$ populations is stable, then it can be characterized by an equilibrium point. 
+Near that equilibrium, the dynamics of the system will be described by linear equations, and stability of these equations requires that all eigenvalues of the matrix $A$ have negative real part. 
+But if $A$ takes the form $A = B - I$, where $B$ consists of independent random interactions with zero mean and variance $\sigma^2$, and $-I$ corresponds to stabilizing self-interactions on some natural timescale, then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
+This condition places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as ``complexity begets instability''. 
+(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, to incomplete connectivity, or to correlated interactions.)
+
+May's prediction is intriging. On the one hand, his argument is simple and compelling, suggesting general applicability. 
+On the other, it is clear that at least some large, complex systems are stable, else science would be impossible. 
 If anything, the opposite relationship between complexity and stability seems to hold empirically: rich, strongly coupled ecosystems (eg.
@@
 
-May's argument goes as follows.
-A system with $N$ populations $x_i$ is stable is one that can be characterized by an equilibrium point $x^*$.
-Near that equilibrium $x = x^* + \delta x$, the dynamics of the system will be described by linear equations $d(\delta x)/dt = A (\delta x)$, and stability of these equations requires that all eigenvalues of $A$ have negative real part.
-But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions with zero mean and variance $\sigma^2$, and $-I$ corresponds to stabilizing self-interactions on some natural timescale, then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
-This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability".
-(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, to incomplete connectivity, or to correlated interactions.)
-
-In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients: spatial or food-web structure, time delays, etc.
-Here, we show that May's argument is incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
+In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients, such as spatial or food-web structure, time delays, etc.
+Here, we show that May's argument is itself incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
 
@@
 \begin{equation}\label{dynamics}
-    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j).
+    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j) \, .
 \end{equation}
-Here $f(x_i)$ represents the contribution of a population $i$ to its own growth (its "production function"), while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations.
-That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp.
-$a_{ji} < 0$) implies a positive (resp.
-negative) effect of $j$ on the growth of $i$.
-In the following we call $g(x_i)$ the "response function".
+Here $f(x_i)$ represents the contribution of a population $i$ to its own growth, while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. 
+That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp. $a_{ji} < 0$) implies a positive (resp. negative) effect of $j$ on the growth of $i$. 
+In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''.
 
-The classic generalized Lotka-Volterra competition model corresponds to $f$, $g$, $h$ all linear.
+The classic generalized Lotka-Volterra (GLV) competition model corresponds to $f$, $g$, $h$ all linear. 
 However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$. 
@@
 
-Under what condition does \eqref{dynamics} admit a linearly stable equilibrium?
+Under what condition is \eqref{dynamics} linearly stable? 
 We begin by considering the simple case where all self-interactions have the same strength $a_{ii} = a_{\textrm{s}}$, and similarly for cross-interactions $a_{ij} = a_{\textrm{c}}$ ($i\neq j$). 
-Under these assumptions, a straightforward computation shows that an equilibrium $x_*$ will be linearly stable if
+In that case, a straightforward computation shows that an equilibrium $x_*$ will be linearly stable if  
 \begin{equation}\label{homogeneous-general}
-    \left(\frac{f'_*}{f_*} - \frac{g'_*}{g_*}\right) + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g_*h'_*}{f_*} < 0. 
+    \left(\frac{f'_*}{f_*} - \frac{g'_*}{g_*}\right) + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g_*h'_*}{f_*} < 0 \, . 
 \end{equation}
 Thus, the stability of a competitive equilibrium depends on the relative strength of self- and cross-interactions ($a_{\textrm{s}} - a_{\textrm{c}}$), but also on the relative convexity of the production and response functions ($f'_*/f_* - g'_*/g_*$).
-With power laws, \eqref{homogeneous-general} evaluates to
-\begin{equation}
-    (\alpha - \beta)(N-1) < \gamma(a_{\textrm{s}}/a_{\textrm{c}}- 1) - (\alpha - \beta)(a_{\textrm{s}}/a_{\textrm{c}}),
-\end{equation}
-leading to three different regimes:
+With power laws, \eqref{homogeneous-general} evaluates to $(\alpha - \beta)(N-1) < \gamma(a_{\textrm{s}}/a_{\textrm{c}}- 1) - (\alpha - \beta)(a_{\textrm{s}}/a_{\textrm{c}})$, leading to three different regimes:
 \begin{itemize}
@@
 
-With more general interactions coefficients $A = (a_{ij})$, the Jacobian of the system \eqref{dynamics} evaluated at a putative fixed point $x_*$ reads
-\end{eqnarray}
+Consider now \eqref{dynamics} with i.i.d. interactions $a_{ij}$ with mean $\mu$ and standard deviation $\sigma$
+and let us specialize to the power laws case, with $f(x_i)=x_i^{\alpha}$, $g(x_i)=-x_i^{\beta}$ and $h(x_i)=x_i^{\gamma}$. 
+Additional proportionality constants can be reabsorbed in the statistics of $a$ and by a rescaling of time, moreover we chose to explicitly factor out a minus sign in order to more straightforwardly connect with competitive models.
+The evolution equations read
 \begin{equation}
-    J_* = \mathcal{D}(f'_* - f_*g'_*/g_*) + \mathcal{D}(g_*)A\mathcal{D}(h'_*)
+    \dot{x_i}=x_i^{\alpha} - x_i^{\beta}\sum_{j}a_{ij}x_j^{\gamma} \, .
+\label{eq: full abg}
 \end{equation}
-Let us now consider the case where $a_{ij}$ are i.i.d. with mean $\mu$ and standard deviation $\sigma$.
+Following, e.g., Ref.~[Roy2019] we can derive, for large $N$,
+a dynamical mean field theory (DMFT) which provide a stochastic 
+differential equation describing the entire ensemble of $N$ variables
+thorugh the evolution of the representtive random variables $x$
+\begin{equation}
+    \dot{x} = x^{\alpha}-x^{\beta}\left( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\right) \, ,
+\label{eq: dmft}
+\end{equation}
+where $\langle . \rangle$ stands for expectation value
+and $\eta$ is a gaussian variable
+with zero mean and correlation 
+$\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$. 
+The averages and correlations have to be computed self-consistently.
+The equilibrium solution can be written as
+\begin{equation}
+    x^* = \left( \mu N \langle (x^*)^{\gamma}\rangle + \sigma \sqrt{N\langle(x^*)^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, ,
+\end{equation}
+where $\xi$ is a standard normal random variable (i.e. $\xi\sim\mathcal{N}(0,1)$).
+The equilibrium probability distribution function for $x^*$, $P(x)$,
+can be obtained as the pushforward of the distribition of $\xi$.
+The change of measure gives
+\begin{equation}
+    \left|\frac{d \xi}{d x}\right|=\frac{|\alpha-\beta|x^{\alpha-\beta-1}}{\sigma \sqrt{N\langle(x^*)^{2\gamma}\rangle}} \, ,
+\end{equation}
+therefore we obtain
+\begin{equation}
+    P(x)=\frac{|\alpha-\beta|x^{\alpha-\beta-1}}{\sqrt{2\pi}\sigma \sqrt{N\langle(x^*)^{2\gamma}\rangle}}
+    \exp{\left\{-\frac{(x^{\alpha-\beta}-\mu N\langle(x^*)^{\gamma}\rangle)^2}{2\sigma^2N\langle(x^*)^{2\gamma}\rangle}\right\}} \, ,
+\end{equation}
+where the expectations have to be computed self-consistently.
 

### 2023-03-15 scripts/phase-diagram.jl

@@ -12,36 +12,31 @@ foreach(includet, glob("*.jl", srcdir()))
 using ProgressMeter, ThreadsX
 using Plots, LaTeXStrings
 
-P = Dict{Symbol, Any}(
-        :scaled => false,
-        :S => 20,
-        :μ => .01:.1:2.,
-        :σ => .001:.05:.5,
-        :k => .75,
-        :a => 1,
-        :b => 1,
-        :b0 => 10.,
-        :threshold => false,
-        :z => 0.1,
-        :K => 10.,
-        :order => 2,
-        :dist => "normal",
-        :N => 4,
-        :symm => false,
-    );
+P = Dict{Symbol,Any}(
+    :scaled => false,
+    :S => 100,
+    :μ => 0.01,
+    :σ => 0.0001, 
+    :α => 0.5:0.1:1.5, 
+    :β => 0.5:0.1:1.5, 
+    :γ => 1.0, 
+    :b0 => 1.0, 
+    :threshold => false, 
+    :z => 0.0,
+    :K => 0.001,
+    :order => 2,
+    :dist => "normal",
+    :N => 10,
+    :symm => false,
+);
+ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
 
-begin
-    ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
+sublinear = heatmap(P[:α], P[:β], reshape(ϕ, length(P[:α]), length(P[:β]))', dpi=500, alpha=1.0, grid=false,
+    xlabel=L"\alpha",
+    ylabel=L"\beta",
+    title=L"\textrm{Stability \,\, in \,\, parameter \,\, space}"
+)
+plot!(
+    x -> x
+)
 
-    sublinear = heatmap(
-        P[:μ], 
-        P[:σ],
-        reshape(ϕ, length(P[:μ]), length(P[:σ]))',
-        dpi = 500,
-        alpha = 1.,
-        grid = false,
-        xlabel = L"\mu",
-        ylabel = L"\sigma",
-        title = L"\textrm{Stability \,\, in \,\, parameter \,\, space}"
-    )
-end

### 2023-03-15 scripts/single-run.jl

@@ -5,17 +5,17 @@ foreach(includet, glob("*.jl", srcdir()))
 using Plots, LaTeXStrings
 
 p = Dict{Symbol, Any}(
-    :scaled => false,
+    :scaled => false,    
     :S => 50,
-    :μ => 0.,
-    :σ => .1,
-    :k => .75,
-    :a => 1,
-    :b => 1,
+    :μ => 0.1,
+    :σ => 0.05,
+    :α => 1.,
+    :β => 1,
+    :γ => 1,
     :b0 => .01,
     :threshold => true,
-    :z => 0.1,
-    :K => 1.,
+    :z => 0.,
+    :K => 1e4,
     :order => 4,
     :dist => "normal",
     :N => 1,
@@ -33,5 +33,8 @@ plot(
     linewidth = 1,
     legend = false,
     grid = false,
-    palette = :blues
+    palette = :blues,
+    # yaxis = :log
 )
+
+p[:diversity]

### 2023-03-15 src/dynamics.jl

@@ -7,9 +7,8 @@ using ForwardDiff
 
 function F!(f, x, p)
     x .= ppart.(x)
-
     pop = p[:r] .* x.^p[:α] - p[:z] .* x
-    pop[x .< p[:x0]] .= 0
+    pop[x .< p[:b0]] .= 0
     comm = - x.^p[:β] .* contract(p[:A], x.^p[:γ])
 
     f .= pop + comm
@@ -18,7 +17,7 @@ end
 function F(x, p)
     x .= ppart.(x)
     pop = p[:r] .* x.^p[:α] - p[:z] .* x
-    pop[x .< p[:x0]] .= 0
+    pop[x .< p[:b0]] .= 0
     comm = - x.^p[:β] .* contract(p[:A], x.^p[:γ])
 
     return pop + comm
@@ -46,7 +45,7 @@ function evolve!(p; trajectory=false)
     if !haskey(p, :r)
         add_growth_rates!(p)
     end
-    if !haskey(p, :x0)
+    if !haskey(p, :b0)
         add_initial_condition!(p)
     end
 
@@ -66,8 +65,8 @@ function evolve!(p; trajectory=false)
     )
     p[:equilibrium] = sol.retcode == :Terminated ? sol.u[end] : NaN
     p[:converged] = (sol.retcode == :Terminated && maximum(p[:equilibrium]) < MAX_ABUNDANCE)
-    p[:richness] = sum(sol.u[end] .> p[:x0] * p[:threshold])
-    p[:diversity] = p[:richness] == 0 ? 0 : Ω(sol.u[end] .* (sol.u[end] .> p[:x0] * p[:threshold]))
+    p[:richness] = sum(sol.u[end] .> p[:b0] * p[:threshold])
+    p[:diversity] = p[:richness] == 0 ? 0 : Ω(sol.u[end] .* (sol.u[end] .> p[:b0] * p[:threshold]))
 
     if trajectory
         p[:trajectory] = sol
@@ -76,7 +75,7 @@ function evolve!(p; trajectory=false)
 end
 
 function add_initial_condition!(p)
-    p[:x0] = rand(p[:rng], Uniform(2, 10), p[:S])
+    p[:b0] = rand(p[:rng], Uniform(2, 10), p[:S])
 end
 
 function equilibria!(p)
@@ -100,7 +99,7 @@ function equilibria!(p)
                 (f, x, p, t) -> F!(f, x, p); #in-place F faster
[…]

### 2023-03-17 scripts/phase-diagram.jl
small cleanup

@@ -24,7 +24,7 @@ function plot_contour_line!(plt, ϕ, P; level = 0.9, label)
     
     for cl in levels(c)
         # for line in lines(cl)
-            plot!(plt, coordinates(lines(cl)[1])..., lw = 2, ls = :dash, label = label)
+            plot!(plt, coordinates(lines(cl)[1])..., lw = 2, color = :black, label = label)
         # end
     end
     return current()

### 2023-03-17 scripts/phase-diagram.jl
phase digram = alpha, beta plane

@@ -24,7 +24,7 @@ function plot_contour_line!(plt, ϕ, P; level = 0.9, label)
     
     for cl in levels(c)
         # for line in lines(cl)
-            plot!(plt, coordinates(lines(cl)[1])..., lw = 2, color = :black, label = label)
+            plot!(plt, coordinates(lines(cl)[1])..., lw = 2, ls = :dash, label = label)
         # end
     end
     return current()
@@ -33,36 +33,44 @@ end
 
 P = Dict{Symbol, Any}(
         :scaled => false,
-        :S => 20,
-        :μ => .01:.1:2.,
-        :σ => .001:.05:.5,
-        :k => .75,
-        :a => 1,
-        :b => 1,
-        :b0 => 10.,
-        :threshold => false,
-        :z => 0.1,
-        :K => 10.,
-        :order => 2,
+        :S => 100,
+        :μ => 1e-2,
+        # :μₛ => 1e-2,
+        :σ => 1e-4,
+        :α => 0.1:.1:1.5,
+        :β => 0.1:.1:1.5,
+        :γ => 1.,
+        :extinction_threshold => 1e-4,
+        :z => 0.,
         :dist => "normal",
-        :N => 4,
-        :symm => false,
+        :N => 10,
     );
 
-begin
-    ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
+ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
+plot()
 
-    sublinear = heatmap(
-        P[:μ], 
-        P[:σ],
-        reshape(ϕ, length(P[:μ]), length(P[:σ]))',
-        dpi = 500,
-        alpha = 1.,
-        grid = false,
-        xlabel = L"\mu",
-        ylabel = L"\sigma",
-        title = L"\textrm{Stability \,\, in \,\, parameter \,\, space}"
-    )
+heatmap(
+    P[:α],
+    P[:β],
+    Matrix(reshape(ϕ, length(P[:α]), length(P[:β]))'),
+    xlabel = L"\alpha",
+    ylabel = L"\beta"
+)
+plot!(x -> x)
+
+P[:α] = 1
+P[:β] = .6
+
+evolve!(P; trajectory = true)
+plot(P[:trajectory], legend = false, palette = :blues)
+
+plt = plot(x->x, xlims = (0, 1), color = :black, lw = 2)
[…]

### 2023-03-17 run-sims.sh
slurm script

@@ -1,0 +1,20 @@
+#!/bin/sh
+#SBATCH --job-name=phase-diagram
+#SBATCH --time=2-00:00:00  
+#SBATCH --mail-user=matteo.smerlak@gmail.com --mail-type=ALL
+#SBATCH --clusters=mesopsl1
+#SBATCH --partition=def
+#SBATCH --qos=mesopsl1_def_long
+#SBATCH --account=msmerlak
+#SBATCH --nodes=12 
+#SBATCH --ntasks-per-node=16
+
+module load ma_liste_de_modules
+
+RUNDIR=/travail/msmerlak/job.${SLURM_JOB_ID}
+mkdir -p ${RUNDIR}
+cd ${RUNDIR}
+
+mpirun -np ${SLURM_NTASKS} /travail/msmerlak/general-stability/scripts/test-slurm.jl
+
+exit 0

### 2023-03-17 scripts/test-slurm.jl
test script

@@ -1,0 +1,17 @@
+using Pkg
+Pkg.add("DrWatson")
+
+using DrWatson
+@quickactivate
+Pkg.instantiate()
+
+using Glob, Revise
+foreach(includet, glob("*.jl", srcdir()))
+
+using ProgressMeter, ThreadsX
+using Plots, LaTeXStrings
+gr(label = false)
+using Contour
+
+plot(x -> x^2)
+savefig(plotsdir("x2"))

### 2023-03-21 scripts/alpha-beta.jl
rename alpha beta

@@ -108,3 +108,14 @@ end
 plot(B, σₘ, 
 lw = 4,
 lc = :black)
+
+plt = plot(x->x, xlims = (0, 1), color = :black, lw = 2)
+xlabel!(L"\alpha")
+ylabel!(L"\beta")
+for σ ∈ (1e-3, 5e-3, 1e-2) 
+    P[:σ] = σ
+    ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
+    plot_contour_line!(plt, ϕ, P; label = "σ = $(P[:σ])")
+end
+current()
+savefig(plotsdir("phase-diag"))

### 2023-03-21 scripts/phase-diagram.jl

@@ -1,70 +1,0 @@
-using DrWatson, Glob, Revise
-@quickactivate
-foreach(includet, glob("*.jl", srcdir()))
-
-using ProgressMeter, Suppressor, ThreadsX
-using Plots, LaTeXStrings
-
-p = Dict{Symbol, Any}(
-        :scaled => true,
-        :S => 50,
-        :μ => .01:.1:1.21,
-        :μₛ => true,
-        :σ => .01:.1:1.21,
-        :α => .5,
-        :β => 3.,
-        :γ => 1,
-        :x0 => .1,
-        :threshold => false,
-        :dist => "normal",
-        :N => 1,
-        :symm => false,
-        :seed => rand(UInt)
-    );
-
-#begin
-    ϕ = ThreadsX.collect(diversity!(q) for q in expand(p));
-
-    sublinear = heatmap(
-        p[:μ], 
-        p[:σ],
-        reshape(ϕ, length(p[:μ]), length(p[:σ]))',
-        dpi = 500,
-        alpha = 1.,
-        grid = false,
-        xlabel = L"\mu N",
-        ylabel = L"\sigma \sqrt{N}",
-        c=cgrad(:Greys, rev=true),
-        ylims=[.0,.4],
-        xlims=[.0,1.2],
-    )
-#end
-
-#= critical line for gaussian approximation =#
-    μ_critical, σ_critical = critical_line_gauss(p, μ_range=(.01:.25:1.26))
-    plot!(μ_critical, σ_critical,
-    labels = false,
-    linewidth = 4,
-    linecolor = :red,
-    grid = false,
-    ylims=[.0,1.2],
-    xlims=[.0,1.2],
-    yticks = [0, 0.6, 1.2],
-    xticks = [0, 0.6, 1.2],
-    size = (600,400),
-    )
-
-#= critical line =#
-    μ_critical, σ_critical = critical_line(p, μ_range=(.01:.2:1.25),
-    n_max = 1e1)
-    plot!(μ_critical, σ_critical,
-    labels = false,
-    linewidth = 4,
-    linecolor = :green,
-    grid = false,
-    ylims=[.0,1.2],
-    xlims=[.0,1.2],
-    yticks = [0, 0.6, 1.2],
-    xticks = [0, 0.6, 1.2],
-    size = (600,400),
-    )

### 2023-03-21 scripts/alpha-beta.jl
rename alpha beta

@@ -14,21 +14,18 @@ gr(label = false)
 using Contour
 
 function plot_contour_line!(plt, ϕ, P; level = 0.9, label)
-
     c = Contour.contours(
     P[:α], 
     P[:β],
     Matrix(reshape(ϕ, length(P[:α]), length(P[:β]))'),
     [level]
     )
-    
     for cl in levels(c)
         # for line in lines(cl)
             plot!(plt, coordinates(lines(cl)[1])..., lw = 2, ls = :dash, label = label)
         # end
     end
     return current()
-
 end
 
 P = Dict{Symbol, Any}(
@@ -45,24 +42,6 @@ P = Dict{Symbol, Any}(
         :dist => "normal",
         :N => 10,
     );
-
-ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
-plot()
-
-heatmap(
-    P[:α],
-    P[:β],
-    Matrix(reshape(ϕ, length(P[:α]), length(P[:β]))'),
-    xlabel = L"\alpha",
-    ylabel = L"\beta"
-)
-plot!(x -> x)
-
-P[:α] = 1
-P[:β] = .6
-
-evolve!(P; trajectory = true)
-plot(P[:trajectory], legend = false, palette = :blues)
 
 plt = plot(x->x, xlims = (0, 1), color = :black, lw = 2)
 xlabel!(L"\alpha")

### 2023-03-21 scripts/alpha-beta.jl
minor

@@ -52,4 +52,4 @@ for σ ∈ (1e-3, 5e-3, 1e-2)
     plot_contour_line!(plt, ϕ, P; label = "σ = $(P[:σ])")
 end
 current()
-savefig(plotsdir("phase-diag"))
+savefig(plotsdir("alpha-beta"))

### 2023-03-21 scripts/alpha-beta.jl
minor

@@ -118,4 +118,4 @@ for σ ∈ (1e-3, 5e-3, 1e-2)
     plot_contour_line!(plt, ϕ, P; label = "σ = $(P[:σ])")
 end
 current()
-savefig(plotsdir("phase-diag"))
+savefig(plotsdir("alpha-beta"))

### 2023-03-21 scripts/ahmadian.jl
begin ahmadian

@@ -1,0 +1,10 @@
+using LinearAlgebra
+
+function ahmadian(μ, σ)
+    N = size(A.M, 1)
+    X = inv(A.L)
+    Y = inv(A.R)
+    M(z) = X*(z*I-A.M)*Y
+    τ(x, y) = tr(M(x + y*im)*M(x + y*im)')
+    plot(eigen(A.M + A.L * ))
+end

### 2023-03-21 papers/general-stability.tex
update draft

--- 2023-03-15 papers/general-stability.tex
@@
 \usepackage{colortbl}
-\usepackage{graphicx} 
-\usepackage{booktabs}
-\usepackage{caption}
 \newcommand{\red}{\textcolor{red}}
@@
 Fifty years ago, Robert May predicted that large, complex systems are unlikely to be stable.
-Here, we revisit May's argument to show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems, depending on the relative convexity of the response to self- vs.
+Here, we revisit May's argument to show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems: in some cases, strong interaction and large diversity are, indeed, destabilizing; in other cases, they are stabilizing.
+Which class a given system falls into depends on the relative convexity of the response to self- vs.
 cross-interactions.
@@
 
-Few mathematical arguments have influenced biological thinking like May's prediction that large, complex systems cannot be stable.
-It is, on the face of it, a perplexing conclusion.
-On the one hand, May's argument is so simple and compelling that it hard not to conclude that it must hold universally.
+Few mathematical arguments have influenced biological thinking like May's 1970 prediction that large, complex systems cannot be stable.
+It is, indeed, a perplexing prediction.
+On the one hand, the argument is so simple and compelling that it is hard not to conclude that it must hold universally.
 On the other, it is clear that at least some large, complex systems are stable---or else which stable patterns would biology be studying in the first place?
@@
 
-May's argument goes as follows.
-A system with $N$ populations $x_i$ is stable is one that can be characterized by an equilibrium point $x^*$.
-Near that equilibrium $x = x^* + \delta x$, the dynamics of the system will be described by linear equations $d(\delta x)/dt = A (\delta x)$, and stability of these equations requires that all eigenvalues of $A$ have negative real part.
+May's argument can be paraphrased as follows.
+If a system with $N$ populations $x_i$ is stable, then it can be characterized by an equilibrium $x^*$.
+Near that equilibrium $x = x^* + y$, the dynamics of the system will be described by linear equations $\delta\dot{y} = A y$, and stability of these equations requires that all eigenvalues of $A$ have negative real part.
 But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions with zero mean and variance $\sigma^2$, and $-I$ corresponds to stabilizing self-interactions on some natural timescale, then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
@@
 
-In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients, such as spatial or food-web structure, time delays, etc.
-Here, we show that May's argument is itself incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
+In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients: spatial or food-web structure, time delays, etc.
+Here, we show that May's argument is incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
 
@@
 \end{equation}
-Here $f(x_i)$ represents the contribution of a population $i$ to its own growth, while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. 
-That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp. $a_{ji} < 0$) implies a positive (resp. negative) effect of $j$ on the growth of $i$. 
-In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''.
+Here $f(x_i)$ represents the contribution of a population $i$ to its own growth (its "production function"), while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations.
+That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp.
+$a_{ji} < 0$) implies a positive (resp.
+negative) effect of $j$ on the growth of $i$.
+In the following we call $g(x_i)$ the "response function".
 
-The classic generalized Lotka-Volterra (GLV) competition model corresponds to $f$, $g$, $h$ all linear. 
+The classic generalized Lotka-Volterra competition model corresponds to $f$, $g$, $h$ all linear.
 However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
@@
 
-Under what condition does \eqref{dynamics} admit a linearly stable equilibrium?
+Under what condition is \eqref{dynamics} linearly stable?
 We begin by considering the simple case where all self-interactions have the same strength $a_{ii} = a_{\textrm{s}}$, and similarly for cross-interactions $a_{ij} = a_{\textrm{c}}$ ($i\neq j$).
-Under these assumptions, a straightforward computation shows that an equilibrium $x_*$ will be linearly stable if
+In that case, a straightforward computation shows that an equilibrium $x_*$ will be linearly stable if
 \begin{equation}\label{homogeneous-general}
-    \left(\frac{f'_*}{f_*} - \frac{g'_*}{g_*}\right) + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g_*h'_*}{f_*} < 0. 
+    \left(\frac{f'_*}{f_*} - \frac{g'_*}{g_*}\right) + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g_*h'_*}{f_*} < 0 \, . 
 \end{equation}
 Thus, the stability of a competitive equilibrium depends on the relative strength of self- and cross-interactions ($a_{\textrm{s}} - a_{\textrm{c}}$), but also on the relative convexity of the production and response functions ($f'_*/f_* - g'_*/g_*$).
-With power laws, \eqref{homogeneous-general} evaluates to
-\begin{equation}
-    (\alpha - \beta)(N-1) < \gamma(a_{\textrm{s}}/a_{\textrm{c}}- 1) - (\alpha - \beta)(a_{\textrm{s}}/a_{\textrm{c}}),
-\end{equation}
-leading to three different regimes:
+With power laws, \eqref{homogeneous-general} evaluates to $(\alpha - \beta)(N-1) < \gamma(a_{\textrm{s}}/a_{\textrm{c}}- 1) - (\alpha - \beta)(a_{\textrm{s}}/a_{\textrm{c}})$, leading to three different regimes:
 \begin{itemize}
@@
 This is the usual conclusion drawn from the Lotka-Volterra model.
-    \item If $\alpha > \beta$, stability places an upper bound on $N$: the more complex the system, the less likely to be stable. 
-    We can call this "May" behavior.
+    \item If $\alpha > \beta$, stability places an upper bound on $N$: the more complex the system, the less likely to be stable. We can call this "May" behavior.
     \item If $\alpha < \beta$, stability places an lower bound on $N$: the more complex the system, the more likely to be stable. This is "anti-May" behavior.
@@
 It reduces indeed to the homogenous case for $\sigma\to 0$.
-\red{Discuss better.}
+\red{To be discussed.}
 
@@
 \end{equation}
+\red{There are things to clarify and discuss.}
 This expression allows us to obtain insights on the stability properties of
@@
 The change of measure
-$\left|d \xi/d x_*\right|=|\alpha-\beta|x_*^{\alpha-\beta-1}/\sigma \sqrt{N\langle x_*^{2\gamma}\rangle}$,
+\begin{equation}
+    \left|\frac{d \xi}{d x_*}\right|=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sigma \sqrt{N\langle x_*^{2\gamma}\rangle}} \, ,
+\end{equation}
 leads to
@@
 
-Let us focus to the case $\alpha=1$, $\beta=1.5$ and $\gamma=1$, with
+Let us focus to the case $\alpha=1$, $\beta=2$ and $\gamma=1$, with
 $a_s=0$ in order to leverage the equilibrium solution in Eq.~\eqref{eq: dist general}.
-which is tested against simulations in Fig.~\ref{fig: cavity sol.}.
 
@@
 
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
-        \caption{...}
-        \label{fig: cavity sol.}
-    \end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10.pdf}
-    \caption{...}
-    \label{fig: stability line + sims}
-\end{figure}
+\begin{itemize}
+    \item Equilibrium distribution and histogram for "anti-May" example
+    \item ($\mu N$, $\sigma\sqrt{N}$) plane for "May" and "anti-May" examples
+        with stability lines.
+    \item ($\alpha$, $\beta$) plane with homogeneous interactions line.
+    \item Others?
+\end{itemize}
 

### 2023-03-21 src/cavity.jl
update cavity

@@ -1,5 +1,3 @@
-#!!! DA AGGIUSTARE PER CASO GENERAL !!!#
-
 using StatsBase
 using Random, Distributions, Statistics
 using DifferentialEquations
@@ -9,43 +7,34 @@ using HCubature
 using NLsolve
 using SpecialFunctions
 
-# Expression for density distribution with gaussian approximation (γ=1)
+#= expression for density distribution with gaussian approximation (γ=1) =#
 function P_gauss(p)
     @unpack μ, σ, α, β, S = p
-    μ = !p[:scaled] ? μ*S : μ
-    σ = !p[:scaled] ? σ*sqrt(S) : σ 
+    μ = p[:scaled] ? μ : μ*S
+    σ = p[:scaled] ? σ : σ*sqrt(S) 
     e1 = μ^(1/(α-β-1))
     e2 = e1^2/(1-(1/(α-β))^2*(μ*e1)^(2*(1-(α-β))/(α-β))*σ^2)
     #ϕ = e2>0 ? (1+erf((e1)/sqrt(2*(e2-e1^2))))/2 : 1
     return e2>0 ? Normal(e1, sqrt(e2-e1^2)) : Normal(e1, 0) #(e1, e2>0 ? e2 : 0, ϕ, e2>0 ? Normal(e1,sqrt(e2-e1^2)) : Normal(e1,0))
 end
 
-# Expression for density distribution in sublinear case
+#= expression for density distribution (γ=1) =#
 function P_n(n, e1_n, e2_n, p)
-    @unpack μ, σ, k, S = p
-    if !p[:scaled] μ = μ*S end
-    if !p[:scaled] σ = σ*sqrt(S) end
-    if !haskey(p, :dist_r) # unique r
-        r = haskey(p, :r) ? first(p[:r]) : 1
-        return (1-k)*n^(k-2)/(sqrt(2*π*σ^2*e2_n/r^2))*
-        exp(-(n^(k-1)-μ*e1_n/r)^2/(2*σ^2*e2_n/r^2))
-    else # uniformely distributed r
-        a, b = p[:dist_r].a, p[:dist_r].b
-        return (1-k)*n^(k-2)/(sqrt(2*π*e2_n*σ^2)*(b-a)*2*n^(2*k-2))*
-        sqrt(2*π*e2_n*σ^2)*((exp(-(a*n^(k-1)-μ*e1_n)^2/(2*σ^2*e2_n))-
[…]

### 2023-03-21 scripts/phase-diagram.jl
update phase-diagram

@@ -1,15 +1,8 @@
-using Pkg
-Pkg.add("DrWatson")
-
-using DrWatson
+using DrWatson, Glob, Revise
 @quickactivate
+foreach(include, glob("*.jl", srcdir()))
 
-Pkg.instantiate()
-
-using Glob, Revise
-foreach(includet, glob("*.jl", srcdir()))
-
-using ProgressMeter, ThreadsX
+using ProgressMeter, Suppressor, ThreadsX
 using Plots, LaTeXStrings
 
 P = Dict{Symbol, Any}(
@@ -30,14 +23,39 @@ begin
     ϕ = ThreadsX.collect(stability!(p) for p in expand(P));
 
     sublinear = heatmap(
-        P[:μ], 
-        P[:σ],
-        reshape(ϕ, length(P[:μ]), length(P[:σ]))',
+        p[:μ], 
+        p[:σ],
+        reshape(ϕ, length(p[:μ]), length(p[:σ]))',
         dpi = 500,
         alpha = 1.,
         grid = false,
-        xlabel = L"\mu",
-        ylabel = L"\sigma",
-        title = L"\textrm{Stability \,\, in \,\, parameter \,\, space}"
+        xlabel = L"\mu N",
+        ylabel = L"\sigma \sqrt{N}",
+        c=cgrad(:Reds, rev=true),
+        ylims=[.0,.5],
+        xlims=[.0,1.25],
     )
-end
+#end
+
+#= critical line for gaussian approximation =#
+    μ_critical, σ_critical = critical_line_gauss(p, μ_range=(.01:.25:1.26))
+    plot!(μ_critical, σ_critical,
+    labels = false,
+    linewidth = 4,
+    linecolor = :black,
+    grid = false,
+    ylims=[.0,.5],
+    xlims=[.01,1.25]
+    )
+
+#= critical line =#
+    μ_critical, σ_critical = critical_line(p, μ_range=(.05:.2:1.25),
+    n_max = 1e2)
+    plot!(μ_critical, σ_critical,
+    labels = false,
+    linewidth = 4,
+    linecolor = :black,
+    grid = false,
+    ylims=[.0,.5],
+    xlims=[.01,1.25]
+    )

### 2023-03-21 scripts/phase-diagram.jl
update phase-diagram

@@ -1,47 +1,64 @@
-using Pkg
-Pkg.add("DrWatson")
-
-using DrWatson
+using DrWatson, Glob, Revise
 @quickactivate
+foreach(include, glob("*.jl", srcdir()))
 
-Pkg.instantiate()
-
-using Glob, Revise
-foreach(includet, glob("*.jl", srcdir()))
-
-using ProgressMeter, ThreadsX
+using ProgressMeter, Suppressor, ThreadsX
 using Plots, LaTeXStrings
 
-P = Dict{Symbol, Any}(
-        :scaled => false,
-        :S => 20,
-        :μ => .01:.1:2.,
-        :σ => .001:.05:.5,
-        :k => .75,
-        :a => 1,
-        :b => 1,
-        :b0 => 10.,
+p = Dict{Symbol, Any}(
+        :scaled => true,
+        :S => 50,
+        :μ => .01:.01:1.25,
+        :μₛ => 0,
+        :σ => .01:.01:.5,
+        :α => 1,
+        :β => 1.5,
+        :γ => 1,
+        :x0 => .1,
         :threshold => false,
-        :z => 0.1,
-        :K => 10.,
-        :order => 2,
         :dist => "normal",
-        :N => 4,
+        :N => 1,
         :symm => false,
+        :seed => rand(UInt)
     );
 
-begin
-    ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
+#begin
+    ϕ = ThreadsX.collect(diversity!(q) for q in expand(p));
 
     sublinear = heatmap(
-        P[:μ], 
-        P[:σ],
-        reshape(ϕ, length(P[:μ]), length(P[:σ]))',
+        p[:μ], 
+        p[:σ],
+        reshape(ϕ, length(p[:μ]), length(p[:σ]))',
         dpi = 500,
         alpha = 1.,
         grid = false,
-        xlabel = L"\mu",
-        ylabel = L"\sigma",
-        title = L"\textrm{Stability \,\, in \,\, parameter \,\, space}"
+        xlabel = L"\mu N",
+        ylabel = L"\sigma \sqrt{N}",
+        c=cgrad(:Reds, rev=true),
+        ylims=[.0,.5],
+        xlims=[.0,1.25],
     )
-end
+#end
+
+#= critical line for gaussian approximation =#
+    μ_critical, σ_critical = critical_line_gauss(p, μ_range=(.01:.25:1.26))
[…]

### 2023-03-23 scripts/ahmadian.jl
reworking

@@ -1,10 +1,0 @@
-using LinearAlgebra
-
-function ahmadian(μ, σ)
-    N = size(A.M, 1)
-    X = inv(A.L)
-    Y = inv(A.R)
-    M(z) = X*(z*I-A.M)*Y
-    τ(x, y) = tr(M(x + y*im)*M(x + y*im)')
-    plot(eigen(A.M + A.L * ))
-end

### 2023-03-23 scripts/single-run.jl

@@ -21,8 +21,6 @@ p = Dict{Symbol, Any}(
     :seed => rand(UInt)
 );
 
-@time diversity!(p)
-
 evolve!(p; trajectory = true);
 
 p[:converged]

### 2023-03-23 src/constants.jl

@@ -1,13 +1,0 @@
-using Colors, ColorSchemes
-
-COLOR_LOG = :turquoise3
-COLOR_SUB = :darkgoldenrod1
-
-COLOR_LOG49 = "#009af9"
-COLOR_SUB49 = "#fb8d00"
-
-COLOR_LOG35 = "#0070b2"
-COLOR_SUB35 = "#b26300" 
-
-K = 100.
-CUTOFF = 1.

### 2023-03-23 src/dynamics.jl

@@ -1,23 +1,23 @@
+using DrWatson; @quickactivate
+include(srcdir("utils.jl"))
+
 using DifferentialEquations
 using Random, Distributions
 using LinearAlgebra
 
-#= dynamical system =#
+## dynamical system
+function F(x, p)
+    x .= ppart.(x)
+    pop = x.^p[:α]
+    comm = - x.^p[:β] .* (p[:A] * x.^p[:γ])
+    return pop + comm
+end
 
 function F!(f, x, p)
     x .= ppart.(x)
     pop = x.^p[:α]
-    comm = - x.^p[:β] .* contract(p[:A], x.^p[:γ])
-
+    comm = - x.^p[:β] .* (p[:A] * x.^p[:γ])
     f .= pop + comm
-end
-
-function F(x, p)
-    x .= ppart.(x)
-    pop = x.^p[:α]
-    comm = - x.^p[:β] .* contract(p[:A], x.^p[:γ])
-
-    return pop + comm
 end
 
 J(x, p) = ForwardDiff.jacobian(y -> F(y, p), x)
@@ -27,9 +27,8 @@ J(x, p) = ForwardDiff.jacobian(y -> F(y, p), x)
 MAX_TIME = 1e5
 MAX_ABUNDANCE = 1e5
 
-converged(ϵ = 1e-5) = TerminateSteadyState(ϵ)
-
-blowup() = DiscreteCallback((u, t, integrator) -> maximum(u) > MAX_ABUNDANCE, terminate!)
+converged(ϵ = 1e-4) = TerminateSteadyState(ϵ)
+blowup(max_abundance = MAX_ABUNDANCE) = DiscreteCallback((u, t, integrator) -> maximum(u) > max_abundance, terminate!)
 
 function evolve!(p; trajectory=false)
 
@@ -49,20 +48,19 @@ function evolve!(p; trajectory=false)
 
     pb = ODEProblem(
         ODEFunction(
-            (f, x, p, t) -> F!(f, x, p)
+            (f, x, p, t) -> F!(f, x, p) 
         ),
-        ones(p[:S]).*p[:x0], #initial condition
+        p[:x0], 
         (0.0, MAX_TIME),
         p
     )
 
     sol = solve(pb,
         callback=CallbackSet(converged(), blowup()),
-        save_on=trajectory #don't save whole trajectory, only endpoint
+        save_on=trajectory 
     )
-
[…]

### 2023-03-23 src/spectrum.jl

@@ -1,12 +1,0 @@
-using ForwardDiff
-J(x, p) = ForwardDiff.jacobian(y -> F(y, p), x)
-
-function spectrum(p)
-    if p[:converged]
-        return eigen(J(p[:equilibrium], p)).values
-    else
-        @warn "Didn't converge!"
-    end
-end
-
-dominant(p) = maximum(real, spectrum(p))

### 2023-03-23 src/utils.jl

@@ -1,14 +1,7 @@
 ppart(x) = max(x, 0)
 
-using TensorOperations
-contract(A::Array{Float64, 2}, x::Vector) = @tensor y[i] := A[i, j] * x[j]
-contract(A::Array{Float64, 3}, x::Vector) = @tensor y[i] := A[i, j, k] * x[j] * x[k]
-contract(A::Array{Float64, 4}, x::Vector) = @tensor y[i] := A[i, j, k, l] * x[j] * x[k] * x[l]
-
-
-function offdiag(A::AbstractMatrix)
-    [A[ι] for ι in CartesianIndices(A) if ι[1] ≠ ι[2]]
-end 
+normal(m, s) = Normal(m, s)
+gamma(m, s) = Gamma(m^2 / s^2, s^2 / m)
 
 function expand(c::AbstractDict)
 

### 2023-03-23 tests/jacobians.jl

@@ -1,0 +1,23 @@
+using Test
+using DrWatson, Glob, Revise
+@quickactivate
+foreach(includet, glob("*.jl", srcdir()))
+
+p = Dict{Symbol, Any}(
+    :scaled => false,    
+    :S => 10,
+    :μ => 1e-3,
+    :μₛ => 1e-4,
+    :σ => 0.05,
+    :α => .5,
+    :β => 1.,
+    :γ => 1.,
+    :extinction_threshold => 1e-2,
+    :z => 0.,
+    :dist => "gamma",
+    :N => 1,
+    :seed => rand(UInt)
+);
+
+evolve!(p; trajectory = false);
+@test J_analytical(p) ≈ J_numerical(p) atol=1e-3

### 2023-03-23 src/ahmadian.jl
reworking

@@ -1,0 +1,77 @@
+using DrWatson; @quickactivate
+
+using ForwardDiff
+J_numerical(p) = ForwardDiff.jacobian(y -> F(y, p), p[:equilibrium])
+
+d(f) = x -> ForwardDiff.derivative(f, x)
+function J_analytical(p)
+    f(x) = x^p[:α]
+    g(x) = - x^p[:β]
+    h(x) = x^p[:γ]
+    @assert p[:converged]
+
+    X = p[:equilibrium]
+    L = @. g(X)
+    R = @. d(h)(X)
+    M = @. d(f)(X) - d(g)(X)*f(X)/g(X) + (p[:μₛ] - p[:μ])*g(X)*d(h)(X)
+    S = p[:A] - (p[:μₛ] - p[:μ])*I
+
+    return diagm(M) + diagm(L)*S*diagm(R)
+end
+
+using LinearAlgebra
+function spectrum(p)
+    if p[:converged]
+        return eigen(J_numerical(p)).values
+    else
+        @warn "Didn't converge!"
+    end
+end
+dominant(p) = maximum(real, spectrum(p))
+
+function ahmadian(p)
+    f(x) = x^p[:α]
+    g(x) = - x^p[:β]
+    h(x) = x^p[:γ]
+    @assert p[:converged]
+    X = p[:equilibrium]
+    L = @. g(X)
+    R = @. d(h)(X)
+    M = @. d(f)(X) - d(g)(X)*f(X)/g(X) + (p[:μₛ] - p[:μ])*g(X)*d(h)(X)
+    S = p[:A] - (p[:μₛ] - p[:μ])*I
+    
+    Σ = spectrum(p)
+    τ(x, y) = sum(@. (L*R)^2/abs(x + y*im - M)^2 )
+    X = range(minimum(real.(Σ)), maximum(real.(Σ)), length = 100)
+    Y = range(minimum(imag.(Σ)), maximum(imag.(Σ)), length = 100)
+    contour(1.5*X, 1.5*Y; levels = [1/var(S), 2/var(S)], lw = 2)
+    
+    scatter!(spectrum(p); alpha = .2, label = false)
+end
+
+include(srcdir("dynamics.jl"))
+p = Dict{Symbol, Any}(
+    :scaled => false,    
+    :S => 50,
+    :μ => 1e-2,
+    :μₛ => 1e-3,
+    :σ => 0.005,
+    :α => .7,
+    :β => 1.,
+    :γ => 1.,
[…]

### 2023-03-23 src/dynamics.jl

@@ -1,23 +1,23 @@
+using DrWatson; @quickactivate
+include(srcdir("utils.jl"))
+
 using DifferentialEquations
 using Random, Distributions
 using LinearAlgebra
 
 ## dynamical system
+function F(x, p)
+    x .= ppart.(x)
+    pop = x.^p[:α]
+    comm = - x.^p[:β] .* (p[:A] * x.^p[:γ])
+    return pop + comm
+end
 
 function F!(f, x, p)
     x .= ppart.(x)
-    pop = p[:r] .* x.^p[:α] - p[:z] .* x
-    comm = - x.^p[:β] .* contract(p[:A], x.^p[:γ])
-
+    pop = x.^p[:α]
+    comm = - x.^p[:β] .* (p[:A] * x.^p[:γ])
     f .= pop + comm
-end
-
-function F(x, p)
-    x .= ppart.(x)
-    pop = p[:r] .* x.^p[:α] - p[:z] .* x
-    comm = - x.^p[:β] .* contract(p[:A], x.^p[:γ])
-
-    return pop + comm
 end
 
 ## solving
@@ -25,7 +25,7 @@ end
 MAX_TIME = 1e3
 MAX_ABUNDANCE = 1e3
 
-converged(ϵ = 1e-5) = TerminateSteadyState(ϵ)
+converged(ϵ = 1e-4) = TerminateSteadyState(ϵ)
 blowup(max_abundance = MAX_ABUNDANCE) = DiscreteCallback((u, t, integrator) -> maximum(u) > max_abundance, terminate!)
 
 function evolve!(p; trajectory=false)
@@ -38,28 +38,25 @@ function evolve!(p; trajectory=false)
     if !haskey(p, :A)
         add_interactions!(p)
     end
-    if !haskey(p, :r)
-        add_growth_rates!(p)
-    end
     if !haskey(p, :x0)
         add_initial_condition!(p)
     end
 
     pb = ODEProblem(
         ODEFunction(
-            (f, x, p, t) -> F!(f, x, p) #in-place 
+            (f, x, p, t) -> F!(f, x, p) 
         ),
-        p[:x0], #initial condition
+        p[:x0], 
         (0.0, MAX_TIME),
         p
     )
 
     sol = solve(pb,
         callback=CallbackSet(converged(), blowup()),
-        save_on=trajectory #don't save whole trajectory, only endpoint
+        save_on=trajectory 
     )
[…]

### 2023-03-23 src/richness.jl

@@ -1,1 +1,0 @@
-survival(p) = sum(p[:equilibrium] .> p[:b0])/p[:S]

### 2023-03-23 tests/jacobians.jl

@@ -1,25 +1,23 @@
+using Test
 using DrWatson, Glob, Revise
 @quickactivate
 foreach(includet, glob("*.jl", srcdir()))
 
-using Plots, LaTeXStrings
-
 p = Dict{Symbol, Any}(
-    :scaled => true,
-    :S => 100,
-    :μ => 0.5,
-    :μₛ => 0.5,
-    :σ => 9.,
-    :α => .7,
+    :scaled => false,    
+    :S => 10,
+    :μ => 1e-3,
+    :μₛ => 1e-4,
+    :σ => 0.05,
+    :α => .5,
     :β => 1.,
     :γ => 1.,
     :extinction_threshold => 1e-2,
+    :z => 0.,
     :dist => "gamma",
     :N => 1,
     :seed => rand(UInt)
 );
 
-evolve!(p; trajectory = true);
-plot(p[:trajectory], label = false)
-p[:converged]
-p[:A]
+evolve!(p; trajectory = false);
+@test J_analytical(p) ≈ J_numerical(p) atol=1e-3

### 2023-03-23 src/dynamics.jl
una mattina

@@ -25,7 +25,7 @@ J(x, p) = ForwardDiff.jacobian(y -> F(y, p), x)
 #= solving =#
 
 MAX_TIME = 1e5
-MAX_ABUNDANCE = 1e5
+MAX_ABUNDANCE = 1e3
 
 converged(ϵ = 1e-4) = TerminateSteadyState(ϵ)
 blowup(max_abundance = MAX_ABUNDANCE) = DiscreteCallback((u, t, integrator) -> maximum(u) > max_abundance, terminate!)
@@ -62,13 +62,11 @@ function evolve!(p; trajectory=false)
     p[:equilibrium] = sol.retcode == SciMLBase.ReturnCode.Terminated ? sol.u[end] : NaN
     p[:converged] = sol.retcode == SciMLBase.ReturnCode.Terminated && maximum(p[:equilibrium]) < MAX_ABUNDANCE
 
-    survivors = sol.u[end] .> p[:threshold]
-
+    survivors = sol.u[end] .> p[:extinction_threshold]
     p[:richness] = sum(survivors)
-    p[:diversity] = p[:richness] == 0 ? 0 : Ω(sol.u[end] .* (sol.u[end] .> p[:threshold]))
 
-#=  q = p[:equilibrium]/sum(p[:equilibrium])
-    p[:diversity] = 1/(p[:S] * sum(q.^2))  =#
+    q = p[:equilibrium]/sum(p[:equilibrium])
+    p[:diversity] = 1/(p[:S] * sum(q.^2))
 
     if trajectory
         p[:trajectory] = sol
@@ -91,13 +89,12 @@ function add_interactions!(p)
     end
     
     A = rand(p[:rng], dist, (p[:S], p[:S]))
-    @show A
 
     if haskey(p, :symm) && p[:symm]
         A = (A + A')/2
     end
     
-    A[diagind(A)] .= p[:μₛ] 
+    A[diagind(A)] .= p[:scaled] ? p[:μₛ] / p[:S] : p[:μₛ]
     p[:A] = A
 end
 

### 2023-03-23 src/metrics.jl

@@ -5,7 +5,6 @@ function diversity!(p)
 
     for i in 1:p[:N]
         add_interactions!(p)
-        add_growth_rates!(p)
         add_initial_condition!(p)
         evolve!(p)
         
@@ -23,7 +22,6 @@ function stability!(p)
 
     for i in 1:p[:N]
         add_interactions!(p)
-        add_growth_rates!(p)
         add_initial_condition!(p)
         evolve!(p)
         stability[i] = p[:converged] && p[:richness] == p[:S] ? 1 : 0

### 2023-03-23 scripts/single-run.jl
una mattina

@@ -9,11 +9,11 @@ p = Dict{Symbol, Any}(
     :S => 100,
     :μ => 0.5,
     :μₛ => 0.5,
-    :σ => 9.,
-    :α => .7,
+    :σ => .05,
+    :α => .5,
     :β => 1.,
     :γ => 1.,
-    :extinction_threshold => 1e-2,
+    :extinction_threshold => 1e-4,
     :dist => "gamma",
     :N => 1,
     :seed => rand(UInt)
@@ -22,4 +22,8 @@ p = Dict{Symbol, Any}(
 evolve!(p; trajectory = true);
 plot(p[:trajectory], label = false)
 p[:converged]
-p[:A]
+p[:equilibrium]
+
+
+p[:N] = 10
+stability!(p)

### 2023-03-23 src/dynamics.jl

@@ -22,7 +22,7 @@ end
 
 ## solving
 
-MAX_TIME = 1e3
+MAX_TIME = 1e5
 MAX_ABUNDANCE = 1e3
 
 converged(ϵ = 1e-4) = TerminateSteadyState(ϵ)
@@ -59,8 +59,8 @@ function evolve!(p; trajectory=false)
     p[:converged] = sol.retcode == SciMLBase.ReturnCode.Terminated && maximum(p[:equilibrium]) < MAX_ABUNDANCE
 
     survivors = sol.u[end] .> p[:extinction_threshold]
-    
     p[:richness] = sum(survivors)
+
     q = p[:equilibrium]/sum(p[:equilibrium])
     p[:diversity] = 1/(p[:S] * sum(q.^2))
 
@@ -85,13 +85,12 @@ function add_interactions!(p)
     end
     
     A = rand(p[:rng], dist, (p[:S], p[:S]))
-    @show A
 
     if haskey(p, :symm) && p[:symm]
         A = (A + A')/2
     end
     
-    A[diagind(A)] .= p[:μₛ] 
+    A[diagind(A)] .= p[:scaled] ? p[:μₛ] / p[:S] : p[:μₛ]
     p[:A] = A
 end
 

### 2023-03-23 src/metrics.jl

@@ -1,4 +1,3 @@
-
 function diversity!(p)
     # run N simulates and append results to p
     diversity = Vector{Float64}(undef, p[:N])
@@ -6,7 +5,6 @@ function diversity!(p)
 
     for i in 1:p[:N]
         add_interactions!(p)
-        add_growth_rates!(p)
         add_initial_condition!(p)
         evolve!(p)
         
@@ -17,19 +15,17 @@ function diversity!(p)
     return mean(diversity)
 end
 
-function full_coexistence!(p)
+function stability!(p)
     # run N simulates and append results to p
-    full_coexistence = Vector{Float64}(undef, p[:N])
+    stability = Vector{Float64}(undef, p[:N])
     p[:rng] = MersenneTwister()
 
     for i in 1:p[:N]
         add_interactions!(p)
-        add_growth_rates!(p)
         add_initial_condition!(p)
         evolve!(p)
-        # full_coexistence[i] = p[:converged] && p[:richness] == p[:S] ? 1 : 0
-        full_coexistence[i] = p[:richness] == p[:S] ? 1 : 0
+        stability[i] = p[:converged] && p[:richness] == p[:S] ? 1 : 0
     end
 
-    return mean(full_coexistence)
+    return mean(stability)
 end

### 2023-03-23 src/metrics.jl
metrics

@@ -1,4 +1,3 @@
-
 function diversity!(p)
     # run N simulates and append results to p
     diversity = Vector{Float64}(undef, p[:N])
@@ -8,28 +7,25 @@ function diversity!(p)
         add_interactions!(p)
         add_initial_condition!(p)
         evolve!(p)
-       
-        diversity[i] = p[:diversity]
         
-    #=     q = p[:equilibrium]/sum(p[:equilibrium])
-        diversity[i] = 1/(sum(q.^2) * p[:S]) =#
+        q = p[:equilibrium]/sum(p[:equilibrium])
+        diversity[i] = 1/(sum(q.^2) * p[:S])
     end
 
     return mean(diversity)
 end
 
-function full_coexistence!(p)
+function stability!(p)
     # run N simulates and append results to p
-    full_coexistence = Vector{Float64}(undef, p[:N])
+    stability = Vector{Float64}(undef, p[:N])
     p[:rng] = MersenneTwister()
 
     for i in 1:p[:N]
         add_interactions!(p)
         add_initial_condition!(p)
         evolve!(p)
-        # full_coexistence[i] = p[:converged] && p[:richness] == p[:S] ? 1 : 0
-        full_coexistence[i] = p[:richness] == p[:S] ? 1 : 0
+        stability[i] = p[:converged] && p[:richness] == p[:S] ? 1 : 0
     end
 
-    return mean(full_coexistence)
+    return mean(stability)
 end

### 2023-03-23 src/dynamics.jl
phase diagram

@@ -80,6 +80,8 @@ end
 
 function add_interactions!(p)
     (m, s) = p[:scaled] ? (p[:μ] / p[:S], p[:σ] / sqrt(p[:S])) : (p[:μ], p[:σ])
+
+
     if p[:dist] == "normal"
         dist = normal(m, s)
     elseif p[:dist] == "uniform"
@@ -93,7 +95,8 @@ function add_interactions!(p)
     if haskey(p, :symm) && p[:symm]
         A = (A + A')/2
     end
-    
+
+    !haskey(p, :μₛ) && (p[:μₛ] = p[:μ])
     A[diagind(A)] .= p[:scaled] ? p[:μₛ] / p[:S] : p[:μₛ]
     p[:A] = A
 end

### 2023-03-23 scripts/phase-diagram.jl
phase diagram

@@ -1,0 +1,43 @@
+using Pkg
+Pkg.add("DrWatson")
+
+using DrWatson
+@quickactivate
+
+Pkg.instantiate()
+
+using Glob, Revise
+foreach(includet, glob("*.jl", srcdir()))
+
+using ProgressMeter, ThreadsX
+using Plots, LaTeXStrings
+
+P = Dict{Symbol, Any}(
+    :scaled => true,
+    :S => 100,
+    :μ => .01:.1:1.,
+    :σ => .01:.1:1.,
+    :α => .5,
+    :β => 1.,
+    :γ => 1.,
+    :extinction_threshold => 1e-4,
+    :dist => "gamma",
+    :N => 5,
+    :seed => rand(UInt)
+);
+
+begin
+    ϕ = ThreadsX.collect(stability!(p) for p in expand(P));
+
+    sublinear = heatmap(
+        P[:μ], 
+        P[:σ],
+        reshape(ϕ, length(P[:μ]), length(P[:σ]))',
+        dpi = 500,
+        alpha = 1.,
+        grid = false,
+        xlabel = L"\mu",
+        ylabel = L"\sigma",
+        title = L"\textrm{Stability \,\, in \,\, parameter \,\, space}"
+    )
+end

### 2023-03-23 scripts/single-run.jl

@@ -8,7 +8,6 @@ p = Dict{Symbol, Any}(
     :scaled => true,
     :S => 100,
     :μ => 0.5,
-    :μₛ => 0.5,
     :σ => .05,
     :α => .5,
     :β => 1.,

### 2023-03-23 src/dynamics.jl

@@ -76,6 +76,8 @@ end
 
 function add_interactions!(p)
     (m, s) = p[:scaled] ? (p[:μ] / p[:S], p[:σ] / sqrt(p[:S])) : (p[:μ], p[:σ])
+
+
     if p[:dist] == "normal"
         dist = normal(m, s)
     elseif p[:dist] == "uniform"
@@ -89,7 +91,8 @@ function add_interactions!(p)
     if haskey(p, :symm) && p[:symm]
         A = (A + A')/2
     end
-    
+
+    !haskey(p, :μₛ) && (p[:μₛ] = p[:μ])
     A[diagind(A)] .= p[:scaled] ? p[:μₛ] / p[:S] : p[:μₛ]
     p[:A] = A
 end

### 2023-03-23 papers/general-stability.tex
reworking
update draft

@@
 \usepackage{colortbl}
+\usepackage{graphicx} 
+\usepackage{booktabs}
+\usepackage{caption}
 \newcommand{\red}{\textcolor{red}}
@@
 It reduces indeed to the homogenous case for $\sigma\to 0$.
-\red{To be discussed.}
+\red{Discuss better.}
 
@@
 \end{equation}
-\red{There are things to clarify and discuss.}
 This expression allows us to obtain insights on the stability properties of
@@
 The change of measure
-\begin{equation}
-    \left|\frac{d \xi}{d x_*}\right|=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sigma \sqrt{N\langle x_*^{2\gamma}\rangle}} \, ,
-\end{equation}
+$\left|d \xi/d x_*\right|=|\alpha-\beta|x_*^{\alpha-\beta-1}/\sigma \sqrt{N\langle x_*^{2\gamma}\rangle}$,
 leads to
@@
 
-Let us focus to the case $\alpha=1$, $\beta=2$ and $\gamma=1$, with
+Let us focus to the case $\alpha=1$, $\beta=1.5$ and $\gamma=1$, with
 $a_s=0$ in order to leverage the equilibrium solution in Eq.~\eqref{eq: dist general}.
+which is tested against simulations in Fig.~\ref{fig: cavity sol.}.
 
@@
 
-\begin{itemize}
-    \item Equilibrium distribution and histogram for "anti-May" example
-    \item ($\mu N$, $\sigma\sqrt{N}$) plane for "May" and "anti-May" examples
-        with stability lines.
-    \item ($\alpha$, $\beta$) plane with homogeneous interactions line.
-    \item Others?
-\end{itemize}
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
+        \caption{...}
+        \label{fig: cavity sol.}
+    \end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10.pdf}
+    \caption{...}
+    \label{fig: stability line + sims}
+\end{figure}
 

### 2023-03-23 scripts/alpha-beta.jl
add and modify alpha-beta

@@ -1,0 +1,110 @@
+using DrWatson, Glob, Revise
+@quickactivate
+foreach(includet, glob("*.jl", srcdir()))
+
+using ProgressMeter, ThreadsX
+using Plots, LaTeXStrings
+gr(label = false, grid = false)
+using Contour
+
+function plot_contour_line!(plt, ϕ, P; level = 0.5, label)
+    c = Contour.contours(
+    P[:α], 
+    P[:β],
+    Matrix(reshape(ϕ, length(P[:α]), length(P[:β]))'),
+    [level]
+    )
+    for cl in levels(c)
+        # for line in lines(cl)
+            plot!(plt, coordinates(lines(cl)[1])..., lw = 2, ls = :dash, label = label)
+        # end
+    end
+    return current()
+end
+
+P = Dict{Symbol, Any}(
+        :scaled => false,
+        :S => 100,
+        :μ => 1e-2,
+        :μₛ => true,
+        :σ => 1e-4,
+        :α => 0.25:.25:2.5,
+        :β => 0.25:.25:2.5,
+        :γ => 1.,
+        :threshold => false,
+        :dist => "normal",
+        :N => 1,
+    );
+
+plt = plot(x->x, xlims = (0, 2.5), color = :black, lw = 2)
+xlabel!(L"\alpha")
+ylabel!(L"\beta")
+for σ ∈ (1e-3, 5e-3, 1e-2) 
+    P[:σ] = σ
+    ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
+    plot_contour_line!(plt, ϕ, P; label = "σ = $(P[:σ])")
+end
+current()
+#savefig(plotsdir("alpha-beta"))
+
+#= alternative  =#
+P = Dict{Symbol, Any}(
+        :scaled => true,
+        :S => 50,
+        :μ => 1.,
+        :μₛ => true,
+        :σ => .1,
+        :α => 1.,
+        :β => 1.25,
+        :γ => 1.,
+        :threshold => false,
+        :dist => "normal",
+        :N => 1,
+    );
+
+plt = plot(x->x, xlims = (0, 2), color = :black, lw = 4,     label = latexstring("σ_c = 0"))
+xlabel!(L"\alpha")
+ylabel!(L"\beta")
+for β ∈ 1.5:.5:2.5 
+    P[:β] = β
[…]

### 2023-03-23 src/metrics.jl
add metrics

@@ -1,0 +1,35 @@
+
+function diversity!(p)
+    # run N simulates and append results to p
+    diversity = Vector{Float64}(undef, p[:N])
+    p[:rng] = MersenneTwister() # no seed here
+
+    for i in 1:p[:N]
+        add_interactions!(p)
+        add_initial_condition!(p)
+        evolve!(p)
+       
+        diversity[i] = p[:diversity]
+        
+    #=     q = p[:equilibrium]/sum(p[:equilibrium])
+        diversity[i] = 1/(sum(q.^2) * p[:S]) =#
+    end
+
+    return mean(diversity)
+end
+
+function full_coexistence!(p)
+    # run N simulates and append results to p
+    full_coexistence = Vector{Float64}(undef, p[:N])
+    p[:rng] = MersenneTwister()
+
+    for i in 1:p[:N]
+        add_interactions!(p)
+        add_initial_condition!(p)
+        evolve!(p)
+        # full_coexistence[i] = p[:converged] && p[:richness] == p[:S] ? 1 : 0
+        full_coexistence[i] = p[:richness] == p[:S] ? 1 : 0
+    end
+
+    return mean(full_coexistence)
+end

### 2023-03-23 src/cavity.jl
remove normalization

@@ -44,11 +44,11 @@ function Cavity(p;
     e2_n = e2_init*ones(Iter+1) #<n²>
 
     for i in 1:Iter
-        ϕ_new = first(quadgk(x -> P_n(x, e1_n[i], e2_n[i], p) , n_min, n_max, rtol=tol))
-        e1_new = first(quadgk(x -> x*P_n(x, e1_n[i], e2_n[i], p)/ϕ_n[i] , n_min, n_max, rtol=tol))
-        e2_new = first(quadgk(x -> x*x*P_n(x, e1_n[i], e2_n[i], p)/ϕ_n[i] , n_min, n_max, rtol=tol))
+    #    ϕ_new = first(quadgk(x -> P_n(x, e1_n[i], e2_n[i], p) , n_min, n_max, rtol=tol))
+        e1_new = first(quadgk(x -> x*P_n(x, e1_n[i], e2_n[i], p) , n_min, n_max, rtol=tol))
+        e2_new = first(quadgk(x -> x*x*P_n(x, e1_n[i], e2_n[i], p) , n_min, n_max, rtol=tol))
 
-        ϕ_n[i+1] = (1-rela)*ϕ_n[i] + rela*ϕ_new
+    #    ϕ_n[i+1] = (1-rela)*ϕ_n[i] + rela*ϕ_new
         e1_n[i+1] = (1-rela)*e1_n[i] + rela*e1_new
         e2_n[i+1] = (1-rela)*e2_n[i] + rela*e2_new
     end
@@ -74,14 +74,14 @@ function σ_crit(p;
     σc = σc_init*ones(Iter+1) 
     for i in 1:Iter
         p[:σ] = σc[i]
-        ϕ_new = first(quadgk(x -> P_n(x, e1_n[i], e2_n[i], p) , n_min, n_max, rtol=tol))
-        e1_new = first(quadgk(x -> x*P_n(x, e1_n[i], e2_n[i], p)/ϕ_n[i] , n_min, n_max, rtol=tol))
-        e2_new = first(quadgk(x -> x*x*P_n(x, e1_n[i], e2_n[i], p)/ϕ_n[i] , n_min, n_max, rtol=tol))
+#        ϕ_new = first(quadgk(x -> P_n(x, e1_n[i], e2_n[i], p) , n_min, n_max, rtol=tol))
[…]

### 2023-03-23 scripts/phase-diagram.jl
updates

@@ -1,6 +1,6 @@
 using DrWatson, Glob, Revise
 @quickactivate
-foreach(include, glob("*.jl", srcdir()))
+foreach(includet, glob("*.jl", srcdir()))
 
 using ProgressMeter, Suppressor, ThreadsX
 using Plots, LaTeXStrings
@@ -8,11 +8,11 @@ using Plots, LaTeXStrings
 p = Dict{Symbol, Any}(
         :scaled => true,
         :S => 50,
-        :μ => .01:.01:1.25,
-        :μₛ => 0,
-        :σ => .01:.01:.5,
-        :α => 1,
-        :β => 1.5,
+        :μ => .01:.1:1.21,
+        :μₛ => true,
+        :σ => .01:.1:1.21,
+        :α => .5,
+        :β => 3.,
         :γ => 1,
         :x0 => .1,
         :threshold => false,
@@ -34,9 +34,9 @@ p = Dict{Symbol, Any}(
         grid = false,
         xlabel = L"\mu N",
         ylabel = L"\sigma \sqrt{N}",
-        c=cgrad(:Reds, rev=true),
-        ylims=[.0,.5],
-        xlims=[.0,1.25],
+        c=cgrad(:Greys, rev=true),
+        ylims=[.0,.4],
+        xlims=[.0,1.2],
     )
 #end
 
@@ -45,20 +45,26 @@ p = Dict{Symbol, Any}(
     plot!(μ_critical, σ_critical,
     labels = false,
     linewidth = 4,
-    linecolor = :black,
+    linecolor = :red,
     grid = false,
-    ylims=[.0,.5],
-    xlims=[.01,1.25]
+    ylims=[.0,1.2],
+    xlims=[.0,1.2],
+    yticks = [0, 0.6, 1.2],
+    xticks = [0, 0.6, 1.2],
+    size = (600,400),
     )
 
 #= critical line =#
-    μ_critical, σ_critical = critical_line(p, μ_range=(.05:.2:1.25),
-    n_max = 1e2)
+    μ_critical, σ_critical = critical_line(p, μ_range=(.01:.2:1.25),
+    n_max = 1e1)
     plot!(μ_critical, σ_critical,
     labels = false,
     linewidth = 4,
-    linecolor = :black,
+    linecolor = :green,
     grid = false,
-    ylims=[.0,.5],
-    xlims=[.01,1.25]
+    ylims=[.0,1.2],
[…]

### 2023-03-23 scripts/single-run.jl

@@ -5,19 +5,17 @@ foreach(includet, glob("*.jl", srcdir()))
 using Plots, LaTeXStrings
 
 p = Dict{Symbol, Any}(
-    :scaled => false,
-    :S => 50,
-    :μ => 0.,
+    :scaled => true,
+    :S => 100,
+    :μ => 1.,
+    :μₛ => false,
     :σ => .1,
-    :k => .75,
-    :a => 1,
-    :b => 1,
-    :b0 => .01,
-    :threshold => true,
-    :z => 0.1,
-    :K => 1.,
-    :order => 4,
-    :dist => "normal",
+    :α => 1,
+    :β => 1.5,
+    :γ => 1,
+    :x0 => .1,
+    :threshold => false,
+    :dist => "gamma",
     :N => 1,
     :symm => false,
     :seed => rand(UInt)
@@ -25,13 +23,50 @@ p = Dict{Symbol, Any}(
 
 evolve!(p; trajectory = true);
 
-#trajectories
+p[:converged]
+
+#= trajectories =#
 plot(
     p[:trajectory],
-    ylabel = L"B_i(t)",
+    ylabel = L"x_i(t)",
     xlabel = L"t",
+    yscale = :log,
     linewidth = 1,
     legend = false,
     grid = false,
-    palette = :blues
+)
+
+#= density distribution =#
+p_eq = p[:trajectory].u[end]
+histogram(p_eq,
+    normalize = true,
+    alpha = .5,
+    color = :green,
+    ylabel = L"P(x_*)",
+    xlabel = L"x_*",
+    label = false,
+    grid = false,
+)
+
+#= cavity solution with gaussian approximation =#
+X=[x for x in .9*minimum(p_eq):.0001:1.1*maximum(p_eq)]
+plot!(X,[pdf(P_gauss(p), x) for x in X],
+labels = false, #"cavity - gaussian",
+linewidth = 2,
+alpha = 1,
+linecolor = :black,
+linestyle = :dash,
+grid = false
+)
+
+a, b, c = Cavity(p)
+
+X=[x for x in .9*minimum(p_eq):.01:1.1*maximum(p_eq)]
+plot!(X,[P_n(x, b, c, p) for x in X],
+labels = false, #"cavity - gaussian",
+linewidth = 2,
+alpha = 1,
+linecolor = :black,
+grid = false,
+size = (600,200)
 )

### 2023-03-23 src/dynamics.jl

@@ -3,13 +3,11 @@ using Random, Distributions
 using LinearAlgebra
 using ForwardDiff
 
-## dynamical system
+#= dynamical system =#
 
 function F!(f, x, p)
     x .= ppart.(x)
-
-    pop = p[:r] .* x.^p[:α] - p[:z] .* x
-    pop[x .< p[:x0]] .= 0
+    pop = x.^p[:α]
     comm = - x.^p[:β] .* contract(p[:A], x.^p[:γ])
 
     f .= pop + comm
@@ -17,8 +15,7 @@ end
 
 function F(x, p)
     x .= ppart.(x)
-    pop = p[:r] .* x.^p[:α] - p[:z] .* x
-    pop[x .< p[:x0]] .= 0
+    pop = x.^p[:α]
     comm = - x.^p[:β] .* contract(p[:A], x.^p[:γ])
 
     return pop + comm
@@ -26,12 +23,12 @@ end
 
 J(x, p) = ForwardDiff.jacobian(y -> F(y, p), x)
 
-## solving
+#= solving =#
 
-MAX_TIME = 1e3
-MAX_ABUNDANCE = 1e3
+MAX_TIME = 1e5
+MAX_ABUNDANCE = 1e5
 
-converged(ϵ = 1e-3) = TerminateSteadyState(ϵ)
+converged(ϵ = 1e-5) = TerminateSteadyState(ϵ)
 
 blowup() = DiscreteCallback((u, t, integrator) -> maximum(u) > MAX_ABUNDANCE, terminate!)
 
@@ -43,9 +40,6 @@ function evolve!(p; trajectory=false)
     if !haskey(p, :A)
         add_interactions!(p)
     end
-    if !haskey(p, :r)
-        add_growth_rates!(p)
-    end
     if !haskey(p, :x0)
         add_initial_condition!(p)
     end
@@ -55,7 +49,7 @@ function evolve!(p; trajectory=false)
             (f, x, p, t) -> F!(f, x, p) #in-place F faster
             # jac=(j, x, p, t) -> J!(j, x, p) #specify jacobian speeds things up
         ),
-        ones(p[:S]), #initial condition
+        ones(p[:S]).*p[:x0], #initial condition
         (0.0, MAX_TIME),
         p
     )
@@ -64,10 +58,17 @@ function evolve!(p; trajectory=false)
         callback=CallbackSet(converged(), blowup()),
[…]

### 2023-03-23 scripts/phase-diagram.jl
updates

@@ -1,6 +1,6 @@
 using DrWatson, Glob, Revise
 @quickactivate
-foreach(include, glob("*.jl", srcdir()))
+foreach(includet, glob("*.jl", srcdir()))
 
 using ProgressMeter, Suppressor, ThreadsX
 using Plots, LaTeXStrings
@@ -31,9 +31,9 @@ begin
         grid = false,
         xlabel = L"\mu N",
         ylabel = L"\sigma \sqrt{N}",
-        c=cgrad(:Reds, rev=true),
-        ylims=[.0,.5],
-        xlims=[.0,1.25],
+        c=cgrad(:Greys, rev=true),
+        ylims=[.0,.4],
+        xlims=[.0,1.2],
     )
 #end
 
@@ -42,20 +42,26 @@ begin
     plot!(μ_critical, σ_critical,
     labels = false,
     linewidth = 4,
-    linecolor = :black,
+    linecolor = :red,
     grid = false,
-    ylims=[.0,.5],
-    xlims=[.01,1.25]
+    ylims=[.0,1.2],
+    xlims=[.0,1.2],
+    yticks = [0, 0.6, 1.2],
+    xticks = [0, 0.6, 1.2],
+    size = (600,400),
     )
 
 #= critical line =#
-    μ_critical, σ_critical = critical_line(p, μ_range=(.05:.2:1.25),
-    n_max = 1e2)
+    μ_critical, σ_critical = critical_line(p, μ_range=(.01:.2:1.25),
+    n_max = 1e1)
     plot!(μ_critical, σ_critical,
     labels = false,
     linewidth = 4,
-    linecolor = :black,
+    linecolor = :green,
     grid = false,
-    ylims=[.0,.5],
-    xlims=[.01,1.25]
+    ylims=[.0,1.2],
+    xlims=[.0,1.2],
+    yticks = [0, 0.6, 1.2],
+    xticks = [0, 0.6, 1.2],
+    size = (600,400),
     )

### 2023-03-23 scripts/single-run.jl

@@ -70,3 +70,51 @@ linecolor = :black,
 grid = false,
 size = (600,200)
 )
+
+p[:converged]
+
+#= trajectories =#
+plot(
+    p[:trajectory],
+    ylabel = L"x_i(t)",
+    xlabel = L"t",
+    yscale = :log,
+    linewidth = 1,
+    legend = false,
+    grid = false,
+)
+
+#= density distribution =#
+p_eq = p[:trajectory].u[end]
+histogram(p_eq,
+    normalize = true,
+    alpha = .5,
+    color = :green,
+    ylabel = L"P(x_*)",
+    xlabel = L"x_*",
+    label = false,
+    grid = false,
+)
+
+#= cavity solution with gaussian approximation =#
+X=[x for x in .9*minimum(p_eq):.0001:1.1*maximum(p_eq)]
+plot!(X,[pdf(P_gauss(p), x) for x in X],
+labels = false, #"cavity - gaussian",
+linewidth = 2,
+alpha = 1,
+linecolor = :black,
+linestyle = :dash,
+grid = false
+)
+
+a, b, c = Cavity(p)
+
+X=[x for x in .9*minimum(p_eq):.01:1.1*maximum(p_eq)]
+plot!(X,[P_n(x, b, c, p) for x in X],
+labels = false, #"cavity - gaussian",
+linewidth = 2,
+alpha = 1,
+linecolor = :black,
+grid = false,
+size = (600,200)
+)

### 2023-03-23 src/dynamics.jl

@@ -20,8 +20,6 @@ function F!(f, x, p)
     f .= pop + comm
 end
 
-J(x, p) = ForwardDiff.jacobian(y -> F(y, p), x)
-
 #= solving =#
 
 MAX_TIME = 1e5

### 2023-03-23 src/dynamics.jl
prova

@@ -99,3 +99,5 @@ function add_interactions!(p)
     p[:A] = A
 end
 
+
+prova

### 2023-03-23 scripts/alpha-beta.jl
minor

@@ -1,14 +1,10 @@
-using Pkg
-Pkg.add("DrWatson")
-
 using DrWatson
 @quickactivate
-Pkg.instantiate()
 
 using Glob, Revise
 foreach(includet, glob("*.jl", srcdir()))
 
-using ProgressMeter, ThreadsX
+using ThreadsX
 using Plots, LaTeXStrings
 gr(label = false)
 using Contour
@@ -29,26 +25,34 @@ function plot_contour_line!(plt, ϕ, P; level = 0.9, label)
 end
 
 P = Dict{Symbol, Any}(
-        :scaled => false,
-        :S => 100,
-        :μ => 1e-2,
-        # :μₛ => 1e-2,
-        :σ => 1e-4,
+        :scaled => true,
+        :S => 10,
+        :μ => .5,
+        :σ => .7,
         :α => 0.1:.1:1.5,
         :β => 0.1:.1:1.5,
         :γ => 1.,
         :extinction_threshold => 1e-4,
-        :z => 0.,
-        :dist => "normal",
-        :N => 10,
+        :dist => "gamma",
+        :N => 5,
     );
+
+ϕ = ThreadsX.collect(stability!(p) for p in expand(P));
+contour(P[:α], 
+P[:β],
+    Matrix(reshape(ϕ, length(P[:α]), length(P[:β]))')
+    )
+
+
+plt = plot()    
+plot_contour_line!(plt, ϕ, P; label = "σ = $(P[:σ])")
 
 plt = plot(x->x, xlims = (0, 1), color = :black, lw = 2)
 xlabel!(L"\alpha")
 ylabel!(L"\beta")
-for σ ∈ (1e-3, 5e-3, 1e-2) 
+for σ ∈ (0.01, .3, .5) 
     P[:σ] = σ
-    ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
+    ϕ = ThreadsX.collect(stability!(p) for p in expand(P));
     plot_contour_line!(plt, ϕ, P; label = "σ = $(P[:σ])")
 end
 current()

### 2023-03-23 scripts/phase-diagram.jl

@@ -9,7 +9,7 @@ Pkg.instantiate()
 using Glob, Revise
 foreach(includet, glob("*.jl", srcdir()))
 
-using ProgressMeter, ThreadsX
+using ThreadsX
 using Plots, LaTeXStrings
 
 P = Dict{Symbol, Any}(

### 2023-03-23 src/dynamics.jl

@@ -22,7 +22,7 @@ end
 
 ## solving
 
-MAX_TIME = 1e5
+MAX_TIME = 1e3
 MAX_ABUNDANCE = 1e3
 
 converged(ϵ = 1e-4) = TerminateSteadyState(ϵ)

### 2023-03-23 src/dynamics.jl
clean dynamics

@@ -98,6 +98,3 @@ function add_interactions!(p)
     A[diagind(A)] .= p[:scaled] ? p[:μₛ] / p[:S] : p[:μₛ]
     p[:A] = A
 end
-
-
-prova

### 2023-03-23 papers/general-stability.tex

@@
 
-May's argument can be paraphrased as follows.
-If a system with $N$ populations $x_i$ is stable, then it can be characterized by an equilibrium $x^*$.
-Near that equilibrium $x = x^* + y$, the dynamics of the system will be described by linear equations $\delta\dot{y} = A y$, and stability of these equations requires that all eigenvalues of $A$ have negative real part.
-But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions with zero mean and variance $\sigma^2$, and $-I$ corresponds to stabilizing self-interactions on some natural timescale, then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
-This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability".
-(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, to incomplete connectivity, or to correlated interactions.)
+May's prediction is intriging. On the one hand, his argument is simple and compelling, suggesting general applicability. 
+On the other, it is clear that at least some large, complex systems are stable, else science would be impossible. 
+If anything, the opposite relationship between complexity and stability seems to hold empirically: rich, strongly coupled ecosystems (eg.
+rainforests) tend to be stable over time, while sparser ones (eg.
+arctic environments) often exhibit large fluctuations, exctinctions, and invasions.
 
-In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients: spatial or food-web structure, time delays, etc.
-Here, we show that May's argument is incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
+In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients, such as spatial or food-web structure, time delays, etc.
+Here, we show that May's argument is itself incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
 

### 2023-03-23 scripts/alpha-beta.jl
let's time alpha-beta

@@ -1,121 +1,31 @@
-using DrWatson, Glob, Revise
+using DrWatson
 @quickactivate
+
+using Glob, Revise
 foreach(includet, glob("*.jl", srcdir()))
 
-using ProgressMeter, ThreadsX
 using Plots, LaTeXStrings
 gr(label = false, grid = false)
-using Contour
 
-function plot_contour_line!(plt, ϕ, P; level = 0.5, label)
-    c = Contour.contours(
-    P[:α], 
-    P[:β],
-    Matrix(reshape(ϕ, length(P[:α]), length(P[:β]))'),
-    [level]
-    )
-    for cl in levels(c)
-        # for line in lines(cl)
-            plot!(plt, coordinates(lines(cl)[1])..., lw = 2, ls = :dash, label = label)
-        # end
+function σ_critical(α, β = 1)
+    p = Dict{Symbol, Any}(
+        :scaled => true,
+        :S => 50,
+        :μ => .5,
+        :α => α,
+        :β => β,
+        :γ => 1.,
+        :extinction_threshold => 1e-4,
+        :dist => "gamma",
+        :N => 1,
+    );
+
+    for σ in 0.01:.1:5.
+        p[:σ] = σ
+        s = stability!(p)
+        s < .9 && return σ
     end
-    return current()
 end
-
-P = Dict{Symbol, Any}(
-        :scaled => false,
-        :S => 100,
-        :μ => 1e-2,
-        :μₛ => true,
-        :σ => 1e-4,
-        :α => 0.25:.25:2.5,
-        :β => 0.25:.25:2.5,
-        :γ => 1.,
-        :threshold => false,
-        :dist => "normal",
-        :N => 1,
-    );
-
-plt = plot(x->x, xlims = (0, 2.5), color = :black, lw = 2)
-xlabel!(L"\alpha")
-ylabel!(L"\beta")
-for σ ∈ (1e-3, 5e-3, 1e-2) 
-    P[:σ] = σ
-    ϕ = ThreadsX.collect(diversity!(p) for p in expand(P));
-    plot_contour_line!(plt, ϕ, P; label = "σ = $(P[:σ])")
-end
-current()
-#savefig(plotsdir("alpha-beta"))
-
-#= alternative  =#
-P = Dict{Symbol, Any}(
-        :scaled => true,
-        :S => 50,
[…]

### 2023-03-23 scripts/single-run.jl

@@ -4,22 +4,22 @@ foreach(includet, glob("*.jl", srcdir()))
 
 using Plots, LaTeXStrings
 
-p = Dict{Symbol, Any}(
+p = Dict{Symbol,Any}(
     :scaled => true,
     :S => 100,
     :μ => 0.5,
-    :σ => .05,
-    :α => .5,
-    :β => 1.,
-    :γ => 1.,
+    :σ => 0.05,
+    :α => 0.5,
+    :β => 1.0,
+    :γ => 1.0,
     :extinction_threshold => 1e-4,
     :dist => "gamma",
     :N => 1,
     :seed => rand(UInt)
 );
 
-evolve!(p; trajectory = true);
-plot(p[:trajectory], label = false)
+evolve!(p; trajectory=true);
+plot(p[:trajectory], label=false)
 p[:converged]
 p[:equilibrium]
 
@@ -29,45 +29,45 @@ p[:converged]
 #= trajectories =#
 plot(
     p[:trajectory],
-    ylabel = L"x_i(t)",
-    xlabel = L"t",
-    yscale = :log,
-    linewidth = 1,
-    legend = false,
-    grid = false,
+    ylabel=L"x_i(t)",
+    xlabel=L"t",
+    yscale=:log,
+    linewidth=1,
+    legend=false,
+    grid=false,
 )
 
 #= density distribution =#
 p_eq = p[:trajectory].u[end]
 histogram(p_eq,
-    normalize = true,
-    alpha = .5,
-    color = :green,
-    ylabel = L"P(x_*)",
-    xlabel = L"x_*",
-    label = false,
-    grid = false,
+    normalize=true,
+    alpha=0.5,
+    color=:green,
+    ylabel=L"P(x_*)",
+    xlabel=L"x_*",
+    label=false,
+    grid=false,
 )
 
 #= cavity solution with gaussian approximation =#
-X=[x for x in .9*minimum(p_eq):.0001:1.1*maximum(p_eq)]
-plot!(X,[pdf(P_gauss(p), x) for x in X],
-labels = false, #"cavity - gaussian",
-linewidth = 2,
-alpha = 1,
-linecolor = :black,
-linestyle = :dash,
-grid = false
+X = [x for x in 0.9*minimum(p_eq):0.0001:1.1*maximum(p_eq)]
+plot!(X, [pdf(P_gauss(p), x) for x in X],
+    labels=false, #"cavity - gaussian",
[…]

### 2023-03-23 src/cavity.jl
remove cavity here

@@ -1,140 +1,0 @@
-using StatsBase
-using Random, Distributions, Statistics
-using DifferentialEquations
-using LinearAlgebra
-using QuadGK
-using HCubature
-using NLsolve
-using SpecialFunctions
-
-#= expression for density distribution with gaussian approximation (γ=1) =#
-function P_gauss(p)
-    @unpack μ, σ, α, β, S = p
-    μ = p[:scaled] ? μ : μ*S
-    σ = p[:scaled] ? σ : σ*sqrt(S) 
-    e1 = μ^(1/(α-β-1))
-    e2 = e1^2/(1-(1/(α-β))^2*(μ*e1)^(2*(1-(α-β))/(α-β))*σ^2)
-    #ϕ = e2>0 ? (1+erf((e1)/sqrt(2*(e2-e1^2))))/2 : 1
-    return e2>0 ? Normal(e1, sqrt(e2-e1^2)) : Normal(e1, 0) #(e1, e2>0 ? e2 : 0, ϕ, e2>0 ? Normal(e1,sqrt(e2-e1^2)) : Normal(e1,0))
-end
-
-#= expression for density distribution (γ=1) =#
-function P_n(n, e1_n, e2_n, p)
-    @unpack μ, σ, α, β = p
-    μ = p[:scaled] ? μ : μ*S
-    σ = p[:scaled] ? σ : σ*sqrt(S)
-    return abs(α-β)*n^(α-β-1)/(sqrt(2*π*σ^2*e2_n))*
-        exp(-(n^(α-β)-μ*e1_n)^2/(2*σ^2*e2_n))
-end
-
-#= cavity calculation for fraction of positive equilibriums 
-and first two moments of the equilibrium distribution given μ and σ. =#
-function Cavity(p;
-    Iter = 2000, #number of iteration
-    rela = .01, #relax parameter for fixed point
-    tol = 1e-9, #requested tolerance for numerical integrator
-    n_max = 1e4, #upper bound for integration
-    n_min = p[:threshold], #lower bound for integration
-    e1_init = (p[:scaled] ? 1/p[:μ] : 1/p[:μ]/p[:S]), #initial guess for e1_n
-    e2_init = (p[:scaled] ? 1/p[:μ] : 1/p[:μ]/p[:S]) #initial guess for e1_n
-    )
-
-    ϕ_n = ones(Iter+1) #fraction of survivals
[…]

### 2023-03-23 src/ahmadian.jl
remove script from ahmadian

@@ -48,30 +48,3 @@ function ahmadian(p)
     
     scatter!(spectrum(p); alpha = .2, label = false)
 end
-
-include(srcdir("dynamics.jl"))
-p = Dict{Symbol, Any}(
-    :scaled => false,    
-    :S => 50,
-    :μ => 1e-2,
-    :μₛ => 1e-3,
-    :σ => 0.005,
-    :α => .7,
-    :β => 1.,
-    :γ => 1.,
-    :extinction_threshold => 1e-2,
-    :dist => "gamma",
-    :N => 1,
-    :seed => rand(UInt)
-);
-evolve!(p; trajectory = true);
-
-using Plots
-traj = plot(
-    p[:trajectory],
-    linewidth = 1,
-    legend = false,
-    grid = false,
-    palette = :blues,
-)
-plot(traj, ahmadian(p))

### 2023-03-23 scripts/alpha-beta.jl
boh

@@ -26,6 +26,8 @@ function σ_critical(α, β = 1)
         s < .9 && return σ
     end
 end
- 
+
+σ_critical(.3)
+
 α_range = .1:.1:1.
 @time map(σ_critical, α_range)

### 2023-03-27 scripts/phase-diagram-plot.jl
prepare mu-sigma for cluster

@@ -1,0 +1,44 @@
+
+
+begin
+    
+    sublinear = heatmap(
+        P[:μ],
+        P[:σ],
+        Φ,
+        dpi=500,
+        grid=false,
+        xlabel=L"\mu N",
+        ylabel=L"\sigma \sqrt{N}",
+        #c=cgrad(:Greys, rev=true)
+    )
+    #end
+
+    #= critical line for gaussian approximation =#
+    μ_critical, σ_critical = critical_line_gauss(P, μ_range=(0.01:0.25:1.26))
+    plot!(μ_critical, σ_critical,
+        labels=false,
+        linewidth=4,
+        linecolor=:red,
+        grid=false,
+        yticks=[0, 0.6, 1.2],
+        xticks=[0, 0.6, 1.2],
+        size=(600, 400),
+    )
+
+    #= critical line =#
+    μ_critical, σ_critical = critical_line(P, μ_range=(0.01:0.2:1.25),
+        n_max=1e1)
+    plot!(μ_critical, σ_critical,
+        labels=false,
+        linewidth=4,
+        linecolor=:green,
+        grid=false,
+        ylims=[0.0, 1.2],
+        xlims=[0.0, 1.2],
+        yticks=[0, 0.6, 1.2],
+        xticks=[0, 0.6, 1.2],
+        size=(600, 400),
+    )
+    savefig(plotsdir("mu-sigma-$dist"))
+end

### 2023-03-27 scripts/mu-sigma.jl
mu-sigma on cluster

@@ -1,5 +1,7 @@
-using DrWatson, Glob, Revise
+using DrWatson
 @quickactivate
+
+using Glob, Revise
 foreach(includet, glob("*.jl", srcdir()))
 
 using ThreadsX

### 2023-03-27 papers/general-stability.tex
update draft

@@
 
-\title{Beyond May: Complexity-stability relationships in high-dimensional dynamical systems}
+\title{Beyond May: Complexity-stability relationships \\
+in high-dimensional dynamical systems}
 
@@
 
-Few mathematical arguments have influenced biological thinking like May's 1970 prediction that large, complex systems cannot be stable.
-It is, indeed, a perplexing prediction.
-On the one hand, the argument is so simple and compelling that it is hard not to conclude that it must hold universally.
+Few mathematical arguments have influenced biological thinking like May's prediction that large, complex systems cannot be stable.
+It is, on the face of it, a perplexing conclusion.
+On the one hand, May's argument is so simple and compelling that it hard not to conclude that it must hold universally.
 On the other, it is clear that at least some large, complex systems are stable---or else which stable patterns would biology be studying in the first place?
@@
 
-May's prediction is intriging. On the one hand, his argument is simple and compelling, suggesting general applicability. 
-On the other, it is clear that at least some large, complex systems are stable, else science would be impossible. 
-If anything, the opposite relationship between complexity and stability seems to hold empirically: rich, strongly coupled ecosystems (eg.
-rainforests) tend to be stable over time, while sparser ones (eg.
-arctic environments) often exhibit large fluctuations, exctinctions, and invasions.
+May's argument goes as follows.
+A system with $N$ populations $x_i$ is stable is one that can be characterized by an equilibrium point $x^*$.
+Near that equilibrium $x = x^* + \delta x$, the dynamics of the system will be described by linear equations $d(\delta x)/dt = A (\delta x)$, and stability of these equations requires that all eigenvalues of $A$ have negative real part.
+But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions with zero mean and variance $\sigma^2$, and $-I$ corresponds to stabilizing self-interactions on some natural timescale, then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
+This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability".
+(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, to incomplete connectivity, or to correlated interactions.)
 
@@
 
-Under what condition is \eqref{dynamics} linearly stable? 
+Under what condition does \eqref{dynamics} admit a linearly stable equilibrium?
 We begin by considering the simple case where all self-interactions have the same strength $a_{ii} = a_{\textrm{s}}$, and similarly for cross-interactions $a_{ij} = a_{\textrm{c}}$ ($i\neq j$).
-In that case, a straightforward computation shows that an equilibrium $x_*$ will be linearly stable if  
+Under these assumptions, a straightforward computation shows that an equilibrium $x_*$ will be linearly stable if
 \begin{equation}\label{homogeneous-general}
-    \left(\frac{f'_*}{f_*} - \frac{g'_*}{g_*}\right) + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g_*h'_*}{f_*} < 0 \, . 
+    \left(\frac{f'_*}{f_*} - \frac{g'_*}{g_*}\right) + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g_*h'_*}{f_*} < 0. 
 \end{equation}
 Thus, the stability of a competitive equilibrium depends on the relative strength of self- and cross-interactions ($a_{\textrm{s}} - a_{\textrm{c}}$), but also on the relative convexity of the production and response functions ($f'_*/f_* - g'_*/g_*$).
-With power laws, \eqref{homogeneous-general} evaluates to $(\alpha - \beta)(N-1) < \gamma(a_{\textrm{s}}/a_{\textrm{c}}- 1) - (\alpha - \beta)(a_{\textrm{s}}/a_{\textrm{c}})$, leading to three different regimes:
+With power laws, \eqref{homogeneous-general} evaluates to
+\begin{equation}
+    (\alpha - \beta)(N-1) < \gamma(a_{\textrm{s}}/a_{\textrm{c}}- 1) - (\alpha - \beta)(a_{\textrm{s}}/a_{\textrm{c}}),
+\end{equation}
+leading to three different regimes:
 \begin{itemize}
@@
 \centering
-\includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10.pdf}
+\includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
     \caption{...}
     \label{fig: stability line + sims}
+\end{figure}
+
+\section{Conclusions}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=.45\textwidth]{figs/alpha-beta.pdf}
+        \caption{...}
+        \label{fig: alpha-beta}
     \end{figure}

### 2023-03-29 papers/general-stability.tex
update draft

@@
 which appears again in the denominator of the summand.
-It reduces indeed to the homogenous case for $\sigma\to 0$.
-\red{Discuss better.}
 
@@
 where $\xi$ is a standard normal random variable (i.e. $\xi\sim\mathcal{N}(0,1)$).
-The equation above can be solved for $x_*$ in if $(\alpha-\beta)/\gamma\in\mathbb{N}$,
-or in general, if we set $a_s=0$, which amounts to include
+The equation above can be solved for $x_*$ for specific values of
+$(\alpha-\beta)/\gamma$, or, in general, if we set $a_s=0$, which amounts to include
 all the self-interaction in the nonlinear produciton term
-$x_*^{\alpha}$.
+$x_*^{\alpha}$. In the following, in studying a system which
+follows in the 'anti-May' phase, we show that numerical results
+obtained for the case in which the diagonal part of the
+interaction matrix is non-vanishing and equal to the
+off-diagonal part are well approximated by the results obtained
+setting $a_s=0$.
 In this case the stationary solution is given by 
-\begin{equation}
+\begin{equation} \label{eq: cavity solution}
     x_* = \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
@@
 \end{equation}
-where the expectations have to be computed self-consistently, 
-taking also in account normalization constraint.
+where the expectations have to be computed self-consistently.
 
@@
 $a_s=0$ in order to leverage the equilibrium solution in Eq.~\eqref{eq: dist general}.
-which is tested against simulations in Fig.~\ref{fig: cavity sol.}.
-
-\dots
-
-\subsection{Simulations}
+In this case the first and second moment to be self-consistently
+computed formally diverges. This is not truly porblematic in practice
+because we are dealing with large, but \emph{finite} populations.
+Following~[Cui2020,Us2023] and assuming
+$\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$,
+we can Taylor expand to the first-order the cavity 
+solution~\eqref{eq: cavity solution} and obtain a
+gaussian approximation to Eq.~\eqref{eq: dist general}
+with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and
+$\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.
+This is plotted against simulations in 
+Fig.~\ref{fig: cavity sol.}, alongside the exact solution~\eqref{eq: dist general} 
+where integration for the moments has been performed up to a maximum
+value above which no equilibirum value should be found for a given $N$.
+Simulations correspond to $a_s=\mu$, showing that the choice
+of vanishing self-interaction for the analytical results
+is suited to describe this case as well.
 
@@
 
+\subsection{Simulations}
+
+Equipped with the equilibirum distribution, we can compute
+the critical value of heterogeneity 
+$\sigma_c$, for a given $N$ and $\mu$,
+defined by taking the equality in 
+Eq.~\eqref{eq: random-stability int}. 
+For $\sigma$ above this value the system becomes
+unstable. In Fig~\ref{fig: stability line + sims}
+the result is portrayed in the $\sigma \sqrt{N} - \mu N$ plane
+compared with results from simulations where stability is defined
+as stable, full coexistence of the system.
+Everything else left fixed, an increase in the number of
+degrees of freedom $N$ will move the system along a square-root
+trajectory, bringing it into the stable regime or keeping it
+in it if already stable. This confirms we are in the
+'anti-May' phase.
+
 \begin{figure}[h!]
@@
 \end{figure}
-
-\section{Conclusions}
 

### 2023-04-21 papers/general-stability.tex
update draft

@@
-\documentclass[%
+\documentclass[
  reprint,
@@
 \usepackage{caption}
+\usepackage{notes2bib}
 \newcommand{\red}{\textcolor{red}}
@@
 
-Few mathematical arguments have influenced biological thinking like May's prediction that large, complex systems cannot be stable.
+Few mathematical arguments have influenced biological thinking like May's
+prediction that large, complex systems cannot be stable~\cite{May1972}. 
 It is, on the face of it, a perplexing conclusion. 
-On the one hand, May's argument is so simple and compelling that it hard not to conclude that it must hold universally.
-On the other, it is clear that at least some large, complex systems are stable---or else which stable patterns would biology be studying in the first place?
-If anything, the opposite relationship between complexity and stability seems to hold empirically: rich, strongly coupled ecosystems (eg.
-rainforests) tend to be  stable over time, while sparser ones (eg.
-arctic environments) often exhibit large fluctuations, exctinctions, and invasions.
+On the one hand, May's argument is so simple and compelling
+that it hard not to conclude that it must hold universally,
+and beyond biology~\cite{Haldane2011,Moran2019}. 
+On the other, it is clear that at least some large, 
+complex systems are stable---or else which stable patterns 
+would ecology be studying in the first place? If anything, 
+the opposite relationship between complexity and stability 
+seems to hold empirically: rich, strongly coupled ecosystems 
+(eg. rainforests) tend to be  stable over time, while sparser ones 
+(eg. arctic environments) often exhibit large fluctuations, 
+exctinctions, and invasions~\cite{Loreau2022,McCann2000,Hutchinson1959,Odum1959,MacArthur1955}.
 
-May's argument goes as follows.
-A system with $N$ populations $x_i$ is stable is one that can be characterized by an equilibrium point $x^*$.
-Near that equilibrium $x = x^* + \delta x$, the dynamics of the system will be described by linear equations $d(\delta x)/dt = A (\delta x)$, and stability of these equations requires that all eigenvalues of $A$ have negative real part.
-But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions with zero mean and variance $\sigma^2$, and $-I$ corresponds to stabilizing self-interactions on some natural timescale, then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
-This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability".
-(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, to incomplete connectivity, or to correlated interactions.)
+May's argument goes as follows. A system with $N$ populations $x_i$ is stable 
+is one that can be characterized by an equilibrium point $x^*$. 
+Near that equilibrium $x = x^* + \delta x$, the 
+dynamics of the system will be described by linear 
+equations $d(\delta x)/dt = A (\delta x)$, and stability of 
+these equations requires that all eigenvalues of $A$ 
+have negative real part. But if $A$ can be represented as $A = B - I$, 
+where $B$ consists of random, independent interactions with zero 
+mean and variance $\sigma^2$, and $-I$ corresponds to 
+stabilizing self-interactions on some natural timescale, 
+then the circular law of random matrix theory implies that all 
+eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$. 
+This places a sharp constraint on both diversity $N$ and interaction 
+strengh $\sigma$, often referred to as "complexity begets instability". 
+(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, 
+to incomplete connectivity, or to correlated interactions.)
 
-In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients, such as spatial or food-web structure, time delays, etc.
-Here, we show that May's argument is itself incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
+In ecology, many authors have sought to ease the tension between May's 
+prediction and empirical observation by invoking effects not captured by 
+dynamical systems with random 
+coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}. 
+Here, we show that May's argument is itself incomplete: 
+random dynamical systems do not necessarily imply that 
+stability decreases with diversity or interaction 
+strength---the opposite behavior is also possible, 
+without the need for special or additional structure.
 
@@
 \end{equation}
-Here $f(x_i)$ represents the contribution of a population $i$ to its own growth, while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. 
+Here $f(x_i)$ represents the self-dynamics of a population $i$, while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. 
 That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp. $a_{ji} < 0$) implies a positive (resp. negative) effect of $j$ on the growth of $i$. 
 In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''. 
+This rather general setting has been used to study universality\cite{Barzel2013} and to construct 
+minimal models for complex dynamics\cite{Barzel2015} ranging from biological~\cite{Alon2006,Karlebach2008,Skinner2012} 
+to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
 
@@
 From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
-Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
-For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level. 
-It is also common to model predator-prey interactions with a square root laws ($g(x) \sim h(x) \sim x^{1/2}$), for example.
+Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
+For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level~\cite{Hatton2015,Hatton2023}. 
+It has recently been shown appropriate to model predator-prey interactions with a square root laws ($g(x) \sim h(x) \sim x^{1/2}$)~\cite{Barbier2021,Mazzarisi2023}, for example.
 
@@
 
-\subsection{Random interactions}
+\subsection{Random interactions: general reults}
 
@@
 In order to investigates its spectral properties, 
-we follow Stone~[Stone2018] and use a
-result in random matrix theory~[Ahmadian2015]
-which generalize the classical `circular law'~[Circular].
+we follow Stone~\cite{Stone2018} and use a
+result in random matrix theory~\cite{Ahmadian2015}
+which generalize the classical `circular law'~\cite{Potters2020}.
 
-In Ref.~[Ahmadian2015] Ahmadian \emph{et al.} 
+In Ref.~\cite{Ahmadian2015} Ahmadian \emph{et al.} 
 consider matrices of the form $M + LSR$, where $M$,  
@@
 eigenspectrum of large matrices of this form is defined by
-\begin{equation}
-    \textrm{Tr}[(M_\zeta M_\zeta^\dagger)^{-1}]\geq \sigma^{-2} \, , 
-\end{equation}
+$\textrm{Tr}[(M_\zeta M_\zeta^\dagger)^{-1}]\geq \sigma^{-2}$ , 
 where $M_\zeta = L^{-1}(\zeta I - M)R^{-1}$ and $\zeta\in\mathbb{C}$. 
@@
 $a = (\mu_s-\mu)I + S$ by relying on 
-rank-one perturbation theory, we would only neglect outliers
-uninfluential for the linear stbaility properties of the system by doing so
-(see Ref.~[Stone2018]).
+rank-one perturbation theory: we would only neglect outliers
+uninfluential for the linear stbaility properties of the system (see Ref.~\cite{Stone2018}).
 The Jacobian in Eqs.~\eqref{eq: jac off-diag} and \eqref{eq: jac diag} assumes therefore the form $M + LJR$ with
@@
     +(\mu_s-\mu)g_*(\mathbf x)h_*'(\mathbf x)\right) \ , \\
-    L &= \textrm{diag}(g_*(\mathbf x)) \ , \\
-    R &= \textrm{diag}(h_*'(\mathbf x)) \ .
+    L &= \textrm{diag}(g_*(\mathbf x)) \ , \qquad  
+    R = \textrm{diag}(h_*'(\mathbf x)) \ . \nonumber
 \end{align}
@@
 This expression can be considered the
-random interaction case equivalent of
-the expression in Eq.~\eqref{homogeneous-general} 
+generalization to random interaction of
+the expression in Eq.~\eqref{homogeneous-general},
 which appears again in the denominator of the summand.
 
-There are straightforward results 
-that follows from this expression.
+There are straightforward results that follows.
 For generalized Lotka-Volterra (GLV) models,
@@
 the expression in Eq.~\eqref{eq: random-stability}
-becomes independent on the equilibria and we easily 
+becomes independent on the equilibria and we 
 recover the linear stability condition
-\begin{equation}
-    \sigma\sqrt{N} < (\mu_s-\mu) \ .
-\end{equation}
+$\sigma\sqrt{N} < (\mu_s-\mu)$.
 Moreover we can define an equivalence class of models
@@
 In the general case in which the dependence on the equilibria
-$\mathbf{x}_*$ does not cancel out, is not possible to
-obtain a clean complexity-stability relationship as before.
+$\mathbf{x}_*$ does not cancel out, it is not possible to
+obtain an explicit and simple complexity-stability relationship.
 We can make progress in the large $N$ limit by
@@
 
+\subsection{Random interactions: power laws case}
 In order to investigate the different regimes
@@
 for the system. This can be achieved in different ways,
-depending among other things on specific constraint that
-one can impose on the system. 
-One approach is reported below, then we specialize to a choice of
-the power laws exponens.
+depending, among other things, on specific constraints 
+imposed on the system. 
+One approach is reported below, 
+then we specialize to a specific choice of 
+the power laws exponents.
 
-Following, e.g., Ref.~[Roy2019] we can derive from 
+Following, e.g., Ref.~\cite{Roy2019} we can derive from 
 the evolution equation
@@
 The averages and correlations have to be computed self-consistently.
-If the system admits an equilibrium (i.e. is feasible), for 
+If the system admits an equilibrium, for 
 $t\to\infty$ each of the initial $N$ degrees 
-of freedom will reach a final value, 
-and the effective random variable describing their
+of freedom will reach a final finite value. 
+The effective random variable describing their statistics
 also become time-independent,
@@
 all the self-interaction in the nonlinear produciton term
-$x_*^{\alpha}$. In the following, in studying a system which
-follows in the 'anti-May' phase, we show that numerical results
-obtained for the case in which the diagonal part of the
-interaction matrix is non-vanishing and equal to the
+$x_*^{\alpha}$. In the following, for cases which
+falls in the "anti-May" phase, we show that numerical results
+for non-vanishing diagonal part of the
+interaction matrix and equal to the
 off-diagonal part are well approximated by the results obtained
 setting $a_s=0$.
+
 In this case the stationary solution is given by 
@@
 \end{equation}
-
 The equilibrium probability distribution function for $x_*$, $P(x_*)$,
@@
 
-Let us focus to the case $\alpha=1$, $\beta=1.5$ and $\gamma=1$, with
+Let us focus on the specific choice 
+$\alpha=1$, $\beta=1.5$ and $\gamma=1$, with
 $a_s=0$ in order to leverage the equilibrium solution in Eq.~\eqref{eq: dist general}.
-In this case the first and second moment to be self-consistently
-computed formally diverges. This is not truly porblematic in practice
+In this case the first and second moments, 
+to be self-consistently computed, formally diverges. This is not truly porblematic in practice
 because we are dealing with large, but \emph{finite} populations.
-Following~[Cui2020,Us2023] and assuming
+Following~\cite{Cui2020,Hatton2023} and assuming
 $\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$,
-we can Taylor expand to the first-order the cavity 
-solution~\eqref{eq: cavity solution} and obtain a
-gaussian approximation to Eq.~\eqref{eq: dist general}
+we can Taylor expand to the first-order in 
+$\sigma \sqrt{N\langle x_*^2\rangle}$ the cavity 
+solution~\eqref{eq: cavity solution} and 
+approximate to Eq.~\eqref{eq: dist general} with
+a gaussian distribuiton
 with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and
 $\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.
-This is plotted against simulations in 
-Fig.~\ref{fig: cavity sol.}, alongside the exact solution~\eqref{eq: dist general} 
+
+The result is plotted against simulations in 
+Fig.~\ref{fig: cavity sol.}, alongside the exact 
+solution~\eqref{eq: dist general} 
 where integration for the moments has been performed up to a maximum
-value above which no equilibirum value should be found for a given $N$.
+value above which no equilibirum value should 
+be found, for a given $N$.
 Simulations correspond to $a_s=\mu$, showing that the choice
@@
     \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
-    \caption{...}
+    \caption{The equilibrium distribution for a realization
+    of the system described by Eq.~\eqref{eq: full abg} 
+    is accurately reproduced by the cavity solution in 
+    Eq.\eqref{eq: dist general} and its gaussian approximation.
+    For the simulation we used  $\alpha=1$, $\beta=1.5$,
+    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=10^{-2}/2$.}
     \label{fig: cavity sol.}
 \end{figure}
-
-\subsection{Simulations}
 
@@
 For $\sigma$ above this value the system becomes
-unstable. In Fig~\ref{fig: stability line + sims}
-the result is portrayed in the $\sigma \sqrt{N} - \mu N$ plane
+unstable. In Fig.~\ref{fig: stability line + sims}
+the result is portrayed in the $(\sigma \sqrt{N},\mu N)$ plane,
 compared with results from simulations where stability is defined
-as stable, full coexistence of the system.
+as stable, full, coexistence of the system.
 Everything else left fixed, an increase in the number of
@@
 trajectory, bringing it into the stable regime or keeping it
-in it if already stable. This confirms we are in the
-'anti-May' phase.
+in it if already stable. In other words, we are in the
+"anti-May" phase. The stability boundary for GLV, by contrast, is touched
+at $\sigma \sqrt{N} = \mu N/N + \mu_s$, which, for $N\to \infty$
+gives an horizontal line in the $(\sigma \sqrt{N},\mu N)$ plane,
+defined by $\sigma_c\sqrt{N} = \mu_s$.
 
@@
     \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
-    \caption{...}
+    \caption{For a system in the "anti-May" phase an increase in $N$,
+    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane,
+    will be stabilizing. Here we show results from simulation of system~\eqref{eq: full abg},
+    where stability is defined as full stable coexistence, overleyed with the critical line
+    obtained through the cavity method starting from Eq.\eqref{eq: random-stability int} 
+    (see main text). The values of the parameters are $\alpha=1$, $\beta=1.5$,
+    $\gamma=1$, $N=50$ and $10$ replicates for the simulations.}
     \label{fig: stability line + sims}
 \end{figure}
+
+Figure~\ref{fig: alpha-beta} portrays results of simulaitons 
+for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, 
+at fixed $\mu N$ and $\gamma = 1$ (further details in caption), 
+defined as the value above which
+full stable coexistence (i.e. 'stability'
+as defined for Fig.\ref{fig: stability line + sims}) 
+is less probable than $50\%$.
+In these simulations we are using $\mu_s = \mu$, which result
+in unstable outcomes for $\beta<\alpha$ while
+a certain degree of heterogeneity in the interactions
+admits admits stability for $\beta>\alpha$, the higher
+the larger is the difference between $\beta$ and $\alpha$.
+Notice that the existence of a $\sigma_c\sqrt{N}$ above which
+instability is triggered does not imply that increasing $N$ will
+destabilize the system, because also $\mu N$ would change.
+Rather, for all the value of $\beta$ and $\alpha$ in the
+upper triangle, the picture is the same as the one depicted in 
+Fig.~\ref{fig: stability line + sims}: we are in "anti-May"
+phase.
 
@@
     \includegraphics[width=.45\textwidth]{figs/alpha-beta.pdf}
-    \caption{...}
+    \caption{The "May" and "anti-May" phases described for the
+    case of uniform interactions are robust for random interactions.
+    Here we simulated system~\eqref{eq: full abg} for a fixed value of
+    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates 
+    and evaluated $\sigma_c$, defined here as
+    the value above which full stable coexistence is less probable than 50\%,
+    at varying $\alpha$ and $\beta$. No amount of heterogeneity is allowed in the
+    lower triangle because we have $\mu=\mu_s$, while an increasing value of $\sigma_c$
+    is found as $\beta$ becomes bigger than $\alpha$.
+    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which
+    instability is triggered also in the upper triangle,
+    does not imply that increasing $N$ will
+    destabilize the system, because also $\mu N$ would change (see main text).
+    }
     \label{fig: alpha-beta}
@@
 
+\clearpage
+
+\bibliography{general-stability-bib}
+
+\bibliographystyle{unsrt}
+
 \end{document}

### 2023-04-23 papers/general-stability.tex

@@
-\documentclass[
+\documentclass[%
  reprint,
@@
 \usepackage{caption}
-\usepackage{notes2bib}
 \newcommand{\red}{\textcolor{red}}
@@
 
-Few mathematical arguments have influenced biological thinking like May's
-prediction that large, complex systems cannot be stable~\cite{May1972}. 
+Few mathematical arguments have influenced biological thinking like May's prediction that large, complex systems cannot be stable.
 It is, on the face of it, a perplexing conclusion.
-On the one hand, May's argument is so simple and compelling
-that it hard not to conclude that it must hold universally,
-and beyond biology~\cite{Haldane2011,Moran2019}. 
-On the other, it is clear that at least some large, 
-complex systems are stable---or else which stable patterns 
-would ecology be studying in the first place? If anything, 
-the opposite relationship between complexity and stability 
-seems to hold empirically: rich, strongly coupled ecosystems 
-(eg. rainforests) tend to be  stable over time, while sparser ones 
-(eg. arctic environments) often exhibit large fluctuations, 
-exctinctions, and invasions~\cite{Loreau2022,McCann2000,Hutchinson1959,Odum1959,MacArthur1955}.
+On the one hand, May's argument is so simple and compelling that it hard not to conclude that it must hold universally.
+On the other, it is clear that at least some large, complex systems are stable---or else which stable patterns would biology be studying in the first place?
+If anything, the opposite relationship between complexity and stability seems to hold empirically: rich, strongly coupled ecosystems (eg.
+rainforests) tend to be  stable over time, while sparser ones (eg.
+arctic environments) often exhibit large fluctuations, exctinctions, and invasions.
 
-May's argument goes as follows. A system with $N$ populations $x_i$ is stable 
-is one that can be characterized by an equilibrium point $x^*$. 
-Near that equilibrium $x = x^* + \delta x$, the 
-dynamics of the system will be described by linear 
-equations $d(\delta x)/dt = A (\delta x)$, and stability of 
-these equations requires that all eigenvalues of $A$ 
-have negative real part. But if $A$ can be represented as $A = B - I$, 
-where $B$ consists of random, independent interactions with zero 
-mean and variance $\sigma^2$, and $-I$ corresponds to 
-stabilizing self-interactions on some natural timescale, 
-then the circular law of random matrix theory implies that all 
-eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$. 
-This places a sharp constraint on both diversity $N$ and interaction 
-strengh $\sigma$, often referred to as "complexity begets instability". 
-(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, 
-to incomplete connectivity, or to correlated interactions.)
+May's argument goes as follows.
+A system with $N$ populations $x_i$ is stable is one that can be characterized by an equilibrium point $x^*$.
+Near that equilibrium $x = x^* + \delta x$, the dynamics of the system will be described by linear equations $d(\delta x)/dt = A (\delta x)$, and stability of these equations requires that all eigenvalues of $A$ have negative real part.
+But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions with zero mean and variance $\sigma^2$, and $-I$ corresponds to stabilizing self-interactions on some natural timescale, then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
+This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability".
+(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, to incomplete connectivity, or to correlated interactions.)
 
-In ecology, many authors have sought to ease the tension between May's 
-prediction and empirical observation by invoking effects not captured by 
-dynamical systems with random 
-coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}. 
-Here, we show that May's argument is itself incomplete: 
-random dynamical systems do not necessarily imply that 
-stability decreases with diversity or interaction 
-strength---the opposite behavior is also possible, 
-without the need for special or additional structure.
+In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients, such as spatial or food-web structure, time delays, etc.
+Here, we show that May's argument is itself incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
 
@@
 \end{equation}
-Here $f(x_i)$ represents the self-dynamics of a population $i$, while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. 
+Here $f(x_i)$ represents the contribution of a population $i$ to its own growth, while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. 
 That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp. $a_{ji} < 0$) implies a positive (resp. negative) effect of $j$ on the growth of $i$. 
 In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''.
-This rather general setting has been used to study universality\cite{Barzel2013} and to construct 
-minimal models for complex dynamics\cite{Barzel2015} ranging from biological~\cite{Alon2006,Karlebach2008,Skinner2012} 
-to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
 
@@
 From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
-Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
-For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level~\cite{Hatton2015,Hatton2023}. 
-It has recently been shown appropriate to model predator-prey interactions with a square root laws ($g(x) \sim h(x) \sim x^{1/2}$)~\cite{Barbier2021,Mazzarisi2023}, for example.
+Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
+For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level. 
+It is also common to model predator-prey interactions with a square root laws ($g(x) \sim h(x) \sim x^{1/2}$), for example.
 
@@
 In order to investigates its spectral properties, 
-we follow Stone~\cite{Stone2018} and use a
-result in random matrix theory~\cite{Ahmadian2015}
-which generalize the classical `circular law'~\cite{Potters2020}.
+we follow Stone~[Stone2018] and use a
+result in random matrix theory~[Ahmadian2015]
+which generalize the classical `circular law'~[Circular].
 
-In Ref.~\cite{Ahmadian2015} Ahmadian \emph{et al.} 
+In Ref.~[Ahmadian2015] Ahmadian \emph{et al.} 
 consider matrices of the form $M + LSR$, where $M$,  
@@
 rank-one perturbation theory: we would only neglect outliers
-uninfluential for the linear stbaility properties of the system (see Ref.~\cite{Stone2018}).
+uninfluential for the linear stbaility properties of the system (see Ref.~[Stone2018]).
 The Jacobian in Eqs.~\eqref{eq: jac off-diag} and \eqref{eq: jac diag} assumes therefore the form $M + LJR$ with
@@
 
-Following, e.g., Ref.~\cite{Roy2019} we can derive from 
+Following, e.g., Ref.~[Roy2019] we can derive from 
 the evolution equation
@@
 $x_*^{\alpha}$. In the following, for cases which
-falls in the "anti-May" phase, we show that numerical results
+falls in the 'anti-May' phase, we show that numerical results
 for non-vanishing diagonal part of the
@@
 because we are dealing with large, but \emph{finite} populations.
-Following~\cite{Cui2020,Hatton2023} and assuming
+Following~[Cui2020,Us2023] and assuming
 $\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$,
@@
     \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
+    \label{fig: cavity sol.}
     \caption{The equilibrium distribution for a realization
@@
     $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=10^{-2}/2$.}
-    \label{fig: cavity sol.}
 \end{figure}
@@
 compared with results from simulations where stability is defined
-as stable, full, coexistence of the system.
+as stable, full coexistence of the system.
 Everything else left fixed, an increase in the number of
@@
 in it if already stable. In other words, we are in the
-"anti-May" phase. The stability boundary for GLV, by contrast, is touched
+'anti-May' phase. The stability boundary for GLV, by contrast, is touched
 at $\sigma \sqrt{N} = \mu N/N + \mu_s$, which, for $N\to \infty$
@@
     \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
-    \caption{For a system in the "anti-May" phase an increase in $N$,
-    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane,
-    will be stabilizing. Here we show results from simulation of system~\eqref{eq: full abg},
-    where stability is defined as full stable coexistence, overleyed with the critical line
-    obtained through the cavity method starting from Eq.\eqref{eq: random-stability int} 
-    (see main text). The values of the parameters are $\alpha=1$, $\beta=1.5$,
-    $\gamma=1$, $N=50$ and $10$ replicates for the simulations.}
+    \caption{... $\alpha=1$, $\beta=1.5$,
+    $\gamma=1$, $N=50$ and $10$ replicates.}
     \label{fig: stability line + sims}
@@
 as defined for Fig.\ref{fig: stability line + sims}) 
-is less probable than $50\%$.
+is less probable then $50\%$.
 In these simulations we are using $\mu_s = \mu$, which result
@@
 upper triangle, the picture is the same as the one depicted in 
-Fig.~\ref{fig: stability line + sims}: we are in "anti-May"
+Fig.~\ref{fig: stability line + sims}: we are in 'anti-May'
 phase.
@@
     \includegraphics[width=.45\textwidth]{figs/alpha-beta.pdf}
-    \caption{The "May" and "anti-May" phases described for the
-    case of uniform interactions are robust for random interactions.
-    Here we simulated system~\eqref{eq: full abg} for a fixed value of
-    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates 
-    and evaluated $\sigma_c$, defined here as
-    the value above which full stable coexistence is less probable than 50\%,
-    at varying $\alpha$ and $\beta$. No amount of heterogeneity is allowed in the
-    lower triangle because we have $\mu=\mu_s$, while an increasing value of $\sigma_c$
-    is found as $\beta$ becomes bigger than $\alpha$.
-    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which
-    instability is triggered also in the upper triangle,
-    does not imply that increasing $N$ will
-    destabilize the system, because also $\mu N$ would change (see main text).
-    }
+    \caption{...
+    $\mu=\mu_s=10^{-2}$, $N=100$ and $100$ replicates.}
     \label{fig: alpha-beta}
@@
 
-\clearpage
-
-\bibliography{general-stability-bib}
-
-\bibliographystyle{unsrt}
-
 \end{document}

### 2023-04-29 papers/beyond-may.tex
reorganize and edit draft

\documentclass[
 prl,
 twocolumn,
 amsmath,
 amssymb,
 aps,
]{revtex4-2}

\usepackage{colortbl}
\usepackage{graphicx} 
\usepackage{booktabs}
\usepackage{caption}
\usepackage{notes2bib}
\newcommand{\red}{\textcolor{red}}
\usepackage[justification=raggedright]{caption}

\begin{document}

\title{Beyond May: Complexity-stability relationships \\
in disordered dynamical systems}

\author{Onofrio Mazzarisi}
\affiliation{Department of Ecology and Evolutionary Biology, University of Kansas, Haworth Hall, 
1200 Sunnyside Avenue
Lawrence, KS 66045, USA}

\author{Matteo Smerlak}
\affiliation{Laboratoire Biophysique et Evolution, ESPCI ParisTech, 10 rue Vauquelin, 75005 Paris, France}
\affiliation{Capital Fund Management, 23 Rue de l'Université, 75007 Paris, France}

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified

\begin{abstract}

\end{abstract}
\maketitle

\section*{Introduction}

\section*{Results}

\paragraph*{\bf Model.}
\input{model.tex}

\paragraph*{\bf Homogeneous interactions.}
\input{homogeneous.tex}

\paragraph*{\bf Random interactions: general reults.}
\input{random-general.tex}
\paragraph*{\bf Random interactions: power laws case.}
\input{random-power.tex}

\section{Conclusion}
\input{conclusion.tex}

\begin{acknowledgments}
We gratefully acknowledge multiple discussions with Ian Hatton, Matthieu Barbier and Ada Altieri on ecological power laws and the diversity-stability debate. 
\end{acknowledgments}

\clearpage

\bibliography{general-stability}
\bibliographystyle{unsrt}

\end{document}

### 2023-04-29 papers/conclusion.tex
reorganize and edit draft

The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle.

We reconsidered the relationship between diversity and stability in complex systems, a classic problem since May's 1973's puzzling prediction that "complexity begets instability".

Considering dynamical models more general than the classical Lotka-Volterra model, we found that the relation can in fact go both ways:  depending on the relative convexity of self- and cross-interactions,

### 2023-04-29 papers/homogeneous.tex
reorganize and edit draft

--- 2023-03-29 papers/general-stability.tex
@@
-\documentclass[%
- reprint,
-%superscriptaddress,
-%groupedaddress,
-%unsortedaddress,
-%runinaddress,
-%frontmatterverbose, 
-%preprint,
-%preprintnumbers,
-%nofootinbib,
-%nobibnotes,
-%bibnotes,
- amsmath,amssymb,
- aps,
-%pra,
-%prb,
-%rmp,
-%prstab,
-%prstper,
-%floatfix,
-]{revtex4-2}
-
-\usepackage{colortbl}
-\usepackage{graphicx} 
-\usepackage{booktabs}
-\usepackage{caption}
-\newcommand{\red}{\textcolor{red}}
-
-\begin{document}
-
-\title{Beyond May: Complexity-stability relationships \\
-in high-dimensional dynamical systems}
-
-\author{Onofrio Mazzarisi}
-\author{Matteo Smerlak}
-
- \affiliation{MPI MiS}
-
-\date{\today}% It is always \today, today,
-             %  but any date may be explicitly specified
-
-\begin{abstract}
-Fifty years ago, Robert May predicted that large, complex systems are unlikely to be stable.
-Here, we revisit May's argument to show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems, depending on the relative convexity of the response to self- vs.
-cross-interactions.
-We illustrate the transition between "May" (complexity begets instability) and "anti-May" (complexity begets stability) behavior with a non-linear generalization of the Lotka-Volterra model.
-\end{abstract}
-
-\maketitle
-
-\section{Introduction}
-
-Few mathematical arguments have influenced biological thinking like May's prediction that large, complex systems cannot be stable.
-It is, on the face of it, a perplexing conclusion.
-On the one hand, May's argument is so simple and compelling that it hard not to conclude that it must hold universally.
-On the other, it is clear that at least some large, complex systems are stable---or else which stable patterns would biology be studying in the first place?
-If anything, the opposite relationship between complexity and stability seems to hold empirically: rich, strongly coupled ecosystems (eg.
-rainforests) tend to be  stable over time, while sparser ones (eg.
-arctic environments) often exhibit large fluctuations, exctinctions, and invasions.
-
-May's argument goes as follows.
-A system with $N$ populations $x_i$ is stable is one that can be characterized by an equilibrium point $x^*$.
-Near that equilibrium $x = x^* + \delta x$, the dynamics of the system will be described by linear equations $d(\delta x)/dt = A (\delta x)$, and stability of these equations requires that all eigenvalues of $A$ have negative real part.
-But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions with zero mean and variance $\sigma^2$, and $-I$ corresponds to stabilizing self-interactions on some natural timescale, then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
-This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability".
-(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, to incomplete connectivity, or to correlated interactions.)
-
-In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients, such as spatial or food-web structure, time delays, etc.
-Here, we show that May's argument is itself incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
-
-\section{Results}
-
-\subsection{Model}
-
-Consider the dynamical system in $N$ variables
-\begin{equation}\label{dynamics}
-    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j) \, .
-\end{equation}
-Here $f(x_i)$ represents the contribution of a population $i$ to its own growth, while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. 
-That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp. $a_{ji} < 0$) implies a positive (resp. negative) effect of $j$ on the growth of $i$. 
-In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''.
-
-The classic generalized Lotka-Volterra (GLV) competition model corresponds to $f$, $g$, $h$ all linear. 
-However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$. 
-From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
-Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
-For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level. 
-It is also common to model predator-prey interactions with a square root laws ($g(x) \sim h(x) \sim x^{1/2}$), for example.
-
-\subsection{Homogeneous interactions}
-
 Under what condition does \eqref{dynamics} admit a linearly stable equilibrium? 
 We begin by considering the simple case where all self-interactions have the same strength $a_{ii} = a_{\textrm{s}}$, and similarly for cross-interactions $a_{ij} = a_{\textrm{c}}$ ($i\neq j$).
-Under these assumptions, a straightforward computation shows that an equilibrium $x_*$ will be linearly stable if
+Under these assumptions, a straightforward computation shows that a homogeneous equilibrium $\mathbf x_*$ will be linearly stable if  
 \begin{equation}\label{homogeneous-general}
-    \left(\frac{f'_*}{f_*} - \frac{g'_*}{g_*}\right) + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g_*h'_*}{f_*} < 0. 
+   \frac{f'(x^*)}{f(x^*)} - \frac{g'(x^*)}{g(x^*)} + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g(x^*)h'(x^*)}{f(x^*)} < 0.
 \end{equation}
-Thus, the stability of a competitive equilibrium depends on the relative strength of self- and cross-interactions ($a_{\textrm{s}} - a_{\textrm{c}}$), but also on the relative convexity of the production and response functions ($f'_*/f_* - g'_*/g_*$).
+Thus, stability of an equilibrium depends on the relative strength of self- and cross-interactions ($a_{\textrm{s}} - a_{\textrm{c}}$), but also on the relative growth rate of the production and response functions near that equilibrium ($f'_*/f_* - g'_*/g_*$).
 With power laws, \eqref{homogeneous-general} evaluates to 
@@
     We can call this "May" behavior.
-    \item If $\alpha < \beta$, stability places an lower bound on $N$: the more complex the system, the more likely to be stable. This is "anti-May" behavior.
+    \item If $\alpha < \beta$, stability places an lower bound on $N$: the more complex the system, the more likely to be stable.
+    This is "anti-May" behavior.
 \end{itemize}
-
-\subsection{Random interactions}
-
-Consider now Eq.~\eqref{dynamics} with i.i.d. random 
-interactions $a_{ij}$ with mean $\mu$ and standard 
-deviation $\sigma$, with the diagonal elements $a_{ii}$ 
-extracted from a distribution with mean $\mu_s$ 
-possibly different from $\mu$ and same standard deviation
-as for the off-diagonal elements.  
-We can compute the Jacobian matrix at equilibrium
-\begin{align}
-    J_{ij}^* & = a_{ij}g_*(x_i)h_*'(x_j) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
-    J_{ii}^* & = f_*'(x_i) - \frac{g_*'(x_i)f_*(x_i)}{g_*(x_i)} - a_{ii}g_*(x_i)h_*'(x_j) \ , \label{eq: jac diag}
-\end{align}
-where we used $\sum_{j}a_{ij}g_*(x_i)h_*(x_j)=-f_*(x_i)/g_*(x_i)$.
-In order to investigates its spectral properties, 
-we follow Stone~[Stone2018] and use a
-result in random matrix theory~[Ahmadian2015]
-which generalize the classical `circular law'~[Circular].
-
-In Ref.~[Ahmadian2015] Ahmadian \emph{et al.} 
-consider matrices of the form $M + LSR$, where $M$,  
-$L$ and $R$ are deterministic matrices, and $S$ 
-is a random matrix with i.i.d. coefficients, 
-zero mean and variance $\sigma^2$. They 
-show that in the complex plane the boundary of the 
-eigenspectrum of large matrices of this form is defined by
-\begin{equation}
-    \textrm{Tr}[(M_\zeta M_\zeta^\dagger)^{-1}]\geq \sigma^{-2} \, , 
-\end{equation}
-where $M_\zeta = L^{-1}(\zeta I - M)R^{-1}$ and $\zeta\in\mathbb{C}$. 
-If $L$, $R$ and $M$ are all diagonal $N\times N$ matrices, 
-this condition simplifies to
-\begin{equation}
-    \sum_{i=1}^N\frac{(L_{i}R_{i})^2}{ \vert \zeta - M_{i}\vert^2 }\geq \sigma^{-2} \ .
-\label{eq: domain}
-\end{equation}
-
-Now we decompose the interaction matrix as
-$a = \mu \mathbf{1} + (\mu_s-\mu)I + S$,
-with $\mathbf{1}$ the rank-one matrix with all entries equal to $1$
-and $S$ a random matrix as defined above.
-We can limit ourselves to consider 
-$a = (\mu_s-\mu)I + S$ by relying on 
-rank-one perturbation theory, we would only neglect outliers
-uninfluential for the linear stbaility properties of the system by doing so
-(see Ref.~[Stone2018]).
-The Jacobian in Eqs.~\eqref{eq: jac off-diag} and \eqref{eq: jac diag} assumes therefore the form $M + LJR$ with
-\begin{align}
-    M & = \textrm{diag}\left(f_*'(\mathbf x) -
-    \frac{g_*'(\mathbf x)f_*(\mathbf x)}{g_*(\mathbf x)}
-    +(\mu_s-\mu)g_*(\mathbf x)h_*'(\mathbf x)\right) \ , \\
-    L &= \textrm{diag}(g_*(\mathbf x)) \ , \\
-    R &= \textrm{diag}(h_*'(\mathbf x)) \ .
-\end{align}
-The domain of the eigenvalues of $J^*$, defined by
-Eq.~\eqref{eq: domain}, 
-first touches the right half-plane at $\zeta = 0$, 
-hence linear stability requires   
-\begin{equation}\label{eq: random-stability}
-    \sum_{i=1}^N \cfrac{\left(\cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)}\right)^2}{
-        \Bigl |\left(\cfrac{f_*'(x_i)}{f_*(x_i)} -
-        \cfrac{g_*'(x_i)}{g_*(x_i)}\right)
-        +(\mu_s-\mu) \cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)} \Bigl |^2}
-    < \sigma^{-2}. 
-\end{equation}
-This expression can be considered the
-random interaction case equivalent of
-the expression in Eq.~\eqref{homogeneous-general} 
-which appears again in the denominator of the summand.
-
-There are straightforward results 
-that follows from this expression.
-For generalized Lotka-Volterra (GLV) models,
-i.e. $f(x_i)=x_i$, $g(x_i)=x_i$ and $h(x_i)=x_i$,
-the expression in Eq.~\eqref{eq: random-stability}
-becomes independent on the equilibria and we easily 
-recover the linear stability condition
-\begin{equation}
-    \sigma\sqrt{N} < (\mu_s-\mu) \ .
-\end{equation}
-Moreover we can define an equivalence class of models
-which share the same simple condition
-for linear stability defined by the equivalence relation
-\begin{equation}\label{eq: equiv class glv}
-    \left(\cfrac{f_*'(x_i)}{f_*(x_i)} -
-        \cfrac{g_*'(x_i)}{g_*(x_i)}\right)\Bigl/
-        \cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)}=c \ ,
-\end{equation}
-where $c$ is a constant,
-resulting in the condition $\sigma\sqrt{N} < (\mu_s-\mu) +c$.
-
-In the general case in which the dependence on the equilibria
-$\mathbf{x}_*$ does not cancel out, is not possible to
-obtain a clean complexity-stability relationship as before.
-We can make progress in the large $N$ limit by
-trasforming the sum in an integral 
-$\sum_{i=1}^N\to N\int dx_*P(x_*)$, where $P(x_*)$ 
-is the equilibrium distribution of the system
-depending on the statistic of the interactions,
-and write
-\begin{equation}\label{eq: random-stability int}
-    \int dx_* \cfrac{NP(x_*) \left(\cfrac{g_*(x)h_*'(x)}{f_*(x)}\right)^2}{
-        \Bigl |\left(\cfrac{f_*'(x)}{f_*(x)} -
-        \cfrac{g_*'(x)}{g_*(x)}\right)
-        +(\mu_s-\mu) \cfrac{g_*(x)h_*'(x)}{f_*(x)} \Bigl |^2}
-    < \sigma^{-2} \ .
-\end{equation}
-This expression allows us to obtain insights on the stability properties of
-the system in the case in which we are able to compute or estimate 
-the probability distribution function $P(x_*)$.
-
-In order to investigate the different regimes
-of complexity-stability behaviour,
-let us specialize to the power laws case, with 
-\begin{equation}
-    f(x_i)=x_i^{\alpha} \ , \quad g(x_i)=-x_i^{\beta} \ , \quad h(x_i)=x_i^{\gamma} \ .
-\end{equation} 
-Additional proportionality constants can be reabsorbed in the statistics
-of $a$ and by a rescaling of time, moreover we chose to explicitly
-factor out a minus sign in order to more straightforwardly connect
-with competitive models.
-
-We wish to consider explictly cases
-for "May" and "anti-May" regimes.
-An example of the former is given by GLV.
-In order to explore the latter, we 
-need to resort to Eq.~\eqref{eq: random-stability int} and therefore
-to compute the equilibrium probability distribution function
-for the system. This can be achieved in different ways,
-depending among other things on specific constraint that
-one can impose on the system. 
-One approach is reported below, then we specialize to a choice of
-the power laws exponens.
-
-Following, e.g., Ref.~[Roy2019] we can derive from 
-the evolution equation
-\begin{equation}
-    \dot{x_i}=x_i^{\alpha} - x_i^{\beta}\sum_{j}a_{ij}x_j^{\gamma} \, ,
-\label{eq: full abg}
-\end{equation}
-for large $N$, a dynamical mean field theory (DMFT) 
-describing the entire ensemble of $N$ variables
-thorugh the evolution of the representative random variable $x$
-\begin{equation}
-    \dot{x} = x^{\alpha}-a_sx^{\beta+\gamma}-x^{\beta}\left( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\right) \, ,
-\label{eq: dmft}
-\end{equation}
-where $a_s$ is a random variable with the statistics of $a_{ii}$,
-$\langle . \rangle$ stands for expectation value
-and $\eta$ is a gaussian variable
-with zero mean and correlation 
-$\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$. 
-The averages and correlations have to be computed self-consistently.
-If the system admits an equilibrium (i.e. is feasible), for 
-$t\to\infty$ each of the initial $N$ degrees 
-of freedom will reach a final value, 
-and the effective random variable describing their
-also become time-independent,
-$\lim_{t\to\infty}x(t)=x_*$ and we can write
-\begin{equation}
-    x_*^{\alpha-\beta} - a_s x_*^{\gamma}= \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\right) \, ,
-\end{equation} 
-where $\xi$ is a standard normal random variable (i.e. $\xi\sim\mathcal{N}(0,1)$).
-The equation above can be solved for $x_*$ for specific values of
-$(\alpha-\beta)/\gamma$, or, in general, if we set $a_s=0$, which amounts to include
-all the self-interaction in the nonlinear produciton term
-$x_*^{\alpha}$. In the following, in studying a system which
-follows in the 'anti-May' phase, we show that numerical results
-obtained for the case in which the diagonal part of the
-interaction matrix is non-vanishing and equal to the
-off-diagonal part are well approximated by the results obtained
-setting $a_s=0$.
-In this case the stationary solution is given by 
-\begin{equation} \label{eq: cavity solution}
-    x_* = \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
-\end{equation}
-
-The equilibrium probability distribution function for $x_*$, $P(x_*)$,
-can be obtained as the pushforward of the distribition of $\xi$.
-The change of measure
-$\left|d \xi/d x_*\right|=|\alpha-\beta|x_*^{\alpha-\beta-1}/\sigma \sqrt{N\langle x_*^{2\gamma}\rangle}$,
-leads to
-\begin{equation}\label{eq: dist general}
-    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2 N\langle x_*^{2\gamma}\rangle}}
-    \exp{\left\{-\frac{(x_*^{\alpha-\beta}-\mu N\langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
-\end{equation}
-where the expectations have to be computed self-consistently.
-
-Let us focus to the case $\alpha=1$, $\beta=1.5$ and $\gamma=1$, with
-$a_s=0$ in order to leverage the equilibrium solution in Eq.~\eqref{eq: dist general}.
-In this case the first and second moment to be self-consistently
-computed formally diverges. This is not truly porblematic in practice
-because we are dealing with large, but \emph{finite} populations.
-Following~[Cui2020,Us2023] and assuming
-$\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$,
-we can Taylor expand to the first-order the cavity 
-solution~\eqref{eq: cavity solution} and obtain a
-gaussian approximation to Eq.~\eqref{eq: dist general}
-with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and
-$\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.
-This is plotted against simulations in 
-Fig.~\ref{fig: cavity sol.}, alongside the exact solution~\eqref{eq: dist general} 
-where integration for the moments has been performed up to a maximum
-value above which no equilibirum value should be found for a given $N$.
-Simulations correspond to $a_s=\mu$, showing that the choice
-of vanishing self-interaction for the analytical results
-is suited to describe this case as well.
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
-    \caption{...}
-    \label{fig: cavity sol.}
-\end{figure}
-
-\subsection{Simulations}
-
-Equipped with the equilibirum distribution, we can compute
-the critical value of heterogeneity 
-$\sigma_c$, for a given $N$ and $\mu$,
-defined by taking the equality in 
-Eq.~\eqref{eq: random-stability int}. 
-For $\sigma$ above this value the system becomes
-unstable. In Fig~\ref{fig: stability line + sims}
-the result is portrayed in the $\sigma \sqrt{N} - \mu N$ plane
-compared with results from simulations where stability is defined
-as stable, full coexistence of the system.
-Everything else left fixed, an increase in the number of
-degrees of freedom $N$ will move the system along a square-root
-trajectory, bringing it into the stable regime or keeping it
-in it if already stable. This confirms we are in the
-'anti-May' phase.
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
-    \caption{...}
-    \label{fig: stability line + sims}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=.45\textwidth]{figs/alpha-beta.pdf}
-    \caption{...}
-    \label{fig: alpha-beta}
-\end{figure}
-
-\section{Conclusions}
-
-\end{document}
-%
-% ****** End of file apssamp.tex ******

### 2023-04-29 papers/intro.tex
reorganize and edit draft

Few mathematical arguments have influenced biological thinking like May's prediction that large, complex ecosystems cannot be stable~\cite{May1972}.
It is a perplexing conclusion.
On the one hand, May's mathematical argument is simple and seemingly model-free, suggesting universal applicability, also outside biology \cite{Haldane2011, Moran2019}.
On the other hand, it is clear that at least some large, complex systems are stable---else which regularities would biology by studying in the first place? Empirically, the opposite relationship between complexity and stability seems to hold across ecosystems: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, exctinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}. 
The tension between May's theoretical argument and  observation is at the center of the longlasting "diversity-stability debate" in ecology \cite{McCann2000, Loreau2022}.

May's argument can be summarized as follows.
A system with $N$ populations $x_i$ is stable is one that can be characterized by an equilibrium point $\mathbf x^*$.
Near that equilibrium $\mathbf x = \mathbf x^* + \delta \mathbf x$, the dynamics of the system is described by linear equations $d(\delta \mathbf x)/dt = A (\delta \mathbf x)$, and the stability of these equations requires that all eigenvalues of $A$ have negative real part.
But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions (with zero 
mean and variance $\sigma^2$), and $-I$ corresponds to stabilizing self-interactions on some natural timescale, 
then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability".
(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, incomplete connectivity, to correlated interactions, see e.g. \cite{allesina2015stability}.)

In ecology, many authors have sought to ease the tension between May's 
prediction and empirical observation by invoking effects not captured by 
dynamical systems with random 
coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}. In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: 
random dynamical systems do not necessarily imply that 
stability decreases with dimensionality and interaction 
strength---the opposite behavior is also possible, 
without the need for special or additional structure.

### 2023-04-29 papers/model.tex
reorganize and edit draft

Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-interaction of a population $i$, while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations. That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp.
$a_{ji} < 0$) implies a positive (resp.
negative) effect of $j$ on the growth of $i$.
In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''.
This rather general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models for complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008,Skinner2012} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.

The classic generalized Lotka-Volterra (GLV) competition model studied recently by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear. 
However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level~\cite{Hatton2015,Hatton2023}.
From a different perspective, it has been suggested recently that predator-prey interactions can be modelled with a square root law ($g(x) \sim h(x) \sim x^{1/2}$)~\cite{Barbier2021,Mazzarisi2023}.

### 2023-04-29 papers/random-general.tex
reorganize and edit draft

Consider now Eq.~\eqref{dynamics} with i.i.d. random interactions $a_{ij}$ drawn from a distribution with mean $\mu$ and standard deviation $\sigma$; we assume the diagonal elements $a_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) but the same standard deviation $\sigma$.

We compute the Jacobian matrix at equilibrium
\begin{align}
    J_{ij}^* & = a_{ij}g(x_i^*)h'(x_j^*) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
    J_{ii}^* & = f'(x_i^*) - \frac{g'(x_i^*)f(x_i^*)}{g(x_i^*)} - a_{ii}g(x_i^*)h'(x_j^*) \ , \label{eq: jac diag}
\end{align}
where we used $\sum_{j}a_{ij}g(x_i^*)h(x_j^*)=-f(x_i^*)/g(x_i^*)$.
In order to investigate the spectral properties of $J^*$, 
we follow Stone~\cite{Stone2018} and use a result in random matrix theory~\cite{Ahmadian2015} which generalize the classical `circular law'~\cite{Potters2020}.

In Ref.~\cite{Ahmadian2015} Ahmadian \emph{et al.} consider matrices of the form $M + LSR$, where $M$,  
$L$ and $R$ are deterministic matrices, and $S$ is a random matrix with i.i.d. coefficients, zero mean and variance $\sigma^2$.
They show that eigenvalues of large matrices of this form are contained in the complex domain with equation
$\textrm{Tr}[(M_\zeta M_\zeta^\dagger)^{-1}]\geq \sigma^{-2}$, where $M_\zeta = L^{-1}(\zeta I - M)R^{-1}$ and $\zeta\in\mathbb{C}$.
If $L$, $R$ and $M$ are all diagonal $N\times N$ matrices (identified with the vector of their diagonal elements), this condition simplifies to
\begin{equation}
    \sum_{i=1}^N\frac{(L_{i}R_{i})^2}{ \vert \zeta - M_{i}\vert^2 }\geq \frac{1}{\sigma^{2}} \ .
\label{eq: domain}
\end{equation}

To use this result, we decompose the interaction matrix as $a = \mu \mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Using rank-one perturbation theory, we can show that the $\mathbf{1}$ term does not impact stability properties and can be neglected \cite{Stone2018}.
The Jacobian in Eqs.~\eqref{eq: jac off-diag} and \eqref{eq: jac diag} thus takes the form $M + LJR$ with diagonal matrices
\begin{equation}
    M = f(\mathbf x^*)\psi(\mathbf x^*),\quad L = g(\mathbf x^*), \quad R = h'(\mathbf x^*),
\end{equation}
where $\psi(x)$ is the function defined by the LHS of \eqref{homogeneous-general}. \red{Assuming all $M_i < 0$}, the domain \eqref{eq: domain} containing the eigenvalues of $J^*$ first touches the right half-plane at $\zeta = 0$, 
hence linear stability requires
\begin{equation}
    \sum_{i=1}^N \left(\frac{g(x_i^*)h'(x_i^*)}{f(x_i^*)\psi(x_i^*)}\right)^2
    < \frac{1}{\sigma^{2}}
    \label{eq: random-stability}
\end{equation}
This expression generalizes \eqref{homogeneous-general} to random interactions.

Straightforward results follow from condition \eqref{eq: random-stability}.
For GLV models, i.e. $f(x_i)=x_i$, $g(x_i)=x_i$ and $h(x_i)=x_i$, \eqref{eq: random-stability} is independent of the equilibria $x^*$ and we recover the linear stability condition
$\sigma\sqrt{N} < (\mu_s-\mu)$.
More generally, any model satisfying
\begin{equation}\label{eq: equiv class glv}
    \left(\cfrac{f'(x_i^*)}{f(x_i^*)} -
        \cfrac{g'(x_i^*)}{g(x_i^*)}\right)\Bigl /
        \cfrac{g(x_i^*)h'(x_i^*)}{f(x_i^*)} = c \ ,
\end{equation}
for some constant $c$ has the same simple linear stability condition as GLV models, namely $\sigma\sqrt{N} < (\mu_s-\mu) +c$.

When the dependence on $\mathbf{x}_*$ does not cancel out, it is not possible to obtain an explicit complexity-stability relationship. 
However, introducing the distribution of equilibrium $P(x^*)$ with corresponding equilibrium average $\langle\cdot\rangle$, we can write the stability condition as
\begin{equation}\label{eq: random-stability int}
    N\Big\langle\frac{gh'}{f\psi}\Big\rangle < \sigma^{-2} \ .
\end{equation}
We now discuss how $P(x_*)$ can be estimated from the statistics of interactions, focusing on the power-law case.

### 2023-04-29 scripts/mu-sigma-plot.jl
housekeeping

@@ -1,0 +1,78 @@
+using DrWatson; @quickactivate
+
+using Revise, Glob
+foreach(includet, glob("*.jl", srcdir()))
+
+using DelimitedFiles
+using Plots, LaTeXStrings
+
+dist = "normal"
+P = Dict{Symbol,Any}(
+    :scaled => true,
+    :S => 100,
+    :μ => 0.01:0.02:1.0,
+    :σ => 0.01:0.02:0.4,
+    :α => 0.5,
+    :β => 1.0,
+    :γ => 1.0,
+    :extinction_threshold => 1e-4,
+    :dist => dist,
+    :N => 100,
+    :seed => rand(UInt)
+)
+
+
+Φ = readdlm(datadir("mu-sigma-gamma.csv"), ',')
+sublinear = heatmap(
+        P[:μ],
+        P[:σ],
+        Φ,
+        dpi=500,
+        grid=false,
+        xlabel=L"\mu N",
+        ylabel=L"\sigma \sqrt{N}",
+        #c=cgrad(:Greys, rev=true)
+    )
+
+begin
+    Φ = readdlm(datadir("mu-sigma-normal.csv"), ',')
+    sublinear = heatmap(
+        P[:μ],
+        P[:σ],
+        Φ,
+        dpi=500,
+        grid=false,
+        xlabel=L"\mu N",
+        ylabel=L"\sigma \sqrt{N}",
+        #c=cgrad(:Greys, rev=true)
+    )
+end
+
+    #= critical line for gaussian approximation =#
+    μ_critical, σ_critical = critical_line_gauss(P, μ_range=(0.01:0.25:1.26))
+    plot!(μ_critical, σ_critical,
+        labels=false,
+        linewidth=4,
+        linecolor=:red,
+        grid=false,
+        yticks=[0, 0.6, 1.2],
+        xticks=[0, 0.6, 1.2],
+        size=(600, 400),
+    )
+
+    #= critical line =#
+    μ_critical, σ_critical = critical_line(P, μ_range=(0.01:0.2:1.25),
+        n_max=1e1)
+    plot!(μ_critical, σ_critical,
+        labels=false,
+        linewidth=4,
+        linecolor=:green,
+        grid=false,
+        ylims=[0.0, 1.2],
+        xlims=[0.0, 1.2],
+        yticks=[0, 0.6, 1.2],
+        xticks=[0, 0.6, 1.2],
+        size=(600, 400),
+    )
+    savefig(plotsdir("mu-sigma-$dist"))
+end

### 2023-04-29 scripts/single-run.jl

@@ -18,14 +18,6 @@ p = Dict{Symbol,Any}(
     :seed => rand(UInt)
 );
 
-evolve!(p; trajectory=true);
-plot(p[:trajectory], label=false)
-p[:converged]
-p[:equilibrium]
-
-
-p[:converged]
-
 #= trajectories =#
 plot(
     p[:trajectory],

### 2023-04-29 papers/abstract.tex
abstract

Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
However, this seemingly general conclusion is not supported by empirical observation: from cells to biomes, biological systems are large, complex---and, by and large, stable.
In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 
Using a non-linear generalization of the Lotka-Volterra model, we show that there are in fact two kinds of complexity-stability relationships in disordered dynamical systems: 
if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
Our result shows that May's principle that "complexity begets instability" is not a general property of complex systems; instead, it is a property of a class of weakly cross-regulated complex systems.

### 2023-04-30 papers/beyond-may.tex
housekeeping
edit draft

@@
 
-\date{\today}% It is always \today, today,
-             %  but any date may be explicitly specified
+\date{\today}
 
 \begin{abstract}
-
+\input{abstract.tex}
 \end{abstract}
@@
 
-\section*{Introduction}
-
-\section*{Results}
+\paragraph*{\bf Introduction.}
+\input{intro.tex}
 
@@
 
-\paragraph*{\bf Random interactions: general reults.}
-\input{random-general.tex}
-\paragraph*{\bf Random interactions: power laws case.}
-\input{random-power.tex}
+\paragraph*{\bf Random interactions: stability condition.}
+\input{stability.tex}
 
-\section{Conclusion}
+\paragraph*{\bf Random interactions: equilibrium distribution.}
+\input{equilibrium.tex}
+
+\paragraph*{\bf Conclusion.}
 \input{conclusion.tex}
 
+\medskip
+
 \begin{acknowledgments}
-We gratefully acknowledge multiple discussions with Ian Hatton, Matthieu Barbier and Ada Altieri on ecological power laws and the diversity-stability debate. 
+We thank Ian Hatton, Matthieu Barbier and Ada Altieri for many productive discussions on ecological power laws and the diversity-stability debate. 
 \end{acknowledgments}
 
-\clearpage
+\medskip
 
-\bibliography{general-stability}
+\bibliography{beyond-may}
 \bibliographystyle{unsrt}

### 2023-04-30 papers/conclusion.tex
edit draft

@@
-The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle.
-
-We reconsidered the relationship between diversity and stability in complex systems, a classic problem since May's 1973's puzzling prediction that "complexity begets instability".
-
-Considering dynamical models more general than the classical Lotka-Volterra model, we found that the relation can in fact go both ways:  depending on the relative convexity of self- and cross-interactions,
+The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and in other fields. 
+In this letter, we have showed that the condition $\sigma\sqrt{N}< \mu_s - \mu$ (popularized by May's slogan ``complexity begets instability") does not provide a complete picture of the relationship between complexity and stability.
+In particular, we have seen that generalized Lotka-Volterra model, often cited in support of May's prediction, corresponds to a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
+In models where the self-dynamics is more strongly regulated than the cross-interactions ($\alpha < \beta$, which might be more appropriate for ecological modeling \cite{Hatton2023}), the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.

### 2023-04-30 papers/equilibrium.tex
edit draft

Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=-x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time. 
We factor out a minus sign in order to connect more directly with competition models.)

\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
    \caption{The equilibrium distribution with power-law self- and cross-interactions is accurately reproduced by 
    Eq.\eqref{eq: dist general} and its Gaussian approximation.
    For simulations (in green) we used $\alpha=1$, $\beta=1.5$,
    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=5\cdot 10^{-3}$.}
    \label{fig: cavity sol.}
\end{figure}

Following e.g. Ref.~\cite{Roy2019}, we can derive from the evolution equation \eqref{dynamics} a dynamical mean field theory (DMFT) 
which describes the ensemble of $N\gg 1$ variables
by means of a single, self-consistent representative stochastic trajectory
\begin{equation}
    \dot{x} = x^{\alpha}-A_sx^{\beta+\gamma}-x^{\beta}\big( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\eta$ is a Gaussian noise with zero mean and correlation $\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$.

\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases described in the
    case of uniform interactions are robust with respect to random interactions.
    Here we simulated system~\eqref{eq: full abg} for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates 
    and evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the lower triangle because we have $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes bigger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$ will destabilize the system, because $\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}

At equilibrium, we can write
\begin{equation}
    x_*^{\alpha-\beta} - A_s x_*^{\gamma}= \big( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
\end{equation} 
where $\xi$ is now a standard normal random variable.
The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or if $A_s=0$.
(In the following we show that neglecting $A_s$ even when it is not zero does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
\begin{equation} \label{eq: cavity solution}
    x_* = \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
\end{equation}
The equilibrium distribution $P(x_*)$ can then be obtained as the pushforward of the distribution of $\xi$: 
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2 N\langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}-\mu N\langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
where the expectation values must be computed self-consistently.

Let us focus on the choice 
$\alpha=1$, $\beta=3/2$ and $\gamma=1$, with
$A_s=0$. 
In this case the first and second moments formally diverge.
This is not truly problematic because we are dealing with large, but \emph{finite} populations.
Following~\cite{Cui2020,Hatton2023} and assuming
$\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$,
we can expand the  
solution \eqref{eq: cavity solution} to first order in 
$\sigma \sqrt{N\langle x_*^2\rangle}$, and 
approximate \eqref{eq: dist general} with
a Gaussian distribution with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and $\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.

The result is plotted against simulations in 
Fig.~\ref{fig: cavity sol.}. We show both the exact 
solution~\eqref{eq: dist general} 
(where integration for the moments has been performed up to a maximum
value above which no equilibrium value should 
be found) and the Gaussian approximation.
Simulations correspond to $A_s=\mu$, showing that neglecting the $A_s$ term in analytical expressions does not introduce a significant error.

Equipped with the equilibrium distribution $P(x^*)$, we can compute the maximal heterogeneity 
$\sigma_c$, for a given $N$ and $\mu$, for which the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ holds. In Fig.~\ref{fig: stability line + sims} the result in portrayed in the $(\sigma \sqrt{N},\mu N)$ plane,
compared with results from simulations (where stability is defined
as stable, full, coexistence of the system).
Everything else fixed, an increase in the number of
degrees of freedom $N$ will move the system along a square-root
trajectory in that plane, eventually entering the stable region where $\sigma_c > \sigma$. 
In other words, we are in the
``anti-May" phase.
(For comparison the stability boundary for GLV is a horizontal line in the $(\sigma \sqrt{N},\mu N)$ plane, defined by $\sigma_c\sqrt{N} = \mu_s$ \cite{bunin2017ecological}.)

Figure~\ref{fig: alpha-beta} shows results of simulations 
for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, 
at fixed $\mu N$ and $\gamma = 1$, 
defined as the value above which
full stable coexistence is less probable than $50\%$.
(In these simulations we use $\mu_s = \mu$, which would never be stable in the GLV model.) This plot illustrates the transition between a ``May" phase for $\alpha \geq \beta$ (where stability decreases with $N$, $\mu$, and $\sigma$) and an ``anti-May" phase for $\alpha < \beta$, even in the presence of heterogeneous interactions.

### 2023-04-30 papers/homogeneous.tex
edit draft

@@
-Under what condition does \eqref{dynamics} admit a linearly stable equilibrium? 
-We begin by considering the simple case where all self-interactions have the same strength $a_{ii} = a_{\textrm{s}}$, and similarly for cross-interactions $a_{ij} = a_{\textrm{c}}$ ($i\neq j$).
-Under these assumptions, a straightforward computation shows that a homogeneous equilibrium $\mathbf x_*$ will be linearly stable if  
-\begin{equation}\label{homogeneous-general}
-   \frac{f'(x^*)}{f(x^*)} - \frac{g'(x^*)}{g(x^*)} + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g(x^*)h'(x^*)}{f(x^*)} < 0.
+Under what condition does \eqref{dynamics} admit a (linearly) stable equilibrium? 
+We begin with the simple case where all self-interactions have the same strength $A_{ii} = \mu_s$, and similarly for cross-interactions $A_{ij} = \mu$ ($i\neq j$).
+Defining 
+\begin{equation}
+    \psi(x) \equiv (\mu - \mu_s) -  \left(\frac{f'(x)}{f(x)} - \frac{g'(x)}{g(x)}\right)\frac{f(x)}{g(x)h'(x)},
+    \label{eq: psi}
 \end{equation}
-Thus, stability of an equilibrium depends on the relative strength of self- and cross-interactions ($a_{\textrm{s}} - a_{\textrm{c}}$), but also on the relative growth rate of the production and response functions near that equilibrium ($f'_*/f_* - g'_*/g_*$).
-With power laws, \eqref{homogeneous-general} evaluates to 
+we find that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*) > 0$.
+This condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'_*/f_* - g'_*/g_*$).
+
+With power laws, the condition $\psi(x^*)<0$ evaluates to 
 \begin{equation}
-    (\alpha - \beta)(N-1) < \gamma(a_{\textrm{s}}/a_{\textrm{c}}- 1) - (\alpha - \beta)(a_{\textrm{s}}/a_{\textrm{c}}),
+    (\alpha - \beta)(N-1) < \gamma(\mu_s/\mu- 1) - (\alpha - \beta)(\mu_s/\mu),
 \end{equation}
@@
 \begin{itemize}
-    \item If $\alpha = \beta$, stability requires $a_{\textrm{s}} > a_{\textrm{c}}$, i.e.
+    \item If $\alpha = \beta$, stability requires $\mu_s > \mu$, i.e.
     self-interactions must be stronger than cross-interactions.
@@
     \item If $\alpha > \beta$, stability places an upper bound on $N$: the more complex the system, the less likely to be stable.
-    We can call this "May" behavior.
+    We can call this ``May" behavior.
     \item If $\alpha < \beta$, stability places an lower bound on $N$: the more complex the system, the more likely to be stable.
-    This is "anti-May" behavior.
+    This is ``anti-May" behavior.
 \end{itemize}

### 2023-04-30 papers/intro.tex
edit draft

@@
 Few mathematical arguments have influenced biological thinking like May's prediction that large, complex ecosystems cannot be stable~\cite{May1972}.
-It is a perplexing conclusion.
-On the one hand, May's mathematical argument is simple and seemingly model-free, suggesting universal applicability, also outside biology \cite{Haldane2011, Moran2019}.
+It is, on the face of it, a perplexing conclusion.
+On the one hand, May's mathematical argument is simple and seemingly model-free, suggesting universal applicability, also beyond biology \cite{Haldane2011, Moran2019}.
 On the other hand, it is clear that at least some large, complex systems are stable---else which regularities would biology by studying in the first place? 
-Empirically, the opposite relationship between complexity and stability seems to hold across ecosystems: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, exctinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}.
-The tension between May's theoretical argument and  observation is at the center of the longlasting "diversity-stability debate" in ecology \cite{McCann2000, Loreau2022}.
+Empirically, the opposite relationship between complexity and stability seems to hold: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, exctinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}.
+The tension between May's theoretical argument and  observation is at the center of the longstanding "diversity-stability debate" in ecology \cite{McCann2000, Loreau2022}.
 
@@
 This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability".
-(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, incomplete connectivity, to correlated interactions, see e.g. \cite{allesina2015stability}.)
+(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, incomplete connectivity, or correlated interactions, see e.g. \cite{allesina2015stability}.)
 
-In ecology, many authors have sought to ease the tension between May's 
-prediction and empirical observation by invoking effects not captured by 
-dynamical systems with random 
-coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}.
-In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete:
-random dynamical systems do not necessarily imply that 
-stability decreases with dimensionality and interaction 
-strength---the opposite behavior is also possible, 
-without the need for special or additional structure.
+In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}.
+In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: random dynamical systems do not necessarily imply that stability decreases with dimensionality and interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
+For an in-depth discussion of ecological implications of our result, see \cite{Mazzarisi2023}.

### 2023-04-30 papers/model.tex
housekeeping
edit draft

@@
 \begin{equation}\label{dynamics}
-    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j) \, .
+    \dot{x}_i = f(x_i) + \sum_{j}A_{ij}g(x_i)h(x_j) \, .
 \end{equation}
 Here $f(x_i)$ represents the self-interaction of a population $i$, while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations.
-That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp.
-$a_{ji} < 0$) implies a positive (resp.
+That interaction is weighted by a coefficient $A_{ij}$, such that $A_{ij} > 0$ (resp.
+$A_{ji} < 0$) implies a positive (resp.
 negative) effect of $j$ on the growth of $i$. 
-In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''.
+We all assume all of $f$, $g$ and $h$ are increasing functions, and refer to $f$ as the ``production function'' and $g$ as the ``response function''.
+
 This rather general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models for complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008,Skinner2012}
@@
 For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level~\cite{Hatton2015,Hatton2023}.
-From a different perspective, it has been suggested recently that predator-prey interactions can be modelled with a square root law ($g(x) \sim h(x) \sim x^{1/2}$)~\cite{Barbier2021,Mazzarisi2023}.
+In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law ($g(x) \sim h(x) \sim x^{1/2}$)~\cite{Barbier2021,Mazzarisi2023}.

### 2023-04-30 papers/random-gen.tex
edit draft

--- 2023-04-29 papers/random-general.tex
@@
 \begin{equation}
-    M = f(\mathbf x^*)\psi(\mathbf x^*),\quad L = g(\mathbf x^*), \quad R = h'(\mathbf x^*),
+    M_i = f(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*).
 \end{equation}
-where $\psi(x)$ is the function defined by the LHS of \eqref{homogeneous-general}.
 \red{Assuming all $M_i < 0$}, the domain \eqref{eq: domain} containing the eigenvalues of $J^*$ first touches the right half-plane at $\zeta = 0$, 
@@
 
-Straightforward results follow from condition \eqref{eq: random-stability}.
-For GLV models, i.e. $f(x_i)=x_i$, $g(x_i)=x_i$ and $h(x_i)=x_i$, \eqref{eq: random-stability} is independent of the equilibria $x^*$ and we recover the linear stability condition
+For GLV models, i.e. $f(x_i)=x_i$, $g(x_i)=x_i$ and $h(x_i)=x_i$, \eqref{eq: random-stability} is independent of the equilibria $x^*$, and we recover the linear stability condition
 $\sigma\sqrt{N} < (\mu_s-\mu)$.
@@
 When the dependence on $\mathbf{x}_*$ does not cancel out, it is not possible to obtain an explicit complexity-stability relationship. 
-However, introducing the distribution of equilibrium $P(x^*)$ with corresponding equilibrium average $\langle\cdot\rangle$, we can write the stability condition as
+However, introducing the distribution of equilibrium $P(x^*)$ and replacing sums with integrals in the large $N$ limit, we can write the stability condition as
 \begin{equation}\label{eq: random-stability int}
-    N\Big\langle\frac{gh'}{f\psi}\Big\rangle < \sigma^{-2} \ .
+    N\int dx^* frac{g(x^*)h'(x^*)}{f(x^*)\psi(x^*)} < \sigma^{-2} \ .
 \end{equation}

### 2023-04-30 papers/stability.tex
edit draft

We now consider the case of random interactions.
Speficially, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$;
We assume the diagonal elements $A_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) but the same standard deviation $\sigma$.

We compute the Jacobian matrix at equilibrium
\begin{align}
    J_{ij}^* & = A_{ij}g(x_i^*)h'(x_j^*) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
    J_{ii}^* & = f'(x_i^*) - g'(x_i^*)f(x_i^*)/g(x_i^*) + A_{ii}g(x_i^*)h'(x_j^*) \ , \label{eq: jac diag}
\end{align}
where we used $\sum_{j}A_{ij}h(x_j^*)=-f(x_i^*)/g(x_i^*)$.
In order to investigate the spectral properties of $J^*$, 
we follow Stone~\cite{Stone2018} in using a recent generalization of the circular law in random matrix theory~\cite{Ahmadian2015}.

In Ref.~\cite{Ahmadian2015}, Ahmadian \emph{et al.} consider matrices of the form $M + LSR$, where $M$,  
$L$ and $R$ are deterministic matrices, and $S$ is a random matrix with i.i.d. coefficients, zero mean and variance $\sigma^2$.
They show that eigenvalues of large matrices of this form are contained in the complex domain $\mathcal{D} = \{\zeta \in \mathbb{C},\, \textrm{Tr}[(\Psi(\zeta) \Psi(\zeta)^\dagger)^{-1}]\geq \sigma^{-2}\}$, where $\Psi(\zeta) = L^{-1}(M-\zeta I)R^{-1}$. If $L$, $R$ and $M$ are all diagonal $N\times N$ matrices, the equation of $\mathcal{D}$ simplifies to 
\begin{equation}
    \sum_{i=1}^N\Big\vert\frac{M_{ii} - \zeta}{L_{ii}R_{ii}}\Big\vert^{-2}\geq \sigma^{-2}.
\label{eq: domain}
\end{equation}

To use this result, we decompose the interaction matrix as $A = \mu\mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian \eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
\begin{equation}
    M_i = g(x_i^*)h'(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*),
\end{equation}
where $\psi(x)$ is the function defined in \eqref{eq: psi}.

\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane, will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system \eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed obtained using DMFT and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=1.5$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations.}
    \label{fig: stability line + sims}
\end{figure}

Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part). For this, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from \eqref{eq: domain}:
\begin{equation}
    \sum_{i=1}^N \Big(\frac{\sigma}{\psi(x_i^*)}\Big)^{2}
    < 1.
    \label{eq: random-stability}
\end{equation}
In the GLV model, $\psi(x) = \mu-\mu_s$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}. With more general power laws, the dependence on equilibrium values $x_i^*$ does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition \eqref{eq: random-stability}.

### 2023-04-30 src/dynamics.jl
solve last issue

@@ -4,7 +4,6 @@ include(srcdir("utils.jl"))
 using DifferentialEquations
 using Random, Distributions
 using LinearAlgebra
-<<<<<<< HEAD
 
 ## dynamical system
 function F(x, p)
@@ -34,38 +33,6 @@ function evolve!(p; trajectory=false)
     if !haskey(p, :seed)
         p[:seed] = 1234
     end
-=======
-
-## dynamical system
-function F(x, p)
-    x .= ppart.(x)
-    pop = x.^p[:α]
-    comm = - x.^p[:β] .* (p[:A] * x.^p[:γ])
-    return pop + comm
-end
-
-function F!(f, x, p)
-    x .= ppart.(x)
-    pop = x.^p[:α]
-    comm = - x.^p[:β] .* (p[:A] * x.^p[:γ])
-    f .= pop + comm
-end
-
-#= solving =#
-
-MAX_TIME = 1e5
-MAX_ABUNDANCE = 1e3
-
-converged(ϵ = 1e-4) = TerminateSteadyState(ϵ)
-blowup(max_abundance = MAX_ABUNDANCE) = DiscreteCallback((u, t, integrator) -> maximum(u) > max_abundance, terminate!)
-
-function evolve!(p; trajectory=false)
-
-    if !haskey(p, :seed)
-        p[:seed] = rand(UInt32)
-    end
-
->>>>>>> onofrio
     if !haskey(p, :rng)
         p[:rng] = MersenneTwister(p[:seed])
     end
@@ -78,11 +45,7 @@ function evolve!(p; trajectory=false)
 
     pb = ODEProblem(
         ODEFunction(
-<<<<<<< HEAD
             F!
-=======
-            (f, x, p, t) -> F!(f, x, p) 
->>>>>>> onofrio
         ),
         p[:x0], 
         (0.0, MAX_TIME),
@@ -110,11 +73,7 @@ function evolve!(p; trajectory=false)
 end
 
 function add_initial_condition!(p)
-<<<<<<< HEAD
     p[:x0] = rand(p[:rng], Uniform(1, 10), p[:S])
-=======
-    p[:b0] = rand(p[:rng], Uniform(2, 10), p[:S])
->>>>>>> onofrio
 end
 
 function add_interactions!(p)

### 2023-05-02 papers/equilibrium.tex
update draft

@@
-Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=-x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
+\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=-x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
 (Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time. 
@@
 \begin{figure}[t!]
-    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
-    \caption{The equilibrium distribution with power-law self- and cross-interactions is accurately reproduced by 
-    Eq.\eqref{eq: dist general} and its Gaussian approximation.
-    For simulations (in green) we used $\alpha=1$, $\beta=1.5$,
-    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=5\cdot 10^{-3}$.}
-    \label{fig: cavity sol.}
+    \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
+    \caption{For a system in the ``anti-May" phase, an increase in $N$,
+    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane, will be stabilizing rather than destabilizing.
+    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}.
+Stability is defined as full stable coexistence.
+The red line is computed obtained using DMFT and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
+    Parameters values are $\alpha=1$, $\beta=3/2$,
+    $\gamma=1$, $N=50$ and $10$ replicates for the simulations.}
+    \label{fig: stability line + sims}
 \end{figure}
 
-Following e.g. Ref.~\cite{Roy2019}, we can derive from the evolution equation \eqref{dynamics} a dynamical mean field theory (DMFT) 
+Following e.g. Ref.~\cite{Roy2019}, we can derive from the evolution equation~\eqref{dynamics} a dynamical mean field theory (DMFT) 
 which describes the ensemble of $N\gg 1$ variables
@@
     case of uniform interactions are robust with respect to random interactions.
-    Here we simulated system~\eqref{eq: full abg} for a fixed value of
+    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
     $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates 
@@
 we can expand the  
-solution \eqref{eq: cavity solution} to first order in 
+solution~\eqref{eq: cavity solution} to first order in 
 $\sigma \sqrt{N\langle x_*^2\rangle}$, and 
-approximate \eqref{eq: dist general} with
+approximate~\eqref{eq: dist general} with
 a Gaussian distribution with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and $\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.

### 2023-05-02 papers/model.tex
update draft

@@
-Consider the dynamical system in $N$ variables
+\emph{Model.---}Consider the dynamical system in $N$ variables
 \begin{equation}\label{dynamics}
@@
 negative) effect of $j$ on the growth of $i$. 
-We all assume all of $f$, $g$ and $h$ are increasing functions, and refer to $f$ as the ``production function'' and $g$ as the ``response function''.
+We assume all of $f$, $g$ and $h$ are increasing functions, and refer to $f$ as the ``production function'' and $g$ as the ``response function''.
 

### 2023-05-02 papers/stability.tex
update draft

@@
-We now consider the case of random interactions.
+\emph{Random interactions: stability condition.---}We now consider the case of random interactions.
 Speficially, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$;
@@
 with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
-Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian \eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
+Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian~\eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
 \begin{equation}
@@
 \end{equation}
-where $\psi(x)$ is the function defined in \eqref{eq: psi}.
+where $\psi(x)$ is the function defined in~\eqref{eq: psi}.
 
 \begin{figure}[t!]
-    \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
-    \caption{For a system in the ``anti-May" phase, an increase in $N$,
-    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane, will be stabilizing rather than destabilizing.
-    Here we show results from the numerical resolution of the dynamical system \eqref{dynamics}.
-Stability is defined as full stable coexistence.
-The red line is computed obtained using DMFT and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
-    Parameters values are $\alpha=1$, $\beta=1.5$,
-    $\gamma=1$, $N=50$ and $10$ replicates for the simulations.}
-    \label{fig: stability line + sims}
+    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
+    \caption{The equilibrium distribution with power-law self- and cross-interactions is accurately reproduced by 
+    Eq.~\eqref{eq: dist general} and its Gaussian approximation.
+    For simulations (in green) we used $\alpha=1$, $\beta=3/2$,
+    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=5\cdot 10^{-3}$.}
+    \label{fig: cavity sol.}
 \end{figure}
@@
 Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part).
-For this, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from \eqref{eq: domain}:
+For this, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from~\eqref{eq: domain}:
 \begin{equation}
@@
 In the GLV model, $\psi(x) = \mu-\mu_s$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}.
-With more general power laws, the dependence on equilibrium values $x_i^*$ does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition \eqref{eq: random-stability}.
+In general, the dependence on equilibrium values $x_i^*$ (which, in turn, can depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.

### 2023-05-03 papers/conclusion.tex
clean folder

@@
-The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and in other fields. 
+\emph{Conclusion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and in other fields. 
 In this letter, we have showed that the condition $\sigma\sqrt{N}< \mu_s - \mu$ (popularized by May's slogan ``complexity begets instability") does not provide a complete picture of the relationship between complexity and stability.

### 2023-05-03 papers/homogeneous.tex

@@
-Under what condition does \eqref{dynamics} admit a (linearly) stable equilibrium? 
+\emph{Homogeneous interactions.---}Under what condition does~\eqref{dynamics} admit a (linearly) stable equilibrium? 
 We begin with the simple case where all self-interactions have the same strength $A_{ii} = \mu_s$, and similarly for cross-interactions $A_{ij} = \mu$ ($i\neq j$).
@@
 we find that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*) > 0$.
-This condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'_*/f_* - g'_*/g_*$).
+This condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'(x_*)/f(x_*) - g'(x_*)/g(x_*)$).
 

### 2023-05-03 papers/intro.tex

@@
-Few mathematical arguments have influenced biological thinking like May's prediction that large, complex ecosystems cannot be stable~\cite{May1972}.
+\emph{Introduction.---}Few mathematical arguments have influenced biological thinking like May's prediction that large, complex ecosystems cannot be stable~\cite{May1972}.
 It is, on the face of it, a perplexing conclusion.
 On the one hand, May's mathematical argument is simple and seemingly model-free, suggesting universal applicability, also beyond biology \cite{Haldane2011, Moran2019}.
-On the other hand, it is clear that at least some large, complex systems are stable---else which regularities would biology by studying in the first place? 
+On the other hand, it is clear that at least some large, complex systems are stable---else which regularities would biology be studying in the first place? 
 Empirically, the opposite relationship between complexity and stability seems to hold: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, exctinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}.
@@
 May's argument can be summarized as follows.
-A system with $N$ populations $x_i$ is stable is one that can be characterized by an equilibrium point $\mathbf x^*$.
+Consider a system with $N$ populations $x_i$ characterized by an equilibrium point $\mathbf x^*$.
 Near that equilibrium $\mathbf x = \mathbf x^* + \delta \mathbf x$, the dynamics of the system is described by linear equations $d(\delta \mathbf x)/dt = A (\delta \mathbf x)$, and the stability of these equations requires that all eigenvalues of $A$ have negative real part.
@@
 In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}.
-In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: random dynamical systems do not necessarily imply that stability decreases with dimensionality and interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
-For an in-depth discussion of ecological implications of our result, see \cite{Mazzarisi2023}.
+In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: in random dynamical systems stability does not necessarily decrease with dimensionality and interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
+For an in-depth discussion of ecological implications of our result, see \cite{Hatton2023}.

### 2023-05-03 papers/abstract.tex
captions and refs abbreviated

@@
 Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
-However, this seemingly general conclusion is not supported by empirical observation: from cells to biomes, biological systems are large, complex---and, by and large, stable.
+Unfortunately, this general conclusion is not supported by empirical observation: from cells to biomes, biological systems are large, complex---and, by and large, stable.
 In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 
@@
 if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
-Our result shows that May's principle that "complexity begets instability" is not a general property of complex systems; instead, it is a property of a class of weakly cross-regulated complex systems.
+Our result shows that May's principle that "complexity begets instability" is not a general property of complex systems; instead, it is a property of a subclass of weakly cross-regulated disordered systems.

### 2023-05-03 papers/beyond-may.tex
captions and refs abbreviated

@@
 
+\captionsetup{justification=Justified}
+
 \begin{document}
@@
 \author{Onofrio Mazzarisi}
-\affiliation{Department of Ecology and Evolutionary Biology, University of Kansas, Haworth Hall, 
-1200 Sunnyside Avenue
-Lawrence, KS 66045, USA}
+\affiliation{Department of Ecology and Evolutionary Biology, University of Kansas,
+Higuchi Hall, 2101 Constant Avenue, Lawrence, KS 66047, USA}
 
@@
 
-\paragraph*{\bf Introduction.}
+%Introduction
+\paragraph*{}
 \input{intro.tex}
 
-\paragraph*{\bf Model.}
+%Model
+\paragraph*{}
 \input{model.tex}
 
-\paragraph*{\bf Homogeneous interactions.}
+%Homogeneous interactions
+\paragraph*{}
 \input{homogeneous.tex}
 
-\paragraph*{\bf Random interactions: stability condition.}
+%Random interactions: stability condition
+\paragraph*{}
 \input{stability.tex}
 
-\paragraph*{\bf Random interactions: equilibrium distribution.}
+%Random interactions: equilibrium distribution
+\paragraph*{}
 \input{equilibrium.tex}
 
-\paragraph*{\bf Conclusion.}
+%Conclusion
+\paragraph*{}
 \input{conclusion.tex}
@@
 \begin{acknowledgments}
-We thank Ian Hatton, Matthieu Barbier and Ada Altieri for many productive discussions on ecological power laws and the diversity-stability debate. 
+We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate. 
 \end{acknowledgments}

### 2023-05-03 papers/model.tex
captions and refs abbreviated

@@
 
-This rather general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models for complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008,Skinner2012}
+This rather general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models for complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008}
 to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.

### 2023-05-03 papers/intro.tex
captions and refs abbreviated

@@
 On the other hand, it is clear that at least some large, complex systems are stable---else which regularities would biology be studying in the first place? 
-Empirically, the opposite relationship between complexity and stability seems to hold: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, exctinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}.
+In fact, empirical observation suggests the opposite relationship between complexity and stability: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, extinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}.
 The tension between May's theoretical argument and observation is at the center of the longstanding "diversity-stability debate" in ecology \cite{McCann2000, Loreau2022}.
@@
 then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
-This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability".
+This places a sharp constraint on both diversity $N$ and interaction strength $\sigma$, often referred to as "complexity begets instability".
 (This argument generalizes to $\langle B_{ij}\rangle \neq 0$, incomplete connectivity, or correlated interactions, see e.g. \cite{allesina2015stability}.)
@@
 In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}.
-In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: in random dynamical systems stability does not necessarily decrease with dimensionality and interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
+In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: in random dynamical systems, stability does not necessarily decrease with dimensionality and interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
 For an in-depth discussion of ecological implications of our result, see \cite{Hatton2023}.

### 2023-05-05 papers/beyond-may.tex
add link to repo
update draft

@@
 \begin{acknowledgments}
-We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate. 
+We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate and Vadim Karatayev and Daniel Reuman for feedbacks on the manuscript.
+O.M. was partly supported by NSF BIO OCE grant 2023473.
 \end{acknowledgments}

### 2023-05-05 papers/conclusion.tex
update draft

--- 2023-04-30 papers/conclusion.tex
@@
-The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and in other fields. 
+\emph{Discussion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and in other fields. 
 In this letter, we have showed that the condition $\sigma\sqrt{N}< \mu_s - \mu$ (popularized by May's slogan ``complexity begets instability") does not provide a complete picture of the relationship between complexity and stability.
 In particular, we have seen that generalized Lotka-Volterra model, often cited in support of May's prediction, corresponds to a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
-In models where the self-dynamics is more strongly regulated than the cross-interactions ($\alpha < \beta$, which might be more appropriate for ecological modeling \cite{Hatton2023}), the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.
+In models where the self-dynamics is more strongly regulated than the cross-interactions, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.
+
+GLV is special also for another reason: for homogenous interaction, dependence on the number of degrees of freedom is lost in the stability condition.
+We showed that in general this dependence is present, and its effect on the complexity-stablity property is preserved for heterogeneous interaction. 
+The simple condition $\psi(x^*)g(x^*)h'(x^*)>0$ can therefore be employed to gain informations on systems desceribed by Eq.~\eqref{dynamics}.
+Future work can be done in the direction of extending the equations for testing complexity-stability for cases in which we can not decompose the two-body interactions, or to include higher order interaction, generalizing, e.g., the analysis of recent results~\cite{Gibbs2022} based on GLV.
+
+Our general conclusion that for large random dynamical systems, an increase in dimensionality can lead to either stabilization or destabilization, without invoking specific system topology, bares significant implications for the study and control of complex systems.

### 2023-05-05 papers/equilibrium.tex
update draft

@@
-\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=-x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
-(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time. 
-We factor out a minus sign in order to connect more directly with competition models.)
-
+\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
+(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time.)
 \begin{figure}[t!]
-    \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
+    \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity-increase.pdf}
     \caption{For a system in the ``anti-May" phase, an increase in $N$,
-    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane, will be stabilizing rather than destabilizing.
+    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane (green dashed line), will be stabilizing rather than destabilizing.
     Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}.
@@
     Parameters values are $\alpha=1$, $\beta=3/2$,
-    $\gamma=1$, $N=50$ and $10$ replicates for the simulations.}
+    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. For the line representing the increase in $N$ we fixed $\mu=0.1$ and $\sigma=0.75$.}
     \label{fig: stability line + sims}
 \end{figure}
-
 Following e.g. Ref.~\cite{Roy2019}, we can derive from the evolution equation~\eqref{dynamics} a dynamical mean field theory (DMFT) 
@@
 where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\eta$ is a Gaussian noise with zero mean and correlation $\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$. 
-
 \begin{figure}[t!]
@@
 \end{figure}
-
 At equilibrium, we can write
@@
 
-Let us focus on the choice 
-$\alpha=1$, $\beta=3/2$ and $\gamma=1$, with
-$A_s=0$. 
+Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
 In this case the first and second moments formally diverge.
 This is not truly problematic because we are dealing with large, but \emph{finite} populations.
-Following~\cite{Cui2020,Hatton2023} and assuming
-$\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$,
-we can expand the  
-solution~\eqref{eq: cavity solution} to first order in 
-$\sigma \sqrt{N\langle x_*^2\rangle}$, and 
-approximate~\eqref{eq: dist general} with
-a Gaussian distribution with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and $\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.
+Following~\cite{Cui2020,Hatton2023} and assuming $\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$, we can expand the solution~\eqref{eq: cavity solution} to first order in $\sigma \sqrt{N\langle x_*^2\rangle}$, and approximate~\eqref{eq: dist general} with a Gaussian distribution with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and $\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.
 
-The result is plotted against simulations in 
-Fig.~\ref{fig: cavity sol.}. We show both the exact 
-solution~\eqref{eq: dist general} 
-(where integration for the moments has been performed up to a maximum
-value above which no equilibrium value should 
-be found) and the Gaussian approximation.
+The result is plotted against simulations in Fig.~\ref{fig: cavity sol.}.
+We show both the exact solution~\eqref{eq: dist general} (where integration for the moments has been performed up to a maximum value above which no equilibrium value should be found) and the Gaussian approximation.
 Simulations correspond to $A_s=\mu$, showing that neglecting the $A_s$ term in analytical expressions does not introduce a significant error.
 
-Equipped with the equilibrium distribution $P(x^*)$, we can compute the maximal heterogeneity 
-$\sigma_c$, for a given $N$ and $\mu$, for which the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ holds.
-In Fig.~\ref{fig: stability line + sims} the result in portrayed in the $(\sigma \sqrt{N},\mu N)$ plane,
-compared with results from simulations (where stability is defined
-as stable, full, coexistence of the system).
-Everything else fixed, an increase in the number of
-degrees of freedom $N$ will move the system along a square-root
-trajectory in that plane, eventually entering the stable region where $\sigma_c > \sigma$. 
-In other words, we are in the
-``anti-May" phase.
-(For comparison the stability boundary for GLV is a horizontal line in the $(\sigma \sqrt{N},\mu N)$ plane, defined by $\sigma_c\sqrt{N} = \mu_s$ \cite{bunin2017ecological}.)
+Equipped with the equilibrium distribution $P(x^*)$, we can compute the maximal heterogeneity $\sigma_c$, for a given $N$ and $\mu$, for which the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ holds.
+A convenient way to test the complexity-stability properties of the system, consists in studying how this condition compart the $(\sigma \sqrt{N},\mu N)$ plan, conventionally used to summarize the properties of large random dynamical systems~\cite{bunin2017ecological}, in stable and unstable parameter regions.
+Everything else fixed, an increase in the number of degrees of freedom $N$ will move the system along a square-root trajectory in that plane and we can test if this has stabilizing effects on the system or if the opposite is true.
+In Fig.~\ref{fig: stability line + sims} the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ is portrayed in the $(\sigma \sqrt{N},\mu N)$ plane, compared with results from simulations (where stability is defined as stable, full, coexistence of the system).
+An increase in the number of degrees of freedom $N$ will eventually bring the system in the stable region, or preserve stability if already stable. 
+In other words, we are in the ``anti-May" phase.
+(For comparison the stability boundary for GLV is a horizontal line in the $(\sigma \sqrt{N},\mu N)$ plane, defined by $\sigma_c\sqrt{N} = \mu_s$~\cite{bunin2017ecological}.)
 
-Figure~\ref{fig: alpha-beta} shows results of simulations 
-for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, 
-at fixed $\mu N$ and $\gamma = 1$, 
-defined as the value above which
-full stable coexistence is less probable than $50\%$.
+We conclude by testing the robustness of our findings in the case of heterogeneous interactions in the paradigmatic case of power laws.
+Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$, defined as the value above which full stable coexistence is less probable than $50\%$.
 (In these simulations we use $\mu_s = \mu$, which would never be stable in the GLV model.) This plot illustrates the transition between a ``May" phase for $\alpha \geq \beta$ (where stability decreases with $N$, $\mu$, and $\sigma$) and an ``anti-May" phase for $\alpha < \beta$, even in the presence of heterogeneous interactions.

### 2023-05-05 papers/homogeneous.tex
update draft

@@
 \begin{equation}
-    \psi(x) \equiv (\mu - \mu_s) -  \left(\frac{f'(x)}{f(x)} - \frac{g'(x)}{g(x)}\right)\frac{f(x)}{g(x)h'(x)},
+    \psi(x) \equiv (\mu_s - \mu) -  \left(\frac{f'(x)}{f(x)} - \frac{g'(x)}{g(x)}\right)\frac{f(x)}{g(x)h'(x)},
     \label{eq: psi}
 \end{equation}
-we find that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*) > 0$.
-This condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'(x_*)/f(x_*) - g'(x_*)/g(x_*)$).
+we find that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*)g(x^*)h'(x^*) > 0$.
+This reduces to $\psi(x^*)>0$ if we assume (and we will, for simplicity, in the following) $g$ and $h$ are positive, increasing functions.
+This condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'(x^*)/f(x^*) - g'(x^*)/g(x^*)$).
 
-With power laws, the condition $\psi(x^*)<0$ evaluates to 
+With power laws, the condition $\psi(x^*)>0$ evaluates to 
 \begin{equation}

### 2023-05-05 papers/model.tex
add link to repo
update draft

@@
 \begin{equation}\label{dynamics}
-    \dot{x}_i = f(x_i) + \sum_{j}A_{ij}g(x_i)h(x_j) \, .
+    \dot{x}_i = f(x_i) - \sum_{j}A_{ij}g(x_i)h(x_j) \, .
 \end{equation}
@@
 That interaction is weighted by a coefficient $A_{ij}$, such that $A_{ij} > 0$ (resp.
-$A_{ji} < 0$) implies a positive (resp.
-negative) effect of $j$ on the growth of $i$. 
-We assume all of $f$, $g$ and $h$ are increasing functions, and refer to $f$ as the ``production function'' and $g$ as the ``response function''.
+$A_{ji} < 0$) implies a negative (resp.
+positive) effect of $j$ on the growth of $i$. Non-zero diagonal elements $A_{ii}$ accounts for additional self-interaction with the same shape as cross-ones. 
+We refer to $f$ as the ``production function'' and $g$ as the ``response function''.
 
@@
 to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
-
 The classic generalized Lotka-Volterra (GLV) competition model studied recently by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear. 

### 2023-05-05 papers/stability.tex
update draft

@@
 \begin{align}
-    J_{ij}^* & = A_{ij}g(x_i^*)h'(x_j^*) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
-    J_{ii}^* & = f'(x_i^*) - g'(x_i^*)f(x_i^*)/g(x_i^*) + A_{ii}g(x_i^*)h'(x_j^*) \ , \label{eq: jac diag}
+    J_{ij}^* & = - A_{ij}g(x_i^*)h'(x_j^*) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
+    J_{ii}^* & = f'(x_i^*) - g'(x_i^*)f(x_i^*)/g(x_i^*) - A_{ii}g(x_i^*)h'(x_j^*) \ , \label{eq: jac diag}
 \end{align}
-where we used $\sum_{j}A_{ij}h(x_j^*)=-f(x_i^*)/g(x_i^*)$.
+where we used $\sum_{j}A_{ij}h(x_j^*)=f(x_i^*)/g(x_i^*)$.
 In order to investigate the spectral properties of $J^*$, 
@@
 \begin{equation}
-    M_i = g(x_i^*)h'(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*),
+    M_i = - g(x_i^*)h'(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*),
 \end{equation}
 where $\psi(x)$ is the function defined in~\eqref{eq: psi}. 
-
 \begin{figure}[t!]
@@
 \end{figure}
-
 Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part).
@@
 \end{equation}
-In the GLV model, $\psi(x) = \mu-\mu_s$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}.
+In the GLV model, $\psi(x) = \mu_s - \mu$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}.
 In general, the dependence on equilibrium values $x_i^*$ (which, in turn, can depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.

### 2023-05-07 papers/beyond-may.tex
review and edit changes

--- 2023-05-03 papers/beyond-may.tex
@@
 \begin{acknowledgments}
-We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate. 
+We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate, and Vadim Karatayev and Daniel Reuman for feedback on the manuscript.
+Funding for this work was provided by the Alexander von Humboldt Foundation in the framework of the Sofja Kovalevskaja Award endowed by the German Federal Ministry of Education and Research.
+O.M. was partly supported by NSF BIO OCE grant 2023473.
 \end{acknowledgments}

### 2023-05-07 papers/conclusion.tex
review and edit changes

--- 2023-04-30 papers/conclusion.tex
@@
-The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and in other fields. 
+\emph{Discussion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and in other fields. 
 In this letter, we have showed that the condition $\sigma\sqrt{N}< \mu_s - \mu$ (popularized by May's slogan ``complexity begets instability") does not provide a complete picture of the relationship between complexity and stability.
 In particular, we have seen that generalized Lotka-Volterra model, often cited in support of May's prediction, corresponds to a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
-In models where the self-dynamics is more strongly regulated than the cross-interactions ($\alpha < \beta$, which might be more appropriate for ecological modeling \cite{Hatton2023}), the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.
+In models where self-interactions grow faster than cross-interactions, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.
+
+Future work can look at the case where two-body interactions are not separable, or include higher order interaction to generalize the results in Ref. \cite{Gibbs2022} based on the GLV model.

### 2023-05-07 papers/homogeneous.tex
review and edit changes

@@
 \end{equation}
-we find that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*)g(x^*)h'(x^*) > 0$.
-This reduces to $\psi(x^*)>0$ if we assume (and we will, for simplicity, in the following) $g$ and $h$ are positive, increasing functions.
-This condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'(x^*)/f(x^*) - g'(x^*)/g(x^*)$).
+we find that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*)g(x^*)h'(x^*) > 0$, or $\psi(x^*)>0$ if we assume that $g$ and $h$ are positive, increasing functions.
+Note that this stability condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'(x^*)/f(x^*) - g'(x^*)/g(x^*)$).
 

### 2023-05-07 papers/intro.tex
review and edit changes

@@
 In fact, empirical observation suggests the opposite relationship between complexity and stability: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, extinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}.
-The tension between May's theoretical argument and observation is at the center of the longstanding "diversity-stability debate" in ecology \cite{McCann2000, Loreau2022}.
+The tension between May's theoretical argument and observation is at the center of the longstanding ``diversity-stability debate" in ecology \cite{McCann2000, Loreau2022}.
 
@@
 then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
-This places a sharp constraint on both diversity $N$ and interaction strength $\sigma$, often referred to as "complexity begets instability".
+This places a sharp constraint on both diversity $N$ and interaction strength $\sigma$, often referred to as ``complexity begets instability".
 (This argument generalizes to $\langle B_{ij}\rangle \neq 0$, incomplete connectivity, or correlated interactions, see e.g. \cite{allesina2015stability}.)

### 2023-05-07 papers/equilibrium.tex

@@
 
+Equipped with the equilibrium distribution $P(x^*)$, we can compute, for given $N$ and $\mu$, the maximal heterogeneity $\sigma_c$ consistent with the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ holds.
+
 Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
@@
 
-Equipped with the equilibrium distribution $P(x^*)$, we can compute the maximal heterogeneity $\sigma_c$, for a given $N$ and $\mu$, for which the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ holds.
-A convenient way to test the complexity-stability properties of the system, consists in studying how this condition compart the $(\sigma \sqrt{N},\mu N)$ plan, conventionally used to summarize the properties of large random dynamical systems~\cite{bunin2017ecological}, in stable and unstable parameter regions.
-Everything else fixed, an increase in the number of degrees of freedom $N$ will move the system along a square-root trajectory in that plane and we can test if this has stabilizing effects on the system or if the opposite is true.
-In Fig.~\ref{fig: stability line + sims} the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ is portrayed in the $(\sigma \sqrt{N},\mu N)$ plane, compared with results from simulations (where stability is defined as stable, full, coexistence of the system).
-An increase in the number of degrees of freedom $N$ will eventually bring the system in the stable region, or preserve stability if already stable. 
-In other words, we are in the ``anti-May" phase.
-(For comparison the stability boundary for GLV is a horizontal line in the $(\sigma \sqrt{N},\mu N)$ plane, defined by $\sigma_c\sqrt{N} = \mu_s$~\cite{bunin2017ecological}.)
+The properties of large random dynamical systems is often portrayed in often portrayed in the $(\sigma \sqrt{N},\mu N)$ plane \cite{bunin2017ecological}.
+Everything else fixed, an increase of the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane.
+In the GLV model, the boundary between the stable and unstable parameter regions is the horizontal line $\sigma_c\sqrt{N} = \mu_s$ \cite{bunin2017ecological}, hence stability is never possible at large $N$.
 
-We conclude by testing the robustness of our findings in the case of heterogeneous interactions in the paradigmatic case of power laws.
-Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$, defined as the value above which full stable coexistence is less probable than $50\%$.
-(In these simulations we use $\mu_s = \mu$, which would never be stable in the GLV model.) This plot illustrates the transition between a ``May" phase for $\alpha \geq \beta$ (where stability decreases with $N$, $\mu$, and $\sigma$) and an ``anti-May" phase for $\alpha < \beta$, even in the presence of heterogeneous interactions.
+In Fig.~\ref{fig: stability line + sims} we plot the stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ for $\alpha = 1$, $\beta = 3/2$ derived from \eqref{eq: dist general} (solid line) together with results from simulations (shading).
+Here, because the boundary between the stable and unstable phases is a straight line with positive slope, increasing $N$ will eventually bring the system in the stable region.
+This behavior corresponds to the ``anti-May" phase defined previously.
+
+Finally, Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$.
+(Numerically, we define $\sigma_c\sqrt{N}$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations we use $\mu_s = \mu$, which would never be stable in the GLV model.
+Fig.~\ref{fig: alpha-beta} illustrates the transition between a ``May" phase for $\alpha \geq \beta$ and an ``anti-May" phase for $\alpha < \beta$, showing that our analytical results are robust to heterogeneous interactions.

### 2023-05-07 papers/stability.tex

@@
 In the GLV model, $\psi(x) = \mu_s - \mu$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}.
-In general, the dependence on equilibrium values $x_i^*$ (which, in turn, can depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.
+In general, the dependence on equilibrium values $x_i^*$ (which in turn depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.

### 2023-05-14 papers/abstract.tex
more edits of the text and caption

@@
 Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
-Unfortunately, this general conclusion is not supported by empirical observation: from cells to biomes, biological systems are large, complex---and, by and large, stable.
+However, this general conclusion is not supported by empirical observation: from cells to biomes, biological systems are large, complex---and, by and large, stable.
 In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 

### 2023-05-14 papers/conclusion.tex
more edits of the text and caption

@@
 In this letter, we have showed that the condition $\sigma\sqrt{N}< \mu_s - \mu$ (popularized by May's slogan ``complexity begets instability") does not provide a complete picture of the relationship between complexity and stability.
-In particular, we have seen that generalized Lotka-Volterra model, often cited in support of May's prediction, corresponds to a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
+In particular, we have seen that the generalized Lotka-Volterra model, often cited in support of May's general prediction, exemplifies a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
 In models where self-interactions grow faster than cross-interactions, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.

### 2023-05-14 papers/homogeneous.tex
more edits of the text and caption

@@
 \end{equation}
-we find that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*)g(x^*)h'(x^*) > 0$, or $\psi(x^*)>0$ if we assume that $g$ and $h$ are positive, increasing functions.
-Note that this stability condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'(x^*)/f(x^*) - g'(x^*)/g(x^*)$).
+an elementary calculation shows that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*)g(x^*)h'(x^*) > 0$, or $\psi(x^*)>0$ if we assume that $g$ and $h$ are positive, increasing functions.
+This stability condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'(x^*)/f(x^*) - g'(x^*)/g(x^*)$).
 

### 2024-03-16 papers/old/equilibrium.tex
organize for resubmission

--- 2023-05-07 papers/equilibrium.tex
@@
     Parameters values are $\alpha=1$, $\beta=3/2$,
-    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. For the line representing the increase in $N$ we fixed $\mu=0.1$ and $\sigma=0.75$.}
+    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. The green line was plotted for $\mu=0.1$ and $\sigma=0.75$.}
     \label{fig: stability line + sims}
@@
     \includegraphics[width=.45\textwidth]{figs/alpha-beta.pdf}
-    \caption{The ``May" and ``anti-May" phases described in the
-    case of uniform interactions are robust with respect to random interactions.
+    \caption{The ``May" and ``anti-May" phases derived in the
+    case of uniform interactions ($\sigma = 0$) are robust with respect to random interactions ($\sigma > 0$).
     Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
-    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates 
-    and evaluated $\sigma_c$, defined here as
+    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates,  
+    and we evaluated $\sigma_c$, defined here as
     the value above which full stable coexistence is less probable than 50\%,
     at varying $\alpha$ and $\beta$.
-    No amount of heterogeneity is allowed in the lower triangle because we have $\mu=\mu_s$, while an increasing value of $\sigma_c$
-    is found as $\beta$ becomes bigger than $\alpha$.
+    No amount of heterogeneity is allowed in the bottom right triangle because we have $\mu=\mu_s$, while an increasing value of $\sigma_c$
+    is found as $\beta$ becomes larger than $\alpha$.
     We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
@@
 
-Equipped with the equilibrium distribution $P(x^*)$, we can compute, for given $N$ and $\mu$, the maximal heterogeneity $\sigma_c$ consistent with the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ holds.
+Equipped with the equilibrium distribution $P(x^*)$, we can compute, for given $N$ and $\mu$, the maximal heterogeneity $\sigma_c$ compatible with the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$.
 
@@
 
-The properties of large random dynamical systems is often portrayed in often portrayed in the $(\sigma \sqrt{N},\mu N)$ plane \cite{bunin2017ecological}.
-Everything else fixed, an increase of the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane.
+The properties of large random dynamical systems is often portrayed in the $(\sigma \sqrt{N},\mu N)$ plane \cite{bunin2017ecological}.
+Everything else being fixed, an increase of the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane.
 In the GLV model, the boundary between the stable and unstable parameter regions is the horizontal line $\sigma_c\sqrt{N} = \mu_s$ \cite{bunin2017ecological}, hence stability is never possible at large $N$.

### 2024-03-16 papers/old/intro.tex
organize for resubmission

--- 2023-05-07 papers/intro.tex
@@
 
-May's argument can be summarized as follows.
-Consider a system with $N$ populations $x_i$ characterized by an equilibrium point $\mathbf x^*$.
+May's argument can be summarized as follows \cite{May1972}.
+Consider a system with $N$ populations $x_i$, characterized by an equilibrium point $\mathbf x^*$.
 Near that equilibrium $\mathbf x = \mathbf x^* + \delta \mathbf x$, the dynamics of the system is described by linear equations $d(\delta \mathbf x)/dt = A (\delta \mathbf x)$, and the stability of these equations requires that all eigenvalues of $A$ have negative real part.
@@
 This places a sharp constraint on both diversity $N$ and interaction strength $\sigma$, often referred to as ``complexity begets instability".
-(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, incomplete connectivity, or correlated interactions, see e.g. \cite{allesina2015stability}.)
+(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, incomplete connectivity, or correlated interactions \cite{allesina2015stability}.)
 
-In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}.
+Many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}.
 In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: in random dynamical systems, stability does not necessarily decrease with dimensionality and interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
-For an in-depth discussion of ecological implications of our result, see \cite{Hatton2023}.
+For an in-depth discussion of the ecological implications of our result, see \cite{Hatton2023}.

### 2024-03-16 papers/old/model.tex
organize for resubmission

--- 2023-05-05 papers/model.tex
@@
 \end{equation}
-Here $f(x_i)$ represents the self-interaction of a population $i$, while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations.
+Here $f(x_i)$ represents the self-dynamics of a population $i$ (growth and self-regulation), while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations.
 That interaction is weighted by a coefficient $A_{ij}$, such that $A_{ij} > 0$ (resp.
 $A_{ji} < 0$) implies a negative (resp.
-positive) effect of $j$ on the growth of $i$. Non-zero diagonal elements $A_{ii}$ accounts for additional self-interaction with the same shape as cross-ones. 
+positive) effect of $j$ on the growth of $i$. Non-zero diagonal elements $A_{ii}$ accounts for additional self-interactions with the same functional form as cross-interactions. 
 We refer to $f$ as the ``production function'' and $g$ as the ``response function''.
 
-This rather general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models for complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008}
+This general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models of complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008}
 to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
 The classic generalized Lotka-Volterra (GLV) competition model studied recently by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear.
+
 However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
@@
 Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
-For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level~\cite{Hatton2015,Hatton2023}.
-In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law ($g(x) \sim h(x) \sim x^{1/2}$)~\cite{Barbier2021,Mazzarisi2023}.
+For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities~\cite{Hatton2015,Hatton2023}.
+In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law, i.e. $g(x) \sim h(x) \sim x^{1/2}$~\cite{Barbier2021,Mazzarisi2023}.

### 2024-03-16 papers/response.md

# Referee reports

## Report of Referee A -- LN17679/Mazzarisi

The authors discuss the stability properties of a system of
first-order differential equations describing the evolution of coupled
populations in an ecosystem. The model is intended to show that May's
prediction that large complex systems cannot be stable is not a
general property of such systems. The model is actually a modification
of the original model, considered by R.M. May, with the difference
that the authors replace linear functions with power functions in
population dynamics. They show that for a certain combination of power
function exponents, the system becomes more stable as its size
increases, unlike the original May model.

The claim is probably true, but unfortunately it does not describe a
generic situation, because it is very hard to imagine that the
exponents alpha, beta, gamma used in the model are identical for all
populations. Perhaps the model would be more interesting if it
included a spectrum of different exponents, described, for example, by
a probabilistic law.

Minor remarks:

The way of writing the matrix elements in Equations (7) is somewhat
confusing. For clarity, the Kronecker delta can be used on the right
side of these equations.

The mean field equations depend on the expected values of <x_*^\gamma>
and <x_*^{2\gamma}>. The self-consistency condition requires that the
probability distribution (12) has exactly the same expected values.
This should be discussed in detail.

In summary, I do not think that the paper meets PRL criteria. However,
I would recommend the article for submission as a regular article to
Physical Review E, after the authors have softened their conclusions
and improved the discussion of the mean-field approximation.

---

## Report of Referee B -- LN17679/Mazzarisi

Dear editor,
I have reviewed the manuscript entitled "Beyond May: Complexity-stability relationships in
disordered dynamical systems" intended for publication in Physical Review Letters. This study
seeks to address the well-known May's stability-complexity conundrum by incorporating

recent developments in random matrix theory into a non-linear generalization of the Lotka-
Volterra (GLV) model.

Summary:

The authors challenge Robert May's conclusion that large, complex systems inherently lack
stability by using a non-linear generalization of the GLV model, incorporating power-law
functions into the interactions between species and species growth rate. Their findings
suggest that the instability in complex systems as posited by May is not a universal
characteristic but rather specific to a subclass of weakly cross-regulated systems.
General Comments:
The manuscript could be of interest to researchers focusing on ecological systems, especially
those applying disordered system approaches. However, the modelling framework and
ecological implications presented in the paper are predicated on assumptions that I find
somewhat tenuous. The authors' arguments do not convincingly justify the generalization of
the GLV model, as well as the main ecological finding is quite consequential given the model’s
hypotheses. Moreover, I have few major technical and conceptual concerns that from my
point of view should be fixed before publication in any journal.
The relevance of the work appears confined to a niche.

#### Major Concerns:

Modelling Framework Justification:

• The rationale behind the proposed generalization of the GLV model is claimed to be
"natural" based on physical and biological grounds, specifically using power laws.
Nevertheless, the literature extensively demonstrates species abundance growth
being well-described by logistic equations or oscillatory predator-prey equations, that
indeed are linear (except for the carrying capacity term). The justifications provided
are weak, with the provided examples failing to convincingly represent the complex
dynamics of natural systems, such as forest biomass distribution. If we think to a forest
there are various heights of trees, and in general the biomass in not uniformly
distributed, and thus also the growth is not necessary constrained by the surface. The
sublinear scaling of the “production function”, f, if this is really the right conclusion
based on the references of the manuscript, must be the result of a cooperative
behavior of the interacting system and as such it should not appear as hand-added
like the authors do. Furthermore, in reference 27 the sublinear scaling with k=3/4
refers to the scaling of the single individual growth rather than to a community.

Feasibility Condition Neglected:

• The authors do not seem to consider the condition for the feasibility of the fixed points
(x* >= 0), which is critical. This oversight, especially evident in Equations 11 and 12,
i.e., Eq. 12 or its Gaussian approximation are normalized from 0 or from -infty to infty?
I think this is a very relevant issue calling into question the appropriateness of the
results.

Lack of Proper Scaling:

• There is no proper scaling of the interactions, so the thermodynamic limit is not well
defined. And this is not discussed or compared with other models in the literature
applying similar disputable choice (see E Mallmin, A Traulsen, S De Monte - arXiv
preprint arXiv:2306.11031, 2023). Relatedly, the comparison with GLV with quenched
disorder (e.g. “The properties of large random dynamical systems is often portrayed
in the ( sigma sqrt(N) , mu N ) plane [25] is not appropriate as such model scales the
interactions strengths with the system size.

Gaussian Approximation:

• The justification for employing a Gaussian approximation is unclear, given that
Equation 12 should be solved self-consistently. Ecologically, Gaussian species

abundance distributions (SADs) are not meaningful due to the prevalence of heavy-
tailed SADs in natural ecosystems.

#### Minor Comments:

Homogenous Case and Stability:

• The manuscript concludes that stability requires stronger self-interactions than cross-
interactions when alpha equals beta. While this is true that self-interactions promote

stability also for the Lotka-Volterra model, any negative mean species interaction
(indicating mutualism) in the homogeneous case leads to instability, irrespective of
system size. Thus, comparing this to the GLV model seems a stretch.

Notation:

• Using S instead of N for species richness would align better with ecological
conventions.

Discussion Quality:

• Given the issues raised, the discussion section of the manuscript needs substantial
improvement to convincingly argue the paper's contributions to the field.
In light of these concerns, before the authors submit their manuscript elsewhere, I
recommend to undertake significant revisions to address the weaknesses in their modelling
framework, the rationale behind their approach, and the ecological implications of their
findings

---

# How to address

## Reviewer 1

### Major comments

- Run simulations with distributed exponents and add to the SM (to be created).

### Minor comments

- Add Kronecker delta to Eq. (7).
- Address the concern of the reviewer in the SM by studying Eq. (11) through the definition of the Gaussian variable $y=x^{\alpha-\beta}$.

## Reviewer 2

### Major comments

- Point to Science for ecological justifications and in general to answer the first point.
- Point to Science paper for feasibility adding some discussion on this in the main text maybe.
- Maybe adjust something in the text to be more clear that the derivation of the results was obtained with rescaling and then we look at the non rescaled value of $\mu$ and $\sigma$.
- Answer to the reviewer that in this kinf of conceptual work we are not necessarily interested in the realistic form of the SAD, indeed usual GLV papers have the Gaussian SAD prediction. In any case the model can still give non gaussian distribution in the case in which we use random growth rates etc... To conclude, the gaussian approximation is used only to more clearly present the results about complexity-stability but it is not necessary.

### Minor comments

- Just engage with the reviewer discussing mutualism etc, but not change anything necessarily.
- We stick to $N$, we don't want to relate to ecological literature, but more to complex systems. Moreover May used $N$ in his paper.
- Just answer something about all the changes that we made.

---

# Response to reviewer

---

# Letter to editor

### 2024-03-16 papers/old/stability.tex

--- 2023-05-07 papers/stability.tex
@@
 \emph{Random interactions: stability condition.---}We now consider the case of random interactions.
-Speficially, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$;
+Specifically, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$;
 We assume the diagonal elements $A_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) but the same standard deviation $\sigma$.
@@
 Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part).
-For this, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from~\eqref{eq: domain}:
+For this to hold, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from~\eqref{eq: domain}:
 \begin{equation}
@@
 In the GLV model, $\psi(x) = \mu_s - \mu$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}.
-In general, the dependence on equilibrium values $x_i^*$ (which in turn depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.
+In general, however, the dependence on equilibrium values $x_i^*$ (which in turn depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.

### 2024-03-16 papers/submission/beyond-may.tex

\documentclass[
 prl,
 twocolumn,
 amsmath,
 amssymb,
 aps,
]{revtex4-2}

\usepackage{colortbl}
\usepackage{graphicx} 
\usepackage{booktabs}
\usepackage{caption}
\usepackage{notes2bib}
\usepackage{hyperref}
\newcommand{\red}{\textcolor{red}}
\usepackage[justification=raggedright]{caption}

\captionsetup{justification=Justified}

\begin{document}

\title{Beyond May: Complexity-stability relationships \\
in disordered dynamical systems}

\author{Onofrio Mazzarisi}
\affiliation{The Abdus Salam International Centre for Theoretical Physics (ICTP), Strada Costiera 11, 34014 Trieste, Italy}
\affiliation{National Institute of Oceanography and Applied Geophysics (OGS), via Beirut 2, 34014 Trieste, Italy}

\author{Matteo Smerlak}
\affiliation{Laboratoire Biophysique et Evolution, ESPCI ParisTech, 10 rue Vauquelin, 75005 Paris, France}
\affiliation{Capital Fund Management, 23 Rue de l'Université, 75007 Paris, France}

\date{\today}

\begin{abstract}
    Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
    However, this general conclusion is not supported by empirical observation: from cells to biomes, biological systems are large, complex---and, by and large, stable.
    In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 
    Using a non-linear generalization of the Lotka-Volterra model, we show that there are in fact two kinds of complexity-stability relationships in disordered dynamical systems: 
    if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
    Our result shows that May's principle that "complexity begets instability" is not a general property of complex systems; instead, it is a property of a subclass of weakly cross-regulated disordered systems.
\end{abstract}

\maketitle

%Introduction
\paragraph*{}
\emph{Introduction.---}Few mathematical arguments have influenced biological thinking like May's prediction that large, complex ecosystems cannot be stable~\cite{May1972}.
It is, on the face of it, a perplexing conclusion.
On the one hand, May's mathematical argument is simple and seemingly model-free, suggesting universal applicability, also beyond biology \cite{Haldane2011, Moran2019}.
On the other hand, it is clear that at least some large, complex systems are stable---else which regularities would biology be studying in the first place? 
In fact, empirical observation suggests the opposite relationship between complexity and stability: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, extinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}. 
The tension between May's theoretical argument and observation is at the center of the longstanding ``diversity-stability debate" in ecology \cite{McCann2000, Loreau2022}.

May's argument can be summarized as follows \cite{May1972}.
Consider a system with $N$ populations $x_i$, characterized by an equilibrium point $\mathbf x^*$.
Near that equilibrium $\mathbf x = \mathbf x^* + \delta \mathbf x$, the dynamics of the system is described by linear equations $d(\delta \mathbf x)/dt = A (\delta \mathbf x)$, and the stability of these equations requires that all eigenvalues of $A$ have negative real part.
But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions (with zero 
mean and variance $\sigma^2$), and $-I$ corresponds to stabilizing self-interactions on some natural timescale, 
then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
This places a sharp constraint on both diversity $N$ and interaction strength $\sigma$, often referred to as ``complexity begets instability".
(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, incomplete connectivity, or correlated interactions \cite{allesina2015stability}.)

Many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}. 
In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: in random dynamical systems, stability does not necessarily decrease with dimensionality and interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
For an in-depth discussion of the ecological implications of our result, see \cite{Hatton2023}.

%Model
\paragraph*{}
\emph{Model.---}Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) - \sum_{j}A_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-dynamics of a population $i$ (growth and self-regulation), while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations. That interaction is weighted by a coefficient $A_{ij}$, such that $A_{ij} > 0$ (resp.
$A_{ji} < 0$) implies a negative (resp.
positive) effect of $j$ on the growth of $i$. Non-zero diagonal elements $A_{ii}$ accounts for additional self-interactions with the same functional form as cross-interactions. 
We refer to $f$ as the ``production function'' and $g$ as the ``response function''.

This general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models of complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
The classic generalized Lotka-Volterra (GLV) competition model studied recently by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear.

However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities~\cite{Hatton2015,Hatton2023}.
In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law, i.e. $g(x) \sim h(x) \sim x^{1/2}$~\cite{Barbier2021,Mazzarisi2023}.

%Homogeneous interactions
\paragraph*{}
\emph{Homogeneous interactions.---}Under what condition does~\eqref{dynamics} admit a (linearly) stable equilibrium? 
We begin with the simple case where all self-interactions have the same strength $A_{ii} = \mu_s$, and similarly for cross-interactions $A_{ij} = \mu$ ($i\neq j$).
Defining 
\begin{equation}
    \psi(x) \equiv (\mu_s - \mu) -  \left(\frac{f'(x)}{f(x)} - \frac{g'(x)}{g(x)}\right)\frac{f(x)}{g(x)h'(x)},
    \label{eq: psi}
\end{equation}
an elementary calculation shows that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*)g(x^*)h'(x^*) > 0$, or $\psi(x^*)>0$ if we assume that $g$ and $h$ are positive, increasing functions. This stability condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'(x^*)/f(x^*) - g'(x^*)/g(x^*)$).

With power laws, the condition $\psi(x^*)>0$ evaluates to 
\begin{equation}
    (\alpha - \beta)(N-1) < \gamma(\mu_s/\mu- 1) - (\alpha - \beta)(\mu_s/\mu),
\end{equation}
leading to three different regimes:
\begin{itemize}
    \item If $\alpha = \beta$, stability requires $\mu_s > \mu$, i.e.
    self-interactions must be stronger than cross-interactions.
    This is the usual conclusion drawn from the Lotka-Volterra model.
    \item If $\alpha > \beta$, stability places an upper bound on $N$: the more complex the system, the less likely to be stable.
    We can call this ``May" behavior.
    \item If $\alpha < \beta$, stability places an lower bound on $N$: the more complex the system, the more likely to be stable.
    This is ``anti-May" behavior.
\end{itemize}

%Random interactions: stability condition
\paragraph*{}
\emph{Random interactions: stability condition.---}We now consider the case of random interactions.
Specifically, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$;
We assume the diagonal elements $A_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) but the same standard deviation $\sigma$.

We compute the Jacobian matrix at equilibrium
\begin{align}
    J_{ij}^* & = - A_{ij}g(x_i^*)h'(x_j^*) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
    J_{ii}^* & = f'(x_i^*) - g'(x_i^*)f(x_i^*)/g(x_i^*) - A_{ii}g(x_i^*)h'(x_j^*) \ , \label{eq: jac diag}
\end{align}
where we used $\sum_{j}A_{ij}h(x_j^*)=f(x_i^*)/g(x_i^*)$.
In order to investigate the spectral properties of $J^*$, 
we follow Stone~\cite{Stone2018} in using a recent generalization of the circular law in random matrix theory~\cite{Ahmadian2015}.

In Ref.~\cite{Ahmadian2015}, Ahmadian \emph{et al.} consider matrices of the form $M + LSR$, where $M$,  
$L$ and $R$ are deterministic matrices, and $S$ is a random matrix with i.i.d. coefficients, zero mean and variance $\sigma^2$.
They show that eigenvalues of large matrices of this form are contained in the complex domain $\mathcal{D} = \{\zeta \in \mathbb{C},\, \textrm{Tr}[(\Psi(\zeta) \Psi(\zeta)^\dagger)^{-1}]\geq \sigma^{-2}\}$, where $\Psi(\zeta) = L^{-1}(M-\zeta I)R^{-1}$. If $L$, $R$ and $M$ are all diagonal $N\times N$ matrices, the equation of $\mathcal{D}$ simplifies to 
\begin{equation}
    \sum_{i=1}^N\Big\vert\frac{M_{ii} - \zeta}{L_{ii}R_{ii}}\Big\vert^{-2}\geq \sigma^{-2}.
\label{eq: domain}
\end{equation}

To use this result, we decompose the interaction matrix as $A = \mu\mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian~\eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
\begin{equation}
    M_i = - g(x_i^*)h'(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*),
\end{equation}
where $\psi(x)$ is the function defined in~\eqref{eq: psi}. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{cavity.pdf}
    \caption{The equilibrium distribution with power-law self- and cross-interactions is accurately reproduced by 
    Eq.~\eqref{eq: dist general} and its Gaussian approximation.
    For simulations (in green) we used $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=5\cdot 10^{-3}$.}
    \label{fig: cavity sol.}
\end{figure}
Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part). For this to hold, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from~\eqref{eq: domain}:
\begin{equation}
    \sum_{i=1}^N \Big(\frac{\sigma}{\psi(x_i^*)}\Big)^{2}
    < 1.
    \label{eq: random-stability}
\end{equation}
In the GLV model, $\psi(x) = \mu_s - \mu$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}. In general, however, the dependence on equilibrium values $x_i^*$ (which in turn depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.

%Random interactions: equilibrium distribution
\paragraph*{}
\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time.)
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{beta1_5-S50-N10-diversity-increase.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane (green dashed line), will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed obtained using DMFT and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. The green line was plotted for $\mu=0.1$ and $\sigma=0.75$.}
    \label{fig: stability line + sims}
\end{figure}
Following e.g. Ref.~\cite{Roy2019}, we can derive from the evolution equation~\eqref{dynamics} a dynamical mean field theory (DMFT) 
which describes the ensemble of $N\gg 1$ variables
by means of a single, self-consistent representative stochastic trajectory
\begin{equation}
    \dot{x} = x^{\alpha}-A_sx^{\beta+\gamma}-x^{\beta}\big( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\eta$ is a Gaussian noise with zero mean and correlation $\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases derived in the
    case of uniform interactions ($\sigma = 0$) are robust with respect to random interactions ($\sigma > 0$).
    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates,  
    and we evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the bottom right triangle because we have $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes larger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$ will destabilize the system, because $\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}
At equilibrium, we can write
\begin{equation}
    x_*^{\alpha-\beta} - A_s x_*^{\gamma}= \big( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
\end{equation} 
where $\xi$ is now a standard normal random variable.
The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or if $A_s=0$.
(In the following we show that neglecting $A_s$ even when it is not zero does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
\begin{equation} \label{eq: cavity solution}
    x_* = \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
\end{equation}
The equilibrium distribution $P(x_*)$ can then be obtained as the pushforward of the distribution of $\xi$: 
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2 N\langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}-\mu N\langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
where the expectation values must be computed self-consistently.

Equipped with the equilibrium distribution $P(x^*)$, we can compute, for given $N$ and $\mu$, the maximal heterogeneity $\sigma_c$ compatible with the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$.

Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
In this case the first and second moments formally diverge.
This is not truly problematic because we are dealing with large, but \emph{finite} populations.
Following~\cite{Cui2020,Hatton2023} and assuming $\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$, we can expand the solution~\eqref{eq: cavity solution} to first order in $\sigma \sqrt{N\langle x_*^2\rangle}$, and approximate~\eqref{eq: dist general} with a Gaussian distribution with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and $\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.

The result is plotted against simulations in Fig.~\ref{fig: cavity sol.}. We show both the exact solution~\eqref{eq: dist general} (where integration for the moments has been performed up to a maximum value above which no equilibrium value should be found) and the Gaussian approximation. Simulations correspond to $A_s=\mu$, showing that neglecting the $A_s$ term in analytical expressions does not introduce a significant error.

The properties of large random dynamical systems is often portrayed in the $(\sigma \sqrt{N},\mu N)$ plane \cite{bunin2017ecological}. Everything else being fixed, an increase of the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane. In the GLV model, the boundary between the stable and unstable parameter regions is the horizontal line $\sigma_c\sqrt{N} = \mu_s$ \cite{bunin2017ecological}, hence stability is never possible at large $N$.

In Fig.~\ref{fig: stability line + sims} we plot the stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ for $\alpha = 1$, $\beta = 3/2$ derived from \eqref{eq: dist general} (solid line) together with results from simulations (shading). Here, because the boundary between the stable and unstable phases is a straight line with positive slope, increasing $N$ will eventually bring the system in the stable region. This behavior corresponds to the ``anti-May" phase defined previously.

Finally, Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$. (Numerically, we define $\sigma_c\sqrt{N}$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations we use $\mu_s = \mu$, which would never be stable in the GLV model. Fig.~\ref{fig: alpha-beta} illustrates the transition between a ``May" phase for $\alpha \geq \beta$ and an ``anti-May" phase for $\alpha < \beta$, showing that our analytical results are robust to heterogeneous interactions.

%Conclusion
\paragraph*{}
\emph{Discussion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and in other fields. 
In this letter, we have showed that the condition $\sigma\sqrt{N}< \mu_s - \mu$ (popularized by May's slogan ``complexity begets instability") does not provide a complete picture of the relationship between complexity and stability. 
In particular, we have seen that the generalized Lotka-Volterra model, often cited in support of May's general prediction, exemplifies a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
In models where self-interactions grow faster than cross-interactions, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.

Future work can look at the case where two-body interactions are not separable, or include higher order interaction to generalize the results in Ref. \cite{Gibbs2022} based on the GLV model.

\medskip

The Julia code used for simulations is available at \url{www.github.com/msmerlak/beyond-may}.

\medskip

\begin{acknowledgments}
We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate, and Vadim Karatayev and Daniel Reuman for feedback on the manuscript.
Funding for this work was provided by the Alexander von Humboldt Foundation in the framework of the Sofja Kovalevskaja Award endowed by the German Federal Ministry of Education and Research.
O.M. was partly supported by NSF BIO OCE grant 2023473.
\end{acknowledgments}

\medskip

\bibliography{beyond-may}
\bibliographystyle{unsrt}

\end{document}

### 2024-03-16 papers/response.md
Update response.md

@@
 
+## Overall
+
+- Try to pitch the paper from a slightly different point of view, more "honest" and aware of what we already did in the Science and putting more stress on what is new/clearer in this letter.
+
 ## Reviewer 1
@@
 
-## Reviewer 2
+## Referee B
 
@@
 - Maybe adjust something in the text to be more clear that the derivation of the results was obtained with rescaling and then we look at the non rescaled value of $\mu$ and $\sigma$.
-- Answer to the reviewer that in this kinf of conceptual work we are not necessarily interested in the realistic form of the SAD, indeed usual GLV papers have the Gaussian SAD prediction.
+- Answer to the reviewer that in this kind of conceptual work we are not necessarily interested in the realistic form of the SAD, indeed usual GLV papers have the Gaussian SAD prediction.
 In any case the model can still give non gaussian distribution in the case in which we use random growth rates etc...
@@
 ### Minor comments
-
 - Just engage with the reviewer discussing mutualism etc, but not change anything necessarily.
@@
 
+## Referee A
+
+We thank the referee for the thoughtful review of our work and for raising important points, addressed below, which we think helped us improve the paper.
+
+- *The claim is probably true, but unfortunately it does not describe a generic situation, because it is very hard to imagine that the exponents alpha, beta, gamma used in the model are identical for all populations.
+Perhaps the model would be more interesting if it included a spectrum of different exponents, described, for example, by a probabilistic law.*
+
+We agree that indeed a spectrum of different exponents is probably more realistic, and thank the reviewer to pointing that out.
+
+We can consider the case in which the exponents are not identical for every variable $x_i$ but instead we have them from Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and, for simplicity, all with the same standard deviation $\sigma_e$.
+
+Simulations show that all our results are robust as long as $\sigma_e$ is small enough that $\displaystyle{\min_i\beta_i>\max_i\alpha_i}$.
+
+As an example, we show in the plot below the probability of stability vs. $\sigma_e$ for a system with $S=100$, $\mu=\mu_s=\sigma=0.01$, $\alpha=\gamma=1$ and $\beta=3/2$.
+The probability of stability is obtained as the fraction of stable systems out of $100$ realizations for each value of $\sigma_e$.
+![](../plots/stoch-exp.png)
+
+We decided, however, to not include this analysis in the letter because we would like to keep the focus on the existence of the two complexity-stability regimes rather than exploring more detailed ecological scenarios.
+We stick to the case in which $f$, $g$ and $h$ are independent from the variable index, following references [18] and [19].
+
+- *The way of writing the matrix elements in Equations (7) is somewhat confusing. For clarity, the Kronecker delta can be used on the right side of these equations.*
+
+We thank the referee for this comment, we modified Equations (7) accoridngly.
+
+- *The mean field equations depend on the expected values of <x_^\gamma> and <x_^{2\gamma}>.
+The self-consistency condition requires that the probability distribution (12) has exactly the same expected values.
+This should be discussed in detail.*
+
+We agree with the referee that the discussion about the expected values and self-consistency conditions needed more clarifications.
+
+In some cases, for example $\alpha=1, \beta=3, \gamma=1$, both expectations converge to finite values and we can safely compute the self-consistency equations.
+However in potential case of interests, and in the example case $\alpha=1, \beta=3/2, \gamma=1$ characterized by a distribution with a power-law tail with exponent $-3/2$, the moments diverge.
+We now modified the discussion removing the gaussian approximation and considering the introduction of a cut-off.
+The idea behind this cut-off can be described by the example below.
+
+Consider, as an example which allows for a complete analytical treatment, the case of a power law distribution 
+$$
+P(x)=\frac{x^{-\alpha}}{\mathcal{Z}} ,
+$$
+defined from 1 to $\infty$ and with 
+$$
+\mathcal{Z}=\int_1^{\infty}x^{-\alpha}.
+$$
+
+Consider the case $\alpha=3/2$. The distribution is normalized with $\mathcal{Z}=2$, but the mean diverges. However, we would like to be able to describe the behaviour of the sample mean 
+$$
+\bar{x}_N\equiv\frac{1}{N}\sum_{i=1}^Nx_i,
+$$
+where the $x_i$ are extracted from $P(x)$.
+For this purpose, we can define the quantity
+$$
+\langle x\rangle_{\Lambda}\equiv\int_1^{\Lambda}dxP(x)x,
+$$
+with the cut-off $\Lambda$ defined such that $\int_1^{\infty}dxP(x)=1/N$, i.e., such that there is statistically less than 1 variable with value above $\Lambda$ out of $N$ extracted variables.
+For the case $\alpha=3/2$ we have
+$$
+\frac{1}{2}\int_{\Lambda}^{\infty}dxx^{-3/2}=\Lambda^{-1/2},
+$$
+and therefore $\Lambda=N^2$. We have for the mean
+$$
+\langle x\rangle_{\Lambda}=\frac{1}{2}\int_1^{N^2}dxx^{-1/2} = N-1.
+$$
+The result is plotted below, alongside sample mean for extractions of $N=10,$ $10^2,$ $10^3,$ $10^4,$ $10^5,$ $10^6$.
+
+![](../plots/cut-off-justification.png)
+
+## Referee B
+
+We thank the referee for the thorough and critical review.
+We address each of the comments below point by point.
+We believe the results of our paper are now more transparent thanks to the work needed to address the weaknesses hihglighted by the referee.
+
+- *Modelling Framework Justification:*
+
+     *The rationale behind the proposed generalization of the GLV model is claimed to be "natural" based on physical and biological grounds, specifically using power laws.
+Nevertheless, the literature extensively demonstrates species abundance growth being well-described by logistic equations or oscillatory predator-prey equations, that indeed are linear (except for the carrying capacity term).
+The justifications provided are weak, with the provided examples failing to convincingly represent the complex dynamics of natural systems, such as forest biomass distribution.
+If we think to a forest there are various heights of trees, and in general the biomass in not uniformly distributed, and thus also the growth is not necessary constrained by the surface.
+The sublinear scaling of the “production function”, f, if this is really the right conclusion based on the references of the manuscript, must be the result of a cooperative behavior of the interacting system and as such it should not appear as hand-added like the authors do.
+Furthermore, in reference 27 the sublinear scaling with k=3/4 refers to the scaling of the single individual growth rather than to a community.*
+
+We understand the criticism of the referee, pointing out the need for better justification of models of this kind in ecology.
+Our goal in this letter, however, is not to propose a model for a specific ecological scenario.
+Rather, we employ this model for complex systems (following references [18] and [19]) to show in a minimal setting the existence of the two opposite complexity-stability relationships in disordered dynamical systems, possibly with applications beyond ecology.
+
+Nevertheless, the examples we provide are indicative that models of this class can be relevant in ecology (see reference [17], where a discussion of possible mechanisms is present; even though admittedly more work is required in that direction, which is beyond the scope of our paper).
+
+Finally, we are aware of the fact that reference [27] refers to the scaling of single individual, indeed we mention it to introduce the scaling at the higher level of organization:
+
+"Biologically, the growth of organisms (populations of cells) has long been known to scale like $f (x) \sim x^k$ with $k\simeq 3/4$ [27], which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
+For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities [17, 28]".
+
+- *Feasibility Condition Neglected:*
+
+    *The authors do not seem to consider the condition for the feasibility of the fixed points (x\* >= 0), which is critical.
+This oversight, especially evident in Equations 11 and 12, i.e., Eq.
+12 or its Gaussian approximation are normalized from 0 or from -infty to infty?
+I think this is a very relevant issue calling into question the appropriateness of the results.*
+
+We thank the referee for highlighting this oversight.
+We have now specified that the distribution in Eq.(11) is defined form zero to infinity and we point to reference [17] where a discussion on feasibility for systems described by this kind of distribution is present.
+
+- *Lack of Proper Scaling:*
+
+    *There is no proper scaling of the interactions, so the thermodynamic limit is not well defined.
+And this is not discussed or compared with other models in the literature applying similar disputable choice (see E Mallmin, A Traulsen, S De Monte - arXiv preprint arXiv:2306.11031, 2023).
+Relatedly, the comparison with GLV with quenched disorder (e.g.
+“The properties of large random dynamical systems is often portrayed in the ( sigma sqrt(N) , mu N ) plane [25] is not appropriate as such model scales the interactions strengths with the system size.*
+
+The referee is absolutely correct in pointing out the lack of proper scaling in the derivation of the cavity results, we are grateful for pointing it out.
+We have now fixed that.
+However, we argue that, after a proper derivation of the analytical results, it is appropriate fix the the strength and heterogeneity of the interaction ($mean(A)=\mu$ and $var(A)=\sigma^2$) if our goal is to isolate and study the effects of changes in the number of degrees of freedom.
+We now discuss this, pointing out that it amount to be in a strong-interactions regime, and compare with reference (E Mallmin, A Traulsen, S De Monte *PNAS* 2024, now reference [34] in the manuscript) for an ecological example in which this choice is made to study chaotic turnover of species with broad abundance distributions.
+
+- *Gaussian Approximation:*
+
+    *The justification for employing a Gaussian approximation is unclear, given that Equation 12 should be solved self-consistently.
+Ecologically, Gaussian species abundance distributions (SADs) are not meaningful due to the prevalence of heavy-tailed SADs in natural ecosystems.*
+
+We thank the referee for this important comment, we now removed the Gaussian approximation in the discussion and clarified how to treat the case in which self-consistently computing the moment in Equation (12) can be problematic, as explained above in the answers to referee A.
+
+- *Homogenous Case and Stability:*
+     *The manuscript concludes that stability requires stronger self-interactions than cross-interactions when alpha equals beta.
+While this is true that self-interactions promote stability also for the Lotka-Volterra model, any negative mean species interaction (indicating mutualism) in the homogeneous case leads to instability, irrespective of system size.
+Thus, comparing this to the GLV model seems a stretch.*
+
+We are grateful to the referee for this comment, it helps to clarify the scope of our work.
+We have now specified everywhere in the manuscript, including Abstract, Introduction and Discussion, that we compare out results to the *competitive* Lotka-Volterra model and in general that our results are relevant for predominantly competitive systems.
+
+- *Notation:*
+
+     *Using S instead of N for species richness would align better with ecological conventions.*
+
+We decided to keep $N$ to indicate the number of degrees of freedom because our goal is not to allign with ecological literature, rather we would like to speak to the community of scientists working in complex systems.
+We made this choice in the first place to align with Robert May notation in reference [1].
+
+- *Discussion Quality:*
+
+    *Given the issues raised, the discussion section of the manuscript needs substantial improvement to convincingly argue the paper's contributions to the field.
+In light of these concerns, before the authors submit their manuscript elsewhere, I recommend to undertake significant revisions to address the weaknesses in their modelling framework, the rationale behind their approach, and the ecological implications of their findings*
+
+We are thankful for the critical reveiw overall. We extended the discussion and improved the paper in different aspects following suggestions of the referee.
+
 ---

### 2024-03-18 papers/submission/beyond-may.tex

@@
 In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: in random dynamical systems, stability does not necessarily decrease with dimensionality and interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
-For an in-depth discussion of the ecological implications of our result, see \cite{Hatton2023}.
+For an in-depth discussion of the ecological implications of our result, see \cite{Hatton2024}.
 
@@
 Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
-For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities~\cite{Hatton2015,Hatton2023}.
+For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities~\cite{Hatton2015,Hatton2024}.
 In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law, i.e. $g(x) \sim h(x) \sim x^{1/2}$~\cite{Barbier2021,Mazzarisi2023}.
@@
 This is not truly problematic because we are dealing with large, but \emph{finite} populations.
-Following~\cite{Cui2020,Hatton2023} and assuming $\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$, we can expand the solution~\eqref{eq: cavity solution} to first order in $\sigma \sqrt{N\langle x_*^2\rangle}$, and approximate~\eqref{eq: dist general} with a Gaussian distribution with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and $\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.
+Following~\cite{Cui2020,Hatton2024} and assuming $\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$, we can expand the solution~\eqref{eq: cavity solution} to first order in $\sigma \sqrt{N\langle x_*^2\rangle}$, and approximate~\eqref{eq: dist general} with a Gaussian distribution with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and $\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.
 
@@
 We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate, and Vadim Karatayev and Daniel Reuman for feedback on the manuscript.
-Funding for this work was provided by the Alexander von Humboldt Foundation in the framework of the Sofja Kovalevskaja Award endowed by the German Federal Ministry of Education and Research.
-O.M. was partly supported by NSF BIO OCE grant 2023473.
+Funding for this work was provided by the Alexander von Humboldt Foundation in the framework of the Sofja Kovalevskaja Award endowed by the German Federal Ministry of Education and Research, by the Trieste Laboratory on Quantitative Sustainability - TLQS and by NSF BIO OCE grant 2023473.
 \end{acknowledgments}

### 2024-03-18 papers/response.md
Update response.md
updates for resub

--- 2024-03-16 papers/response.md
@@
 
+## Overall
+
+- Try to pitch the paper from a slightly different point of view, more "honest" and aware of what we already did in the Science and putting more stress on what is new/clearer in this letter.
+
+- Should we remove "Beyond May" from the title?
+
 ## Reviewer 1

### 2024-03-18 scripts/single-run.jl

@@ -1,34 +1,38 @@
 using DrWatson, Glob, Revise
 @quickactivate
-foreach(includet, glob("*.jl", srcdir()))
+foreach(include, glob("*.jl", srcdir()))
 
 using Plots, LaTeXStrings
 
+pippo = rand(Uniform(1.1,1.5),10)
+pippo[1] = 1.
+
 p = Dict{Symbol,Any}(
     :scaled => false,
-    :S => 100,
-    :μ => 0.5,
-    :μₛ => 0.0,
-    :σ => 0.15,
-    :α => 1.5,
-    :β => 1.0,
+    :S => 10,
+    :μ => 0.75,
+    :μₛ => .75,
+    :σ => 0.,#15,
+    :α => 1.0,
+    :β => pippo,
     :γ => 1.0,
-    :extinction_threshold => 1e-5,
-    :dist => "gamma",
+    :x0 => .1*ones(10),
+    :extinction_threshold => 1e-6,
+    :dist => "normal",
     :N => 1,
     :seed => rand(UInt)
 );
 
 evolve!(p; trajectory=true)
 
-p[:trajectory].t
+p[:trajectory].u[end]
 
 #= trajectories =#
 plot(
     p[:trajectory],
     ylabel=L"x_i(t)",
     xlabel=L"t",
-    yscale=:log,
+    #yscale=:log,
     linewidth=1,
     legend=false,
     grid=false,

### 2024-03-18 papers/submission/beyond-may.tex

@@
 Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian~\eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
-\begin{equation}
-    M_i = - g(x_i^*)h'(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*),
-\end{equation}
-where $\psi(x)$ is the function defined in~\eqref{eq: psi}. 
+\begin{align}
+    M_{ij} = - g(x_i^*)&h'(x_i^*)\psi(x_i^*)\delta_{ij}, \nonumber \\ 
+    L_{ij} = g(x_i^*)\delta_{ij}, & \quad R_{ij} = h'(x_i^*)\delta_{ij}, 
+\end{align}
+where $\delta_{ij}$ is the Kronecker delta and $\psi(x)$ is the function defined in~\eqref{eq: psi}.
 \begin{figure}[t!]

### 2024-03-28 papers/submission/beyond-may.tex
small rewordings + discussion paragraph

@@
 
-\title{Beyond May: Complexity-stability relationships \\
+\title{Complexity-stability relationships
 in disordered dynamical systems}
@@
     Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
-    However, this general conclusion is not supported by empirical observation: from cells to biomes, biological systems are large, complex---and, by and large, stable.
+    However, this general conclusion is not always supported by empirical observation: from cells to biomes, biological systems are large, complex and, often, stable.
     In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 
@@
     if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
-    Our result shows that May's principle that "complexity begets instability" is not a general property of complex systems; instead, it is a property of a subclass of weakly cross-regulated disordered systems.
+    Our result shows that May's principle that "complexity begets instability" is not a general property of complex systems.
 \end{abstract}
@@
 Stability is defined as full stable coexistence.
-The red line is computed obtained using DMFT and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
+The red line is computed using DMFT and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
     Parameters values are $\alpha=1$, $\beta=3/2$,
@@
 Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
-In this case the first and second moments formally diverge.
+In this case, the first and second moments formally diverge.
 This is not truly problematic because we are dealing with large, but \emph{finite} populations.
@@
 
-The properties of large random dynamical systems is often portrayed in the $(\sigma \sqrt{N},\mu N)$ plane \cite{bunin2017ecological}.
-Everything else being fixed, an increase of the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane.
+The properties of large random dynamical systems are often portrayed in the $(\sigma \sqrt{N},\mu N)$ plane \cite{bunin2017ecological}.
+If we keep the mean and variance of interactions fixed, an increase in the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane.
 In the GLV model, the boundary between the stable and unstable parameter regions is the horizontal line $\sigma_c\sqrt{N} = \mu_s$ \cite{bunin2017ecological}, hence stability is never possible at large $N$.
@@
 Finally, Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$.
-(Numerically, we define $\sigma_c\sqrt{N}$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations we use $\mu_s = \mu$, which would never be stable in the GLV model.
+(Numerically, we define $\sigma_c\sqrt{N}$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations, we use $\mu_s = \mu$, which would never be stable in the GLV model.
 Fig.~\ref{fig: alpha-beta} illustrates the transition between a ``May" phase for $\alpha \geq \beta$ and an ``anti-May" phase for $\alpha < \beta$, showing that our analytical results are robust to heterogeneous interactions.
@@
 \paragraph*{}
-\emph{Discussion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and in other fields. 
-In this letter, we have showed that the condition $\sigma\sqrt{N}< \mu_s - \mu$ (popularized by May's slogan ``complexity begets instability") does not provide a complete picture of the relationship between complexity and stability.
+\emph{Discussion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and other fields. 
+In this letter, we have shown that the condition $\sigma\sqrt{N}< \mu_s - \mu$ does not provide a complete picture of the relationship between complexity and stability. 
 In particular, we have seen that the generalized Lotka-Volterra model, often cited in support of May's general prediction, exemplifies a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
-In models where self-interactions grow faster than cross-interactions, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.
+In models where self-interactions grow faster than cross-interactions~\cite{Hatton2024,samadder2024interconnection}, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.
+
+A modification of our model~\eqref{dynamics} of general ecological relevance describing the case in which per capita growth rates depend only on the sum of the other densities (weighted by interactions) is $\dot{x_i}/x_i=1-F(\sum_jA_{ij}x_j)$, where $F$ is an arbitrary function.
+General models of this kind are used, for example, to study the growth of a clonal population in the presence of a limiting factor~\cite{mazzolini2023universality}.
+All the models in this class are destabilized by diversity.
+The Jacobian matrix at equilibrium is $J_{ij}^*=-A_{ij}x_i^*F'(\sum_jA_{ij}x_j^*) \quad \forall i,j$ leading to the same simplification of the stability condition~\eqref{eq: random-stability} possible for GLV.
 
@@
 \begin{acknowledgments}
-We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate, and Vadim Karatayev and Daniel Reuman for feedback on the manuscript.
+We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate, and Jacopo Grilli, Vadim Karatayev and Daniel Reuman for feedback on the manuscript.
 Funding for this work was provided by the Alexander von Humboldt Foundation in the framework of the Sofja Kovalevskaja Award endowed by the German Federal Ministry of Education and Research, by the Trieste Laboratory on Quantitative Sustainability - TLQS and by NSF BIO OCE grant 2023473.

### 2024-03-28 papers/submission/beyond-may.tex
proper scaling of mu sigma

@@
 \emph{Random interactions: stability condition.---}We now consider the case of random interactions.
-Specifically, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$;
-We assume the diagonal elements $A_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) but the same standard deviation $\sigma$.
+Specifically, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with scaled mean $\mu=\tilde{\mu}/N$ and standard deviation $\sigma=\tilde{\sigma}/\sqrt{N}$~\cite{bunin2017ecological}.
+We assume the diagonal elements $A_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) and standard deviation $\sigma$.
 
@@
 \end{equation}
-In the GLV model, $\psi(x) = \mu_s - \mu$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}.
+In the GLV model, $\psi(x) = \mu_s - \tilde{\mu}/N$ (independently of $x$) and for large $N$ we recover the classical condition $\tilde{\sigma} = \sigma\sqrt{N} < \mu_s$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}.
 In general, however, the dependence on equilibrium values $x_i^*$ (which in turn depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.
@@
 \begin{equation}
-    \dot{x} = x^{\alpha}-A_sx^{\beta+\gamma}-x^{\beta}\big( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\big) \, ,
+    \dot{x} = x^{\alpha}-A_sx^{\beta+\gamma}-x^{\beta}\big( \tilde{\mu} \langle x^{\gamma}\rangle + \tilde{\sigma} \eta\big) \, ,
 \label{eq: dmft}
@@
     at varying $\alpha$ and $\beta$.
-    No amount of heterogeneity is allowed in the bottom right triangle because we have $\mu=\mu_s$, while an increasing value of $\sigma_c$
+    No amount of heterogeneity is allowed in the bottom right triangle because we have set $\mu=\mu_s$, while an increasing value of $\sigma_c$
     is found as $\beta$ becomes larger than $\alpha$.
     We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
-    does not imply that increasing $N$ will destabilize the system, because $\mu N$ would also change.
+    does not imply that increasing $N$, at fixed $\mu$ and $\sigma$, will destabilize the system, because $\tilde{\mu}=\mu N$ would also change.
     }
@@
 \begin{equation}
-    x_*^{\alpha-\beta} - A_s x_*^{\gamma}= \big( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
+    x_*^{\alpha-\beta} - A_s x_*^{\gamma}= \big( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
 \end{equation} 
@@
 $(\alpha-\beta)/\gamma$, or if $A_s=0$.
-(In the following we show that neglecting $A_s$ even when it is not zero does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
+(In the following we show that neglecting $A_s$, when it is of the order of $\mu$, does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
 \begin{equation} \label{eq: cavity solution}
-    x_* = \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
+    x_* = \left( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
 \end{equation}
@@
 \begin{equation}\label{eq: dist general}
-    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2 N\langle x_*^{2\gamma}\rangle}}
-    \exp{\left\{-\frac{(x_*^{\alpha-\beta}-\mu N\langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
+    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\tilde{\sigma}^2 \langle x_*^{2\gamma}\rangle}}
+    \exp{\left\{-\frac{(x_*^{\alpha-\beta}- \tilde{\mu} \langle x_*^{\gamma}\rangle)^2}{2\tilde{\sigma}^2\langle x_*^{2\gamma}\rangle}\right\}} \, ,
 \end{equation}
@@
 This is not truly problematic because we are dealing with large, but \emph{finite} populations.
-Following~\cite{Cui2020,Hatton2024} and assuming $\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$, we can expand the solution~\eqref{eq: cavity solution} to first order in $\sigma \sqrt{N\langle x_*^2\rangle}$, and approximate~\eqref{eq: dist general} with a Gaussian distribution with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and $\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.
+Following~\cite{Cui2020,Hatton2024} and assuming $\tilde{\sigma} \sqrt{\langle x_*^2\rangle}\ll \tilde{\mu}  \langle x_* \rangle$, we can expand the solution~\eqref{eq: cavity solution} to first order in $\tilde{\sigma} \sqrt{\langle x_*^2\rangle}$, and approximate~\eqref{eq: dist general} with a Gaussian distribution with moments $\langle x_*\rangle=\tilde{\mu}^{-2/3}$ and $\langle x_*^2\rangle=\tilde{\mu}^{-4/3}/(1-4\tilde{\mu}^{-2}\tilde{\sigma}^2)$.
 
@@
 The properties of large random dynamical systems are often portrayed in the $(\sigma \sqrt{N},\mu N)$ plane \cite{bunin2017ecological}.
-If we keep the mean and variance of interactions fixed, an increase in the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane.
+If we keep the mean $\mu$ and standard deviation $\sigma$ of interactions fixed, an increase in the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane.
 In the GLV model, the boundary between the stable and unstable parameter regions is the horizontal line $\sigma_c\sqrt{N} = \mu_s$ \cite{bunin2017ecological}, hence stability is never possible at large $N$.
@@
 
-Finally, Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$.
+Finally, Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\tilde{\mu}=\mu N$ and $\gamma = 1$.
 (Numerically, we define $\sigma_c\sqrt{N}$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations, we use $\mu_s = \mu$, which would never be stable in the GLV model.
@@
 \emph{Discussion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and other fields. 
-In this letter, we have shown that the condition $\sigma\sqrt{N}< \mu_s - \mu$ does not provide a complete picture of the relationship between complexity and stability. 
+In this letter, we have shown that the condition $\sigma\sqrt{N}< \mu_s$ does not provide a complete picture of the relationship between complexity and stability. 
 In particular, we have seen that the generalized Lotka-Volterra model, often cited in support of May's general prediction, exemplifies a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.

### 2024-03-29 papers/submission/beyond-may.tex

@@
 Stability is defined as full stable coexistence.
-The red line is computed using DMFT and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
+The red line is computed using the cavity solution and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
     Parameters values are $\alpha=1$, $\beta=3/2$,
-    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. The green line was plotted for $\mu=0.1$ and $\sigma=0.75$.}
+    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. The green line is plotted for $\mu=0.1$ and $\sigma=0.75$.}
     \label{fig: stability line + sims}
 \end{figure}
-Following e.g. Ref.~\cite{Roy2019}, we can derive from the evolution equation~\eqref{dynamics} a dynamical mean field theory (DMFT) 
-which describes the ensemble of $N\gg 1$ variables
-by means of a single, self-consistent representative stochastic trajectory
+Following e.g.
+Ref.~\cite{Roy2019}, we can derive from Eq.~\eqref{dynamics} a cavity solution which describes the ensemble of the stationary value of $N\gg 1$ variables by means of a single representative radnom variable $x_*$
 \begin{equation}
-    \dot{x} = x^{\alpha}-A_sx^{\beta+\gamma}-x^{\beta}\big( \tilde{\mu} \langle x^{\gamma}\rangle + \tilde{\sigma} \eta\big) \, ,
+    0 = x_*^{\alpha}-A_sx_*^{\beta+\gamma}-x_*^{\beta}\big( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
 \label{eq: dmft}
 \end{equation}
-where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\eta$ is a Gaussian noise with zero mean and correlation $\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$. 
+where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\xi$ is a standard normal random variable. 
 \begin{figure}[t!]
@@
 \end{figure}
-At equilibrium, we can write
-\begin{equation}
-    x_*^{\alpha-\beta} - A_s x_*^{\gamma}= \big( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
-\end{equation} 
-where $\xi$ is now a standard normal random variable.
 The equation above can be solved for $x_*$ for specific values of

### 2024-04-10 scripts/single-run.jl
add self-consistent computation for cut-off

@@ -4,28 +4,23 @@ foreach(include, glob("*.jl", srcdir()))
 
 using Plots, LaTeXStrings
 
-pippo = rand(Uniform(1.1,1.5),100)
-pippo[3] = 1.
-
 p = Dict{Symbol,Any}(
     :scaled => false,
     :S => 100,
-    :μ => 0.75,
-    :μₛ => 1.,
-    :σ => 0.15,
+    :μ => .01,
+    :μₛ => .01,
+    :σ => .01,
     :α => 1.0,
-    :β => pippo,
+    :β => 1.5,
     :γ => 1.0,
-    :x0 => .1*ones(100),
+    :x0 => 1. *ones(100),
     :extinction_threshold => 1e-6,
-    :dist => "normal",
+    :dist => "gamma",
     :N => 1,
-    :seed => rand(UInt)
+    :seed => 7#rand(UInt)
 );
 
 evolve!(p; trajectory=true)
-
-p[:trajectory].u[end]
 
 #= trajectories =#
 plot(
@@ -40,7 +35,7 @@ plot(
 
 #= density distribution =#
 p_eq = p[:trajectory].u[end]
-histogram(p_eq,
+stephist(p_eq,
     normalize=true,
     alpha=0.5,
     color=:green,
@@ -48,6 +43,8 @@ histogram(p_eq,
     xlabel=L"x_*",
     label=false,
     grid=false,
+    fill=true,
+    #bins = minimum(p_eq):.005:maximum(p_eq),
 )
 
 #= cavity solution with gaussian approximation =#
@@ -61,17 +58,19 @@ plot!(X, [pdf(P_gauss(p), x) for x in X],
     grid=false
 )
 
-a, b, c = Cavity(p)
+a, b, c, Λ = Cavity_Λ(p)
 
-X=[x for x in .9*minimum(p_eq):.01:1.1*maximum(p_eq)]
+X=[x for x in .9*minimum(p_eq):.001:1.1*maximum(p_eq)]
 plot!(X,[P_n(x, b, c, p) for x in X],
 labels = false, #"cavity - gaussian",
 linewidth = 2,
 alpha = 1,
 linecolor = :black,
 grid = false,
-size = (600,200)
+size = (600,200),
 )
+
+(100*exp(-p[:μ]^2*100^2*1.04^2/(2*p[:σ]^2*100*1.42))/(sqrt(2*pi*p[:σ]^2*100*1.42)))^(2)
 
 p[:converged]
 

### 2024-04-10 src/cavity.jl

@@ -4,31 +4,59 @@ using DifferentialEquations
 using LinearAlgebra
 using QuadGK
 using HCubature
-using NLsolve
+using NLsolve, Roots
 using SpecialFunctions
-
-#= expression for density distribution with gaussian approximation (γ=1) =#
-function P_gauss(p)
-    @unpack μ, σ, α, β, S = p
-    μ = p[:scaled] ? μ : μ*S
-    σ = p[:scaled] ? σ : σ*sqrt(S) 
-    e1 = μ^(1/(α-β-1))
-    e2 = e1^2/(1-(1/(α-β))^2*(μ*e1)^(2*(1-(α-β))/(α-β))*σ^2)
-    #ϕ = e2>0 ? (1+erf((e1)/sqrt(2*(e2-e1^2))))/2 : 1
-    return e2>0 ? Normal(e1, sqrt(e2-e1^2)) : Normal(e1, 0) #(e1, e2>0 ? e2 : 0, ϕ, e2>0 ? Normal(e1,sqrt(e2-e1^2)) : Normal(e1,0))
-end
 
 #= expression for density distribution (γ=1) =#
 function P_n(n, e1_n, e2_n, p)
-    @unpack μ, σ, α, β = p
+    @unpack μ, σ, α, β, S = p
     μ = p[:scaled] ? μ : μ*S
     σ = p[:scaled] ? σ : σ*sqrt(S)
-    return abs(α-β)*n^(α-β-1)/(sqrt(2*π*σ^2*e2_n))*
-        exp(-(n^(α-β)-μ*e1_n)^2/(2*σ^2*e2_n))
+    x = max(n,1e-20)
+    return abs(α-β)*x^(α-β-1)/(sqrt(2*π*σ^2*e2_n))*exp(-(x^(α-β)-μ*e1_n)^2/(2*σ^2*e2_n))
 end
 
 #= cavity calculation for fraction of positive equilibriums 
 and first two moments of the equilibrium distribution given μ and σ. =#
+function Cavity_Λ(p;
+    Iter = 2000, #number of iteration
+    rela = .01, #relax parameter for fixed point
+    tol = 1e-9, #requested tolerance for numerical integrator
+    n_min = p[:extinction_threshold], #lower bound for integration
+    Λ_init = 1, #initial guess for cut-off
[…]

### 2024-04-10 scripts/single-run.jl
clean single run

@@ -17,7 +17,7 @@ p = Dict{Symbol,Any}(
     :extinction_threshold => 1e-6,
     :dist => "gamma",
     :N => 1,
-    :seed => 7#rand(UInt)
+    :seed => rand(UInt)
 );
 
 evolve!(p; trajectory=true)
@@ -44,18 +44,6 @@ stephist(p_eq,
     label=false,
     grid=false,
     fill=true,
-    #bins = minimum(p_eq):.005:maximum(p_eq),
-)
-
-#= cavity solution with gaussian approximation =#
-X = [x for x in 0.9*minimum(p_eq):0.0001:1.1*maximum(p_eq)]
-plot!(X, [pdf(P_gauss(p), x) for x in X],
-    labels=false, #"cavity - gaussian",
-    linewidth=2,
-    alpha=1,
-    linecolor=:black,
-    linestyle=:dash,
-    grid=false
 )
 
 a, b, c, Λ = Cavity_Λ(p)
@@ -70,9 +58,19 @@ grid = false,
 size = (600,200),
 )
 
-(100*exp(-p[:μ]^2*100^2*1.04^2/(2*p[:σ]^2*100*1.42))/(sqrt(2*pi*p[:σ]^2*100*1.42)))^(2)
 
 p[:converged]
+
+#= cavity solution with gaussian approximation =#
+X = [x for x in 0.9*minimum(p_eq):0.0001:1.1*maximum(p_eq)]
+plot!(X, [pdf(P_gauss(p), x) for x in X],
+    labels=false, #"cavity - gaussian",
+    linewidth=2,
+    alpha=1,
+    linecolor=:black,
+    linestyle=:dash,
+    grid=false
+)
 
 #= trajectories =#
 plot(

### 2024-04-10 papers/submission/beyond-may.tex
update discussion on cavity (temporary, still needs modifications)

@@
     \includegraphics[width=.45\textwidth]{cavity.pdf}
-    \caption{The equilibrium distribution with power-law self- and cross-interactions is accurately reproduced by 
-    Eq.~\eqref{eq: dist general} and its Gaussian approximation.
+    \caption{The equilibrium distribution for the case of power-law self- and cross-interactions is accurately reproduced by Eq.~\eqref{eq: dist general}.
     For simulations (in green) we used $\alpha=1$, $\beta=3/2$,
-    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=5\cdot 10^{-3}$.}
+    $\gamma=1$, $N=100$, $\mu=\mu_s=\sigma=10^{-2}$.}
     \label{fig: cavity sol.}
@@
 
-Equipped with the equilibrium distribution $P(x^*)$, we can compute, for given $N$ and $\mu$, the maximal heterogeneity $\sigma_c$ compatible with the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$.
+Equipped with the equilibrium distribution $P(x_*)$, we can compute, for given $N$ and $\mu$, the maximal heterogeneity $\sigma_c$ compatible with the linear stability condition $N\langle (\sigma/\psi(x_*))^2\rangle < 1$.
 
@@
 This is not truly problematic because we are dealing with large, but \emph{finite} populations.
+
+Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$, it diverges for $N\to\infty$.
+For finite $N$ we have $\bar{x}_N\neq\langle x_*\rangle=\infty$.
+What we really have is the convergence of $\bar{x}_N \approx \int_0^{\Lambda}dx_*x_*P(x_*)$ where the cut-off $\Lambda$ dependence on $N$ is such that $1-\int_0^{\Lambda}dx_*P(x_*)=1/N$.
+In orther words such that there is less 1 variable with density above $\Lambda$.
+
 Following~\cite{Cui2020,Hatton2024} and assuming $\tilde{\sigma} \sqrt{\langle x_*^2\rangle}\ll \tilde{\mu}  \langle x_* \rangle$, we can expand the solution~\eqref{eq: cavity solution} to first order in $\tilde{\sigma} \sqrt{\langle x_*^2\rangle}$, and approximate~\eqref{eq: dist general} with a Gaussian distribution with moments $\langle x_*\rangle=\tilde{\mu}^{-2/3}$ and $\langle x_*^2\rangle=\tilde{\mu}^{-4/3}/(1-4\tilde{\mu}^{-2}\tilde{\sigma}^2)$.
@@
 The result is plotted against simulations in Fig.~\ref{fig: cavity sol.}.
-We show both the exact solution~\eqref{eq: dist general} (where integration for the moments has been performed up to a maximum value above which no equilibrium value should be found) and the Gaussian approximation.
+We show both the exact solution~\eqref{eq: dist general} (where integration for the moments has been performed up to a cut-off above which no equilibrium value should be found) and the Gaussian approximation.
 Simulations correspond to $A_s=\mu$, showing that neglecting the $A_s$ term in analytical expressions does not introduce a significant error.

### 2024-04-16 src/cavity.jl
update cavity.jl with cut-off

@@ -54,14 +54,15 @@ end
 
 #= cavity calculation for fraction of positive equilibriums 
 and first two moments of the equilibrium distribution given μ and σ.
-In this case we keep Λ=n_max constant to speed up computations and vaoid to 
-find the right ansatz for initial condition case by case.
+In this case we keep Λ = S^(β-α) constant to speed up computations and avoid to find the right ansatz for initial condition case by case.
+The choice is an analitical approximation of the scaling of Λ with S.
+It can also be chosen to be a fixed constant 10^2 or 10^3.
 Results are not affected by this approximation.=#
 function Cavity(p;
     Iter = 2000, #number of iteration
     rela = .01, #relax parameter for fixed point
     tol = 1e-9, #requested tolerance for numerical integrator
-    n_max = 1e4, #upper bound for integration
+    Λ = p[:S]^(p[:β]-p[:α]), #Cut-off for integration
     n_min = p[:extinction_threshold], #lower bound for integration
     e1_init = (p[:scaled] ? 1/p[:μ] : 1/p[:μ]/p[:S]), #initial guess for e1_n
     e2_init = (p[:scaled] ? 1/p[:μ] : 1/p[:μ]/p[:S]) #initial guess for e1_n
@@ -72,9 +73,9 @@ function Cavity(p;
     e2_n = e2_init*ones(Iter+1) #<n²>
 
     for i in 1:Iter
-    #    ϕ_new = first(quadgk(x -> P_n(x, e1_n[i], e2_n[i], p) , n_min, n_max, rtol=tol))
-        e1_new = first(quadgk(x -> x*P_n(x, e1_n[i], e2_n[i], p) , n_min, n_max, rtol=tol))
-        e2_new = first(quadgk(x -> x*x*P_n(x, e1_n[i], e2_n[i], p) , n_min, n_max, rtol=tol))
+    #    ϕ_new = first(quadgk(x -> P_n(x, e1_n[i], e2_n[i], p) , n_min, Λ, rtol=tol))
+        e1_new = first(quadgk(x -> x*P_n(x, e1_n[i], e2_n[i], p) , n_min, Λ, rtol=tol))
[…]

### 2024-04-16 papers/response.md
update response

@@
 
-## Reviewer 1
+## Referee A
 
@@
 
+## Referee A
+
+- *The claim is probably true, but unfortunately it does not describe a generic situation, because it is very hard to imagine that the exponents alpha, beta, gamma used in the model are identical for all populations.
+Perhaps the model would be more interesting if it included a spectrum of different exponents, described, for example, by a probabilistic law.*
+
+...
+
+- *The way of writing the matrix elements in Equations (7) is somewhat confusing. For clarity, the Kronecker delta can be used on the right side of these equations.*
+
+We thank the referee for this comment, we modified Equations (7) accoridngly.
+
+- *The mean field equations depend on the expected values of <x_^\gamma> and <x_^{2\gamma}>.
+The self-consistency condition requires that the probability distribution (12) has exactly the same expected values.
+This should be discussed in detail.*
+
+We agree with the referee that the discussion about the expected values and self-consistency conditions needed more clarifications.
+
+In some cases, for example $\alpha=1, \beta=3, \gamma=1$, both expectations converge to finite values and we can safely compute the self-consistency equations.
+However in potential case of interests, and in the example case $\alpha=1, \beta=3/2, \gamma=1$ we consider resulting in a distribution with a power-law tail with exponent $-3/2$, the moments diverge.
+We now modified the discussion removing the gaussian approximation and considering the introduction of a cut-off.
+The idea behind this cut-off can be described by the example below.
+
+Consider, as an example which allows for a complete analytical treatment, the case of a power law distribution 
+$$
+P(x)=\frac{x^{-\alpha}}{\mathcal{Z}} ,
+$$
+defined from 1 to $\infty$ and with 
+$$
+\mathcal{Z}=\int_1^{\infty}x^{-\alpha}.
+$$
+
+Consider the case $\alpha=3/2$. The distribution is normalized with $\mathcal{Z}=2$, but the mean diverges. However, we would like to be able to describe the behaviour of the sample mean 
+$$
+\bar{x}_N\equiv\frac{1}{N}\sum_{i=1}^Nx_i,
+$$
+where the $x_i$ are extracted from $P(x)$.
+For this purpose, we can define the quantity
+$$
+\langle x\rangle_{\Lambda}\equiv\int_1^{\Lambda}dxP(x)x,
+$$
+with the cut-off $\Lambda$ defined such that $\int_1^{\infty}dxP(x)=1/N$, i.e., such that there is statistically less than 1 variable with value above $\Lambda$ out of $N$ extracted variables.
+For the case $\alpha=3/2$ we have
+$$
+\frac{1}{2}\int_{\Lambda}^{\infty}dxx^{-3/2}=\Lambda^{-1/2},
+$$
+and therefore $\Lambda=N^2$. We have for the mean
+$$
+\langle x\rangle_{\Lambda}=\frac{1}{2}\int_1^{N^2}dxx^{-1/2} = N-1.
+$$
+The result is plotted below, alongside sample mean for extractions of $N=10,$ $10^2,$ $10^3,$ $10^4,$ $10^5,$ $10^6$.
+
+![](../plots/cut-off-justification.png)
+
+## Referee B
+
+We thank the referee...
+
+- *Modelling Framework Justification:*
+
+     *The rationale behind the proposed generalization of the GLV model is claimed to be "natural" based on physical and biological grounds, specifically using power laws.
+Nevertheless, the literature extensively demonstrates species abundance growth being well-described by logistic equations or oscillatory predator-prey equations, that indeed are linear (except for the carrying capacity term).
+The justifications provided are weak, with the provided examples failing to convincingly represent the complex dynamics of natural systems, such as forest biomass distribution.
+If we think to a forest there are various heights of trees, and in general the biomass in not uniformly distributed, and thus also the growth is not necessary constrained by the surface.
+The sublinear scaling of the “production function”, f, if this is really the right conclusion based on the references of the manuscript, must be the result of a cooperative behavior of the interacting system and as such it should not appear as hand-added like the authors do.
+Furthermore, in reference 27 the sublinear scaling with k=3/4 refers to the scaling of the single individual growth rather than to a community.*
+
+...
+
+- *Feasibility Condition Neglected:*
+
+    *The authors do not seem to consider the condition for the feasibility of the fixed points (x\* >= 0), which is critical.
+This oversight, especially evident in Equations 11 and 12, i.e., Eq.
+12 or its Gaussian approximation are normalized from 0 or from -infty to infty?
+I think this is a very relevant issue calling into question the appropriateness of the results.*
+
+...
+
+- *Lack of Proper Scaling:*
+
+    *There is no proper scaling of the interactions, so the thermodynamic limit is not well defined.
+And this is not discussed or compared with other models in the literature applying similar disputable choice (see E Mallmin, A Traulsen, S De Monte - arXiv preprint arXiv:2306.11031, 2023).
+Relatedly, the comparison with GLV with quenched disorder (e.g.
+“The properties of large random dynamical systems is often portrayed in the ( sigma sqrt(N) , mu N ) plane [25] is not appropriate as such model scales the interactions strengths with the system size.*
+
+...
+
+- *Gaussian Approximation:*
+
+    *The justification for employing a Gaussian approximation is unclear, given that Equation 12 should be solved self-consistently.
+Ecologically, Gaussian species abundance distributions (SADs) are not meaningful due to the prevalence of heavy-tailed SADs in natural ecosystems.*
+
+We thank the referee for this comment, we now removed the Gaussian approximation in the discussion and clarified how to treat the case in which self-consistently computing the moment in Equation (12) can be problematic, as explained in the answers to referee A.
+
+- *Homogenous Case and Stability:*
+     *The manuscript concludes that stability requires stronger self-interactions than cross-interactions when alpha equals beta.
+While this is true that self-interactions promote stability also for the Lotka-Volterra model, any negative mean species interaction (indicating mutualism) in the homogeneous case leads to instability, irrespective of system size.
+Thus, comparing this to the GLV model seems a stretch.*
+
+...
+
+- *Notation:*
+
+     *Using S instead of N for species richness would align better with ecological conventions.*
+
+...
+
+- *Discussion Quality:*
+
+    *Given the issues raised, the discussion section of the manuscript needs substantial improvement to convincingly argue the paper's contributions to the field.
+In light of these concerns, before the authors submit their manuscript elsewhere, I recommend to undertake significant revisions to address the weaknesses in their modelling framework, the rationale behind their approach, and the ecological implications of their findings*
+
+...
+
 ---

### 2024-04-16 papers/submission/beyond-may.tex

@@
     In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 
-    Using a non-linear generalization of the Lotka-Volterra model, we show that there are in fact two kinds of complexity-stability relationships in disordered dynamical systems: 
+    Using a non-linear generalization of the competitive Lotka-Volterra model, we show that there are in fact two kinds of complexity-stability relationships in disordered dynamical systems:
     if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
@@
 Here $f(x_i)$ represents the self-dynamics of a population $i$ (growth and self-regulation), while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations.
-That interaction is weighted by a coefficient $A_{ij}$, such that $A_{ij} > 0$ (resp.
-$A_{ji} < 0$) implies a negative (resp.
-positive) effect of $j$ on the growth of $i$. Non-zero diagonal elements $A_{ii}$ accounts for additional self-interactions with the same functional form as cross-interactions. 
+That interaction is weighted by a coefficient $A_{ij}>0$, such that it implies a negative effect of $j$ on the growth of $i$.
+Non-zero diagonal elements $A_{ii}$ accounts for additional self-interactions with the same functional form as cross-interactions.
 We refer to $f$ as the ``production function'' and $g$ as the ``response function''.
@@
 to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
-The classic generalized Lotka-Volterra (GLV) competition model studied recently by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear.
+The classic generalized Lotka-Volterra (GLV) competition model studied by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear.
 
@@
 For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities~\cite{Hatton2015,Hatton2024}.
-In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law, i.e. $g(x) \sim h(x) \sim x^{1/2}$~\cite{Barbier2021,Mazzarisi2023}.
+In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law, i.e. $g(x) \sim h(x) \sim x^{1/2}$~\cite{Barbier2021,mazzarisi2024}.
 
@@
 In this case, the first and second moments formally diverge.
-This is not truly problematic because we are dealing with large, but \emph{finite} populations.
-
-Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$, it diverges for $N\to\infty$.
-For finite $N$ we have $\bar{x}_N\neq\langle x_*\rangle=\infty$.
-What we really have is the convergence of $\bar{x}_N \approx \int_0^{\Lambda}dx_*x_*P(x_*)$ where the cut-off $\Lambda$ dependence on $N$ is such that $1-\int_0^{\Lambda}dx_*P(x_*)=1/N$.
+It is not uncommon, for example in ecology, to find empirical species abundance distributions characterized by heavy-tails, potentially with diverging moments.
+This is problem can be overcome considering that we are dealing with large, but \emph{finite} populations. 
+We are therefore interested in predicting the sample mean and variance for a given $N$.
+Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$.
+We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$ where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$.
 In orther words such that there is less 1 variable with density above $\Lambda$.
 
-Following~\cite{Cui2020,Hatton2024} and assuming $\tilde{\sigma} \sqrt{\langle x_*^2\rangle}\ll \tilde{\mu}  \langle x_* \rangle$, we can expand the solution~\eqref{eq: cavity solution} to first order in $\tilde{\sigma} \sqrt{\langle x_*^2\rangle}$, and approximate~\eqref{eq: dist general} with a Gaussian distribution with moments $\langle x_*\rangle=\tilde{\mu}^{-2/3}$ and $\langle x_*^2\rangle=\tilde{\mu}^{-4/3}/(1-4\tilde{\mu}^{-2}\tilde{\sigma}^2)$.
-
-The result is plotted against simulations in Fig.~\ref{fig: cavity sol.}.
-We show both the exact solution~\eqref{eq: dist general} (where integration for the moments has been performed up to a cut-off above which no equilibrium value should be found) and the Gaussian approximation.
+Following, e.g.,~\cite{Cui2020,Hatton2024} we can compute self-consistently $\Lambda$, $\langle x_*\rangle_{\Lambda}$ and $\langle x_*^2\rangle_{\Lambda}\equiv\int_0^{\Lambda}dx_*x_*^2P(x_*)$.
+The resulting distribution is plotted against simulations in Fig.~\ref{fig: cavity sol.}.
 Simulations correspond to $A_s=\mu$, showing that neglecting the $A_s$ term in analytical expressions does not introduce a significant error.
@@
 
-In Fig.~\ref{fig: stability line + sims} we plot the stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ for $\alpha = 1$, $\beta = 3/2$ derived from \eqref{eq: dist general} (solid line) together with results from simulations (shading).
+In Fig.~\ref{fig: stability line + sims} we plot the stability condition $N\langle (\sigma/\psi(x_*))^2\rangle < 1$ for $\alpha = 1$, $\beta = 3/2$ derived from \eqref{eq: dist general} (solid line) together with results from simulations (shading).
 Here, because the boundary between the stable and unstable phases is a straight line with positive slope, increasing $N$ will eventually bring the system in the stable region.
@@
 
-A modification of our model~\eqref{dynamics} of general ecological relevance describing the case in which per capita growth rates depend only on the sum of the other densities (weighted by interactions) is $\dot{x_i}/x_i=1-F(\sum_jA_{ij}x_j)$, where $F$ is an arbitrary function.
+A modification of our model~\eqref{dynamics} of general ecological relevance describing the case in which per capita growth rates depend only on the sum of the other densities (weighted by interaction strength) is $\dot{x_i}/x_i=1-F(\sum_jA_{ij}x_j)$, where $F$ is an arbitrary function.
 General models of this kind are used, for example, to study the growth of a clonal population in the presence of a limiting factor~\cite{mazzolini2023universality}.

### 2024-04-30 papers/response-referee.md
update responses

--- 2024-03-16 papers/response.md
@@
-# Referee reports
-
-## Report of Referee A -- LN17679/Mazzarisi
-
-The authors discuss the stability properties of a system of
-first-order differential equations describing the evolution of coupled
-populations in an ecosystem. The model is intended to show that May's
-prediction that large complex systems cannot be stable is not a
-general property of such systems. The model is actually a modification
-of the original model, considered by R.M. May, with the difference
-that the authors replace linear functions with power functions in
-population dynamics. They show that for a certain combination of power
-function exponents, the system becomes more stable as its size
-increases, unlike the original May model.
-
-The claim is probably true, but unfortunately it does not describe a
-generic situation, because it is very hard to imagine that the
-exponents alpha, beta, gamma used in the model are identical for all
-populations. Perhaps the model would be more interesting if it
-included a spectrum of different exponents, described, for example, by
-a probabilistic law.
-
-Minor remarks:
-
-The way of writing the matrix elements in Equations (7) is somewhat
-confusing. For clarity, the Kronecker delta can be used on the right
-side of these equations.
-
-The mean field equations depend on the expected values of <x_*^\gamma>
-and <x_*^{2\gamma}>. The self-consistency condition requires that the
-probability distribution (12) has exactly the same expected values.
-This should be discussed in detail.
-
-In summary, I do not think that the paper meets PRL criteria. However,
-I would recommend the article for submission as a regular article to
-Physical Review E, after the authors have softened their conclusions
-and improved the discussion of the mean-field approximation.
-
----
-
-## Report of Referee B -- LN17679/Mazzarisi
-
-Dear editor,
-I have reviewed the manuscript entitled "Beyond May: Complexity-stability relationships in
-disordered dynamical systems" intended for publication in Physical Review Letters. This study
-seeks to address the well-known May's stability-complexity conundrum by incorporating
-
-recent developments in random matrix theory into a non-linear generalization of the Lotka-
-Volterra (GLV) model.
-
-Summary:
-
-The authors challenge Robert May's conclusion that large, complex systems inherently lack
-stability by using a non-linear generalization of the GLV model, incorporating power-law
-functions into the interactions between species and species growth rate. Their findings
-suggest that the instability in complex systems as posited by May is not a universal
-characteristic but rather specific to a subclass of weakly cross-regulated systems.
-General Comments:
-The manuscript could be of interest to researchers focusing on ecological systems, especially
-those applying disordered system approaches. However, the modelling framework and
-ecological implications presented in the paper are predicated on assumptions that I find
-somewhat tenuous. The authors' arguments do not convincingly justify the generalization of
-the GLV model, as well as the main ecological finding is quite consequential given the model’s
-hypotheses. Moreover, I have few major technical and conceptual concerns that from my
-point of view should be fixed before publication in any journal.
-The relevance of the work appears confined to a niche.
-
-#### Major Concerns:
-
-Modelling Framework Justification:
-
-• The rationale behind the proposed generalization of the GLV model is claimed to be
-"natural" based on physical and biological grounds, specifically using power laws.
-Nevertheless, the literature extensively demonstrates species abundance growth
-being well-described by logistic equations or oscillatory predator-prey equations, that
-indeed are linear (except for the carrying capacity term). The justifications provided
-are weak, with the provided examples failing to convincingly represent the complex
-dynamics of natural systems, such as forest biomass distribution. If we think to a forest
-there are various heights of trees, and in general the biomass in not uniformly
-distributed, and thus also the growth is not necessary constrained by the surface. The
-sublinear scaling of the “production function”, f, if this is really the right conclusion
-based on the references of the manuscript, must be the result of a cooperative
-behavior of the interacting system and as such it should not appear as hand-added
-like the authors do. Furthermore, in reference 27 the sublinear scaling with k=3/4
-refers to the scaling of the single individual growth rather than to a community.
-
-Feasibility Condition Neglected:
-
-• The authors do not seem to consider the condition for the feasibility of the fixed points
-(x* >= 0), which is critical. This oversight, especially evident in Equations 11 and 12,
-i.e., Eq. 12 or its Gaussian approximation are normalized from 0 or from -infty to infty?
-I think this is a very relevant issue calling into question the appropriateness of the
-results.
-
-Lack of Proper Scaling:
-
-• There is no proper scaling of the interactions, so the thermodynamic limit is not well
-defined. And this is not discussed or compared with other models in the literature
-applying similar disputable choice (see E Mallmin, A Traulsen, S De Monte - arXiv
-preprint arXiv:2306.11031, 2023). Relatedly, the comparison with GLV with quenched
-disorder (e.g. “The properties of large random dynamical systems is often portrayed
-in the ( sigma sqrt(N) , mu N ) plane [25] is not appropriate as such model scales the
-interactions strengths with the system size.
-
-Gaussian Approximation:
-
-• The justification for employing a Gaussian approximation is unclear, given that
-Equation 12 should be solved self-consistently. Ecologically, Gaussian species
-
-abundance distributions (SADs) are not meaningful due to the prevalence of heavy-
-tailed SADs in natural ecosystems.
-
-#### Minor Comments:
-
-Homogenous Case and Stability:
-
-• The manuscript concludes that stability requires stronger self-interactions than cross-
-interactions when alpha equals beta. While this is true that self-interactions promote
-
-stability also for the Lotka-Volterra model, any negative mean species interaction
-(indicating mutualism) in the homogeneous case leads to instability, irrespective of
-system size. Thus, comparing this to the GLV model seems a stretch.
-
-Notation:
-
-• Using S instead of N for species richness would align better with ecological
-conventions.
-
-Discussion Quality:
-
-• Given the issues raised, the discussion section of the manuscript needs substantial
-improvement to convincingly argue the paper's contributions to the field.
-In light of these concerns, before the authors submit their manuscript elsewhere, I
-recommend to undertake significant revisions to address the weaknesses in their modelling
-framework, the rationale behind their approach, and the ecological implications of their
-findings
-
----
-
-# How to address
-
-## Overall
-
-- Try to pitch the paper from a slightly different point of view, more "honest" and aware of what we already did in the Science and putting more stress on what is new/clearer in this letter.
-
-## Reviewer 1
-
-### Major comments
-
-- Run simulations with distributed exponents and add to the SM (to be created).
-
-### Minor comments
-
-- Add Kronecker delta to Eq. (7). 
-- Address the concern of the reviewer in the SM by studying Eq. (11) through the definition of the Gaussian variable $y=x^{\alpha-\beta}$.
-
-## Referee B
-
-### Major comments
-
-- Point to Science for ecological justifications and in general to answer the first point.
-- Point to Science paper for feasibility adding some discussion on this in the main text maybe.
-- Maybe adjust something in the text to be more clear that the derivation of the results was obtained with rescaling and then we look at the non rescaled value of $\mu$ and $\sigma$.
-- Answer to the reviewer that in this kind of conceptual work we are not necessarily interested in the realistic form of the SAD, indeed usual GLV papers have the Gaussian SAD prediction.
-In any case the model can still give non gaussian distribution in the case in which we use random growth rates etc...
-To conclude, the gaussian approximation is used only to more clearly present the results about complexity-stability but it is not necessary.
-
-### Minor comments
-- Just engage with the reviewer discussing mutualism etc, but not change anything necessarily.
-- We stick to $N$, we don't want to relate to ecological literature, but more to complex systems. Moreover May used $N$ in his paper.
-- Just answer something about all the changes that we made.
-
----
-
-# Response to reviewer
-
 ## Referee A
@@
 We are thankful for the critical reveiw overall. We extended the discussion and improved the paper in different aspects following suggestions of the referee.
-
----
-
-# Letter to editor

### 2024-04-30 papers/submission/beyond-may.tex
update manuscript

@@
 
-\title{Complexity-stability relationships
-in disordered dynamical systems}
+\title{Complexity-stability relationships disordered dynamical systems}
 
@@
     In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 
-    Using a non-linear generalization of the competitive Lotka-Volterra model, we show that there are in fact two kinds of complexity-stability relationships in disordered dynamical systems:
+    Using a non-linear generalization of the competitive Lotka-Volterra model, we show that there are in fact two kinds of complexity-stability relationships in disordered dynamical systems dominated by competition:
     if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
@@
 \end{equation}
-where the expectation values must be computed self-consistently.
+defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently (see~\cite{Hatton2024} for a discussion on feasibility for systems described by this kind of distributions).
 
@@
 Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
-In this case, the first and second moments formally diverge.
+In this case, the distribution can be normalized but the first and second moments formally diverge.
 It is not uncommon, for example in ecology, to find empirical species abundance distributions characterized by heavy-tails, potentially with diverging moments.
-This is problem can be overcome considering that we are dealing with large, but \emph{finite} populations. 
+This is problem can be overcome considering that we are dealing with large, but \emph{finite}, populations. 
 We are therefore interested in predicting the sample mean and variance for a given $N$.
@@
 The properties of large random dynamical systems are often portrayed in the $(\sigma \sqrt{N},\mu N)$ plane \cite{bunin2017ecological}.
+In particular, we are interested in the case in which the number of degrees of freedom $N$ is the only parameter changing in the system and the strength and heterogeneity of the interactions is fixed.
+This choice, equivalently expressed as having $\tilde{\mu}$ and $\tilde{\sigma}^2$ increasing with $N$, reflects a strong-interaction regime and has been also employed, for example, to study ecological scenarios whith large differences in species abundance and chaotic turnover~\cite{mallmin2024chaotic}.
 If we keep the mean $\mu$ and standard deviation $\sigma$ of interactions fixed, an increase in the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane.
@@
 In this letter, we have shown that the condition $\sigma\sqrt{N}< \mu_s$ does not provide a complete picture of the relationship between complexity and stability. 
-In particular, we have seen that the generalized Lotka-Volterra model, often cited in support of May's general prediction, exemplifies a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
+In particular, we have seen that the competitive Lotka-Volterra model, often cited in support of May's general prediction, exemplifies a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
 In models where self-interactions grow faster than cross-interactions~\cite{Hatton2024,samadder2024interconnection}, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.
@@
 \begin{acknowledgments}
-We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate, and Jacopo Grilli, Vadim Karatayev and Daniel Reuman for feedback on the manuscript.
+We thank Ada Altieri, Matthieu Barbier and Ian Hatton for productive discussions on ecological power laws and the diversity-stability debate and Jacopo Grilli, Vadim Karatayev and Daniel Reuman for feedback on the manuscript.
 Funding for this work was provided by the Alexander von Humboldt Foundation in the framework of the Sofja Kovalevskaja Award endowed by the German Federal Ministry of Education and Research, by the Trieste Laboratory on Quantitative Sustainability - TLQS and by NSF BIO OCE grant 2023473.

### 2024-04-30 scripts/single-run.jl

@@ -4,23 +4,40 @@ foreach(include, glob("*.jl", srcdir()))
 
 using Plots, LaTeXStrings
 
-p = Dict{Symbol,Any}(
+σₑ=.075:.001:.2
+for (i,_) in enumerate(σₑ)
+
+    p = Dict{Symbol,Any}(
     :scaled => false,
     :S => 100,
     :μ => .01,
     :μₛ => .01,
     :σ => .01,
-    :α => 1.0,
-    :β => 1.5,
-    :γ => 1.0,
+    :α => rand(Normal(1., σₑ[i]),100),
+    :β => rand(Normal(1.5,σₑ[i]),100),
+    :γ => rand(Normal(1., σₑ[i]),100),
     :x0 => 1. *ones(100),
     :extinction_threshold => 1e-6,
     :dist => "gamma",
-    :N => 1,
+    :N => 100,
     :seed => rand(UInt)
-);
+    );
+
+    push!(stab, stability!(p))
+
+end
+
+plot(σₑ, stab,
+lw = 2,
+legend = false,
+xlabel = L"\sigma_e",
+ylabel = L"\textrm{Probability \ of \ stability}",
+)
+
+stab = Float64[]
 
 evolve!(p; trajectory=true)
+
 
 #= trajectories =#
 plot(

### 2024-04-30 scripts/single-run.jl
REMEMBER TO REMOVE "Unstructured"

@@ -4,6 +4,10 @@ foreach(include, glob("*.jl", srcdir()))
 
 using Plots, LaTeXStrings
 
+#= Unstructured =#
+
+stab = Float64[]
+
 σₑ=.075:.001:.2
 for (i,_) in enumerate(σₑ)
 
@@ -34,7 +38,23 @@ xlabel = L"\sigma_e",
 ylabel = L"\textrm{Probability \ of \ stability}",
 )
 
-stab = Float64[]
+#= --- =#
+
+p = Dict{Symbol,Any}(
+    :scaled => false,
+    :S => 100,
+    :μ => .01,
+    :μₛ => .01,
+    :σ => .01,
+    :α => 1.,
+    :β => 1.5,
+    :γ => 1.,
+    #:x0 => 1.,
+    :extinction_threshold => 1e-6,
+    :dist => "gamma",
+    :N => 100,
+    :seed => rand(UInt)
+    );
 
 evolve!(p; trajectory=true)
 

### 2024-05-06 papers/response-referee.md
correct typos and highlight missing bits

@@
 $$
-\mathcal{Z}=\int_1^{\infty}x^{-\alpha}.
+\mathcal{Z}=\int_1^{\infty}dxx^{-\alpha}.
 $$
@@
 $$
-with the cut-off $\Lambda$ defined such that $\int_1^{\infty}dxP(x)=1/N$, i.e., such that there is statistically less than 1 variable with value above $\Lambda$ out of $N$ extracted variables.
+with the cut-off $\Lambda$ defined such that $\int_{\Lambda}^{\infty}dxP(x)=1/N$, i.e., such that there is statistically less than 1 variable with value above $\Lambda$ out of $N$ extracted variables.
 For the case $\alpha=3/2$ we have
@@
 
-We thank the referee for highlighting this oversight.
+We thank the referee for raising this point.
 We have now specified that the distribution in Eq.(11) is defined form zero to infinity and we point to reference [17] where a discussion on feasibility for systems described by this kind of distribution is present.

### 2024-05-06 papers/submission/beyond-may.tex

@@
     J_{ij}^* & = - A_{ij}g(x_i^*)h'(x_j^*) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
-    J_{ii}^* & = f'(x_i^*) - g'(x_i^*)f(x_i^*)/g(x_i^*) - A_{ii}g(x_i^*)h'(x_j^*) \ , \label{eq: jac diag}
+    J_{ii}^* & = f'(x_i^*) - g'(x_i^*)f(x_i^*)/g(x_i^*) - A_{ii}g(x_i^*)h'(x_i^*) \ , \label{eq: jac diag}
 \end{align}
@@
 Following e.g.
-Ref.~\cite{Roy2019}, we can derive from Eq.~\eqref{dynamics} a cavity solution which describes the ensemble of the stationary value of $N\gg 1$ variables by means of a single representative radnom variable $x_*$
+Ref.~\cite{Roy2019}, we can derive from Eq.~\eqref{dynamics} a cavity solution which describes the ensemble of the stationary value of $N\gg 1$ variables by means of a single representative random variable $x_*$
 \begin{equation}
@@
 \end{equation}
-defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently (see~\cite{Hatton2024} for a discussion on feasibility for systems described by this kind of distributions).
+defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently. XXX feasibility XXX
 
@@
 In particular, we have seen that the competitive Lotka-Volterra model, often cited in support of May's general prediction, exemplifies a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
-In models where self-interactions grow faster than cross-interactions~\cite{Hatton2024,samadder2024interconnection}, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.
+In models where self-interactions grow slower than cross-interactions~\cite{Hatton2024,samadder2024interconnection}, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.
 

### 2024-06-12 papers/response-referee.md
update response to referee

@@
 
-We agree that indeed a spectrum of different exponents is probably more realistic, and thank the reviewer to pointing that out.
+We agree that indeed a spectrum of different exponents is probably more realistic, and thank the reviewer for pointing that out.
 
@@
 We decided, however, to not include this analysis in the letter because we would like to keep the focus on the existence of the two complexity-stability regimes rather than exploring more detailed ecological scenarios.
-We stick to the case in which $f$, $g$ and $h$ are independent from the variable index, following references [18] and [19].
+We stick to the case in which $f$, $g$ and $h$ are independent of the variable index, following references [18] and [19].
 
@@
 
-We thank the referee for this comment, we modified Equations (7) accoridngly.
+We thank the referee for this comment, we modified Equations (7) accordingly.
 
@@
 
-We agree with the referee that the discussion about the expected values and self-consistency conditions needed more clarifications.
+We agree with the referee that the discussion about the expected values and self-consistency conditions needed more clarification.
 
-In some cases, for example $\alpha=1, \beta=3, \gamma=1$, both expectations converge to finite values and we can safely compute the self-consistency equations.
-However in potential case of interests, and in the example case $\alpha=1, \beta=3/2, \gamma=1$ characterized by a distribution with a power-law tail with exponent $-3/2$, the moments diverge.
-We now modified the discussion removing the gaussian approximation and considering the introduction of a cut-off.
+In some cases, for example, $\alpha=1, \beta=3, \gamma=1$, both expectations converge to finite values and we can safely compute the self-consistency equations.
+However in potential cases of interest, and in the example case $\alpha=1, \beta=3/2, \gamma=1$ characterized by a distribution with a power-law tail with exponent $-3/2$, the moments diverge.
+We now modified the discussion removing the Gaussian approximation and considering the introduction of a cut-off.
 The idea behind this cut-off can be described by the example below.
 
-Consider, as an example which allows for a complete analytical treatment, the case of a power law distribution 
+Consider, as an example that allows for a complete analytical treatment, the case of a power law distribution 
 $$
@@
 
-Consider the case $\alpha=3/2$. The distribution is normalized with $\mathcal{Z}=2$, but the mean diverges. However, we would like to be able to describe the behaviour of the sample mean 
+Consider the case $\alpha=3/2$. The distribution is normalized with $\mathcal{Z}=2$, but the mean diverges. However, we would like to be able to describe the behavior of the sample mean 
 $$
@@
 $$
-The result is plotted below, alongside sample mean for extractions of $N=10,$ $10^2,$ $10^3,$ $10^4,$ $10^5,$ $10^6$.
+The result is plotted below, alongside the sample mean for extractions of $N=10,$ $10^2,$ $10^3,$ $10^4,$ $10^5,$ $10^6$.
 
@@
 
-Finally, we are aware of the fact that reference [27] refers to the scaling of single individual, indeed we mention it to introduce the scaling at the higher level of organization:
+Finally, we are aware of the fact that reference [27] refers to the scaling of single individuals, indeed we mention it to introduce the scaling at the higher level of organization:
 
@@
 We thank the referee for raising this point.
-We have now specified that the distribution in Eq.(11) is defined form zero to infinity and we point to reference [17] where a discussion on feasibility for systems described by this kind of distribution is present.
+We have now specified that the distribution in Eq.(11) is defined from zero to infinity and that in the specific case we analyze the system is always feasible.
+This happens because, given that $\alpha<\beta$, at low enough density each degree of freedom effectively decouples from the others and only feels self-interactions and therefore never crosses zero.
+We also now point to reference [17] where a discussion on the feasibility of ecological systems described by this kind of distributions when a realistic lower threshold beyond which a species can be considered extinct is present.
 
@@
 
-The referee is absolutely correct in pointing out the lack of proper scaling in the derivation of the cavity results, we are grateful for pointing it out.
-We have now fixed that.
-However, we argue that, after a proper derivation of the analytical results, it is appropriate fix the the strength and heterogeneity of the interaction ($mean(A)=\mu$ and $var(A)=\sigma^2$) if our goal is to isolate and study the effects of changes in the number of degrees of freedom.
-We now discuss this, pointing out that it amount to be in a strong-interactions regime, and compare with reference (E Mallmin, A Traulsen, S De Monte *PNAS* 2024, now reference [34] in the manuscript) for an ecological example in which this choice is made to study chaotic turnover of species with broad abundance distributions.
+This is a delicate point and we thank the referee for raising it. This gives us the opportunity to elaborate more on the choice of not scaling the interactions.
+
+As also discussed in reference [25] (end of "I. MODEL DEFINITION"), the choice of scaling or not the interaction depends on the ecological setting of interest. 
+We believe that it is appropriate to fix the strength and heterogeneity of the interaction ($mean(A)=\mu$ and $var(A)=\sigma^2$) if our goal is to study the effects of changes in the number of degrees of freedom.
+In this setting, while the full limit in which $N\to\infty$ leads to the homogeneous interaction case, the cavity approximation does a good job at describing the empirical distributions obtained from simulations as long as $\sigma\sqrt{N}$ is not negligible with respect to $\mu N$.
+
+We now discuss this and point out that it amounts to consider a strong-interactions regime and compare with the reference mentioned by the referee (E Mallmin, A Traulsen, S De Monte *PNAS* 2024, now reference [35] in the manuscript) for an ecological example in which this choice is made to study chaotic turnover of species with broad abundance distributions.
 
@@
 
-We decided to keep $N$ to indicate the number of degrees of freedom because our goal is not to allign with ecological literature, rather we would like to speak to the community of scientists working in complex systems.
+We decided to keep $N$ to indicate the number of degrees of freedom because our goal is not to align with ecological literature, rather we would like to speak to the community of scientists working in complex systems.
 We made this choice in the first place to align with Robert May notation in reference [1].
@@
 
-We are thankful for the critical reveiw overall. We extended the discussion and improved the paper in different aspects following suggestions of the referee.
+We are thankful for the critical review overall. We extended the discussion and improved the paper in different aspects following the suggestions of the referee.

### 2024-06-12 scripts/single-run.jl
update code

@@ -6,14 +6,14 @@ using Plots, LaTeXStrings
 
 p = Dict{Symbol,Any}(
     :S => 100,
-    :μ => .01,
+    :μ => .02,
     :μₛ => .01,
     :σ => .01,
     :α => 1.,
     :β => 1.5,
     :γ => 1.,
     #:x0 => 1.,
-    :extinction_threshold => 1e-6,
+    :extinction_threshold => 1e-10,
     :dist => "gamma",
     :N => 1,
     :scaled => false,
@@ -48,7 +48,7 @@ stephist(p_eq,
 
 E1, E2, Λ = Cavity_Λ(p)
 
-X=[x for x in .9*minimum(p_eq):.001:1.1*maximum(p_eq)]
+X=[x for x in .9*minimum(p_eq):.0001:1.1*maximum(p_eq)]
 plot!(X,[P_n(x, E1, E2, p) for x in X],
 labels = false,
 linewidth = 2,

### 2024-06-12 src/cavity.jl

@@ -13,7 +13,7 @@ function P_n(n, e1_n, e2_n, p)
     μ = (p[:scaled] ? μ/S : μ)
     σ = (p[:scaled] ? σ/S : σ)
     x = max(n,1e-20)
-    return abs(α-β)*x^(α-β-1)/(sqrt(2*π*S*((σ^2+μ^2)*e2_n-μ^2*e1_n^2)))*exp(-(x^(α-β)-S*μ*e1_n)^2/(2*S*((σ^2+μ^2)*e2_n-μ^2*e1_n^2)))
+    return abs(α-β)*x^(α-β-1)/(sqrt(2*π*S*((σ^2)*e2_n)))*exp(-(x^(α-β)-S*μ*e1_n)^2/(2*S*((σ^2)*e2_n)))
 end
 
 #= cavity calculation for fraction of positive equilibriums 
@@ -58,7 +58,7 @@ function Cavity(p;
     Iter = 2000, #number of iteration
     rela = .01, #relax parameter for fixed point
     tol = 1e-9, #requested tolerance for numerical integrator
-    Λ = p[:S]^(p[:β]-p[:α]), #Cut-off for integration
+    Λ = .001*p[:S]^(p[:β]-p[:α]), #Cut-off for integration
     n_min = p[:extinction_threshold], #lower bound for integration
     e1_init = (p[:scaled] ? 1/p[:μ] : 1/p[:μ]/p[:S]), #initial guess for e1_n
     e2_init = (p[:scaled] ? 1/p[:μ] : 1/p[:μ]/p[:S]) #initial guess for e1_n

### 2024-06-12 papers/response-referee.md

@@
 We address each of the comments below point by point.
-We believe the results of our paper are now more transparent thanks to the work needed to address the weaknesses hihglighted by the referee.
+We believe the results of our paper are now more transparent thanks to the work needed to address the points raised by the referee.
 
@@
 This happens because, given that $\alpha<\beta$, at low enough density each degree of freedom effectively decouples from the others and only feels self-interactions and therefore never crosses zero.
-We also now point to reference [17] where a discussion on the feasibility of ecological systems described by this kind of distributions when a realistic lower threshold beyond which a species can be considered extinct is present.
+We also now point to reference [17] where a discussion on the feasibility of ecological systems described by this kind of distribution when a realistic lower threshold beyond which a species can be considered extinct is present.
 
@@
 We are grateful to the referee for this comment, it helps to clarify the scope of our work.
-We have now specified everywhere in the manuscript, including Abstract, Introduction and Discussion, that we compare out results to the *competitive* Lotka-Volterra model and in general that our results are relevant for predominantly competitive systems.
+We have now specified everywhere in the manuscript, including Abstract, Introduction and Discussion, that we compare our results to the *competitive* Lotka-Volterra model and in general that our results are relevant for predominantly competitive systems.
 

### 2024-06-12 papers/submission/beyond-may.tex

@@
     However, this general conclusion is not always supported by empirical observation: from cells to biomes, biological systems are large, complex and, often, stable.
-    In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 
-    Using a non-linear generalization of the competitive Lotka-Volterra model, we show that there are in fact two kinds of complexity-stability relationships in disordered dynamical systems dominated by competition:
+    In this paper, we revisit May's argument in light of recent developments in both ecology and random matrix theory. 
+    Using a non-linear generalization of the competitive Lotka-Volterra model, we show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems:
     if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
-    Our result shows that May's principle that "complexity begets instability" is not a general property of complex systems.
 \end{abstract}
@@
 \emph{Random interactions: stability condition.---}We now consider the case of random interactions.
-Specifically, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with scaled mean $\mu=\tilde{\mu}/N$ and standard deviation $\sigma=\tilde{\sigma}/\sqrt{N}$~\cite{bunin2017ecological}.
+Specifically, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$.
 We assume the diagonal elements $A_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) and standard deviation $\sigma$.
@@
 \end{equation}
-In the GLV model, $\psi(x) = \mu_s - \tilde{\mu}/N$ (independently of $x$) and for large $N$ we recover the classical condition $\tilde{\sigma} = \sigma\sqrt{N} < \mu_s$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}.
-In general, however, the dependence on equilibrium values $x_i^*$ (which in turn depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.
+In the GLV model, $\psi(x) = \mu_s - \mu$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < \mu_s - \mu$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}.
+In general, however, the dependence on equilibrium values $x_i^*$ (which in turn depends on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.
 
@@
 Following e.g.
-Ref.~\cite{Roy2019}, we can derive from Eq.~\eqref{dynamics} a cavity solution which describes the ensemble of the stationary value of $N\gg 1$ variables by means of a single representative random variable $x_*$
+Ref.~\cite{Roy2019}, we can derive from Eq.~\eqref{dynamics} a cavity solution that describes the ensemble of the stationary solutions of the system by means of a single representative random variable $x_*$
 \begin{equation}
-    0 = x_*^{\alpha}-A_sx_*^{\beta+\gamma}-x_*^{\beta}\big( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
+    0 = x_*^{\alpha}-A_sx_*^{\beta+\gamma}-x_*^{\beta}\big( \mu N \langle x_*^{\gamma}\rangle + \sigma\sqrt{N} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
 \label{eq: dmft}
@@
     We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
-    does not imply that increasing $N$, at fixed $\mu$ and $\sigma$, will destabilize the system, because $\tilde{\mu}=\mu N$ would also change.
+    does not imply that increasing $N$, at fixed $\mu$ and $\sigma$, will destabilize the system, because $\mu N$ would also change.
     }
@@
 \end{figure}
+We are interested in the case in which the number of degrees of freedom $N$ is the only parameter changing in the system and the strength and heterogeneity of the interactions are fixed~\cite{kessler2015generalized,fried2016communities}.
+This choice reflects a strong-interaction regime and has been employed, for example, to study ecological scenarios with large differences in species abundance and chaotic turnover~\cite{mallmin2024chaotic}.
+As we show below, the cavity approximation does a good job at describing the effects of heterogeneity in the interactions as long as $\sigma \sqrt{N}$ is not negligible relative to $\mu N$, at which point we recover the homogeneous case.
+
 The equation above can be solved for $x_*$ for specific values of
@@
 \begin{equation} \label{eq: cavity solution}
-    x_* = \left( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
+    x_* = \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
 \end{equation}
@@
 \begin{equation}\label{eq: dist general}
-    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\tilde{\sigma}^2 \langle x_*^{2\gamma}\rangle}}
-    \exp{\left\{-\frac{(x_*^{\alpha-\beta}- \tilde{\mu} \langle x_*^{\gamma}\rangle)^2}{2\tilde{\sigma}^2\langle x_*^{2\gamma}\rangle}\right\}} \, ,
+    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2N \langle x_*^{2\gamma}\rangle}}
+    \exp{\left\{-\frac{(x_*^{\alpha-\beta}- \mu N \langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
 \end{equation}
-defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently. XXX feasibility XXX
+defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently.
 
@@
 Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
-In this case, the distribution can be normalized but the first and second moments formally diverge.
-It is not uncommon, for example in ecology, to find empirical species abundance distributions characterized by heavy-tails, potentially with diverging moments.
-This is problem can be overcome considering that we are dealing with large, but \emph{finite}, populations. 
+In this case the system is always feasible (see Ref.~\cite{Hatton2024} for a discussion on the inclusion of extinction thresholds at low density in an ecological setting) and the distribution can be normalized.
+However, the first and second moments formally diverge.
+This is problem can be overcome considering that we are dealing with a large, but \emph{finite}, number of degrees of freedom. 
 We are therefore interested in predicting the sample mean and variance for a given $N$.
@@
 The properties of large random dynamical systems are often portrayed in the $(\sigma \sqrt{N},\mu N)$ plane \cite{bunin2017ecological}.
-In particular, we are interested in the case in which the number of degrees of freedom $N$ is the only parameter changing in the system and the strength and heterogeneity of the interactions is fixed.
-This choice, equivalently expressed as having $\tilde{\mu}$ and $\tilde{\sigma}^2$ increasing with $N$, reflects a strong-interaction regime and has been also employed, for example, to study ecological scenarios whith large differences in species abundance and chaotic turnover~\cite{mallmin2024chaotic}.
 If we keep the mean $\mu$ and standard deviation $\sigma$ of interactions fixed, an increase in the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane.
-In the GLV model, the boundary between the stable and unstable parameter regions is the horizontal line $\sigma_c\sqrt{N} = \mu_s$ \cite{bunin2017ecological}, hence stability is never possible at large $N$.
+In the GLV model, the boundary between the stable and unstable parameter regions is the horizontal line $\sigma_c\sqrt{N} = \mu_s - \mu N / N \approx \mu_s$ \cite{bunin2017ecological}, hence stability is never possible at large $N$.
 
@@
 
-Finally, Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\tilde{\mu}=\mu N$ and $\gamma = 1$.
-(Numerically, we define $\sigma_c\sqrt{N}$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations, we use $\mu_s = \mu$, which would never be stable in the GLV model.
+Finally, Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$.
+(Numerically, we define $\sigma_c$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations, we use $\mu_s = \mu$, which would never be stable in the GLV model.
 Fig.~\ref{fig: alpha-beta} illustrates the transition between a ``May" phase for $\alpha \geq \beta$ and an ``anti-May" phase for $\alpha < \beta$, showing that our analytical results are robust to heterogeneous interactions.
@@
 
-A modification of our model~\eqref{dynamics} of general ecological relevance describing the case in which per capita growth rates depend only on the sum of the other densities (weighted by interaction strength) is $\dot{x_i}/x_i=1-F(\sum_jA_{ij}x_j)$, where $F$ is an arbitrary function.
+A modification of our model~\eqref{dynamics} of general ecological relevance describing the case in which per capita growth rates depend only on a linear combination of the densities is $\dot{x_i}/x_i=1-F(\sum_jA_{ij}x_j)$, where the function $F$ characterizes the effects of competition.
 General models of this kind are used, for example, to study the growth of a clonal population in the presence of a limiting factor~\cite{mazzolini2023universality}.

### 2024-06-14 papers/submission/beyond-may.tex
update Matteo affil and remove feas. tresh.

@@
 \author{Matteo Smerlak}
-\affiliation{Laboratoire Biophysique et Evolution, ESPCI ParisTech, 10 rue Vauquelin, 75005 Paris, France}
 \affiliation{Capital Fund Management, 23 Rue de l'Université, 75007 Paris, France}
@@
 Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
-In this case the system is always feasible (see Ref.~\cite{Hatton2024} for a discussion on the inclusion of extinction thresholds at low density in an ecological setting) and the distribution can be normalized.
-However, the first and second moments formally diverge.
+In this case the system is always feasible and the distribution can be normalized. However, the first and second moments formally diverge.
 This is problem can be overcome considering that we are dealing with a large, but \emph{finite}, number of degrees of freedom. 

### 2024-06-24 papers/response-referee.md
rebase onto master

@@
 
-We can consider the case in which the exponents are not identical for every variable $x_i$ but instead we have them from Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and, for simplicity, all with the same standard deviation $\sigma_e$.
+We consider now the case in which the exponents are not identical for every variable $x_i$ but instead we have them from Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and, for simplicity, all with the same standard deviation $\sigma_e$.
 
-Simulations show that all our results are robust as long as $\sigma_e$ is small enough that $\displaystyle{\min_i\beta_i>\max_i\alpha_i}$.
-
-As an example, we show in the plot below the probability of stability vs. $\sigma_e$ for a system with $S=100$, $\mu=\mu_s=\sigma=0.01$, $\alpha=\gamma=1$ and $\beta=3/2$.
-The probability of stability is obtained as the fraction of stable systems out of $100$ realizations for each value of $\sigma_e$.
-![](../plots/stoch-exp.png)
-
-We decided, however, to not include this analysis in the letter because we would like to keep the focus on the existence of the two complexity-stability regimes rather than exploring more detailed ecological scenarios.
-We stick to the case in which $f$, $g$ and $h$ are independent of the variable index, following references [18] and [19].
+We now added a Supplemental Material document where report results of simulations showing that our results are robust as long as $\sigma_e$ is small enough that $\displaystyle{\min_i\beta_i>\max_i\alpha_i}$.
+We also added a paragraph in the main text where we discuss this point.
 
@@
 
-We thank the referee for this comment, we modified Equations (7) accordingly.
+We modified Equations (7) accordingly.
 
@@
 
-We agree with the referee that the discussion about the expected values and self-consistency conditions needed more clarification.
+We agree with the referee that the discussion about the expected values and self-consistency conditions needed clarifications.
 
@@
 We now modified the discussion removing the Gaussian approximation and considering the introduction of a cut-off.
-The idea behind this cut-off can be described by the example below.
-
-Consider, as an example that allows for a complete analytical treatment, the case of a power law distribution 
-$$
-P(x)=\frac{x^{-\alpha}}{\mathcal{Z}} ,
-$$
-defined from 1 to $\infty$ and with 
-$$
-\mathcal{Z}=\int_1^{\infty}dxx^{-\alpha}.
-$$
-
-Consider the case $\alpha=3/2$. The distribution is normalized with $\mathcal{Z}=2$, but the mean diverges. However, we would like to be able to describe the behavior of the sample mean 
-$$
-\bar{x}_N\equiv\frac{1}{N}\sum_{i=1}^Nx_i,
-$$
-where the $x_i$ are extracted from $P(x)$.
-For this purpose, we can define the quantity
-$$
-\langle x\rangle_{\Lambda}\equiv\int_1^{\Lambda}dxP(x)x,
-$$
-with the cut-off $\Lambda$ defined such that $\int_{\Lambda}^{\infty}dxP(x)=1/N$, i.e., such that there is statistically less than 1 variable with value above $\Lambda$ out of $N$ extracted variables.
-For the case $\alpha=3/2$ we have
-$$
-\frac{1}{2}\int_{\Lambda}^{\infty}dxx^{-3/2}=\Lambda^{-1/2},
-$$
-and therefore $\Lambda=N^2$. We have for the mean
-$$
-\langle x\rangle_{\Lambda}=\frac{1}{2}\int_1^{N^2}dxx^{-1/2} = N-1.
-$$
-The result is plotted below, alongside the sample mean for extractions of $N=10,$ $10^2,$ $10^3,$ $10^4,$ $10^5,$ $10^6$.
-
-![](../plots/cut-off-justification.png)
+The idea behind this cut-off is described in the main text and in the Supplemental Material.
 
@@
 
-We thank the referee for the thorough and critical review.
+We thank the referee for their thorough and critical review of our work.
 We address each of the comments below point by point.
-We believe the results of our paper are now more transparent thanks to the work needed to address the points raised by the referee.
+We believe the results of our paper are more transparent thanks to these modifications.
 
@@
 
-We understand the criticism of the referee, pointing out the need for better justification of models of this kind in ecology.
-Our goal in this letter, however, is not to propose a model for a specific ecological scenario.
-Rather, we employ this model for complex systems (following references [18] and [19]) to show in a minimal setting the existence of the two opposite complexity-stability relationships in disordered dynamical systems, possibly with applications beyond ecology.
+We recently published a separate paper (cited as Ref.
+[17]) to outline in some detail the ecological motivations and consequences of competitive models with sublinear production.
+The goal of the present letter is different.
+Following Refs.
+[18] and [19], we show in a minimal setting (and for an audience of physicists) the existence of the two opposite complexity-stability relationships in disordered dynamical systems.
+We have reword
 
-Nevertheless, the examples we provide are indicative that models of this class can be relevant in ecology (see reference [17], where a discussion of possible mechanisms is present; even though admittedly more work is required in that direction, which is beyond the scope of our paper).
-
-Finally, we are aware of the fact that reference [27] refers to the scaling of single individuals, indeed we mention it to introduce the scaling at the higher level of organization:
-
-"Biologically, the growth of organisms (populations of cells) has long been known to scale like $f (x) \sim x^k$ with $k\simeq 3/4$ [27], which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
-For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities [17, 28]".
+We are aware that reference [27] refers to the scaling of single individuals. We now mention that explicitly.
 
@@
 This happens because, given that $\alpha<\beta$, at low enough density each degree of freedom effectively decouples from the others and only feels self-interactions and therefore never crosses zero.
-We also now point to reference [17] where a discussion on the feasibility of ecological systems described by this kind of distribution when a realistic lower threshold beyond which a species can be considered extinct is present.
 
@@
 We believe that it is appropriate to fix the strength and heterogeneity of the interaction ($mean(A)=\mu$ and $var(A)=\sigma^2$) if our goal is to study the effects of changes in the number of degrees of freedom.
-In this setting, while the full limit in which $N\to\infty$ leads to the homogeneous interaction case, the cavity approximation does a good job at describing the empirical distributions obtained from simulations as long as $\sigma\sqrt{N}$ is not negligible with respect to $\mu N$.
+In this setting, while the full limit in which $N\to\infty$ leads to the homogeneous interaction case with $x^*\to0$ due to competitive pressure, the cavity approximation does a good job at describing the empirical distributions obtained from simulations as long as $\sigma\sqrt{N}$ is not negligible with respect to $\mu N$.
 

### 2024-06-25 papers/response-referee.md
update draft and response

@@
 
-We modified Equations (7) accordingly.
+We thank the referee for this comment, we modified Equations (7) accordingly.
 
@@
 
-We agree with the referee that the discussion about the expected values and self-consistency conditions needed clarifications.
+We agree with the referee that the discussion about the expected values and self-consistency conditions needed more clarification.
 
@@
 
-We thank the referee for their thorough and critical review of our work.
+We thank the referee for the thorough and critical review.
 We address each of the comments below point by point.
-We believe the results of our paper are more transparent thanks to these modifications.
+We believe the results of our paper are now more transparent thanks to the work needed to address the points raised by the referee.
 
@@
 
-We recently published a separate paper (cited as Ref.
-[17]) to outline in some detail the ecological motivations and consequences of competitive models with sublinear production.
-The goal of the present letter is different.
-Following Refs.
-[18] and [19], we show in a minimal setting (and for an audience of physicists) the existence of the two opposite complexity-stability relationships in disordered dynamical systems.
-We have reword
+We understand the criticism of the referee, pointing out the need for better justification of models of this kind in ecology.
+Our goal in this letter, however, is not to propose a model for a specific ecological scenario.
+Rather, we employ this model for complex systems (following references [18] and [19]) to show in a minimal setting the existence of the two opposite complexity-stability relationships in disordered dynamical systems, possibly with applications beyond ecology.
+
+Nevertheless, the examples we provide are indicative that models of this class can be relevant in ecology (see reference [17], where a discussion of possible mechanisms is present; even though admittedly more work is required in that direction, which is beyond the scope of our paper).
 

### 2024-06-25 papers/submission/beyond-may.tex
update draft and response

@@
 From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
-Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
+Biologically, at the individual organism level (populations of cells), growth has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
 For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities~\cite{Hatton2015,Hatton2024}.
@@
 We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$ where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$.
-In orther words such that there is less 1 variable with density above $\Lambda$.
+In orther words such that there is less 1 variable with density above $\Lambda$ (see Supplemental Material).
 
@@
 
-Finally, Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$.
+Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$.
 (Numerically, we define $\sigma_c$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations, we use $\mu_s = \mu$, which would never be stable in the GLV model.
 Fig.~\ref{fig: alpha-beta} illustrates the transition between a ``May" phase for $\alpha \geq \beta$ and an ``anti-May" phase for $\alpha < \beta$, showing that our analytical results are robust to heterogeneous interactions.
+
+Finally, we check the robustness of our findings in the scenario in which the exponents are not the same for every variable $x_i$.
+As a test, we can extract them from different Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and, for simplicity, all with the same standard deviation $\sigma_e$.
+Simulations show (see Supplemental Material) that our results hold as long as $\sigma_e$ is small enough, in particular, we observe a loss of stability when $\min_i\beta_i>\max_i\alpha_i$.
 

### 2024-06-25 papers/submission/SM-beyond-may.tex
create SM

%\documentclass[aps,prl,reprint]{revtex4-2}
\documentclass[aps,prl,superscriptaddress,twocolumn]{revtex4}
%\documentclass[aps,prl,twocolumn]{revtex4-2}
%\documentclass[pre,amsmath,amssymb,superscriptaddress,showpacs,twocolumn]{revtex4-1}
\usepackage{amsmath,amssymb,graphicx}
%\input epsf
\usepackage{amsbsy}
\usepackage{latexsym}
\usepackage{color}
\usepackage{graphicx}
\usepackage{psfrag}
\usepackage[normalem]{ulem}
\usepackage{bm}
\usepackage{url}
\usepackage{notes2bib}
%\usepackage[inline]{showlabels}
%\usepackage{float}
\usepackage{soul}

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\bea}{\begin{eqnarray}}
\newcommand{\eea}{\end{eqnarray}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bz}{\mathbf{0}}
%\newcommand{\comment}[1]{}
\newcommand{\col}{\textcolor}
\newcommand{\comment}[1]{}

\setcounter{equation}{11}

\begin{document}

\title{\Large Supplemental Material for ``Complexity-stability relationships disordered dynamical systems''}

\author{Onofrio Mazzarisi}
\affiliation{The Abdus Salam International Centre for Theoretical Physics (ICTP), Strada Costiera 11, 34014 Trieste, Italy}
\affiliation{National Institute of Oceanography and Applied Geophysics (OGS), via Beirut 2, 34014 Trieste, Italy}

\author{Matteo Smerlak}
\affiliation{Capital Fund Management, 23 Rue de l'Université, 75007 Paris, France}

\date{\today}

\maketitle

{\it Cut-off.---} In this section we consider a case amenable to analytical treatment to exemplify the argument behind the cut-off $\Lambda$ introduced in the main text to deal with diverging moments distributions.

Consider the case of a power law distribution 
\begin{equation}
  P(x)=\frac{x^{-\beta}}{\mathcal{Z}} ,
\end{equation}
defined from 1 to $\infty$ and with 
\begin{equation}
  \mathcal{Z}=\int_1^{\infty}dxx^{-\beta}.  
\end{equation}

Let us choose $\beta=3/2$. Notice that this power law describes the behavior of the distribution we consider in the main text for our example with $\alpha=\gamma=1$ and $\beta=3/2$.
The distribution is normalized with $\mathcal{Z}=2$, but the mean diverges. However, we would like to be able to describe the behavior of the sample mean 
\begin{equation}
  \bar{x}_N\equiv\frac{1}{N}\sum_{i=1}^Nx_i,
\end{equation}

where the $x_i$ are extracted from $P(x)$.
For this purpose, we can define the quantity
\begin{equation}
  \langle x\rangle_{\Lambda}\equiv\int_1^{\Lambda}dxP(x)x,
\end{equation}
with the cut-off $\Lambda$ defined such that $\int_{\Lambda}^{\infty}dxP(x)=1/N$, i.e., such that there is statistically less than 1 variable with value above $\Lambda$ out of $N$ extracted variables. For the case $\beta=3/2$ we have 
\begin{equation}
  \frac{1}{2}\int_{\Lambda}^{\infty}dxx^{-3/2}=\Lambda^{-1/2},
\end{equation}
and therefore $\Lambda=N^2$. We have for the mean
\begin{equation}
  \langle x\rangle_{\Lambda}=\frac{1}{2}\int_1^{N^2}dxx^{-1/2} = N-1.
\end{equation}
The result is plotted in Fig.~\ref{fig: SM-cut-off}, alongside the sample mean for extractions of $N=10,$ $10^2,$ $10^3,$ $10^4,$ $10^5,$ $10^6$.

\begin{figure}[h!]
  \includegraphics[width=.475\textwidth]{cut-off-justification.pdf}
  \caption{Probability of stability vs. $\sigma_e$ for a system with $S=100$, $\mu=\sigma=0.01$, $\alpha=\gamma=1$ and $\beta=3/2$.
  }
  \label{fig: SM-cut-off}
\end{figure}

{\it Heterogeneous exponents.---} In this section we explore the robustness of our findings in the case in which the exponents characterizing the dynamics of each degree of freedom $x_i$ are not identical for all $i$.

Consider the system
\begin{equation}
  \dot{x}_i = x_i^{\alpha_i} - \sum_jA_{ij}x_i^{\beta_i}x_j^{\gamma_i}
\end{equation}
where the exponents $\alpha_i$, $\beta_i$, $\gamma_i$, are extracted from Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and with the same standard deviation $\sigma_e$. The interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$.
Our results are robust when $\sigma_e$ is small enough and we observe a loss of stability when $\min_i\beta_i>\max_i\alpha_i$.
As an example, we show in the plot in Fig.~\ref{fig: SM-het-exp} the probability of stability vs. $\sigma_e$ for a system with $S=100$, $\mu=\sigma=0.01$, $\alpha=\gamma=1$ and $\beta=3/2$.
The probability of stability is obtained as the fraction of stable systems out of $100$ realizations for each value of $\sigma_e$.

\begin{figure}[h!]
  \includegraphics[width=.475\textwidth]{stoch-exp.pdf}
  \caption{Probability of stability vs. $\sigma_e$ for a system with $S=100$, $\mu=\sigma=0.01$, $\alpha=\gamma=1$ and $\beta=3/2$.
  }
  \label{fig: SM-het-exp}
\end{figure}

\end{document}

### 2024-06-26 papers/submission/beyond-may.tex
minor edits

@@
 
-\title{Complexity-stability relationships disordered dynamical systems}
+\title{Complexity-stability relationships in disordered dynamical systems}
 
@@
 In fact, empirical observation suggests the opposite relationship between complexity and stability: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, extinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}.
-The tension between May's theoretical argument and observation is at the center of the longstanding ``diversity-stability debate" in ecology \cite{McCann2000, Loreau2022}.
+The tension between May's theoretical argument and observation is at the center of the longstanding ``diversity-stability debate" in ecology \cite{McCann2000, Loreau2022,Hatton2024}.
 
@@
 But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions (with zero 
-mean and variance $\sigma^2$), and $-I$ corresponds to stabilizing self-interactions on some natural timescale, 
-then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
-This places a sharp constraint on both diversity $N$ and interaction strength $\sigma$, often referred to as ``complexity begets instability".
+mean and variance $\sigma^2$), and $-I$ corresponds to stabilizing self-interactions on some natural timescale,  the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
+This condition places a sharp constraint on both diversity $N$ and interaction strength $\sigma$, often referred to as ``complexity begets instability".
 (This argument generalizes to $\langle B_{ij}\rangle \neq 0$, incomplete connectivity, or correlated interactions \cite{allesina2015stability}.)
@@
 Many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}.
-In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: in random dynamical systems, stability does not necessarily decrease with dimensionality and interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
-For an in-depth discussion of the ecological implications of our result, see \cite{Hatton2024}.
+In this Letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: in random dynamical systems, stability does not necessarily decrease with dimensionality and interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
+For an in-depth discussion of the ecological implications of our result, see our recent paper \cite{Hatton2024}.
 
@@
 Non-zero diagonal elements $A_{ii}$ accounts for additional self-interactions with the same functional form as cross-interactions.
-We refer to $f$ as the ``production function'' and $g$ as the ``response function''.
+We refer to $f$ as the ``production function'' and $g$ as the ``coupling function''.
 
@@
 to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
-The classic generalized Lotka-Volterra (GLV) competition model studied by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear.
-
+In particular, the classic generalized Lotka-Volterra (GLV) competition model studied by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear.
 However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
@@
 an elementary calculation shows that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*)g(x^*)h'(x^*) > 0$, or $\psi(x^*)>0$ if we assume that $g$ and $h$ are positive, increasing functions.
-This stability condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'(x^*)/f(x^*) - g'(x^*)/g(x^*)$).
+This stability condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and coupling functions near the equilibrium ($f'(x^*)/f(x^*) - g'(x^*)/g(x^*)$).
 
@@
 We are interested in the case in which the number of degrees of freedom $N$ is the only parameter changing in the system and the strength and heterogeneity of the interactions are fixed~\cite{kessler2015generalized,fried2016communities}.
-This choice reflects a strong-interaction regime and has been employed, for example, to study ecological scenarios with large differences in species abundance and chaotic turnover~\cite{mallmin2024chaotic}.
+Compared to the case where $\mu$ and $\sigma$ scale with $N$ (considered in \cite{bunin2017ecological} and follow-up works), our choice reflects a strong-interaction regime; it has been employed, for example, to study ecological scenarios with large differences in species abundance and chaotic turnover~\cite{mallmin2024chaotic}.
 As we show below, the cavity approximation does a good job at describing the effects of heterogeneity in the interactions as long as $\sigma \sqrt{N}$ is not negligible relative to $\mu N$, at which point we recover the homogeneous case.
@@
 defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently.
-
 Equipped with the equilibrium distribution $P(x_*)$, we can compute, for given $N$ and $\mu$, the maximal heterogeneity $\sigma_c$ compatible with the linear stability condition $N\langle (\sigma/\psi(x_*))^2\rangle < 1$.
@@
 In this case the system is always feasible and the distribution can be normalized. However, the first and second moments formally diverge.
-This is problem can be overcome considering that we are dealing with a large, but \emph{finite}, number of degrees of freedom. 
+This problem can be overcome by considering that we are dealing with a large, but \emph{finite}, number of degrees of freedom. 
 We are therefore interested in predicting the sample mean and variance for a given $N$.
 Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$.
-We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$ where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$.
-In orther words such that there is less 1 variable with density above $\Lambda$ (see Supplemental Material).
+We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$, where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$, see Supplemental Material.
 
-Following, e.g.,~\cite{Cui2020,Hatton2024} we can compute self-consistently $\Lambda$, $\langle x_*\rangle_{\Lambda}$ and $\langle x_*^2\rangle_{\Lambda}\equiv\int_0^{\Lambda}dx_*x_*^2P(x_*)$.
+Following, e.g.,~\cite{Cui2020,Hatton2024} we can compute self-consistently: $\Lambda$, $\langle x_*\rangle_{\Lambda}$ and $\langle x_*^2\rangle_{\Lambda}\equiv\int_0^{\Lambda}dx_*x_*^2P(x_*)$.
 The resulting distribution is plotted against simulations in Fig.~\ref{fig: cavity sol.}.
@@
 Finally, we check the robustness of our findings in the scenario in which the exponents are not the same for every variable $x_i$.
-As a test, we can extract them from different Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and, for simplicity, all with the same standard deviation $\sigma_e$.
-Simulations show (see Supplemental Material) that our results hold as long as $\sigma_e$ is small enough, in particular, we observe a loss of stability when $\min_i\beta_i>\max_i\alpha_i$.
+As a test, we can sample exponents from different Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and, for simplicity, all with the same standard deviation $\sigma_e$.
+Simulations show that our results hold as long as $\sigma_e$ is small enough, in particular, we observe a loss of stability when $\min_i\beta_i>\max_i\alpha_i$ (Supplemental Material).
 
@@
 \emph{Discussion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and other fields. 
-In this letter, we have shown that the condition $\sigma\sqrt{N}< \mu_s$ does not provide a complete picture of the relationship between complexity and stability. 
+In this Letter, we have shown that the condition $\sigma\sqrt{N}< \mu_s$ does not provide a complete picture of the relationship between complexity and stability. 
 In particular, we have seen that the competitive Lotka-Volterra model, often cited in support of May's general prediction, exemplifies a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
@@
 
-A modification of our model~\eqref{dynamics} of general ecological relevance describing the case in which per capita growth rates depend only on a linear combination of the densities is $\dot{x_i}/x_i=1-F(\sum_jA_{ij}x_j)$, where the function $F$ characterizes the effects of competition.
+A modification of our model~\eqref{dynamics} of general ecological relevance describing the case in which per capita growth rates depend on a linear combination of the densities is $\dot{x_i}/x_i=1-F(\sum_jA_{ij}x_j)$, where the function $F$ characterizes the effects of competition.
 General models of this kind are used, for example, to study the growth of a clonal population in the presence of a limiting factor~\cite{mazzolini2023universality}.
 All the models in this class are destabilized by diversity.
-The Jacobian matrix at equilibrium is $J_{ij}^*=-A_{ij}x_i^*F'(\sum_jA_{ij}x_j^*) \quad \forall i,j$ leading to the same simplification of the stability condition~\eqref{eq: random-stability} possible for GLV.
+(The Jacobian matrix at equilibrium is $J_{ij}^*=-A_{ij}x_i^*F'(\sum_jA_{ij}x_j^*)$, leading to the same simplification of the stability condition~\eqref{eq: random-stability} as in the GLV model.)
 

### 2024-09-17 papers/response-referee/old/response-referee.md
Refactor response-referee.md and response-referee.tex files

--- 2024-06-24 papers/response-referee.md
@@
+# Response
+## Referee A
+
+We appreciate the positive feedback from the referee. We address the remaining comment below.
+
+*In my opinion, the arguments leading to Eq. (11) and the discussion*  
+*after this equation are not very clear. What is the solution to the*  
+*consistency condition for <x_^gamma> and <x_^2gamma>? Do these two*  
+*quantities depend on N, and if so, how?*
+
+*For example, for the case gamma=1, discussed after Eq. (11), both*  
+*means correspond exactly to the moments <x_> and <x_^2> and, as the*  
+*authors noted, are not well defined for the probability distribution*  
+*(11). This means that conclusions need to be drawn more carefully.*
+
+We have now included the analysis on the cut-off argument previously in the SI in as an appendix in the main text, as we changed the format from letter to regular article. 
+Moreover, we have now modified the paragraph below Eq.
+(11) and conclude with a sentence clarifying that the same "renormalization" procedure we describe for $\alpha = 1$, $\beta = 2/3$, and $\gamma = 1$ can be extended to other choices of exponents in the case they result in diverging expectation values.
+Finally, for clarity, we now write explicitly the self-consistent equations for $\Lambda$, $\langle x_{*}\rangle_{\Lambda}$, and, $\langle x_{*}^2\rangle$.
+
+## Referee B
+
+*The authors clarify most of my previous concerns and clarify the*  
+*scaling issue, and the paper is surely improved. I have still one*  
+*relevant concern, I hope the authors fix before publication.*
+
+We thank the referee for the previous comments, which helped improving the paper, and address the last concern below.
+
+- *In the response to reviewers the authors claim they better highlight*  
+	*that their result are valid only for competitive Lotka Volterra. It is*  
+	*true they added a couple of times the word competitive but in my*  
+	*opinion the paper is not clear enough on this points and the results* *are presented are too general.*
+
+1) *The title "Beyond May: Complexity-stability relationships in*  
+	*disordered dynamical systems" seem to present results that are* 
+	*completely general. I suggest to make a more proper title such as*  
+	*""Beyond May: Complexity-stability relationships in competitive*  
+	*disordered Generalized Lotka Volterra Dynamics", or something similar.*  
+	*Otherwise the title is misleading.*
+
+	 We have now modified the title to make it clear that our work regards competitive disordered interactions only.
+We have additionally modified the abstract to stress our focus on competitive systems "We focus on competitive systems, and, using ...".
+
+2) *Many comparison are made with respect to systems that have not such*  
+	*constraint (positive mu), and this makes confusion.*
+
+	*E.g. 1, " the classic generalized Lotka-Volterra (GLV) competition*  
+	*model studied by Bunin and collaborators [25, 26] corresponds to f, g,*  
+	*h all linear." But this is not true as in Bunin and Biroli works mu*  
+	*can also be negative, is not constrained to assume positive value.*
+
+	We now modified this sentence specifying that the linear case of our model corresponds to GLV only when constrained to competitive interactions.
+
+	*E.g. 2, "Compared to the case where μ and σ scale with N (considered*  
+	*in [25] and follow-up works), our choice reflects a strong-interaction*  
+	*regime; it has been employed, for example, to study ecological*  
+	*scenarios with large differences in species abundance and chaotic*  
+	*turnover [35]". Again in the cited papers mu can be also negative.*
+
+	We have removed the reference to [25] in order to avoid confusion. 
+	We kept reference [35], because in this paper only competitive interactions ($\mu > 0$) are considered, as in our work.
+
+- *I'm not sure it is worth to have a supplementary file. If it is not*  
+	*published as letter and thus there is room it would be suitable to*  
+	*better explain already in the main text the finite size argument to*  
+	*solve self consist equations and the sensitivity to random exponents*
+
+	*Justification for/against Letter:*
+
+	*These calculations hold for a very specific cases and makes many*  
+	*assumption that in general do not hold. Having more space the authors*  
+	*could include the SI in the main text.*
+
+	Following the indication of the referee, we have now included the SI in the main text and changed the format from Letter to Regular article.
+
+---
+# OLD
 ## Referee A
@@
 [18] and [19], we show in a minimal setting (and for an audience of physicists) the existence of the two opposite complexity-stability relationships in disordered dynamical systems.
-We have reword
 

### 2024-09-17 papers/response-referee/response-referee.tex

\documentclass{article}
\usepackage{graphicx} % Required for inserting images

\usepackage{amsmath, amssymb} 
\usepackage{xcolor}

\usepackage{hyperref}
\hypersetup{hidelinks}

\usepackage{etoolbox}
\patchcmd{\thebibliography}{\section*}{\section}{}{}

\newcommand{\pois}{\textrm{Poisson}}
\newcommand{\blue}{\textcolor{blue}}

\begin{document}

\section*{Referee A}

We appreciate the positive feedback from the referee. We address the remaining comment below. 
\\

\blue{In my opinion, the arguments leading to Eq. (11) and the discussion  after this equation are not very clear. 
What is the solution to the consistency condition for $\langle x_*^{\gamma}\rangle$ and $\langle x_*^{2\gamma}\rangle$?
Do these two quantities depend on $N$, and if so, how?} 
\\

\blue{For example, for the case $\gamma=1$, discussed after Eq. (11), both means correspond exactly to the moments $\langle x_*\rangle$ and $\langle x_*^2\rangle$ and, as the authors noted, are not well defined for the probability distribution (11). This means that conclusions need to be drawn more carefully.}
\\

We have now included the analysis on the cut-off argument previously in the SI in as an appendix in the main text, as we changed the format from letter to regular article. We believe this will help the reader to better follow our arguments.

Moreover, we have now modified the paragraph below Eq. (11) and conclude with a sentence clarifying that the same "renormalization" procedure we describe for $\alpha = 1$, $\beta = 2/3$, and $\gamma = 1$ can be extended to other choices of exponents in the case they result in diverging expectation values. The dependence on $N$ of $\langle x_*^{\gamma}\rangle$ and $\langle x_*^{2\gamma}\rangle$ will emerge from the solution of the self-consistency equation for $\Lambda$.

Finally, for clarity, we now write explicitly the self-consistent equations for $\Lambda$, $\langle x_{*}\rangle_{\Lambda}$, and, $\langle x_{*}^2\rangle$ in the main text.

\section*{Referee B}

\blue{The authors clarify most of my previous concerns and clarify the  
scaling issue, and the paper is surely improved. I have still one  
relevant concern, I hope the authors fix before publication.}  
\\

We thank the referee for the previous comments, which helped improving the paper, and address the last concern below.
\\

\blue{In the response to reviewers the authors claim they better highlight  
	that their result are valid only for competitive Lotka Volterra. It is  
	true they added a couple of times the word competitive but in my  
	opinion the paper is not clear enough on this points and the results are presented are too general.} 
\\

\blue{1) The title "Beyond May: Complexity-stability relationships in  
	disordered dynamical systems" seem to present results that are 
	completely general. I suggest to make a more proper title such as  
	"Beyond May: Complexity-stability relationships in competitive  
	disordered Generalized Lotka Volterra Dynamics", or something similar.  
	Otherwise the title is misleading.} 
\\

We have now modified the title to make it clear that our work regards competitive disordered interactions only. We have additionally modified the abstract to stress our focus on competitive systems "We focus on competitive systems, and, using ...".
\\

\blue{2) Many comparison are made with respect to systems that have not such  
	constraint (positive mu), and this makes confusion.} 
\\

\blue{E.g. 1, " the classic generalized Lotka-Volterra (GLV) competition  
	model studied by Bunin and collaborators [25, 26] corresponds to f, g,  
	h all linear." But this is not true as in Bunin and Biroli works mu  
	can also be negative, is not constrained to assume positive value.}   
\\

We now modified this sentence specifying that the linear case of our model corresponds to GLV only when constrained to competitive interactions. 
\\

\blue{E.g. 2, "Compared to the case where $\mu$ and $\sigma$ scale with N (considered  
	in [25] and follow-up works), our choice reflects a strong-interaction  
	regime; it has been employed, for example, to study ecological  
	scenarios with large differences in species abundance and chaotic  
	turnover [35]". Again in the cited papers mu can be also negative.}  
\\

	We have removed the reference to [25] in order to avoid confusion. 
	We kept reference [35], because in this paper only competitive interactions ($\mu > 0$) are considered, as in our work.
\\

    \blue{I'm not sure it is worth to have a supplementary file. If it is not  
	published as letter and thus there is room it would be suitable to  
	better explain already in the main text the finite size argument to  
	solve self consist equations and the sensitivity to random exponents} 
\\

	\blue{Justification for/against Letter:}  
\\

	\blue{These calculations hold for a very specific cases and makes many  
	assumption that in general do not hold. Having more space the authors  
	could include the SI in the main text.}
\\

	Following the indication of the referee, we have now included the SI in the main text and changed the format from Letter to Regular article.

\end{document}

### 2024-09-17 papers/submission/beyond-may.tex
minor edits
minor corrections

@@
 \documentclass[
- prl,
+ pre,
  twocolumn,
@@
 \usepackage[justification=raggedright]{caption}
+\renewcommand{\appendixname}{APPENDIX}
 
@@
 
-\title{Complexity-stability relationships in disordered dynamical systems}
+\title{Complexity-stability relationships in competitive disordered dynamical systems}
 
@@
     In this paper, we revisit May's argument in light of recent developments in both ecology and random matrix theory. 
-    Using a non-linear generalization of the competitive Lotka-Volterra model, we show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems:
+    We focus on competitive systems, and, using a non-linear generalization of the competitive Lotka-Volterra model, we show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems:
     if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
@@
 %Introduction
-\paragraph*{}
-\emph{Introduction.---}Few mathematical arguments have influenced biological thinking like May's prediction that large, complex ecosystems cannot be stable~\cite{May1972}.
+\section{Introduction}
+Few mathematical arguments have influenced biological thinking like May's prediction that large, complex ecosystems cannot be stable~\cite{May1972}.
 It is, on the face of it, a perplexing conclusion.
@@
 %Model
-\paragraph*{}
-\emph{Model.---}Consider the dynamical system in $N$ variables
+\section{Model}
+Consider the dynamical system in $N$ variables
 \begin{equation}\label{dynamics}
@@
 to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
-In particular, the classic generalized Lotka-Volterra (GLV) competition model studied by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear.
+In particular, the classic generalized Lotka-Volterra (GLV) model studied by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally}, when constrained to competitive interactions, corresponds to $f$, $g$, $h$ all linear.
 However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
@@
 
+%Results
+\section{Results}
 %Homogeneous interactions
-\paragraph*{}
-\emph{Homogeneous interactions.---}Under what condition does~\eqref{dynamics} admit a (linearly) stable equilibrium? 
+\subsection{Homogeneous interactions}
+Under what condition does~\eqref{dynamics} admit a (linearly) stable equilibrium? 
 We begin with the simple case where all self-interactions have the same strength $A_{ii} = \mu_s$, and similarly for cross-interactions $A_{ij} = \mu$ ($i\neq j$).
@@
 %Random interactions: stability condition
-\paragraph*{}
-\emph{Random interactions: stability condition.---}We now consider the case of random interactions.
+\subsection{Random interactions: stability condition}
+We now consider the case of random interactions.
 Specifically, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$.
@@
 %Random interactions: equilibrium distribution
-\paragraph*{}
-\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
+\subsection{Random interactions: equilibrium distribution}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
 (Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time.)
@@
 We are interested in the case in which the number of degrees of freedom $N$ is the only parameter changing in the system and the strength and heterogeneity of the interactions are fixed~\cite{kessler2015generalized,fried2016communities}.
-Compared to the case where $\mu$ and $\sigma$ scale with $N$ (considered in \cite{bunin2017ecological} and follow-up works), our choice reflects a strong-interaction regime; it has been employed, for example, to study ecological scenarios with large differences in species abundance and chaotic turnover~\cite{mallmin2024chaotic}.
+Compared to the case where $\mu$ and $\sigma$ scale with $N$, our choice reflects a strong-interaction regime; it has been employed, for example, to study ecological scenarios with large differences in species abundance and chaotic turnover~\cite{mallmin2024chaotic}.
 As we show below, the cavity approximation does a good job at describing the effects of heterogeneity in the interactions as long as $\sigma \sqrt{N}$ is not negligible relative to $\mu N$, at which point we recover the homogeneous case.
@@
 defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently.
-Equipped with the equilibrium distribution $P(x_*)$, we can compute, for given $N$ and $\mu$, the maximal heterogeneity $\sigma_c$ compatible with the linear stability condition $N\langle (\sigma/\psi(x_*))^2\rangle < 1$.
 
 Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
-In this case the system is always feasible and the distribution can be normalized. However, the first and second moments formally diverge.
+In this case the system is always feasible, and the distribution can be normalized. 
+However, the first and second moments formally diverge.
 This problem can be overcome by considering that we are dealing with a large, but \emph{finite}, number of degrees of freedom. 
@@
 Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$.
-We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$, where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$, see Supplemental Material.
+We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$, where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$, i.e., such that there is statistically less than 1 variable out of $N$ with value above $\Lambda$; see Appendix~\ref{appendix a}.
 
-Following, e.g.,~\cite{Cui2020,Hatton2024} we can compute self-consistently: $\Lambda$, $\langle x_*\rangle_{\Lambda}$ and $\langle x_*^2\rangle_{\Lambda}\equiv\int_0^{\Lambda}dx_*x_*^2P(x_*)$.
+Following, e.g.,~\cite{Cui2020,Hatton2024} we can compute self-consistently $\Lambda$, $\langle x_*\rangle_{\Lambda}$ and $\langle x_*^2\rangle_{\Lambda}$:
+\begin{align}
+    \int_{0}^{\Lambda}dx_*P(x_*) &= 1 - \frac{1}{N} , \\    
+    \langle x_* \rangle_{\Lambda} &= \int_0^{\Lambda}dx_*x_*P(x_*) , \\
+    \langle x_*^2\rangle_{\Lambda} &= \int_0^{\Lambda}dx_*x_*^2P(x_*) .
+\end{align}
 The resulting distribution is plotted against simulations in Fig.~\ref{fig: cavity sol.}. 
 Simulations correspond to $A_s=\mu$, showing that neglecting the $A_s$ term in analytical expressions does not introduce a significant error.
+Notice that the same ``renormalization'' procedure described above can be extended to other choices of the exponents $\alpha$, $\beta$, and $\gamma$, if needed, to compute $\langle x_*^{\gamma}\rangle$, and $\langle x_*^{2\gamma}\rangle$.
+
+Equipped with the equilibrium distribution $P(x_*)$, we can compute, for given $N$ and $\mu$, the maximal heterogeneity $\sigma_c$ compatible with the linear stability condition $N\langle (\sigma/\psi(x_*))^2\rangle < 1$.
 
@@
 As a test, we can sample exponents from different Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and, for simplicity, all with the same standard deviation $\sigma_e$.
-Simulations show that our results hold as long as $\sigma_e$ is small enough, in particular, we observe a loss of stability when $\min_i\beta_i>\max_i\alpha_i$ (Supplemental Material).
+Simulations show that our results hold as long as $\sigma_e$ is small enough, in particular, we observe a loss of stability when $\min_i\beta_i>\max_i\alpha_i$ (Appendix~\ref{appendix b}).
 
-%Conclusion
-\paragraph*{}
-\emph{Discussion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and other fields. 
+%Discussion
+\section{Discussion}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and other fields. 
 In this Letter, we have shown that the condition $\sigma\sqrt{N}< \mu_s$ does not provide a complete picture of the relationship between complexity and stability. 
@@
 
-\medskip
-
 \begin{acknowledgments}
@@
 
-\medskip
+\appendix
+
+\section{CUT-OFF}
+\label{appendix a}
+In this appendix we consider a case amenable to analytical treatment to exemplify the argument behind the cut-off $\Lambda$ introduced in the main text to deal with diverging moments distributions.
+
+Consider the case of a power law distribution 
+\begin{equation}
+  P(x)=\frac{x^{-\beta}}{\mathcal{Z}} ,
+\end{equation}
+defined from 1 to $\infty$ and with 
+\begin{equation}
+  \mathcal{Z}=\int_1^{\infty}dxx^{-\beta}.  
+\end{equation}
+
+Let us choose $\beta=3/2$. Notice that this power law describes the behavior of the distribution we consider in the main text for our example with $\alpha=\gamma=1$ and $\beta=3/2$.
+The distribution is normalized with $\mathcal{Z}=2$, but the mean diverges. However, we would like to be able to describe the behavior of the sample mean 
+\begin{equation}
+  \bar{x}_N\equiv\frac{1}{N}\sum_{i=1}^Nx_i,
+\end{equation}
+
+where the $x_i$ are extracted from $P(x)$.
+For this purpose, we can define the quantity
+\begin{equation}
+  \langle x\rangle_{\Lambda}\equiv\int_1^{\Lambda}dxP(x)x,
+\end{equation}
+with the cut-off $\Lambda$ defined such that $\int_{\Lambda}^{\infty}dxP(x)=1/N$, i.e., such that there is statistically less than 1 variable with value above $\Lambda$ out of $N$ extracted variables.
+For the case $\beta=3/2$ we have
+\begin{equation}
+  \frac{1}{2}\int_{\Lambda}^{\infty}dxx^{-3/2}=\Lambda^{-1/2},
+\end{equation}
+and therefore $\Lambda=N^2$. We have for the mean
+\begin{equation}
+  \langle x\rangle_{\Lambda}=\frac{1}{2}\int_1^{N^2}dxx^{-1/2} = N-1.
+  \label{eq: SM-sample-mean}
+\end{equation}
+The result is plotted in Fig.~\ref{fig: SM-cut-off}, alongside the sample mean for extractions of $N=10,$ $10^2,$ $10^3,$ $10^4,$ $10^5,$ $10^6$.
+
+\begin{figure}[h!]
+  \includegraphics[width=.475\textwidth]{cut-off-justification.pdf}
+  \caption{Analytical prediction for the sample mean in Eq.~\eqref{eq: SM-sample-mean} as a function of the number of extractions, $N$, compared with the results from random extractions (dots).
+  }
+  \label{fig: SM-cut-off}
+\end{figure}
+
+\section{HETEROGENEOUS EXPONENTS}
+\label{appendix b}
+In this appendix we explore the robustness of our findings in the case in which the exponents characterizing the dynamics of each degree of freedom $x_i$ are not identical for all $i$.
+
+Consider the system
+\begin{equation}
+  \dot{x}_i = x_i^{\alpha_i} - \sum_jA_{ij}x_i^{\beta_i}x_j^{\gamma_i}
+\end{equation}
+where the exponents $\alpha_i$, $\beta_i$, $\gamma_i$, are extracted from Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and with the same standard deviation $\sigma_e$.
+The interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$.
+Our results are robust when $\sigma_e$ is small enough and we observe a loss of stability when $\min_i\beta_i>\max_i\alpha_i$.
+As an example, we show in the plot in Fig.~\ref{fig: SM-het-exp} the probability of stability vs. $\sigma_e$ for a system with $S=100$, $\mu=\sigma=0.01$, $\alpha=\gamma=1$ and $\beta=3/2$.
+The probability of stability is obtained as the fraction of stable systems out of $100$ realizations for each value of $\sigma_e$.
+
+\begin{figure}[h!]
+  \includegraphics[width=.475\textwidth]{stoch-exp.pdf}
+  \caption{Probability of stability vs. $\sigma_e$ for a system with $S=100$, $\mu=\sigma=0.01$, $\alpha=\gamma=1$ and $\beta=3/2$.
+  }
+  \label{fig: SM-het-exp}
+\end{figure}
 
